### 2018-05-21 submission/manuscript.Rmd
setting up template for writing

---
title: "Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl 
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
source('../code/functions.R')

loadLibs(c("tidyverse", "knitr", "rmarkdown"))
```

```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

```

\vspace{35mm}

Running title: SCFAs and Colorectal Cancer

\vspace{35mm}

Marc A Sze${^1}$, Nicholas A Lesniak${^1}$, Mack T Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

$2$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA

\newpage
\linenumbers

## Abstract

\newpage

## Introduction

\newpage

## Results

\newpage

## Discussion

\newpage

## Conclusions

\newpage

## Materials and Methods

\newpage

## Acknowledgements

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chian fatty acid analysis on the High-Performace Liquid Chromatography machine at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

\newpage

## References

<div id="refs"></div>

\newpage

Insert figure legends with the first sentence in bold, for example:

**Figure 1. Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B). Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).

\newpage

### 2018-05-21 submission/manuscript.tex
setting up template for writing

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal
Cancer}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}
\maketitle

\vspace{35mm}

Running title: SCFAs and Colorectal Cancer

\vspace{35mm}

Marc A Sze\({^1}\), Nicholas A Lesniak\({^1}\), Mack T Ruffin
IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

\(2\) Department of Family Medicine and Community Medicine, Penn State
Hershey Medical Center, Hershey, PA

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

\newpage

\subsection{Introduction}\label{introduction}

\newpage

\subsection{Results}\label{results}

\newpage

\subsection{Discussion}\label{discussion}

\newpage

\subsection{Conclusions}\label{conclusions}

\newpage

\subsection{Materials and Methods}\label{materials-and-methods}

\newpage

\subsection{Acknowledgements}\label{acknowledgements}

The authors thank the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study. We
would also like to thank Kwi Kim and Thomas M Schmidt for their help in
running the short-chian fatty acid analysis on the High-Performace
Liquid Chromatography machine at the University of Michigan. Salary
support for Marc A. Sze came from the Canadian Institute of Health
Research and NIH grant UL1TR002240. Salary support for Patrick D.
Schloss came from NIH grants P30DK034933 and 1R01CA215574.

\newpage

\subsection{References}\label{references}

\hypertarget{refs}{}

\newpage

Insert figure legends with the first sentence in bold, for example:

\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
16S rRNA gene sequences for different OTU definitions and level of
sequencing effort.} Rarefaction curves for different OTU definitions of
Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
environments in Table 1 for Bacteria (C) and Archaea (D).

\newpage

\end{document}

### 2018-05-21 submission/manuscript.Rmd
add to methods

@@
 ```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+# Read in meta data
+metaf <- read_csv("../data/raw/metadata/good_metaf_final.csv")
+metai <- read_csv("../data/raw/metadata/metaI_final.csv")
+geof_samples <- read_tsv("../data/raw/metadata/303030_subset/NexteraXT003Map.tsv", 
+                         col_names = F) %>% filter(X22 != "Negative")
 
@@
 
+**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016].
+In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before and after treatment for their tumor (n=`r length(metaf$initial)`).
+For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
+Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017].
+Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016].
+The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.
+
+**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol [@scfa_measures_venkataraman2016].
+The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations.
+These changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 in this report).
+
+**16s rRNA gene sequencing.** The workflow and processing have been described previously [@Schloss2009; @Kozich2013; @normalization_sze2017].
+The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity used the OptClust algorithm [@opticlust_westcott2017].
+
+**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used.
+Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the normalize_by_copy_number.py function, and subsequent imputed metagenomes created using the predict_metagenomes.py function.
+
+**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq with 125 bp paired end reads using a previously described method [@Hannigan2017].
+Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015].
+Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF gene alignment used the KEGG database [@Kanehisa2015].
+
+**Statistical analysis workflow.**
+
 \newpage

### 2018-05-21 submission/manuscript.tex
add to methods

@@
 
+\textbf{Study design and sampling.} The overall protocol has been
+described in detail previously (1, 2). In brief, this study used fecal
+samples obtained at either a single cross-sectional time point (n=490)
+or from before and after treatment for their tumor (n=67). For patients
+undergoing treatment for their tumor the length of time between their
+initial and follow up sample ranged from 188 - 546 days. Our use of
+treatment has been previously defined as encompassing removal of a tumor
+with or without chemotherapy and radiation (1). Diagnosis of tumor was
+made by colonoscopic examination and histopathological review of
+biopsies obtained (1, 2). The University of Michigan Institutional
+Review Board approved the study and informed consent was obtained from
+all participants in accordance to the guidelines set out by the Helsinki
+Decleration.
+
+\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
+acetate, butyrate, and propionate followed a previously published
+protocol (3). The following changes to this protocol included the use of
+frozen fecal samples suspended in 1ml of PBS instead of fecal
+suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
+weight of fecal samples instead of the average weight for SCFA
+concentration normalizations. These changes did not affect the overall
+median concentrations of these SCFAs between the two studies (see Table
+1 (3) and Figure 1 in this report).
+
+\textbf{16s rRNA gene sequencing.} The workflow and processing have been
+described previously (1, 4, 5). The major differences from these
+previous reports include: the use of version 1.39.5 of the mothur
+software package and clustering Operational Taxonomic Units (OTUs) at
+97\% similarity used the OptClust algorithm (6).
+
+\textbf{Generating imputed metagenomes.} The use of PICRUSt version
+1.1.2 with the recommended standard operating protocol (7) was used.
+Briefly, the mothur shared file and metadata was converted into a biom
+formated table using the biom convert function, the subsequent biom file
+was processed with the normalize\_by\_copy\_number.py function, and
+subsequent imputed metagenomes created using the predict\_metagenomes.py
+function.
+
+\textbf{Obtaining OPFs from metagenomes.} A subset of the
+cross-sectional group (n=490) containing a total of 85 individuals
+(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
+sequenced on an Illumina HiSeq with 125 bp paired end reads using a
+previously described method (8). Briefly, the sequences were quality
+filtered and sequences aligning to the human genome were removed prior
+to contig assembly with MEGAHIT (9). Open Reading Frames (ORFs) were
+identified using Prodigal (10), counts generated using Diamond (11),
+subsequent clustering into Operational Protein Families (OPFs) used
+mmseq2 (12), and OPF gene alignment used the KEGG database (13).
+
+\textbf{Statistical analysis workflow.}
+
 \newpage
@@
 \hypertarget{refs}{}
+\hypertarget{ref-normalization_sze2017}{}
+1. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
+\hypertarget{ref-crc_model_baxter2016}{}
+2. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-scfa_measures_venkataraman2016}{}
+3. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
+Variable responses of human microbiomes to dietary supplementation with
+resistant starch. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
+
+\hypertarget{ref-Schloss2009}{}
+4. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\hypertarget{ref-Kozich2013}{}
+5. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-opticlust_westcott2017}{}
+6. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+improved method for assigning amplicon-based sequence data to
+operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
+\hypertarget{ref-Langille2013}{}
+7. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
+\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
+\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
+Predictive functional profiling of microbial communities using 16S rRNA
+marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
+doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
+
+\hypertarget{ref-Hannigan2017}{}
+8. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
+\& the interactive dynamics of the colorectal cancer virome.
+doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
+
+\hypertarget{ref-Li2015}{}
+9. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane K},
+\textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution for
+large and complex metagenomics assembly via succinct de bruijn graph.
+Bioinformatics \textbf{31}:1674--1676.
+doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.
+
+\hypertarget{ref-Hyatt2010}{}
+10. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
+Prodigal: Prokaryotic gene recognition and translation initiation site
+identification. BMC Bioinformatics \textbf{11}:119.
+doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.
+
+\hypertarget{ref-Buchfink2014}{}
+11. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+and sensitive protein alignment using DIAMOND. Nature Methods
+\textbf{12}:59--60.
+doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.
+
+\hypertarget{ref-Steinegger2017}{}
+12. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+sensitive protein sequence searching for the analysis of massive data
+sets. Nature Biotechnology.
+doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.
+
+\hypertarget{ref-Kanehisa2015}{}
+13. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
+resource for gene and protein annotation. Nucleic Acids Research
+\textbf{44}:D457--D462.
+doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.
 

### 2018-05-21 submission/manuscript.Rmd
complete method first go write up

@@
 
-**Statistical analysis workflow.**
+**Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors.
+These genes were based on pathways from KEGG and can be found in the supplemental material [Table S1].
+
+**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017].
+Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test.
+Models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002].
+Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm.
+The measured SCFA concentrations were first tested for differences between groups.
+The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed.
+Next, the imputed gene counts of important mediators of SCFA creation were tested.
+Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
 

### 2018-05-21 submission/manuscript.tex
complete method first go write up

@@
 
-\textbf{Statistical analysis workflow.}
+\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
+located near the end of the pathways involved in the synthesis of
+acetate, butyrate, and propionate were analyzed for any differences
+between individuals with normal colons and those with tumors. These
+genes were based on pathways from KEGG and can be found in the
+supplemental material {[}Table S1{]}.
+
+\textbf{Statistical analysis workflow.} All analysis was performed using
+the statistical language R (14). Generally, differences between the
+different disease groups used a Kruskal-Walis rank sum test with a
+Dunn's post-hoc test. Models to classify high or low SCFA concentration
+based on 16S rRNA gene sequencing data were created using Random Forest
+(15). Regression models to classify the exact SCFA concentration based
+on 16S rRNA gene sequencing data also were built using the Random Forest
+algorithm. The measured SCFA concentrations were first tested for
+differences between groups. The ability of 16S rRNA gene sequencing to
+classify these concentrations were then assessed. Next, the imputed gene
+counts of important mediators of SCFA creation were tested. Finally, the
+counts generated for OPFs that matched important genes involved with
+SCFA creation were analyzed.
 
@@
 
+\hypertarget{ref-r_citation_2017}{}
+14. \textbf{R Core Team}. 2017. R: A language and environment for
+statistical computing. R Foundation for Statistical Computing, Vienna,
+Austria.
+
+\hypertarget{ref-randomforest_citation_2002}{}
+15. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
+
 \newpage

### 2018-05-21 submission/supplemental.Rmd
add supplemental file -- first commit

---
title: ''
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=0.75in
bibliography: null
---
\pagenumbering{gobble}
```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr"))

```

**Table S1: Breakdown of select genes and corresponding KEGG ID analyzed.**
```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

Kegg_ID <- c("K00929", "K01034", "K01035", "K01908", "K01895", "K00932", "K19697", 
             "K18372", "K00467", "K00156", "K00925", "K01512", "K01067", "K18118", "K01026", 
             "K01905", "K22224", "K01895", "K00128", "K14085", "K00149", "K00138")

Gene_Name <- c("Butyrate Kinase", "Acetate CoA subnunit A", "Acetate CoA subunit B", "Propionyl-CoA Synthetase", 
               "Acetyl-CoA Synthetase", "Propionate Kinase", "Propionate Kinase", "Methyl Acetate Hydrolase", 
               "Lactate 2-Monooxygenase", "Pyruvate Dehydrogenase", "Acetate Kinase", "Acylphosphatase", 
               "Acetyl-CoA Hydrolase", "Succinyl-CoA:Acetate", "Propionate CoA-Transferase", 
               "Acetate---CoA Ligase subunit A", "Acetate---CoA Ligase subunit B", "Acetyl-CoA Synthetase", 
               "Aldehyde Dehydrogenase", "Aldehyde Dehydrogenase Family 7", "Aldehyde Dehydrogenase Family 9", 
               "Aldehyde Dehydrogenase")

sup_table1 <- data_frame(Kegg_ID, Gene_Name)

colnames(sup_table1) <-  c("KEGG IDs", "Gene Name")

kable(sup_table1, format = "markdown", align = 'c', padding = 2)

```

### 2018-05-21 submission/manuscript.tex
add references for scfa gene specific look

@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG and can be found in the
-supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (13,
+14) and can be found in the supplemental material {[}Table S1{]}.
 
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (14). Generally, differences between the
+the statistical language R (15). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
@@
 based on 16S rRNA gene sequencing data were created using Random Forest
-(15). Regression models to classify the exact SCFA concentration based
+(16). Regression models to classify the exact SCFA concentration based
 on 16S rRNA gene sequencing data also were built using the Random Forest
@@
 
+\hypertarget{ref-scfa_baxter2014}{}
+14. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
+colonization with human feces determines colonic tumor burden.
+Microbiome \textbf{2}:20.
+doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
+
 \hypertarget{ref-r_citation_2017}{}
-14. \textbf{R Core Team}. 2017. R: A language and environment for
+15. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-15. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+16. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.

### 2018-05-21 submission/manuscript.Rmd
add references for scfa gene specific look
update methods write up
add figure legend template write up

@@
 **Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors.
-These genes were based on pathways from KEGG and can be found in the supplemental material [Table S1].
+These genes were based on pathways from KEGG as well as previous research [@scfa_baxter2014; @Kanehisa2015] and a list can be found in the supplemental material [Table S1].
 
@@
 Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test.
-Models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002].
+We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification.
+Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002].
 Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm.
@@
 
-Insert figure legends with the first sentence in bold, for example:
+**Figure 1.
+Using HPLC no change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A).
+Butyrate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (B).
+Propionate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (C).
+The black links indicate the median SCFA concentration.
+Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D).
+Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E).
+Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F).
+The black dots and lines represent the median change in SCFA concentration.
 
-**Figure 1.
-Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B).
-Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).
+**Figure 2.
+SCFAs do not improve OTU-based Random Forest models.** Difference between the area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
+Difference between the area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (B).
+The black linke represents the median AUC.
+The dotted line highlights an AUC of 0.5.
+
+**Figure 3.
+No change in butyrate producing genes identified between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A).
+Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B).
+The other genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
 
 \newpage
+
+**Figure S1.
+OTU-based Random Forest models of SCFA concentrations.** Classification Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration (A).
+The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).
+Regression Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on correlation to actual SCFA concentration (C).
+The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (D).

### 2018-05-21 submission/manuscript.tex
update methods write up
add figure legend template write up

@@
 genes were based on pathways from KEGG as well as previous research (13,
-14) and can be found in the supplemental material {[}Table S1{]}.
+14) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 different disease groups used a Kruskal-Walis rank sum test with a
-Dunn's post-hoc test. Models to classify high or low SCFA concentration
-based on 16S rRNA gene sequencing data were created using Random Forest
-(16). Regression models to classify the exact SCFA concentration based
-on 16S rRNA gene sequencing data also were built using the Random Forest
+Dunn's post-hoc test. We next assessed whether SCFAs added to Random
+Forest models built with OTU data significantly improved classification.
+Next, models to classify high or low SCFA concentration based on 16S
+rRNA gene sequencing data were created using Random Forest (16).
+Regression models to classify the exact SCFA concentration based on 16S
+rRNA gene sequencing data also were built using the Random Forest
 algorithm. The measured SCFA concentrations were first tested for
@@
 
-Insert figure legends with the first sentence in bold, for example:
+\textbf{Figure 1. Using HPLC no change in SCFA measurements was observed
+between normal, adenoma, and carcinoma individuals.} Acetate
+concentrations in fecal samples of individuals with normal colons,
+adenomas, and carcinomas (A). Butyrate concentrations in fecal samples
+of individuals with normal colons, adenomas, and carcinomas (B).
+Propionate concentrations in fecal samples of individuals with normal
+colons, adenomas, and carcinomas (C). The black links indicate the
+median SCFA concentration. Acetate concentrations in fecal samples
+before and after treatment for adenoma (yellow) and carcinoma (red) (D).
+Butyrate concentrations in fecal samples before and after treatment for
+adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in
+fecal samples before and after treatment for adenoma (yellow) and
+carcinoma (red) (F). The black dots and lines represent the median
+change in SCFA concentration.
 
-\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
-16S rRNA gene sequences for different OTU definitions and level of
-sequencing effort.} Rarefaction curves for different OTU definitions of
-Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
-environments in Table 1 for Bacteria (C) and Archaea (D).
+\textbf{Figure 2. SCFAs do not improve OTU-based Random Forest models.}
+Difference between the area under the curve of 100 different 80/20 split
+OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
+(A). Difference between the area under the curve of 100 different 80/20
+OTU-based normal versus carcinoma 10-fold CV models with and without
+SCFAs (B). The black linke represents the median AUC. The dotted line
+highlights an AUC of 0.5.
+
+\textbf{Figure 3. No change in butyrate producing genes identified
+between normal, adenoma, and carcinoma individuals.} Imputed gene
+relative abundance of important butyrate pathway genes using PICRUSt
+(A). Counts per million (corrected for size and number of contigs in an
+OPF) for the Butyrate Kinase gene (B). The other genes from the PICRUSt
+analysis did not align to any of the OPFs in the metagenome analysis.
 
@@
 
+\textbf{Figure S1. OTU-based Random Forest models of SCFA
+concentrations.} Classification Random Forest train and tests of 100
+different 80/20 OTU-based models with 10-fold CV based on higher or
+lower than the medain SCFA concentration (A). The top 10 OTUs based on
+mean decrease in accuracy (MDA) for each model, colored by their lowest
+taxonomic identification (B). Regression Random Forest train and tests
+of 100 different 80/20 OTU-based models with 10-fold CV based on
+correlation to actual SCFA concentration (C). The top 10 OTUs based on
+mean decrease in accuracy (MDA) for each model, colored by their lowest
+taxonomic identification (D).
+
 \end{document}

### 2018-05-22 submission/manuscript.Rmd
update methods and add first results section

@@
                          col_names = F) %>% filter(X22 != "Negative")
+
+# Read in analysis of SCFA by HPLC
+hplc_groups_scfa_kruskal <- read_csv("../data/process/tables/acetate_kruskal_crc_groups.csv") %>% 
+  mutate(scfa = "acetate") %>% slice(1) %>% 
+  bind_rows(
+    read_csv("../data/process/tables/butyrate_kruskal_crc_groups.csv") %>% 
+      mutate(scfa = "butyrate") %>% slice(1), 
+    read_csv("../data/process/tables/propionate_kruskal_crc_groups.csv") %>% 
+      mutate(scfa = "propionate") %>% slice(1)) %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH"))
+
+hplc_treat_scfa_kruskal <- read_csv("../data/process/tables/scfa_treatment_comparison_pvalues.csv") %>% 
+  filter(scfa != "isobutyrate")
+
+# Load in Random Forest with and without SCFA summary data
+disease_rf <- read_csv("../data/process/tables/rf_model_ttest_comparisons.csv")
 
@@
 
+**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time.
+There was no difference between individuals with normal colons and patients with either an adenoma or carcinoma for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
+We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma or carcinoma in which we had pre- and post-treatment fecal samples.
+Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
+Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.
+
+**Random Forest models with SCFA concentrations do not classify tumor better.**
+
 \newpage
@@
 **Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016].
-In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before and after treatment for their tumor (n=`r length(metaf$initial)`).
+In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment for their tumor (n=`r length(metaf$initial)`).
 For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
@@
 
-**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq with 125 bp paired end reads using a previously described method [@Hannigan2017].
+**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
 Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015].
-Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF gene alignment used the KEGG database [@Kanehisa2015].
+Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].
 
@@
 Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
+Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995.
 

### 2018-05-22 submission/manuscript.tex
update methods and add first results section

@@
 
+\textbf{Decreased SCFA concentrations are not associated with tumors.}
+There was no difference between individuals with normal colons and
+patients with either an adenoma or carcinoma after multiple comparison
+correction for any of the SCFAs measured (P-value \textgreater{} 0.15)
+{[}Figure 1A - 1C{]}. Although there was a trend for increasing acetate,
+butyrate, and propionate concentrations after treatment for tumors,
+there was no significant difference pre- and post-treatment for
+individuals with adenoma (P-value \textgreater{} 0.20) or carcinoma
+(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
+
 \newpage
@@
 samples obtained at either a single cross-sectional time point (n=490)
-or from before and after treatment for their tumor (n=67). For patients
-undergoing treatment for their tumor the length of time between their
-initial and follow up sample ranged from 188 - 546 days. Our use of
-treatment has been previously defined as encompassing removal of a tumor
-with or without chemotherapy and radiation (1). Diagnosis of tumor was
-made by colonoscopic examination and histopathological review of
-biopsies obtained (1, 2). The University of Michigan Institutional
-Review Board approved the study and informed consent was obtained from
-all participants in accordance to the guidelines set out by the Helsinki
-Decleration.
+or from before (pre-) and after (post-) treatment for their tumor
+(n=67). For patients undergoing treatment for their tumor the length of
+time between their initial and follow up sample ranged from 188 - 546
+days. Our use of treatment has been previously defined as encompassing
+removal of a tumor with or without chemotherapy and radiation (1).
+Diagnosis of tumor was made by colonoscopic examination and
+histopathological review of biopsies obtained (1, 2). The University of
+Michigan Institutional Review Board approved the study and informed
+consent was obtained from all participants in accordance to the
+guidelines set out by the Helsinki Decleration.
 
@@
 (normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
-sequenced on an Illumina HiSeq with 125 bp paired end reads using a
+sequenced on an Illumina HiSeq using 125 bp paired end reads and a
 previously described method (8). Briefly, the sequences were quality
@@
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (12), and OPF gene alignment used the KEGG database (13).
+mmseq2 (12), and OPF alignment used the KEGG database (13).
 
@@
 counts generated for OPFs that matched important genes involved with
-SCFA creation were analyzed.
+SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was
+used to correct for multiple comparisons {[}(17).
 
@@
 
+\hypertarget{ref-benjamini_controlling_1995}{}
+17. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+false discovery rate: A practical and powerful approach to multiple
+testing. Journal of the Royal Statistical Society Series B
+(Methodological) \textbf{57}:289--300.
+
 \newpage

### 2018-05-22 submission/manuscript.Rmd
add second result section update methods

@@
 
+# Load in data for select genes in SCFA pathways
+picrust_select <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv")
+opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")
+
 ```
@@
 
-**Random Forest models with SCFA concentrations do not classify tumor better.**
+**Random Forest models with SCFA concentrations do not classify tumor better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations.
+For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
+Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.
 
@@
 
+**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002].
+This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model.
+The reported AUCs, unless otherwise specified, are for the test sets.
+Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations.
+Regression models were built to assess how well OTUs could be used to get the SCFA concentration within a sample.
+
 **Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017].
 Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test.
-We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification.
-Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002].
-Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm.
-The measured SCFA concentrations were first tested for differences between groups.
-The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed.
+Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995].
+First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma).
+We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment.
+Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models.
 Next, the imputed gene counts of important mediators of SCFA creation were tested.
 Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
-Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995.
+Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest.
+Regression models also were created to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm.
+This was done to assess the ability of 16S rRNA gene sequencing to classify the SCFA concentrations.
 

### 2018-05-22 submission/manuscript.tex
add second result section update methods

@@
 \textbf{Decreased SCFA concentrations are not associated with tumors.}
-There was no difference between individuals with normal colons and
-patients with either an adenoma or carcinoma after multiple comparison
-correction for any of the SCFAs measured (P-value \textgreater{} 0.15)
-{[}Figure 1A - 1C{]}. Although there was a trend for increasing acetate,
-butyrate, and propionate concentrations after treatment for tumors,
-there was no significant difference pre- and post-treatment for
-individuals with adenoma (P-value \textgreater{} 0.20) or carcinoma
-(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
+We used frozen fecal samples from 490 individuals and HPLC to measure
+acetate, butyrate, and propionate concentrations at a cross-sectional
+point in time. There was no difference between individuals with normal
+colons and patients with either an adenoma or carcinoma for any of the
+SCFAs measured after multiple comparison correction (P-value
+\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
+concentration of SCFAs in 67 patients with an adenoma or carcinoma in
+which we had pre- and post-treatment fecal samples. Although there was a
+general trend for increasing acetate, butyrate, and propionate
+concentrations after treatment for tumors, there was no significant
+difference pre- and post-treatment for patients with adenoma (P-value
+\textgreater{} 0.20) or carcinoma (P-value \textgreater{} 0.80)
+{[}Figure 1D - 1F{]}. Even though there was no change in SCFA
+concentrations between individuals with normal colons and those with
+tumors, this information could still be important to help classify
+disease.
+
+\textbf{Random Forest models with SCFA concentrations do not classify
+tumor better.} Using the Random Forest algorithm we built models to
+classify normal versus adenoma and normal versus carcinoma with OTUs or
+OTUs and SCFA concentrations. For both adenoma and carcinoma models,
+there was no difference between the median AUC of models with or without
+SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
+Although there was no added information gained for classification of
+disease by including SCFAs to Random Forest models, it is possible that
+the genes for enzymes involved in SCFA pathways may change due to
+disease status; where a smaller number of microbes are responsible for
+the observed SCFA concentrations.
 
@@
 
+\textbf{Random Forest Models.} The model was first trained on 80\% of
+the data and then tested on the held out 20\% (80/20 split) using the
+Random Forest algorithm for both classification and regression models
+(15). This was repeated on 100 differen 80/20 splits of the data to
+generate a reasonable range for the AUC of the model. The reported AUCs,
+unless otherwise specified, are for the test sets. Classification models
+were built to group normal versus adenoma and normal versus carcinoma or
+high versus low SCFA concentrations. Regression models were built to
+assess how well OTUs could be used to get the SCFA concentration within
+a sample.
+
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (15). Generally, differences between the
+the statistical language R (16). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
-Dunn's post-hoc test. We next assessed whether SCFAs added to Random
-Forest models built with OTU data significantly improved classification.
-Next, models to classify high or low SCFA concentration based on 16S
-rRNA gene sequencing data were created using Random Forest (16).
-Regression models to classify the exact SCFA concentration based on 16S
-rRNA gene sequencing data also were built using the Random Forest
-algorithm. The measured SCFA concentrations were first tested for
-differences between groups. The ability of 16S rRNA gene sequencing to
-classify these concentrations were then assessed. Next, the imputed gene
+Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
+correct for multiple comparisons (17). First, we assessed differences in
+SCFA concentrations measured by HPLC between individuals with normal
+colons and patients with tumors (adenoma or carcinoma). We then analyzed
+whether SCFA concentrations changed in patients with an adenoma or
+carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
+alone or OTUs and SCFAs were better able to classify individuals with
+and without tumor using Random Forest models. Next, the imputed gene
 counts of important mediators of SCFA creation were tested. Finally, the
 counts generated for OPFs that matched important genes involved with
-SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was
-used to correct for multiple comparisons {[}(17).
+SCFA creation were analyzed. Finally, models to classify high or low
+SCFA concentration based on 16S rRNA gene sequencing data were created
+using Random Forest. Regression models also were created to classify the
+exact SCFA concentration based on 16S rRNA gene sequencing data also
+were built using the Random Forest algorithm. This was done to assess
+the ability of 16S rRNA gene sequencing to classify the SCFA
+concentrations.
 
@@
 
+\hypertarget{ref-randomforest_citation_2002}{}
+15. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
+
 \hypertarget{ref-r_citation_2017}{}
-15. \textbf{R Core Team}. 2017. R: A language and environment for
+16. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
 Austria.
-
-\hypertarget{ref-randomforest_citation_2002}{}
-16. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News \textbf{2}:18--22.
 

### 2018-05-22 submission/manuscript.Rmd
add third results section

@@
 
-**Random Forest models with SCFA concentrations do not classify tumor better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations.
+**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations.
 For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
 Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.
+
+**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas.
+First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
+This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
+We next took a subset of these `r length(metai$sample)` fecal samples and performed metagenomic sequencing to confirm these results.
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 3B].
+These observations provide evidence that the gene content does not change due to tumors.
+Overall, there may be some systematic bias present within our results that may account for the lack of associations.
 

### 2018-05-22 submission/manuscript.tex
add third results section

@@
 \textbf{Random Forest models with SCFA concentrations do not classify
-tumor better.} Using the Random Forest algorithm we built models to
+tumors better.} Using the Random Forest algorithm we built models to
 classify normal versus adenoma and normal versus carcinoma with OTUs or
@@
 the observed SCFA concentrations.
+
+\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
+associated with tumors.} Using a list of specific genes that are
+important for the synthesis of SCFAs {[}Table S1{]}, we looked for
+differences in gene abundance based on individuals having normal colons,
+adenomas, or carcinomas. First, using imputed gene relative abundance
+based on 16S rRNA gene sequencing we found no difference in any of the
+genes inolved with acetate, butyrate, or propionate synthesis (P-value
+\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
+highlighted by visualizing genes important in butyrate synthesis
+{[}Figure 3A{]}. We next took a subset of these 490 fecal samples and
+performed metagenomic sequencing to confirm these results. Like the
+imputed gene results, metagenomic analysis found that there was no
+difference in any of the genes involved in SCFA synthesis between
+individuals with normal colons and patients with adenoma or carcinoma
+(P-value \textgreater{} 0.70) {[}Table S3{]}. This lack of difference
+can be highlighted when we visualize the results for butyrate kinase
+{[}Figure 3B{]}. These observations provide evidence that the gene
+content does not change due to tumors. Overall, there may be some
+systematic bias present within our results that may account for the lack
+of associations.
 

### 2018-05-22 submission/supplemental.Rmd
add third results section

@@
 ```
+
+\newpage
+
+**Table S2: Breakdown of Kruskal-Wallis tests using PICRUSt imputed relative abundance of select genes in SCFA pathways.**
+```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+table_s2 <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv") %>% 
+  select(kegg_ortholog, statistic, p.value, bh) %>% 
+  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))
+
+colnames(table_s2) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")
+
+kable(table_s2, format = "markdown", align = 'c', padding = 2)
+# opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")
+```
+
+\newpage
+
+**Table S3: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+table_s3 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
+  select(kegg_id, statistic, p.value, bh) %>% 
+  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))
+
+colnames(table_s3) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")
+
+kable(table_s3, format = "markdown", align = 'c', padding = 2)
+```

### 2018-05-22 submission/manuscript.Rmd
add complete result section template

@@
 **Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time.
-There was no difference between individuals with normal colons and patients with either an adenoma or carcinoma for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
-We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma or carcinoma in which we had pre- and post-treatment fecal samples.
+There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
+We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples.
 Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
@@
 For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
-Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.
+Although including SCFA concentrations did not add extra information for classification of disease using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.
 
@@
 This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
-We next took a subset of these `r length(metai$sample)` fecal samples and performed metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
-This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 3B].
+Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results.
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B].
 These observations provide evidence that the gene content does not change due to tumors.
-Overall, there may be some systematic bias present within our results that may account for the lack of associations.
+Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.
+
+**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations.
+Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A].
+However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A].
+The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B].
+These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate.
+Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations.
+Overall, our results do not support the hypothesis that lower SCFA concentrations are associated with colorectal cancer.
 
@@
 Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations.
-Regression models were built to assess how well OTUs could be used to get the SCFA concentration within a sample.
 
@@
 Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
-Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest.
-Regression models also were created to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm.
-This was done to assess the ability of 16S rRNA gene sequencing to classify the SCFA concentrations.
+Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using the Random Forest algorithm.
 

### 2018-05-22 submission/manuscript.tex
add complete result section template

@@
 point in time. There was no difference between individuals with normal
-colons and patients with either an adenoma or carcinoma for any of the
-SCFAs measured after multiple comparison correction (P-value
-\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
-concentration of SCFAs in 67 patients with an adenoma or carcinoma in
-which we had pre- and post-treatment fecal samples. Although there was a
-general trend for increasing acetate, butyrate, and propionate
-concentrations after treatment for tumors, there was no significant
-difference pre- and post-treatment for patients with adenoma (P-value
-\textgreater{} 0.20) or carcinoma (P-value \textgreater{} 0.80)
-{[}Figure 1D - 1F{]}. Even though there was no change in SCFA
-concentrations between individuals with normal colons and those with
-tumors, this information could still be important to help classify
-disease.
+colons (n=172) and patients with either an adenoma (n=198) or carcinoma
+(n=120) for any of the SCFAs measured after multiple comparison
+correction (P-value \textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next
+measured the concentration of SCFAs in 67 patients with an adenoma
+(n=41) or carcinoma (n=26) in which we had pre- and post-treatment fecal
+samples. Although there was a general trend for increasing acetate,
+butyrate, and propionate concentrations after treatment for tumors,
+there was no significant difference pre- and post-treatment for patients
+with adenoma (P-value \textgreater{} 0.20) or carcinoma (P-value
+\textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even though there was no
+change in SCFA concentrations between individuals with normal colons and
+those with tumors, this information could still be important to help
+classify disease.
 
@@
 SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
-Although there was no added information gained for classification of
-disease by including SCFAs to Random Forest models, it is possible that
-the genes for enzymes involved in SCFA pathways may change due to
+Although including SCFA concentrations did not add extra information for
+classification of disease using Random Forest models, it is possible
+that the genes for enzymes involved in SCFA synthesis may change due to
 disease status; where a smaller number of microbes are responsible for
@@
 highlighted by visualizing genes important in butyrate synthesis
-{[}Figure 3A{]}. We next took a subset of these 490 fecal samples and
-performed metagenomic sequencing to confirm these results. Like the
-imputed gene results, metagenomic analysis found that there was no
-difference in any of the genes involved in SCFA synthesis between
-individuals with normal colons and patients with adenoma or carcinoma
-(P-value \textgreater{} 0.70) {[}Table S3{]}. This lack of difference
-can be highlighted when we visualize the results for butyrate kinase
-{[}Figure 3B{]}. These observations provide evidence that the gene
-content does not change due to tumors. Overall, there may be some
-systematic bias present within our results that may account for the lack
-of associations.
+{[}Figure 3A{]}. Next, we took a subset of these 490 fecal samples
+(n=85) and performed metagenomic sequencing to confirm these results.
+Like the imputed gene results, metagenomic analysis found that there was
+no difference in any of the genes involved in SCFA synthesis between
+individuals with normal colons (n=29) and patients with adenoma (n=28)
+or carcinoma (n=28) (P-value \textgreater{} 0.70) {[}Table S3{]}. This
+lack of difference is highlighted when we visualize the results for
+butyrate kinase {[}Figure 3B{]}. These observations provide evidence
+that the gene content does not change due to tumors. Although these
+results suggest that SCFA concentrations do not change due to tumor,
+there may be some errors present within our results that may account for
+the lack of associations.
+
+\textbf{Expected taxa are associated with higher SCFA concentrations
+regardless of tumor status.} Using OTU data we built Random Forest
+models to classify higher than median and lower than median SCFA
+concentrations. Overall, OTU data had a reasonable ability to classify
+high and low SCFA concentrations {[}Figure S1A{]}. However, these models
+tended to be overfit, suggesting that rarer taxa may be important for
+this classification {[}Figure S1A{]}. The most important OTUs to these
+models (assessed with mean decrease in accuracy (MDA)) were to taxa that
+are normally associated with SCFA production {[}Figure S1B{]}. These
+results highlight that there are no immediately apparent errors in our
+data because SCFA concentrations are associated with taxa known to
+produce acetate, butyrate, and propionate. Additionally, OTUs associated
+with these taxa are the most important to models that can classify high
+and low SCFA concentrations. Overall, our results do not support the
+hypothesis that lower SCFA concentrations are associated with colorectal
+cancer.
 
@@
 were built to group normal versus adenoma and normal versus carcinoma or
-high versus low SCFA concentrations. Regression models were built to
-assess how well OTUs could be used to get the SCFA concentration within
-a sample.
+high versus low SCFA concentrations.
 
@@
 SCFA concentration based on 16S rRNA gene sequencing data were created
-using Random Forest. Regression models also were created to classify the
-exact SCFA concentration based on 16S rRNA gene sequencing data also
-were built using the Random Forest algorithm. This was done to assess
-the ability of 16S rRNA gene sequencing to classify the SCFA
-concentrations.
+using the Random Forest algorithm.
 

### 2018-05-22 code/download_bacterial_genomes.py
first commit

@@ -1,0 +1,86 @@
+#!python
+
+# This code does a diamond blast on a select set of sequences
+# Designed to be used after the opf_seq_picker.py program
+# It uses the the kegg database
+
+
+############## Internal parameters used by all functions is program ########################
+
+# Import needed libraries
+import os, sys, argparse
+
+# Import other code with useful functions that will be used
+from qual_trim import create_samples_to_download
+
+# Set up working directory
+workdir = "data/raw/"
+
+# Set up refrence directory
+refdir = "data/references/"
+
+
+############################################################################################
+
+# Function to create the bacterial genome download directory
+def make_bacterial_genome_directory(directory_name):
+
+	# Create a directory to hold diamond analysis
+	if os.path.exists(directory_name):
+
+		print("%s directory exists" % (directory_name))
+
+	else:
+
+		print("No %s directory found, creating..." % (directory_name))
+
+		os.system("mkdir %s" % (directory_name))
+
+
+
+
+
+
+
+# Run the actual program
+def main(bacteria, data_base, type_of_file, av_genomes, outputPath):
+
+	make_bacterial_genome_directory(outputPath)
+
+
+
+
+
+# Initialized when program starts
+if __name__ == '__main__': 
+	# Command line argument parser with tags for componenets within the program
+	parser = argparse.ArgumentParser(description=__doc__)
+	# Add arguments that will be called from the command line
+	parser.add_argument("-b", "--bacterium", 
+		default="ALL", 
+		type=str, help="Sets the specific bacterium genome to download. \
+		The default is to download all bacteria and the argument is set. \
+		to 'ALL'. Otherwise entries should be in the form of eg. Klebsiella_pneumoniae.\n")
+	parser.add_argument("-d", "--database_used", 
+		default="genbank", choices = ["refseq", "genbank"],  
+		type=str, help="The database to download the genome(s) from. \
+		The default database that is used is set to 'genbank'. \
+		The input must be one of 'genbank' or 'refseq'.")
+	parser.add_argument("-t", "--file_type", 
+		default="fasta",choices = ["genbank", "fasta", "gff", "feature_table"], type=str, 
+		help="The file type that will be downloaded from the database. \
+		The default is set to fasta. There are total of 4 options: \
+		'genbank', 'fasta', 'gff', and 'feature_table'.")
+	parser.add_argument("-g", "--genomes", 
+		choices=["genbank", "refseq"], type=str, 
[…]

### 2018-05-22 code/download_bacterial_genomes.py
update code for download

@@ -1,17 +1,19 @@
 #!python
 
-# This code does a diamond blast on a select set of sequences
-# Designed to be used after the opf_seq_picker.py program
-# It uses the the kegg database
+# Code inspired by:
+	# github repo https://github.com/ryjohnson09/bacteria_genome_pull.git
+
+# The main role for this code is to download full genomes from NCBI to be used to search 
+# subsequent OGUs for closest taxonomy
 
 
 ############## Internal parameters used by all functions is program ########################
 
 # Import needed libraries
-import os, sys, argparse
+import os, sys, argparse, subprocess, fnmatch
 
 # Import other code with useful functions that will be used
-from qual_trim import create_samples_to_download
+from ftplib import FTP
 
 # Set up working directory
 workdir = "data/raw/"
@@ -38,14 +40,87 @@ def make_bacterial_genome_directory(directory_name):
 
 
 
+# Function to find all genomes in the bacteria directory
+def find_genomes(data_base_to_use, output_dir):
+
+	#initialize FTP server 
+	ftp_site = 'ftp.ncbi.nlm.nih.gov'
+	ftp = FTP(ftp_site)
+
+	ftp.login()
+	ftp.cwd("genomes/%s/bacteria" % (data_base_to_use))
+	dirs = ftp.nlst()
+
+	for genome in dirs:
+
+		print(genome)
+
+	sys.exit()
 
 
+# Function to run the actual download
+def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_dir):
+
+	#initialize FTP server 
+	ftp_site = 'ftp.ncbi.nlm.nih.gov'
+	ftp = FTP(ftp_site)
+
+	# set up system calls for downloads of specific file types
+	if file_needed == "fasta":
+		file_type = "--exclude='*cds_from*' --exclude='*rna_from*' --include='*genomic.fna.gz' --exclude='*'"
+	elif file_needed == "genbank":
+		file_type = "--include='*genomic.gbff.gz' --exclude='*'"
+	elif file_needed == "gff":
+		file_type = "--include='*genomic.gff.gz' --exclude='*'"
+	elif args.type == "feature_table":
+		file_type = "--include='*feature_table.txt.gz' --exclude='*'"
+
+	if bacteria_needed == "ALL":
+
+		try:
[…]

### 2018-05-23 submission/manuscript.Rmd
add template discussion

@@
 
+The observations from this study do not support the hypothesis that SCFA concentrations are lower in individuals with tumors.
+Whether we directly measured the SCFA concentration or investigate genes associated with their production no difference could be identified [Figure 1 & 3].
+Further, no benefit could be found in using the concentrations to help classify individuals with and without tumors [Figure 2].
+These observations are in stark contrast to some of the previous literature.
+
+Much of the previous literature on SCFA benefit to human colorectal cancer has been illustrated in model systems [@test_OKeefe2016].
+Many SCFAs are produced through the breakdown of fiber [@Holscher2017].
+A recent study in mice found that fiber supplementation increased SCFA concentration and decreased tumor burden [@Bishehsari2018].
+Additionally, SCFAs such as butyrate have been found to inhibit tumor growth in *in vitro* experiments [@Encarnao2018].
+Yet, observations in humans has been mixed.
+Previous case/control studies have suggested that SCFAs are lower in individuals with carcinoma tumors [@Ohigashi2013].
+However, individual studies and meta-analysis on fiber supplementation to prevent tumor recurrence have found no benefit [@Schatzkin2000; @Yao2017].
+Our results align with what has been reported in randomized-controlled trials that SCFAs do not protect against all types of colorectal tumors.
+It is possible though that there are specific instances where SCFAs may be beneficial.
+
+One limitation of ours and others research into the effect of SCFAs in colorectal cancer has been that all tumors are treated as the same type.
+However, there are known differences in the types of mutations that occur [] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
+Similar to the idea of specific immunotherapy for specific tumors [], SCFAs may have beneficial effects for specific types of colorectal cancers.
+Future research will need to test if this is a valid hypothesis.
+Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective.
+Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [].
+
+The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [].
+Protection could be via a different pathway and other metabolites that have not been extensively studied.
+Alternatively, protection may not occur via a metabolite but instead occur through niche exclusion of mouth-associated microbes (e.g.
+*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* []).
+The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection with chronic inflammation replacing the role of antibiotics.
+Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these observations.
+
 \newpage

### 2018-05-23 submission/manuscript.tex
add template discussion

@@
 
+The observations from this study do not support the hypothesis that SCFA
+concentrations are lower in individuals with tumors. Whether we directly
+measured the SCFA concentration or investigate genes associated with
+their production no difference could be identified {[}Figure 1 \& 3{]}.
+Further, no benefit could be found in using the concentrations to help
+classify individuals with and without tumors {[}Figure 2{]}. These
+observations are in stark contrast to some of the previous literature.
+
+Much of the previous literature on SCFA benefit to human colorectal
+cancer has been illustrated in model systems (1). Many SCFAs are
+produced through the breakdown of fiber (2). A recent study in mice
+found that fiber supplementation increased SCFA concentration and
+decreased tumor burden (3). Additionally, SCFAs such as butyrate have
+been found to inhibit tumor growth in \emph{in vitro} experiments (4).
+Yet, observations in humans has been mixed. Previous case/control
+studies have suggested that SCFAs are lower in individuals with
+carcinoma tumors (5). However, individual studies and meta-analysis on
+fiber supplementation to prevent tumor recurrence have found no benefit
+(6, 7). Our results align with what has been reported in
+randomized-controlled trials that SCFAs do not protect against all types
+of colorectal tumors. It is possible though that there are specific
+instances where SCFAs may be beneficial.
+
+One limitation of ours and others research into the effect of SCFAs in
+colorectal cancer has been that all tumors are treated as the same type.
+However, there are known differences in the types of mutations that
+occur {[}{]} and treating all tumors as equal may actually hide any
+benefit that could be found in specific subsets of individuals. Similar
+to the idea of specific immunotherapy for specific tumors {[}{]}, SCFAs
+may have beneficial effects for specific types of colorectal cancers.
+Future research will need to test if this is a valid hypothesis.
+Although there are limitations with the current research on SCFAs and
+colorectal tumors, evidence is building that these specific metabolites
+may not be protective. Yet, taxa that are associated with SCFA
+production are consistently higher in normal colons than those with
+carcinomas {[}{]}.
+
+The potential protection against colorectal cancers may not be from
+SCFAs even though taxa associated with their production are higher in
+individuals with normal colons versus patients with carcinomas {[}{]}.
+Protection could be via a different pathway and other metabolites that
+have not been extensively studied. Alternatively, protection may not
+occur via a metabolite but instead occur through niche exclusion of
+mouth-associated microbes (e.g. \emph{Fusobacterium},
+\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus}
+{[}{]}). The idea of niche exclusion is similar to how the community
+protects against \emph{Clostridium difficile} infection with chronic
+inflammation replacing the role of antibiotics. Although we did not find
+lower concentrations of SCFAs associated with colorectal tumors, we
+think that there are many exciting new avenues to explore because of
+these observations.
+
 \newpage
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (1, 2). In brief, this study used fecal
+described in detail previously (8, 9). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (1).
+removal of a tumor with or without chemotherapy and radiation (8).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (1, 2). The University of
+histopathological review of biopsies obtained (8, 9). The University of
 Michigan Institutional Review Board approved the study and informed
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (3). The following changes to this protocol included the use of
-frozen fecal samples suspended in 1ml of PBS instead of fecal
+protocol (10). The following changes to this protocol included the use
+of frozen fecal samples suspended in 1ml of PBS instead of fecal
 suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
@@
 median concentrations of these SCFAs between the two studies (see Table
-1 (3) and Figure 1 in this report).
+1 (10) and Figure 1 in this report).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (1, 4, 5). The major differences from these
+described previously (8, 11, 12). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (6).
+97\% similarity used the OptClust algorithm (13).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (7) was used.
+1.1.2 with the recommended standard operating protocol (14) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (8). Briefly, the sequences were quality
+previously described method (15). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (9). Open Reading Frames (ORFs) were
-identified using Prodigal (10), counts generated using Diamond (11),
+to contig assembly with MEGAHIT (16). Open Reading Frames (ORFs) were
+identified using Prodigal (17), counts generated using Diamond (18),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (12), and OPF alignment used the KEGG database (13).
+mmseq2 (19), and OPF alignment used the KEGG database (20).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (13,
-14) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (20,
+21) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 Random Forest algorithm for both classification and regression models
-(15). This was repeated on 100 differen 80/20 splits of the data to
+(22). This was repeated on 100 differen 80/20 splits of the data to
 generate a reasonable range for the AUC of the model. The reported AUCs,
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (16). Generally, differences between the
+the statistical language R (23). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
 Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (17). First, we assessed differences in
+correct for multiple comparisons (24). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 \hypertarget{refs}{}
+\hypertarget{ref-test_OKeefe2016}{}
+1. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+metabolites and colon cancer. Nature Reviews Gastroenterology \&
+Hepatology \textbf{13}:691--706.
+doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
+
+\hypertarget{ref-Holscher2017}{}
+2. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
+gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
+doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.
+
+\hypertarget{ref-Bishehsari2018}{}
+3. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
+M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
+\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
+\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
+composition of gut microbiota, promotes SCFA production, and suppresses
+colon carcinogenesis. Genes \textbf{9}:102.
+doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
+
+\hypertarget{ref-Encarnao2018}{}
+4. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
+\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
+\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
+AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
+that improves irinotecan effect in colon cancer cells. The Journal of
+Nutritional Biochemistry \textbf{56}:183--192.
+doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.
+
+\hypertarget{ref-Ohigashi2013}{}
+5. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
+\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
+\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
+microbiota, short chain fatty acids, and fecal pH in patients with
+colorectal cancer. Digestive Diseases and Sciences
+\textbf{58}:1717--1726.
+doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.
+
+\hypertarget{ref-Schatzkin2000}{}
+6. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
+\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
+\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
+\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
+of effect of a low-fat, high-fiber diet on the recurrence of colorectal
+adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
+doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.
+
+\hypertarget{ref-Yao2017}{}
+7. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao Y},
+\textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre for
+the prevention of recurrent colorectal adenomas and carcinomas. Cochrane
+Database of Systematic Reviews.
+doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.
+
 \hypertarget{ref-normalization_sze2017}{}
-1. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+8. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-2. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+9. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-3. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+10. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-4. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-5. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+12. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-6. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+13. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Langille2013}{}
-7. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+14. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 \hypertarget{ref-Hannigan2017}{}
-8. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+15. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-9. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane K},
-\textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution for
-large and complex metagenomics assembly via succinct de bruijn graph.
-Bioinformatics \textbf{31}:1674--1676.
+16. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
+for large and complex metagenomics assembly via succinct de bruijn
+graph. Bioinformatics \textbf{31}:1674--1676.
 doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.
@@
 \hypertarget{ref-Hyatt2010}{}
-10. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+17. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-11. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+18. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-12. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+19. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-13. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+20. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-14. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+21. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-15. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-r_citation_2017}{}
-16. \textbf{R Core Team}. 2017. R: A language and environment for
+23. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-17. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-05-23 submission/manuscript.Rmd
add references to discussion

@@
 One limitation of ours and others research into the effect of SCFAs in colorectal cancer has been that all tumors are treated as the same type.
-However, there are known differences in the types of mutations that occur [] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
-Similar to the idea of specific immunotherapy for specific tumors [], SCFAs may have beneficial effects for specific types of colorectal cancers.
+However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
+Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal cancers.
 Future research will need to test if this is a valid hypothesis.
 Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective.
-Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [].
+Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 
-The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [].
+The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 Protection could be via a different pathway and other metabolites that have not been extensively studied.
 Alternatively, protection may not occur via a metabolite but instead occur through niche exclusion of mouth-associated microbes (e.g.
-*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* []).
+*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016]).
 The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection with chronic inflammation replacing the role of antibiotics.

### 2018-05-23 submission/manuscript.tex
add references to discussion

@@
 However, there are known differences in the types of mutations that
-occur {[}{]} and treating all tumors as equal may actually hide any
-benefit that could be found in specific subsets of individuals. Similar
-to the idea of specific immunotherapy for specific tumors {[}{]}, SCFAs
-may have beneficial effects for specific types of colorectal cancers.
-Future research will need to test if this is a valid hypothesis.
-Although there are limitations with the current research on SCFAs and
-colorectal tumors, evidence is building that these specific metabolites
-may not be protective. Yet, taxa that are associated with SCFA
-production are consistently higher in normal colons than those with
-carcinomas {[}{]}.
+occur (8) and treating all tumors as equal may actually hide any benefit
+that could be found in specific subsets of individuals. Similar to the
+idea of specific immunotherapy for specific tumors (9), SCFAs may have
+beneficial effects for specific types of colorectal cancers. Future
+research will need to test if this is a valid hypothesis. Although there
+are limitations with the current research on SCFAs and colorectal
+tumors, evidence is building that these specific metabolites may not be
+protective. Yet, taxa that are associated with SCFA production are
+consistently higher in normal colons than those with carcinomas
+(10--12).
 
@@
 SCFAs even though taxa associated with their production are higher in
-individuals with normal colons versus patients with carcinomas {[}{]}.
+individuals with normal colons versus patients with carcinomas (10--12).
 Protection could be via a different pathway and other metabolites that
@@
 mouth-associated microbes (e.g. \emph{Fusobacterium},
-\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus}
-{[}{]}). The idea of niche exclusion is similar to how the community
+\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (13,
+14)). The idea of niche exclusion is similar to how the community
 protects against \emph{Clostridium difficile} infection with chronic
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (8, 9). In brief, this study used fecal
+described in detail previously (15, 16). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (8).
+removal of a tumor with or without chemotherapy and radiation (15).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (8, 9). The University of
-Michigan Institutional Review Board approved the study and informed
+histopathological review of biopsies obtained (15, 16). The University
+of Michigan Institutional Review Board approved the study and informed
 consent was obtained from all participants in accordance to the
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (10). The following changes to this protocol included the use
+protocol (17). The following changes to this protocol included the use
 of frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 median concentrations of these SCFAs between the two studies (see Table
-1 (10) and Figure 1 in this report).
+1 (17) and Figure 1 in this report).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (8, 11, 12). The major differences from these
+described previously (15, 18, 19). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (13).
+97\% similarity used the OptClust algorithm (20).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (14) was used.
+1.1.2 with the recommended standard operating protocol (21) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (15). Briefly, the sequences were quality
+previously described method (22). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (16). Open Reading Frames (ORFs) were
-identified using Prodigal (17), counts generated using Diamond (18),
+to contig assembly with MEGAHIT (23). Open Reading Frames (ORFs) were
+identified using Prodigal (24), counts generated using Diamond (25),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (19), and OPF alignment used the KEGG database (20).
+mmseq2 (26), and OPF alignment used the KEGG database (27).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (20,
-21) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (27,
+28) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 Random Forest algorithm for both classification and regression models
-(22). This was repeated on 100 differen 80/20 splits of the data to
+(29). This was repeated on 100 differen 80/20 splits of the data to
 generate a reasonable range for the AUC of the model. The reported AUCs,
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (23). Generally, differences between the
+the statistical language R (30). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
 Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (24). First, we assessed differences in
+correct for multiple comparisons (31). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 
+\hypertarget{ref-Fearon1990}{}
+8. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+colorectal tumorigenesis. Cell \textbf{61}:759--767.
+doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.
+
+\hypertarget{ref-Thomas2016}{}
+9. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+agents targeting the BCR-ABL tyrosine kinase as philadelphia
+chromosome-positive acute lymphoblastic leukemia treatment. Expert
+Opinion on Drug Discovery \textbf{11}:1061--1070.
+doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
+
+\hypertarget{ref-Shah2017}{}
+10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
+\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
+\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
+sequence-based faecal microbial community survey data to identify a
+composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.
+
+\hypertarget{ref-normalization_Sze2017}{}
+11. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
+\hypertarget{ref-meta_analysis_crc_Sze2018}{}
+12. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors.
+doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.
+
+\hypertarget{ref-Zeller2014}{}
+13. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
+\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Molecular Systems
+Biology \textbf{10}:766--766.
+doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
+
+\hypertarget{ref-Baxter2016}{}
+14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
 \hypertarget{ref-normalization_sze2017}{}
-8. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+15. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-9. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+16. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-10. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+17. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+18. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-12. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-13. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+20. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Langille2013}{}
-14. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 \hypertarget{ref-Hannigan2017}{}
-15. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+22. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-16. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+23. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-17. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+24. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-18. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+25. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-19. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+26. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-20. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+27. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-21. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+28. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+29. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-r_citation_2017}{}
-23. \textbf{R Core Team}. 2017. R: A language and environment for
+30. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+31. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-05-23 submission/manuscript.Rmd
add conclusion statement

@@
 
+Our observations conssitently found no difference in SCFA concentration between individuals with normal colons and patients with either adenoma or carcinoma tumors.
+Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.
+Overall, these results suggest that the SCFAs produced by resident microbes do not protect against tumor.
+By focusing on other types of metabolites, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
+
 \newpage

### 2018-05-23 submission/manuscript.tex
add conclusion statement

@@
 \subsection{Conclusions}\label{conclusions}
+
+Our observations conssitently found no difference in SCFA concentration
+between individuals with normal colons and patients with either adenoma
+or carcinoma tumors. Although these results are different than other
+reports in the literature, they do align with the randomized controlled
+trials that have tested fiber use in preventing colorectal tumor
+recurrence. Overall, these results suggest that the SCFAs produced by
+resident microbes do not protect against tumor. By focusing on other
+types of metabolites, the identification of more promising therapeutic
+options for use in treating colorectal cancer may be found.
 

### 2018-05-23 submission/manuscript.Rmd
add intro template

@@
 
+Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [].
+The envrionment has been attributed as an important risk factor for CRC [].
+This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [] and whether the bacterial community is altered [].
+Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors [].
+Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet [].
+The most extensively studied of these SCFAs are acetate, butyrate, and propionate [].
+Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a viable treatment option.
+
+Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment [].
+In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [].
+Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects.
+Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while increasing SCFA concentrations [].
+Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.
+
+There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
+The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons [].
+Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [].
+These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.
+Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC there is a need for more investigation into this area.
+
+Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations.
+Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors.
+Additionally, by increasing the adenoma tumor cohort it we will be able to identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases.
+It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors.
+This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.
+
+To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups.
+The first group will be at a cross sectional point in time while the second group will be before (pre-) and after (post-)treatment.
+Additionally, we also will assess whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma, use PICRUSt and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyze how well 16S rRNA gene sequencing can predict SCFA concentrations.
+This extensive investigation will provide additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against tumorigenesis.
+
 \newpage
@@
 
-Our observations conssitently found no difference in SCFA concentration between individuals with normal colons and patients with either adenoma or carcinoma tumors.
+Our observations found no difference in SCFA concentration, their utility as a classification tool, and for genes for enzymes involved in SCFA synthesis between individuals with normal colons and patients with either adenoma or carcinoma tumors.
 Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.

### 2018-05-23 submission/manuscript.tex
add intro template

@@
 \subsection{Introduction}\label{introduction}
+
+Colorectal cancer (CRC) is currently the third leading cancer-related
+cause of death within the US {[}{]}. The envrionment has been attributed
+as an important risk factor for CRC {[}{]}. This has lead to the
+investigation of how the microbiota may exacerbate or cause tumorigensis
+{[}{]} and whether the bacterial community is altered {[}{]}. Multiple
+reports in case/control studies on the microbiota and CRC have
+identified taxa commonly associated with the normal colon to be
+decreased in patients with carcinoma tumors {[}{]}. Many of these taxa
+actively produce short-chain fatty acids (SCFAs) from fiber that is part
+of our general diet {[}{]}. The most extensively studied of these SCFAs
+are acetate, butyrate, and propionate {[}{]}. Overall, CRC is a growing
+health problem and the microbiota through SCFAs have shown promise as a
+viable treatment option.
+
+Within model systems SCFAs, in particular butyrate, have shown positive
+results in CRC treatment {[}{]}. In particular, butyrate has been shown
+to inhibit cancer cell growth in \emph{in vitro} systems {[}{]}.
+Additionally, supplementation with food sources that bacteria use to
+create these SCFAs may also be able to confer beneficial effects. Fiber
+supplementation in mouse models of CRC caused an overall reduction in
+tumor burden while increasing SCFA concentrations {[}{]}. Although these
+model systems provide important preliminary evidence towards the ability
+of SCFAs to reduce and treat tumors, the studies reporting benefit in
+humans has been less convincing.
+
+There is a lack of evidence on the benefit of increasing SCFA
+concentrations to protect against CRC in human populations. The initial
+case/control studies that investigated SCFA concentrations in CRC have
+found that patients with carcinomas had lower concentrations of acetate,
+butyrate, and propionate versus patients with adenomas and individuals
+with normal colons {[}{]}. Although this would argue that increasing
+SCFA concentrations could be protective against tumorigenesis, fiber
+supplementation in randomized controlled trials have consistently failed
+to protect against tumor recurrence {[}{]}. These findings would argue
+against the utility of treatments that aim to use SCFAs to reduce or
+protect against tumorigenesis. Given the lack of clear evidence in human
+studies of the benefit of SCFAs in CRC there is a need for more
+investigation into this area.
+
+Our current study will fill some of the current gaps in the literature
+that relate to the study of SCFAs and CRC in human populations.
+Specifically, it will attempt to replicate previous case/control
+findings on SCFA concentrations in individuals with and without tumors.
+Additionally, by increasing the adenoma tumor cohort it we will be able
+to identify if the observation of a continuous gradual reduction in SCFA
+concentrations occurs as tumor severity increases. It also will build
+upon these observations and assess the utility of using SCFAs and
+Operational Taxonomic Units (OTUs) as a potential risk stratification
+tool of colorectal tumors. This study will provide important information
+on the replicability of previous findings in humans by extensively
+studying how SCFAs could be associated with colorectal tumors.
+
+To accomplish this task we will directly measure the concentration of
+acetate, butyrate, and propionate within fecal samples for two different
+groups. The first group will be at a cross sectional point in time while
+the second group will be before (pre-) and after (post-)treatment.
+Additionally, we also will assess whether adding SCFA concentrations to
+OTU data improves the classification of adenoma or carcinoma, use
+PICRUSt and metagenomic sequencing to assess the genes involved in SCFA
+synthesis, and analyze how well 16S rRNA gene sequencing can predict
+SCFA concentrations. This extensive investigation will provide
+additional information as to whether the taxa that are decreased in
+carcionma use SCFAs to help protect against tumorigenesis.
 
@@
 
-Our observations conssitently found no difference in SCFA concentration
-between individuals with normal colons and patients with either adenoma
-or carcinoma tumors. Although these results are different than other
-reports in the literature, they do align with the randomized controlled
-trials that have tested fiber use in preventing colorectal tumor
-recurrence. Overall, these results suggest that the SCFAs produced by
-resident microbes do not protect against tumor. By focusing on other
+Our observations found no difference in SCFA concentration, their
+utility as a classification tool, and for genes for enzymes involved in
+SCFA synthesis between individuals with normal colons and patients with
+either adenoma or carcinoma tumors. Although these results are different
+than other reports in the literature, they do align with the randomized
+controlled trials that have tested fiber use in preventing colorectal
+tumor recurrence. Overall, these results suggest that the SCFAs produced
+by resident microbes do not protect against tumor. By focusing on other
 types of metabolites, the identification of more promising therapeutic

### 2018-05-23 submission/manuscript.Rmd
add references to intro

@@
 
-Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [].
-The envrionment has been attributed as an important risk factor for CRC [].
-This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [] and whether the bacterial community is altered [].
-Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors [].
-Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet [].
-The most extensively studied of these SCFAs are acetate, butyrate, and propionate [].
+Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
+The envrionment has been attributed as an important risk factor for CRC [@Lichtenstein2000].
+This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
+Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011].
+Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet [@Holscher2017].
+The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016].
 Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a viable treatment option.
 
-Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment [].
-In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [].
+Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment [@test_OKeefe2016].
+In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
 Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects.
-Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while increasing SCFA concentrations [].
+Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
 Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.
@@
 There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
-The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons [].
-Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [].
+The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons [@Ohigashi2013].
+Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
 These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.

### 2018-05-23 submission/manuscript.tex
add references to intro

@@
 Colorectal cancer (CRC) is currently the third leading cancer-related
-cause of death within the US {[}{]}. The envrionment has been attributed
-as an important risk factor for CRC {[}{]}. This has lead to the
+cause of death within the US (1, 2). The envrionment has been attributed
+as an important risk factor for CRC (3). This has lead to the
 investigation of how the microbiota may exacerbate or cause tumorigensis
-{[}{]} and whether the bacterial community is altered {[}{]}. Multiple
+(4--6) and whether the bacterial community is altered (7, 8). Multiple
 reports in case/control studies on the microbiota and CRC have
 identified taxa commonly associated with the normal colon to be
-decreased in patients with carcinoma tumors {[}{]}. Many of these taxa
+decreased in patients with carcinoma tumors (9--11). Many of these taxa
 actively produce short-chain fatty acids (SCFAs) from fiber that is part
-of our general diet {[}{]}. The most extensively studied of these SCFAs
-are acetate, butyrate, and propionate {[}{]}. Overall, CRC is a growing
+of our general diet (12). The most extensively studied of these SCFAs
+are acetate, butyrate, and propionate (13). Overall, CRC is a growing
 health problem and the microbiota through SCFAs have shown promise as a
@@
 Within model systems SCFAs, in particular butyrate, have shown positive
-results in CRC treatment {[}{]}. In particular, butyrate has been shown
-to inhibit cancer cell growth in \emph{in vitro} systems {[}{]}.
+results in CRC treatment (14). In particular, butyrate has been shown to
+inhibit cancer cell growth in \emph{in vitro} systems (15).
 Additionally, supplementation with food sources that bacteria use to
@@
 supplementation in mouse models of CRC caused an overall reduction in
-tumor burden while increasing SCFA concentrations {[}{]}. Although these
-model systems provide important preliminary evidence towards the ability
-of SCFAs to reduce and treat tumors, the studies reporting benefit in
-humans has been less convincing.
+tumor burden while also increasing SCFA concentrations (16). Although
+these model systems provide important preliminary evidence towards the
+ability of SCFAs to reduce and treat tumors, the studies reporting
+benefit in humans has been less convincing.
 
@@
 butyrate, and propionate versus patients with adenomas and individuals
-with normal colons {[}{]}. Although this would argue that increasing
-SCFA concentrations could be protective against tumorigenesis, fiber
+with normal colons (17). Although this would argue that increasing SCFA
+concentrations could be protective against tumorigenesis, fiber
 supplementation in randomized controlled trials have consistently failed
-to protect against tumor recurrence {[}{]}. These findings would argue
+to protect against tumor recurrence (18). These findings would argue
 against the utility of treatments that aim to use SCFAs to reduce or
@@
 Much of the previous literature on SCFA benefit to human colorectal
-cancer has been illustrated in model systems (1). Many SCFAs are
-produced through the breakdown of fiber (2). A recent study in mice
+cancer has been illustrated in model systems (14). Many SCFAs are
+produced through the breakdown of fiber (12). A recent study in mice
 found that fiber supplementation increased SCFA concentration and
-decreased tumor burden (3). Additionally, SCFAs such as butyrate have
-been found to inhibit tumor growth in \emph{in vitro} experiments (4).
+decreased tumor burden (16). Additionally, SCFAs such as butyrate have
+been found to inhibit tumor growth in \emph{in vitro} experiments (15).
 Yet, observations in humans has been mixed. Previous case/control
 studies have suggested that SCFAs are lower in individuals with
-carcinoma tumors (5). However, individual studies and meta-analysis on
+carcinoma tumors (17). However, individual studies and meta-analysis on
 fiber supplementation to prevent tumor recurrence have found no benefit
-(6, 7). Our results align with what has been reported in
+(18, 19). Our results align with what has been reported in
 randomized-controlled trials that SCFAs do not protect against all types
@@
 However, there are known differences in the types of mutations that
-occur (8) and treating all tumors as equal may actually hide any benefit
-that could be found in specific subsets of individuals. Similar to the
-idea of specific immunotherapy for specific tumors (9), SCFAs may have
-beneficial effects for specific types of colorectal cancers. Future
-research will need to test if this is a valid hypothesis. Although there
-are limitations with the current research on SCFAs and colorectal
-tumors, evidence is building that these specific metabolites may not be
-protective. Yet, taxa that are associated with SCFA production are
-consistently higher in normal colons than those with carcinomas
-(10--12).
+occur (20) and treating all tumors as equal may actually hide any
+benefit that could be found in specific subsets of individuals. Similar
+to the idea of specific immunotherapy for specific tumors (21), SCFAs
+may have beneficial effects for specific types of colorectal cancers.
+Future research will need to test if this is a valid hypothesis.
+Although there are limitations with the current research on SCFAs and
+colorectal tumors, evidence is building that these specific metabolites
+may not be protective. Yet, taxa that are associated with SCFA
+production are consistently higher in normal colons than those with
+carcinomas (7, 8, 22).
 
@@
 SCFAs even though taxa associated with their production are higher in
-individuals with normal colons versus patients with carcinomas (10--12).
-Protection could be via a different pathway and other metabolites that
-have not been extensively studied. Alternatively, protection may not
-occur via a metabolite but instead occur through niche exclusion of
+individuals with normal colons versus patients with carcinomas (7, 8,
+22). Protection could be via a different pathway and other metabolites
+that have not been extensively studied. Alternatively, protection may
+not occur via a metabolite but instead occur through niche exclusion of
 mouth-associated microbes (e.g. \emph{Fusobacterium},
-\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (13,
-14)). The idea of niche exclusion is similar to how the community
+\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
+10)). The idea of niche exclusion is similar to how the community
 protects against \emph{Clostridium difficile} infection with chronic
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (15, 16). In brief, this study used fecal
+described in detail previously (23, 24). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (15).
+removal of a tumor with or without chemotherapy and radiation (23).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (15, 16). The University
+histopathological review of biopsies obtained (23, 24). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (17). The following changes to this protocol included the use
+protocol (25). The following changes to this protocol included the use
 of frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 median concentrations of these SCFAs between the two studies (see Table
-1 (17) and Figure 1 in this report).
+1 (25) and Figure 1 in this report).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (15, 18, 19). The major differences from these
+described previously (23, 26, 27). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (20).
+97\% similarity used the OptClust algorithm (28).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (21) was used.
+1.1.2 with the recommended standard operating protocol (29) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (22). Briefly, the sequences were quality
+previously described method (30). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (23). Open Reading Frames (ORFs) were
-identified using Prodigal (24), counts generated using Diamond (25),
+to contig assembly with MEGAHIT (31). Open Reading Frames (ORFs) were
+identified using Prodigal (32), counts generated using Diamond (33),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (26), and OPF alignment used the KEGG database (27).
+mmseq2 (34), and OPF alignment used the KEGG database (35).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (27,
-28) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (35,
+36) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 Random Forest algorithm for both classification and regression models
-(29). This was repeated on 100 differen 80/20 splits of the data to
+(37). This was repeated on 100 differen 80/20 splits of the data to
 generate a reasonable range for the AUC of the model. The reported AUCs,
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (30). Generally, differences between the
+the statistical language R (38). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
 Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (31). First, we assessed differences in
+correct for multiple comparisons (39). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 \hypertarget{refs}{}
-\hypertarget{ref-test_OKeefe2016}{}
-1. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
-metabolites and colon cancer. Nature Reviews Gastroenterology \&
-Hepatology \textbf{13}:691--706.
-doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
+\hypertarget{ref-Haggar2009}{}
+1. \textbf{Haggar F}, \textbf{Boushey R}. 2009. Colorectal cancer
+epidemiology: Incidence, mortality, survival, and risk factors. Clinics
+in Colon and Rectal Surgery \textbf{22}:191--197.
+doi:\href{https://doi.org/10.1055/s-0029-1242458}{10.1055/s-0029-1242458}.
 
-\hypertarget{ref-Holscher2017}{}
-2. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
-gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
-doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.
+\hypertarget{ref-Siegel2016}{}
+2. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
+Cancer statistics, 2016. CA: A Cancer Journal for Clinicians
+\textbf{66}:7--30.
+doi:\href{https://doi.org/10.3322/caac.21332}{10.3322/caac.21332}.
 
-\hypertarget{ref-Bishehsari2018}{}
-3. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
-\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
-M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
-\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
-\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
-composition of gut microbiota, promotes SCFA production, and suppresses
-colon carcinogenesis. Genes \textbf{9}:102.
-doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
+\hypertarget{ref-Lichtenstein2000}{}
+3. \textbf{Lichtenstein P}, \textbf{Holm NV}, \textbf{Verkasalo PK},
+\textbf{Iliadou A}, \textbf{Kaprio J}, \textbf{Koskenvuo M},
+\textbf{Pukkala E}, \textbf{Skytthe A}, \textbf{Hemminki K}. 2000.
+Environmental and heritable factors in the causation of cancer analyses
+of cohorts of twins from sweden, denmark, and finland. New England
+Journal of Medicine \textbf{343}:78--85.
+doi:\href{https://doi.org/10.1056/nejm200007133430201}{10.1056/nejm200007133430201}.
 
-\hypertarget{ref-Encarnao2018}{}
-4. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
-\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
-\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
-AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
-that improves irinotecan effect in colon cancer cells. The Journal of
-Nutritional Biochemistry \textbf{56}:183--192.
-doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.
+\hypertarget{ref-Zackular2013}{}
+4. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
+tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
+doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.
 
-\hypertarget{ref-Ohigashi2013}{}
-5. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
-\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
-\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
-microbiota, short chain fatty acids, and fecal pH in patients with
-colorectal cancer. Digestive Diseases and Sciences
-\textbf{58}:1717--1726.
-doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.
+\hypertarget{ref-Baxter2014}{}
+5. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
+colonization with human feces determines colonic tumor burden.
+Microbiome \textbf{2}:20.
+doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
 
-\hypertarget{ref-Schatzkin2000}{}
-6. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
-\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
-\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
-\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
-\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
-of effect of a low-fat, high-fiber diet on the recurrence of colorectal
-adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
-doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.
-
-\hypertarget{ref-Yao2017}{}
-7. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao Y},
-\textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre for
-the prevention of recurrent colorectal adenomas and carcinomas. Cochrane
-Database of Systematic Reviews.
-doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.
-
-\hypertarget{ref-Fearon1990}{}
-8. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
-colorectal tumorigenesis. Cell \textbf{61}:759--767.
-doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.
-
-\hypertarget{ref-Thomas2016}{}
-9. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
-agents targeting the BCR-ABL tyrosine kinase as philadelphia
-chromosome-positive acute lymphoblastic leukemia treatment. Expert
-Opinion on Drug Discovery \textbf{11}:1061--1070.
-doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
+\hypertarget{ref-Zackular2015}{}
+6. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
+role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
+doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.
 
 \hypertarget{ref-Shah2017}{}
-10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
+7. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
 \textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
@@
 
-\hypertarget{ref-normalization_Sze2017}{}
-11. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
-\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
-microbiota in patients after treatment for colonic lesions. Microbiome
-\textbf{5}.
-doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
-
 \hypertarget{ref-meta_analysis_crc_Sze2018}{}
-12. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+8. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-Zeller2014}{}
-13. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+9. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 \hypertarget{ref-Baxter2016}{}
-14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+10. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 
+\hypertarget{ref-Kostic2011}{}
+11. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
+\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
+AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
+\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
+\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
+\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
+identifies association of fusobacterium with colorectal carcinoma.
+Genome Research \textbf{22}:292--298.
+doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.
+
+\hypertarget{ref-Holscher2017}{}
+12. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
+gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
+doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.
+
+\hypertarget{ref-Louis2016}{}
+13. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
+and butyrate by the human colonic microbiota. Environmental Microbiology
+\textbf{19}:29--41.
+doi:\href{https://doi.org/10.1111/1462-2920.13589}{10.1111/1462-2920.13589}.
+
+\hypertarget{ref-test_OKeefe2016}{}
+14. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+metabolites and colon cancer. Nature Reviews Gastroenterology \&
+Hepatology \textbf{13}:691--706.
+doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
+
+\hypertarget{ref-Encarnao2018}{}
+15. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
+\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
+\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
+AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
+that improves irinotecan effect in colon cancer cells. The Journal of
+Nutritional Biochemistry \textbf{56}:183--192.
+doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.
+
+\hypertarget{ref-Bishehsari2018}{}
+16. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
+M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
+\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
+\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
+composition of gut microbiota, promotes SCFA production, and suppresses
+colon carcinogenesis. Genes \textbf{9}:102.
+doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
+
+\hypertarget{ref-Ohigashi2013}{}
+17. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
+\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
+\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
+microbiota, short chain fatty acids, and fecal pH in patients with
+colorectal cancer. Digestive Diseases and Sciences
+\textbf{58}:1717--1726.
+doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.
+
+\hypertarget{ref-Yao2017}{}
+18. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
+Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
+for the prevention of recurrent colorectal adenomas and carcinomas.
+Cochrane Database of Systematic Reviews.
+doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.
+
+\hypertarget{ref-Schatzkin2000}{}
+19. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
+\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
+\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
+\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
+of effect of a low-fat, high-fiber diet on the recurrence of colorectal
+adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
+doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.
+
+\hypertarget{ref-Fearon1990}{}
+20. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+colorectal tumorigenesis. Cell \textbf{61}:759--767.
+doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.
+
+\hypertarget{ref-Thomas2016}{}
+21. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+agents targeting the BCR-ABL tyrosine kinase as philadelphia
+chromosome-positive acute lymphoblastic leukemia treatment. Expert
+Opinion on Drug Discovery \textbf{11}:1061--1070.
+doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
+
+\hypertarget{ref-normalization_Sze2017}{}
+22. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
 \hypertarget{ref-normalization_sze2017}{}
-15. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-16. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+24. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-17. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+25. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-18. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-20. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+28. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Langille2013}{}
-21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+29. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 \hypertarget{ref-Hannigan2017}{}
-22. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+30. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-23. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+31. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-24. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+32. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-25. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+33. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-26. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+34. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-27. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+35. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-28. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+36. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-29. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+37. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-r_citation_2017}{}
-30. \textbf{R Core Team}. 2017. R: A language and environment for
+38. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-31. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+39. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-05-23 code/download_bacterial_genomes.py
add completed code

@@ -77,6 +77,9 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 
 	if bacteria_needed == "ALL":
 
+		print("Downloading all bacterial genomes from %s database..." % 
+			(data_base_to_use))
+
 		try:
 			subprocess.call("rsync -Lrtv %s rsync://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/*/latest_assembly_versions/*/ %s" % 
 				(file_type, data_base_to_use, output_dir), shell=True)
@@ -85,30 +88,28 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 
 	else:
 
+		ftp.login()
+		ftp.cwd('genomes/%s/bacteria' % (data_base_to_use))
+		dirs_list = ftp.nlist()
+
+		genome_matches = fnmatch.filter(dirs_list, bacteria_needed)
+
+		if len(genome_matches) == 0:
+
+			print("%s not found in %s database. Exiting." % 
+				(bacteria_needed, data_base_to_use))
+
+			sys.exit(1)
+
+		print("Downloading %s genome from %s database..." % 
+			(bacteria_needed, data_base_to_use))
+
 		try:
 			subprocess.call("rsync -Lrtv %s rsync://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/%s/latest_assembly_versions/*/ %s" % 
 				(file_type, data_base_to_use, bacteria_needed, output_dir), shell=True)
 		except:
 			raise Exception("Failed to download files")
-
-# Need to implement this as a fail safe 
-#Check that bacterium is in NCBI genbank/refseq
-# if args.bacterium:
-# 	ftp.login()
-# 	ftp.cwd('genomes/%s/bacteria' % NCBI_database)
-# 	dirs = ftp.nlst()
-# 	pattern = args.bacterium
-# 	genome_match = fnmatch.filter(dirs, pattern)
-# 	if len(genome_match) == 0:
-# 		print("Bacterium %s not found in %s -----> Exiting program" % (args.bacterium, NCBI_database))
-# 		sys.exit(1)
-# else:
-# 	print("")
-# 	print("Please provide bacterium name '-b'")
-# 	print("Exiting program")
-# 	print("")
-# 	sys.exit(1)
-		
+	
 
 
 # Run the actual program

### 2018-05-23 code/download_bacterial_genomes.py
add comments

@@ -46,14 +46,19 @@ def find_genomes(data_base_to_use, output_dir):
 	#initialize FTP server 
 	ftp_site = 'ftp.ncbi.nlm.nih.gov'
 	ftp = FTP(ftp_site)
-
+	# gets all bacteria that are contained in the current database
 	ftp.login()
 	ftp.cwd("genomes/%s/bacteria" % (data_base_to_use))
 	dirs = ftp.nlst()
-
+	# Open txt file to write to
+	temp_file = open("%sbacteria_in_database.txt" % (output_dir), 'w')
+	# Output message to stdout
+	print("Identifying all bacterial genomes in %s database..." % 
+		(data_base_to_use))
+	# Create the needed list of all bacteria present
 	for genome in dirs:
 
-		print(genome)
+		temp_file.write(genome+'\n')
 
 	sys.exit()
 
@@ -74,26 +79,26 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 		file_type = "--include='*genomic.gff.gz' --exclude='*'"
 	elif args.type == "feature_table":
 		file_type = "--include='*feature_table.txt.gz' --exclude='*'"
-
+	# Checks if the default to download all bacterial genomes is set
 	if bacteria_needed == "ALL":
 
 		print("Downloading all bacterial genomes from %s database..." % 
 			(data_base_to_use))
-
+		# Attempts to dowload all bacterial genomes and throws exception if it fails
 		try:
 			subprocess.call("rsync -Lrtv %s rsync://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/*/latest_assembly_versions/*/ %s" % 
 				(file_type, data_base_to_use, output_dir), shell=True)
 		except:
 			raise Exception("Failed to download files")
-
+	# If a specific bacterium is needed downloads this from given database
 	else:
-
+		# login into the ftp server
 		ftp.login()
 		ftp.cwd('genomes/%s/bacteria' % (data_base_to_use))
 		dirs_list = ftp.nlist()
-
+		# finds any matches within the given database
 		genome_matches = fnmatch.filter(dirs_list, bacteria_needed)
-
+		# Exits program if there is no matches found
 		if len(genome_matches) == 0:
 
 			print("%s not found in %s database. Exiting." % 
@@ -103,7 +108,7 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 
[…]

### 2018-05-23 code/download_bacterial_genomes.py
update download call

@@ -72,13 +72,13 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 
 	# set up system calls for downloads of specific file types
 	if file_needed == "fasta":
-		file_type = "--exclude='*cds_from*' --exclude='*rna_from*' --include='*genomic.fna.gz' --exclude='*'"
+		file_type = "--reject-regex='*cds_from*' --reject-regex='*rna_from*' --accept-regex='*genomic.fna.gz'"
 	elif file_needed == "genbank":
-		file_type = "--include='*genomic.gbff.gz' --exclude='*'"
+		file_type = "--accept-regex='*genomic.gbff.gz'"
 	elif file_needed == "gff":
-		file_type = "--include='*genomic.gff.gz' --exclude='*'"
+		file_type = "--accept-regex='*genomic.gff.gz'"
 	elif args.type == "feature_table":
-		file_type = "--include='*feature_table.txt.gz' --exclude='*'"
+		file_type = "--accept-regex='*feature_table.txt.gz'"
 	# Checks if the default to download all bacterial genomes is set
 	if bacteria_needed == "ALL":
 
@@ -86,8 +86,8 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 			(data_base_to_use))
 		# Attempts to dowload all bacterial genomes and throws exception if it fails
 		try:
-			subprocess.call("rsync -Lrtv %s rsync://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/*/latest_assembly_versions/*/ %s" % 
-				(file_type, data_base_to_use, output_dir), shell=True)
+			os.system("wget -r -np %s ftp://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/*/latest_assembly_versions/*/ %s" % 
+				(file_type, data_base_to_use, output_dir))
 		except:
 			raise Exception("Failed to download files")
 	# If a specific bacterium is needed downloads this from given database
@@ -110,7 +110,7 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_di
 			(bacteria_needed, data_base_to_use))
 		# If there is a match then trys to downloads genome from desired database or throws an error
 		try:
[…]

### 2018-05-24 submission/manuscript.Rmd
add abstract template

@@
 ## Abstract
+
+**Background.** Certain bacterial community members have been shown to increase colon tumorigenesis while other taxa have been associated with normal colons.
+The some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids which have been shown to inhibit tumor growth in model systems.
+However, the data supporting the importance of SCFAs in humans is less certain.
+Here we test the hypothesis that short-chain fatty acid concentration decreases in individuals with colorectal tumors.
+
+**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate.
+Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.
+
+**Results.** No difference in short-chain fatty acid concentration was found between normal individuals and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
+There was no difference in the ability of classification models with or without short-chain fatty acids added to them in predicting adenoma or carcinoma (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
+Using metagenomic sequencing, there was also no difference in genes involved with short-chain fatty acid synthesis (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
+
+**Conclusions.** Although our data does not support the hypothesis that short-chain fatty acids decrease in individuals that have colorectal tumors, there may be specific scenarios where short-chain fatty acids may still be beneficial for treatment of colorectal cancer.
+Alternatively, there may be other metabolites that have not been thoroughly investigated that are more important to human colorectal cancer.
 

### 2018-05-24 submission/manuscript.tex
add abstract template

@@
 \subsection{Abstract}\label{abstract}
+
+\textbf{Background.} Certain bacterial community members have been shown
+to increase colon tumorigenesis while other taxa have been associated
+with normal colons. The some of the taxa associated with normal colons
+can use fiber to produce short-chain fatty acids which have been shown
+to inhibit tumor growth in model systems. However, the data supporting
+the importance of SCFAs in humans is less certain. Here we test the
+hypothesis that short-chain fatty acid concentration decreases in
+individuals with colorectal tumors.
+
+\textbf{Methods.} We analyzed a cross-sectional (n=490) and pre- and
+post-treatment (n=67) group for their concentrations of acetate,
+butyrate, and propionate. Analysis also included tumor classification
+models using Random Forest, imputed gene relative abundance with
+PICRUSt, and metagenomic sequencing on a subset (n=85) of the total
+cross-sectional group.
+
+\textbf{Results.} No difference in short-chain fatty acid concentration
+was found between normal individuals and patients with adenoma or
+carcinoma (P-value \textgreater{} 0.15). There was no difference in the
+ability of classification models with or without short-chain fatty acids
+added to them in predicting adenoma or carcinoma (P-value \textgreater{}
+0.05). Using metagenomic sequencing, there was also no difference in
+genes involved with short-chain fatty acid synthesis (P-value
+\textgreater{} 0.70).
+
+\textbf{Conclusions.} Although our data does not support the hypothesis
+that short-chain fatty acids decrease in individuals that have
+colorectal tumors, there may be specific scenarios where short-chain
+fatty acids may still be beneficial for treatment of colorectal cancer.
+Alternatively, there may be other metabolites that have not been
+thoroughly investigated that are more important to human colorectal
+cancer.
 

### 2018-05-24 code/make_pi_opf_combined_graph.R
update code for location and treatment tests

@@ -22,10 +22,10 @@ acetate_genes <- c("K18372", "K00467", "K00156", "K00925", "K01512", "K01067", "
   # K01035 - acetate CoA (beta)
 
 # Load in needed metadata and rename sample column
-picrust_data <- read_csv("data/process/selected_scfa_gene_data.csv") %>% 
+picrust_data <- read_csv("data/process/tables/selected_scfa_gene_data.csv") %>% 
   filter(kegg_ortholog %in% butyrate_genes)
 
-opf_data <- read_csv("data/process/select_scfa_opf_data.csv") %>% 
+opf_data <- read_csv("data/process/tables/select_scfa_opf_data.csv") %>% 
   filter(kegg_id %in% butyrate_genes)
 
 imputed_fig <- picrust_data %>% 

### 2018-05-24 code/run_opf_select_scfa_analysis.R

@@ -17,9 +17,9 @@ option_list = list(
               help="opf shared file [default= %default]", metavar="character"), 
   make_option(c("-md", "--metadata"), type="character", default="data/process/sra_meta_conversion.txt", 
               help="metadata of the samples [default= %default]", metavar="character"), 
-  make_option(c("-o", "--output_file"), type="character", default="data/process/select_scfa_opf_kruskal_summary.csv", 
+  make_option(c("-o", "--output_file"), type="character", default="data/process/tables/select_scfa_opf_kruskal_summary.csv", 
               help="Output kruskal summary name [default= %default]", metavar="character"), 
-  make_option(c("-eo", "--extra_output"), type="character", default="data/process/select_scfa_opf_data.csv", 
+  make_option(c("-eo", "--extra_output"), type="character", default="data/process/tables/select_scfa_opf_data.csv", 
               help="Output kruskal summary name [default= %default]", metavar="character"))
 
 opt_parser = OptionParser(option_list=option_list)

### 2018-05-24 code/run_targeted_scfa_picrust.R

@@ -48,6 +48,16 @@ cross_sectional_data <- nzv_full_pi_data %>%
   select(-fit_result) %>% 
   gather("kegg_ortholog", "rel_abund", 4:length(colnames(.)))
 
+
+# Treatment data
+treatment_data <- nzv_full_pi_data %>% 
+  filter(sample_id %in% c(metaF$initial, metaF$followUp)) %>% 
+  left_join(picrust_meta, by = c("sample_id" = "Group")) %>% 
+  select(sample_id, Dx_Bin, dx, everything()) %>% 
+  select(-fit_result) %>% 
+  gather("kegg_ortholog", "rel_abund", 4:length(colnames(.)))
+
+
 # Pull specific KEGG IDs based on pathway and literature search
 # Target Butyrate, Propionate, and Acetate pathway genes 
 # Enzymes at the very end of the pathways
@@ -60,6 +70,13 @@ acetate_genes <- c("K18372", "K00467", "K00156", "K00925", "K01512", "K01067", "
 # Select the specific data from the cleaned data set
 selected_data <- cross_sectional_data %>% 
   filter(kegg_ortholog %in% c(butyrate_genes, propionate_genes, acetate_genes))
+
+treat_selected_data <- treatment_data %>% 
+  filter(kegg_ortholog %in% c(butyrate_genes, propionate_genes, acetate_genes)) %>% 
+  mutate(time_point = ifelse(sample_id %in% metaF$initial, invisible("initial"), invisible("follow"))) %>% 
+  group_by(kegg_ortholog) %>% 
+  slice(match(c(metaF$initial, metaF$followUp), sample_id)) %>% 
+  ungroup()
 
 # Run the Kruskal-wallis analysis
 kruskal_test <- selected_data %>% 
@@ -76,9 +93,28 @@ kruskal_test <- selected_data %>%
   mutate(bh = p.adjust(p.value, method = "BH")) %>% 
   arrange(p.value)
 
-# Write out the results
-write_csv(kruskal_test, "data/process/specific_scfa_kruskal_picrust_summary.csv")
-write_csv(selected_data, "data/process/selected_scfa_gene_data.csv")
 
+# Run the Kruskal-wallis analysis for treatment
+treat_wilcox_test <- treat_selected_data %>% 
+  group_by(kegg_ortholog, dx) %>% 
+  mutate(time_point = factor(time_point, 
+                     levels = c("initial", "follow"), 
[…]

### 2018-05-24 submission/manuscript.Rmd
final updates to first draft

@@
 # Load in data for select genes in SCFA pathways
-picrust_select <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv")
-opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")
+picrust_select <- read_csv("../data/process/tables/specific_scfa_kruskal_picrust_summary.csv")
+picrust_select_treat <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv")
+opf_select <- read_csv("../data/process/tables/select_scfa_opf_kruskal_summary.csv")
 
@@
 However, the data supporting the importance of SCFAs in humans is less certain.
-Here we test the hypothesis that short-chain fatty acid concentration decreases in individuals with colorectal tumors.
+Here we test the hypothesis that short-chain fatty acid concentration changes in individuals with colorectal tumors.
 
@@
 This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
+Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
 Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
 This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B].

### 2018-05-24 submission/manuscript.tex
final updates to first draft

@@
 the importance of SCFAs in humans is less certain. Here we test the
-hypothesis that short-chain fatty acid concentration decreases in
+hypothesis that short-chain fatty acid concentration changes in
 individuals with colorectal tumors.
@@
 highlighted by visualizing genes important in butyrate synthesis
-{[}Figure 3A{]}. Next, we took a subset of these 490 fecal samples
-(n=85) and performed metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was
-no difference in any of the genes involved in SCFA synthesis between
-individuals with normal colons (n=29) and patients with adenoma (n=28)
-or carcinoma (n=28) (P-value \textgreater{} 0.70) {[}Table S3{]}. This
-lack of difference is highlighted when we visualize the results for
-butyrate kinase {[}Figure 3B{]}. These observations provide evidence
-that the gene content does not change due to tumors. Although these
-results suggest that SCFA concentrations do not change due to tumor,
-there may be some errors present within our results that may account for
-the lack of associations.
+{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
+difference in imputed gene relative abundance between pre- and
+post-treatment samples for any genes involved with SCFA synthesis
+(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
+these 490 fecal samples (n=85) and performed metagenomic sequencing to
+confirm these results. Like the imputed gene results, metagenomic
+analysis found that there was no difference in any of the genes involved
+in SCFA synthesis between individuals with normal colons (n=29) and
+patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
+0.70) {[}Table S4{]}. This lack of difference is highlighted when we
+visualize the results for butyrate kinase {[}Figure 3B{]}. These
+observations provide evidence that the gene content does not change due
+to tumors. Although these results suggest that SCFA concentrations do
+not change due to tumor, there may be some errors present within our
+results that may account for the lack of associations.
 

### 2018-05-24 submission/supplemental.Rmd
final updates to first draft

@@
 
-**Table S3: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
+**Table S3: Breakdown of Wilcoxson rank-sum tests for pre- versus post-treatment of tumor using PICRUSt imputed relative abundance of select genes in SCFA pathways.**
 ```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
-table_s3 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
+
+table_s3 <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv") %>% 
+  select(kegg_ortholog, group, statistic, p.value, bh) %>% 
+   mutate_at(vars(p.value:bh), function(x) round(x, digits = 3))
+
+colnames(table_s3) <-  c("KEGG IDs", "Disease Group", "Test Statistic", "P-value", "BH")
+
+kable(table_s3, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S4: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+table_s4 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
   select(kegg_id, statistic, p.value, bh) %>% 
@@
 
-colnames(table_s3) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")
+colnames(table_s4) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")
 
-kable(table_s3, format = "markdown", align = 'c', padding = 2)
+kable(table_s4, format = "markdown", align = 'c', padding = 2)
 ```

### 2018-05-24 code/download_bacterial_genomes.py
re write download that works

@@ -43,92 +43,93 @@ def make_bacterial_genome_directory(directory_name):
 # Function to find all genomes in the bacteria directory
 def find_genomes(data_base_to_use, output_dir):
 
-	#initialize FTP server 
-	ftp_site = 'ftp.ncbi.nlm.nih.gov'
-	ftp = FTP(ftp_site)
-	# gets all bacteria that are contained in the current database
-	ftp.login()
-	ftp.cwd("genomes/%s/bacteria" % (data_base_to_use))
-	dirs = ftp.nlst()
-	# Open txt file to write to
-	temp_file = open("%sbacteria_in_database.txt" % (output_dir), 'w')
-	# Output message to stdout
-	print("Identifying all bacterial genomes in %s database..." % 
-		(data_base_to_use))
-	# Create the needed list of all bacteria present
-	for genome in dirs:
+	os.system("wget ftp://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/assembly_summary.txt -P %s" % 
+		(data_base_to_use, output_dir))
 
-		temp_file.write(genome+'\n')
+	temp_file = open("%sassembly_summary.txt" % (output_dir), 'r')
 
-	sys.exit()
+	temp_link_list = {}
+
+	x = 1
+
+	print("Creating dictionary of %s database to download..." % (data_base_to_use))
+
+	for line in temp_file:
+
+		temp_list = line.split('\t')
+
+		for value in temp_list:
+
+			if "ftp" in value and x > 2:
+
+				temp_link_list[temp_list[7]] = value
+
+		x += 1
+
+	temp_file.close()
+
+	# Write the file paths out to at txt file
+	temp_path_file = open("%scomplete_genomes.txt" % (output_dir), 'w')
+
+	print("Creating a key txt file...")
+
+	for bacterium in temp_link_list:
+
+		temp_path_file.write(bacterium+'\t'+temp_link_list[bacterium])
+
+	temp_path_file.close()
+
+
+	return(temp_link_list)
+
+
+
+
+
 
 
 # Function to run the actual download
-def download_from_ncbi(bacteria_needed, data_base_to_use, file_needed, output_dir):
-
-	#initialize FTP server 
-	ftp_site = 'ftp.ncbi.nlm.nih.gov'
-	ftp = FTP(ftp_site)
-
-	# set up system calls for downloads of specific file types
-	if file_needed == "fasta":
[…]

### 2018-05-24 code/download_bacterial_genomes.py
add comments

@@ -42,55 +42,51 @@ def make_bacterial_genome_directory(directory_name):
 
 # Function to find all genomes in the bacteria directory
 def find_genomes(data_base_to_use, output_dir):
-
+	# Download data on all genomes in database
 	os.system("wget ftp://ftp.ncbi.nlm.nih.gov/genomes/%s/bacteria/assembly_summary.txt -P %s" % 
 		(data_base_to_use, output_dir))
-
+	# Read in the downloaded file
 	temp_file = open("%sassembly_summary.txt" % (output_dir), 'r')
-
+	# Create an empty storage dictionary
 	temp_link_list = {}
-
+	# Create a counter
 	x = 1
 
 	print("Creating dictionary of %s database to download..." % (data_base_to_use))
-
+	# Parse the text file and populate the dictionary
 	for line in temp_file:
-
+		# separate the line by tab
 		temp_list = line.split('\t')
-
+		# iterate through each part of the list
 		for value in temp_list:
-
+			# Find values that are ftp and not in the first two lines
 			if "ftp" in value and x > 2:
-
+				# Add to dictionary key = bacterium, value = link
 				temp_link_list[temp_list[7]] = value
-
+		# Increase the counter
 		x += 1
-
+	# close the file
 	temp_file.close()
 
 	# Write the file paths out to at txt file
 	temp_path_file = open("%scomplete_genomes.txt" % (output_dir), 'w')
 
 	print("Creating a key txt file...")
-
+	# Create a master txt file of the key value pairs
 	for bacterium in temp_link_list:
 
 		temp_path_file.write(bacterium+'\t'+temp_link_list[bacterium])
-
+	# close the writing to file
 	temp_path_file.close()
 
-
+	# return the dictionary for later use
 	return(temp_link_list)
-
-
-
-
 
 
 
 # Function to run the actual download
 def download_from_ncbi(bacteria_needed, data_base_to_use, bacterial_list, output_dir):
-	
+		# Execute full download if default ALL is used
 		if bacteria_needed == "ALL":
 
 			# Execute the download
@@ -101,7 +97,7 @@ def download_from_ncbi(bacteria_needed, data_base_to_use, bacterial_list, output
 --accept='*genomic.fna.gz' %s/*genomic.fna.gz" % 
 			(output_dir, bacterial_list[link]))
 
-
+		# If it is not ALL try to download specific fasta
 		else:
 
 			try:
[…]

### 2018-06-04 submission/manuscript.Rmd
smooth out  abstract

@@
 
-**Background.** Certain bacterial community members have been shown to increase colon tumorigenesis while other taxa have been associated with normal colons.
-The some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids which have been shown to inhibit tumor growth in model systems.
-However, the data supporting the importance of SCFAs in humans is less certain.
-Here we test the hypothesis that short-chain fatty acid concentration changes in individuals with colorectal tumors.
+**Background.** Colorectal cancer (CRC) is a growing health concern with a large percentage of the risk of developing disease being due to environmental factors.
+Certain bacterial community members have been associated to increase CRC while other taxa have been associated with normal colons.
+Some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
+However, the data supporting the importance of SCFAs in human CRC is less certain.
+Here, we test the hypothesis that SCFA concentration changes in individuals with colorectal tumors.
 
@@
 
-**Results.** No difference in short-chain fatty acid concentration was found between normal individuals and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
-There was no difference in the ability of classification models with or without short-chain fatty acids added to them in predicting adenoma or carcinoma (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
-Using metagenomic sequencing, there was also no difference in genes involved with short-chain fatty acid synthesis (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
+**Results.** No difference in SCFA concentration was found between normal individuals and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
+There was no difference in classification models with or without SCFAs in their ability to predict adenoma or carcinoma (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
+Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
 
-**Conclusions.** Although our data does not support the hypothesis that short-chain fatty acids decrease in individuals that have colorectal tumors, there may be specific scenarios where short-chain fatty acids may still be beneficial for treatment of colorectal cancer.
-Alternatively, there may be other metabolites that have not been thoroughly investigated that are more important to human colorectal cancer.
+**Conclusions.** Although our data does not support the hypothesis that SCFAs decrease in individuals that have colorectal tumors, there may be specific scenarios where SCFAs may still be beneficial for treatment of CRC.
+Alternatively, there may be other metabolites that have not been thoroughly investigated that are more important to the development of human CRC.
 

### 2018-06-04 submission/manuscript.tex
smooth out  abstract

--- 2018-05-23 submission/manuscript.tex
@@
 
+\textbf{Background.} Colorectal cancer (CRC) is a growing health concern
+with a large percentage of the risk of developing disease being due to
+environmental factors. Certain bacterial community members have been
+associated to increase CRC while other taxa have been associated with
+normal colons. Some of the taxa associated with normal colons can use
+fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor
+growth in model systems. However, the data supporting the importance of
+SCFAs in human CRC is less certain. Here, we test the hypothesis that
+SCFA concentration changes in individuals with colorectal tumors.
+
+\textbf{Methods.} We analyzed a cross-sectional (n=490) and pre- and
+post-treatment (n=67) group for their concentrations of acetate,
+butyrate, and propionate. Analysis also included tumor classification
+models using Random Forest, imputed gene relative abundance with
+PICRUSt, and metagenomic sequencing on a subset (n=85) of the total
+cross-sectional group.
+
+\textbf{Results.} No difference in SCFA concentration was found between
+normal individuals and patients with adenoma or carcinoma (P-value
+\textgreater{} 0.15). There was no difference in classification models
+with or without SCFAs in their ability to predict adenoma or carcinoma
+(P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
+also no difference in genes involved with SCFA synthesis (P-value
+\textgreater{} 0.70).
+
+\textbf{Conclusions.} Although our data does not support the hypothesis
+that SCFAs decrease in individuals that have colorectal tumors, there
+may be specific scenarios where SCFAs may still be beneficial for
+treatment of CRC. Alternatively, there may be other metabolites that
+have not been thoroughly investigated that are more important to the
+development of human CRC.
+
 \newpage

### 2018-06-04 submission/manuscript.Rmd
smooth out the intro

@@
 Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
-The envrionment has been attributed as an important risk factor for CRC [@Lichtenstein2000].
+Although there is a genetic component to the disease, the envrionment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000].
 This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
-Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011].
-Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet [@Holscher2017].
+Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011].
+Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017].
 The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016].
-Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a viable treatment option.
+Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.
 
-Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment [@test_OKeefe2016].
-In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
+Specific SCFAs, like butyrate, have shown positive results for CRC treatment within model systems [@test_OKeefe2016].
+Butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
 Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects.
-Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
+For example, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
 Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.
@@
 There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
-The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons [@Ohigashi2013].
+The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals with normal colons [@Ohigashi2013].
 Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
 These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.
-Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC there is a need for more investigation into this area.
+Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.
 
@@
 Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors.
-Additionally, by increasing the adenoma tumor cohort it we will be able to identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases.
-It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors.
+Additionally, by increasing the adenoma tumor cohort it we will identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases.
+It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors (adenoma or carcinoma).
 This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.
@@
 To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups.
-The first group will be at a cross sectional point in time while the second group will be before (pre-) and after (post-)treatment.
-Additionally, we also will assess whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma, use PICRUSt and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyze how well 16S rRNA gene sequencing can predict SCFA concentrations.
-This extensive investigation will provide additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against tumorigenesis.
+The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
+Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations.
+This  investigation provides additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against colon tumorigenesis.
 

### 2018-06-04 submission/manuscript.tex
smooth out the intro

@@
 Colorectal cancer (CRC) is currently the third leading cancer-related
-cause of death within the US (1, 2). The envrionment has been attributed
-as an important risk factor for CRC (3). This has lead to the
+cause of death within the US (1, 2). Although there is a genetic
+component to the disease, the envrionment has been attributed as an
+important and large risk factor for CRC (3). This has lead to the
 investigation of how the microbiota may exacerbate or cause tumorigensis
@@
 reports in case/control studies on the microbiota and CRC have
-identified taxa commonly associated with the normal colon to be
+identified taxa commonly associated with the normal colons to be
 decreased in patients with carcinoma tumors (9--11). Many of these taxa
-actively produce short-chain fatty acids (SCFAs) from fiber that is part
-of our general diet (12). The most extensively studied of these SCFAs
-are acetate, butyrate, and propionate (13). Overall, CRC is a growing
-health problem and the microbiota through SCFAs have shown promise as a
-viable treatment option.
+actively produce short-chain fatty acids (SCFAs) from fiber that are a
+part of our general diet (12). The most extensively studied of these
+SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
+growing health problem and the microbiota through SCFAs have shown
+promise as a potential treatment option.
 
-Within model systems SCFAs, in particular butyrate, have shown positive
-results in CRC treatment (14). In particular, butyrate has been shown to
-inhibit cancer cell growth in \emph{in vitro} systems (15).
-Additionally, supplementation with food sources that bacteria use to
-create these SCFAs may also be able to confer beneficial effects. Fiber
+Specific SCFAs, like butyrate, have shown positive results for CRC
+treatment within model systems (14). Butyrate has been shown to inhibit
+cancer cell growth in \emph{in vitro} systems (15). Additionally,
+supplementation with food sources that bacteria use to create these
+SCFAs may also be able to confer beneficial effects. For example, fiber
 supplementation in mouse models of CRC caused an overall reduction in
@@
 concentrations to protect against CRC in human populations. The initial
-case/control studies that investigated SCFA concentrations in CRC have
-found that patients with carcinomas had lower concentrations of acetate,
-butyrate, and propionate versus patients with adenomas and individuals
-with normal colons (17). Although this would argue that increasing SCFA
-concentrations could be protective against tumorigenesis, fiber
-supplementation in randomized controlled trials have consistently failed
-to protect against tumor recurrence (18). These findings would argue
-against the utility of treatments that aim to use SCFAs to reduce or
-protect against tumorigenesis. Given the lack of clear evidence in human
-studies of the benefit of SCFAs in CRC there is a need for more
-investigation into this area.
+case/control studies that investigated SCFA concentrations in CRC found
+that patients with carcinomas had lower concentrations of acetate,
+butyrate, and propionate versus either patients with adenomas or
+individuals with normal colons (17). Although this would argue that
+increasing SCFA concentrations could be protective against
+tumorigenesis, fiber supplementation in randomized controlled trials
+have consistently failed to protect against tumor recurrence (18). These
+findings would argue against the utility of treatments that aim to use
+SCFAs to reduce or protect against tumorigenesis. Given the lack of
+clear evidence in human studies of the benefit of SCFAs in CRC, there is
+a need for more investigation into this area.
 
@@
 findings on SCFA concentrations in individuals with and without tumors.
-Additionally, by increasing the adenoma tumor cohort it we will be able
-to identify if the observation of a continuous gradual reduction in SCFA
+Additionally, by increasing the adenoma tumor cohort it we will identify
+if the observation of a continuous gradual reduction in SCFA
 concentrations occurs as tumor severity increases. It also will build
@@
 Operational Taxonomic Units (OTUs) as a potential risk stratification
-tool of colorectal tumors. This study will provide important information
-on the replicability of previous findings in humans by extensively
-studying how SCFAs could be associated with colorectal tumors.
+tool of colorectal tumors (adenoma or carcinoma). This study will
+provide important information on the replicability of previous findings
+in humans by extensively studying how SCFAs could be associated with
+colorectal tumors.
 
@@
 acetate, butyrate, and propionate within fecal samples for two different
-groups. The first group will be at a cross sectional point in time while
-the second group will be before (pre-) and after (post-)treatment.
-Additionally, we also will assess whether adding SCFA concentrations to
-OTU data improves the classification of adenoma or carcinoma, use
-PICRUSt and metagenomic sequencing to assess the genes involved in SCFA
-synthesis, and analyze how well 16S rRNA gene sequencing can predict
-SCFA concentrations. This extensive investigation will provide
-additional information as to whether the taxa that are decreased in
-carcionma use SCFAs to help protect against tumorigenesis.
+groups. The first group had a sample obtained at a single cross
+sectional point in time while the second group had samples obtained
+before (pre-) and after (post-)treatment for colorectal tumors.
+Additionally, we also assessed whether adding SCFA concentrations to OTU
+data improves the classification of adenoma or carcinoma using the
+Random Forest algorithm (19), used PICRUSt (20) and metagenomic
+sequencing to assess the genes involved in SCFA synthesis, and analyzed
+how well 16S rRNA gene sequencing can predict SCFA concentrations. This
+investigation provides additional information as to whether the taxa
+that are decreased in carcionma use SCFAs to help protect against colon
+tumorigenesis.
 
@@
 fiber supplementation to prevent tumor recurrence have found no benefit
-(18, 19). Our results align with what has been reported in
+(18, 21). Our results align with what has been reported in
 randomized-controlled trials that SCFAs do not protect against all types
@@
 However, there are known differences in the types of mutations that
-occur (20) and treating all tumors as equal may actually hide any
+occur (22) and treating all tumors as equal may actually hide any
 benefit that could be found in specific subsets of individuals. Similar
-to the idea of specific immunotherapy for specific tumors (21), SCFAs
+to the idea of specific immunotherapy for specific tumors (23), SCFAs
 may have beneficial effects for specific types of colorectal cancers.
@@
 production are consistently higher in normal colons than those with
-carcinomas (7, 8, 22).
+carcinomas (7, 8, 24).
 
@@
 individuals with normal colons versus patients with carcinomas (7, 8,
-22). Protection could be via a different pathway and other metabolites
+24). Protection could be via a different pathway and other metabolites
 that have not been extensively studied. Alternatively, protection may
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (23, 24). In brief, this study used fecal
+described in detail previously (25, 26). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (23).
+removal of a tumor with or without chemotherapy and radiation (25).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (23, 24). The University
+histopathological review of biopsies obtained (25, 26). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (25). The following changes to this protocol included the use
+protocol (27). The following changes to this protocol included the use
 of frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 median concentrations of these SCFAs between the two studies (see Table
-1 (25) and Figure 1 in this report).
+1 (27) and Figure 1 in this report).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (23, 26, 27). The major differences from these
+described previously (25, 28, 29). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (28).
+97\% similarity used the OptClust algorithm (30).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (29) was used.
+1.1.2 with the recommended standard operating protocol (20) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (30). Briefly, the sequences were quality
+previously described method (31). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (31). Open Reading Frames (ORFs) were
-identified using Prodigal (32), counts generated using Diamond (33),
+to contig assembly with MEGAHIT (32). Open Reading Frames (ORFs) were
+identified using Prodigal (33), counts generated using Diamond (34),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (34), and OPF alignment used the KEGG database (35).
+mmseq2 (35), and OPF alignment used the KEGG database (36).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (35,
-36) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (36,
+37) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 Random Forest algorithm for both classification and regression models
-(37). This was repeated on 100 differen 80/20 splits of the data to
+(19). This was repeated on 100 differen 80/20 splits of the data to
 generate a reasonable range for the AUC of the model. The reported AUCs,
@@
 
+\hypertarget{ref-randomforest_citation_2002}{}
+19. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
+
+\hypertarget{ref-Langille2013}{}
+20. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
+\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
+\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
+Predictive functional profiling of microbial communities using 16S rRNA
+marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
+doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
+
 \hypertarget{ref-Schatzkin2000}{}
-19. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+21. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
 \textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
@@
 \hypertarget{ref-Fearon1990}{}
-20. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+22. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \hypertarget{ref-Thomas2016}{}
-21. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+23. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
 agents targeting the BCR-ABL tyrosine kinase as philadelphia
@@
 \hypertarget{ref-normalization_Sze2017}{}
-22. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+24. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-normalization_sze2017}{}
-23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+25. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-24. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+26. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-25. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+27. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+29. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-28. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 
-\hypertarget{ref-Langille2013}{}
-29. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
-\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
-\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
-\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
-Predictive functional profiling of microbial communities using 16S rRNA
-marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
-doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
-
 \hypertarget{ref-Hannigan2017}{}
-30. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+31. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-31. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+32. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-32. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+33. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-33. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+34. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-34. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+35. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-35. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+36. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-36. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+37. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
-
-\hypertarget{ref-randomforest_citation_2002}{}
-37. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News \textbf{2}:18--22.
 

### 2018-06-04 submission/manuscript.Rmd
update title

@@
 ---
-title: "Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer"
+title: "Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors"
 bibliography: references.bib
@@
 
-Running title: SCFAs and Colorectal Cancer
+Running title: SCFAs and colorectal tumors
 

### 2018-06-04 submission/manuscript.tex
update title

@@
 \hypersetup{unicode=true,
-            pdftitle={Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer},
+            pdftitle={Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors},
             pdfborder={0 0 0},
@@
   \title{Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal
-Cancer}
+Tumors}
   \pretitle{\vspace{\droptitle}\centering\huge}

### 2018-06-04 submission/manuscript.Rmd
smooth over the results

@@
 
-**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time.
+**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time.
 There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
@@
 
-**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations.
-For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
-Although including SCFA concentrations did not add extra information for classification of disease using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.
+**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models with OTUs only or OTUs and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
+For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
+Although including SCFA concentrations did not add extra information for classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of microbes are responsible for the observed SCFA concentrations in patients with carcinoma.
 
@@
 Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
-Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results.
+Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
 Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
 This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B].
-These observations provide evidence that the gene content does not change due to tumors.
-Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.
+These observations provide evidence that the gene content does not change due to colorectal tumors.
+Although these results suggest that SCFA concentrations do not change due to colorectal tumor, there may be some errors present within our results that may account for the lack of associations.
 
@@
 Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations.
-Overall, our results do not support the hypothesis that lower SCFA concentrations are associated with colorectal cancer.
+Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal cancer.
 

### 2018-06-04 submission/manuscript.tex
smooth over the results

@@
 \textbf{Decreased SCFA concentrations are not associated with tumors.}
-We used frozen fecal samples from 490 individuals and HPLC to measure
-acetate, butyrate, and propionate concentrations at a cross-sectional
-point in time. There was no difference between individuals with normal
-colons (n=172) and patients with either an adenoma (n=198) or carcinoma
-(n=120) for any of the SCFAs measured after multiple comparison
-correction (P-value \textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next
-measured the concentration of SCFAs in 67 patients with an adenoma
-(n=41) or carcinoma (n=26) in which we had pre- and post-treatment fecal
-samples. Although there was a general trend for increasing acetate,
-butyrate, and propionate concentrations after treatment for tumors,
-there was no significant difference pre- and post-treatment for patients
-with adenoma (P-value \textgreater{} 0.20) or carcinoma (P-value
-\textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even though there was no
-change in SCFA concentrations between individuals with normal colons and
-those with tumors, this information could still be important to help
-classify disease.
+We used high-performance liquid chromatography (HPLC) to measure
+acetate, butyrate, and propionate concentrations of frozen fecal samples
+from 490 individuals at a cross-sectional point in time. There was no
+difference between individuals with normal colons (n=172) and patients
+with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
+measured after multiple comparison correction (P-value \textgreater{}
+0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
+in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
+had pre- and post-treatment fecal samples. Although there was a general
+trend for increasing acetate, butyrate, and propionate concentrations
+after treatment for tumors, there was no significant difference pre- and
+post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
+or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even
+though there was no change in SCFA concentrations between individuals
+with normal colons and those with tumors, this information could still
+be important to help classify disease.
 
 \textbf{Random Forest models with SCFA concentrations do not classify
-tumors better.} Using the Random Forest algorithm we built models to
-classify normal versus adenoma and normal versus carcinoma with OTUs or
-OTUs and SCFA concentrations. For both adenoma and carcinoma models,
-there was no difference between the median AUC of models with or without
-SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
-Although including SCFA concentrations did not add extra information for
-classification of disease using Random Forest models, it is possible
-that the genes for enzymes involved in SCFA synthesis may change due to
-disease status; where a smaller number of microbes are responsible for
-the observed SCFA concentrations.
+tumors better.} Using the Random Forest algorithm we built models with
+OTUs only or OTUs and SCFA concentrations to classify normal versus
+adenoma and normal versus carcinoma fecal samples. For adenoma and
+carcinoma models, there was no difference between the median AUC of
+models with or without SCFA concentrations (P-value \textgreater{} 0.05)
+{[}Figure 2{]}. Although including SCFA concentrations did not add extra
+information for classification of colorectal tumors using Random Forest
+models, it is possible that the genes for enzymes involved in SCFA
+synthesis may vary based on the type of colorectal tumor; where a
+smaller number of microbes are responsible for the observed SCFA
+concentrations in patients with carcinoma.
 
@@
 (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
-these 490 fecal samples (n=85) and performed metagenomic sequencing to
+these 490 fecal samples (n=85) and used metagenomic sequencing to
 confirm these results. Like the imputed gene results, metagenomic
@@
 observations provide evidence that the gene content does not change due
-to tumors. Although these results suggest that SCFA concentrations do
-not change due to tumor, there may be some errors present within our
-results that may account for the lack of associations.
+to colorectal tumors. Although these results suggest that SCFA
+concentrations do not change due to colorectal tumor, there may be some
+errors present within our results that may account for the lack of
+associations.
 

### 2018-06-04 submission/manuscript.Rmd
smooth over discussion

@@
 The observations from this study do not support the hypothesis that SCFA concentrations are lower in individuals with tumors.
-Whether we directly measured the SCFA concentration or investigate genes associated with their production no difference could be identified [Figure 1 & 3].
+Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3].
 Further, no benefit could be found in using the concentrations to help classify individuals with and without tumors [Figure 2].
@@
 
-Much of the previous literature on SCFA benefit to human colorectal cancer has been illustrated in model systems [@test_OKeefe2016].
-Many SCFAs are produced through the breakdown of fiber [@Holscher2017].
-A recent study in mice found that fiber supplementation increased SCFA concentration and decreased tumor burden [@Bishehsari2018].
-Additionally, SCFAs such as butyrate have been found to inhibit tumor growth in *in vitro* experiments [@Encarnao2018].
+Much of the previous research on SCFA benefit to human CRC has been illustrated in model systems [@test_OKeefe2016].
+Many SCFAs are produced through the breakdown of fiber [@Holscher2017] and a recent study in mice found that fiber supplementation increased SCFA concentrations and decreased tumor burden [@Bishehsari2018].
+Additionally, SCFAs such as butyrate can inhibit tumor growth in *in vitro* experiments [@Encarnao2018].
 Yet, observations in humans has been mixed.
-Previous case/control studies have suggested that SCFAs are lower in individuals with carcinoma tumors [@Ohigashi2013].
-However, individual studies and meta-analysis on fiber supplementation to prevent tumor recurrence have found no benefit [@Schatzkin2000; @Yao2017].
-Our results align with what has been reported in randomized-controlled trials that SCFAs do not protect against all types of colorectal tumors.
+Previous case/control studies found associations with lower SCFA concentrations in individuals with carcinoma tumors [@Ohigashi2013].
+However, individual randomized controlled trials and a recent meta-analysis on fiber supplementation to prevent tumor recurrence has found no benefit [@Schatzkin2000; @Yao2017].
+Our results align with what has been reported in randomized-controlled trials, that SCFAs do not provide general protection against colorectal tumors.
 It is possible though that there are specific instances where SCFAs may be beneficial.
 
-One limitation of ours and others research into the effect of SCFAs in colorectal cancer has been that all tumors are treated as the same type.
+One limitation of ours and others research into the effect of SCFAs in CRC has been that all tumors are treated as the same type.
 However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
-Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal cancers.
+Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors.
 Future research will need to test if this is a valid hypothesis.
@@
 The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Protection could be via a different pathway and other metabolites that have not been extensively studied.
-Alternatively, protection may not occur via a metabolite but instead occur through niche exclusion of mouth-associated microbes (e.g.
+Protection could be via a different pathway and by extension other metabolites that have not been extensively studied.
+Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g.
 *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016]).
-The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection with chronic inflammation replacing the role of antibiotics.
-Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these observations.
+The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics.
+Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results
 

### 2018-06-04 submission/manuscript.tex
smooth over discussion

@@
 with these taxa are the most important to models that can classify high
-and low SCFA concentrations. Overall, our results do not support the
-hypothesis that lower SCFA concentrations are associated with colorectal
-cancer.
+and low SCFA concentrations. Overall, our results are robust and do not
+support the hypothesis that differences in SCFA concentrations are
+associated with colorectal cancer.
 
@@
 concentrations are lower in individuals with tumors. Whether we directly
-measured the SCFA concentration or investigate genes associated with
-their production no difference could be identified {[}Figure 1 \& 3{]}.
+measured the SCFA concentration or investigated genes associated with
+their production, no difference could be identified {[}Figure 1 \& 3{]}.
 Further, no benefit could be found in using the concentrations to help
@@
 
-Much of the previous literature on SCFA benefit to human colorectal
-cancer has been illustrated in model systems (14). Many SCFAs are
-produced through the breakdown of fiber (12). A recent study in mice
-found that fiber supplementation increased SCFA concentration and
-decreased tumor burden (16). Additionally, SCFAs such as butyrate have
-been found to inhibit tumor growth in \emph{in vitro} experiments (15).
-Yet, observations in humans has been mixed. Previous case/control
-studies have suggested that SCFAs are lower in individuals with
-carcinoma tumors (17). However, individual studies and meta-analysis on
-fiber supplementation to prevent tumor recurrence have found no benefit
+Much of the previous research on SCFA benefit to human CRC has been
+illustrated in model systems (14). Many SCFAs are produced through the
+breakdown of fiber (12) and a recent study in mice found that fiber
+supplementation increased SCFA concentrations and decreased tumor burden
+(16). Additionally, SCFAs such as butyrate can inhibit tumor growth in
+\emph{in vitro} experiments (15). Yet, observations in humans has been
+mixed. Previous case/control studies found associations with lower SCFA
+concentrations in individuals with carcinoma tumors (17). However,
+individual randomized controlled trials and a recent meta-analysis on
+fiber supplementation to prevent tumor recurrence has found no benefit
 (18, 21). Our results align with what has been reported in
-randomized-controlled trials that SCFAs do not protect against all types
-of colorectal tumors. It is possible though that there are specific
-instances where SCFAs may be beneficial.
+randomized-controlled trials, that SCFAs do not provide general
+protection against colorectal tumors. It is possible though that there
+are specific instances where SCFAs may be beneficial.
 
 One limitation of ours and others research into the effect of SCFAs in
-colorectal cancer has been that all tumors are treated as the same type.
-However, there are known differences in the types of mutations that
-occur (22) and treating all tumors as equal may actually hide any
-benefit that could be found in specific subsets of individuals. Similar
-to the idea of specific immunotherapy for specific tumors (23), SCFAs
-may have beneficial effects for specific types of colorectal cancers.
-Future research will need to test if this is a valid hypothesis.
-Although there are limitations with the current research on SCFAs and
-colorectal tumors, evidence is building that these specific metabolites
-may not be protective. Yet, taxa that are associated with SCFA
-production are consistently higher in normal colons than those with
-carcinomas (7, 8, 24).
+CRC has been that all tumors are treated as the same type. However,
+there are known differences in the types of mutations that occur (22)
+and treating all tumors as equal may actually hide any benefit that
+could be found in specific subsets of individuals. Similar to the idea
+of specific immunotherapy for specific tumors (23), SCFAs may have
+beneficial effects for specific types of colorectal tumors. Future
+research will need to test if this is a valid hypothesis. Although there
+are limitations with the current research on SCFAs and colorectal
+tumors, evidence is building that these specific metabolites may not be
+protective. Yet, taxa that are associated with SCFA production are
+consistently higher in normal colons than those with carcinomas (7, 8,
+24).
 
@@
 individuals with normal colons versus patients with carcinomas (7, 8,
-24). Protection could be via a different pathway and other metabolites
-that have not been extensively studied. Alternatively, protection may
-not occur via a metabolite but instead occur through niche exclusion of
-mouth-associated microbes (e.g. \emph{Fusobacterium},
+24). Protection could be via a different pathway and by extension other
+metabolites that have not been extensively studied. Alternatively,
+protection may not occur via a metabolite but instead through niche
+exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
 \emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
 10)). The idea of niche exclusion is similar to how the community
-protects against \emph{Clostridium difficile} infection with chronic
-inflammation replacing the role of antibiotics. Although we did not find
-lower concentrations of SCFAs associated with colorectal tumors, we
-think that there are many exciting new avenues to explore because of
-these observations.
+protects against \emph{Clostridium difficile} infection (25) with
+chronic inflammation replacing the role of antibiotics. Although we did
+not find lower concentrations of SCFAs associated with colorectal
+tumors, we think that there are many exciting new avenues to explore
+because of these results
 
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (25, 26). In brief, this study used fecal
+described in detail previously (26, 27). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (25).
+removal of a tumor with or without chemotherapy and radiation (26).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (25, 26). The University
+histopathological review of biopsies obtained (26, 27). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (27). The following changes to this protocol included the use
+protocol (28). The following changes to this protocol included the use
 of frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 median concentrations of these SCFAs between the two studies (see Table
-1 (27) and Figure 1 in this report).
+1 (28) and Figure 1 in this report).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (25, 28, 29). The major differences from these
+described previously (26, 29, 30). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (30).
+97\% similarity used the OptClust algorithm (31).
 
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (31). Briefly, the sequences were quality
+previously described method (32). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (32). Open Reading Frames (ORFs) were
-identified using Prodigal (33), counts generated using Diamond (34),
+to contig assembly with MEGAHIT (33). Open Reading Frames (ORFs) were
+identified using Prodigal (34), counts generated using Diamond (35),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (35), and OPF alignment used the KEGG database (36).
+mmseq2 (36), and OPF alignment used the KEGG database (37).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (36,
-37) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (37,
+38) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (38). Generally, differences between the
+the statistical language R (39). Generally, differences between the
 different disease groups used a Kruskal-Walis rank sum test with a
 Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (39). First, we assessed differences in
+correct for multiple comparisons (40). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 
+\hypertarget{ref-Theriot2015}{}
+25. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
+the gastrointestinal microbiome and clostridium difficile. Annual Review
+of Microbiology \textbf{69}:445--461.
+doi:\href{https://doi.org/10.1146/annurev-micro-091014-104115}{10.1146/annurev-micro-091014-104115}.
+
 \hypertarget{ref-normalization_sze2017}{}
-25. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+26. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-26. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+27. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-27. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+28. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+29. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-29. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+30. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+31. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Hannigan2017}{}
-31. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+32. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-32. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+33. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-33. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+34. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-34. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+35. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-35. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+36. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-36. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+37. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-37. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+38. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-r_citation_2017}{}
-38. \textbf{R Core Team}. 2017. R: A language and environment for
+39. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-39. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+40. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-06-04 submission/manuscript.Rmd
smooth over the methods

@@
 **Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016].
-In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment for their tumor (n=`r length(metaf$initial)`).
+In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`).
 For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
@@
 
-**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol [@scfa_measures_venkataraman2016].
+**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016].
 The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations.
-These changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 in this report).
+These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).
 
-**16s rRNA gene sequencing.** The workflow and processing have been described previously [@Schloss2009; @Kozich2013; @normalization_sze2017].
-The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity used the OptClust algorithm [@opticlust_westcott2017].
+**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017].
+The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].
 
 **Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used.
-Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the normalize_by_copy_number.py function, and subsequent imputed metagenomes created using the predict_metagenomes.py function.
+Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.
 
-**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
+**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()`, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
 Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015].
@@
 
-**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002].
-This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model.
+**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification models [@randomforest_citation_2002].
+This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model.
 The reported AUCs, unless otherwise specified, are for the test sets.
-Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations.
+The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.
 
 **Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017].
-Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test.
+Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used.
 Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995].
@@
 Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models.
-Next, the imputed gene counts of important mediators of SCFA creation were tested.
-Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
-Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using the Random Forest algorithm.
+Next, the imputed gene counts of important mediators of SCFA synthesis was tested.
+Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
+Finally, models to classify high or low SCFA concentration based on the median of each SCFA and 16S rRNA gene sequencing data was created using the Random Forest algorithm.
 

### 2018-06-04 submission/manuscript.tex
smooth over the methods

@@
 samples obtained at either a single cross-sectional time point (n=490)
-or from before (pre-) and after (post-) treatment for their tumor
+or from before (pre-) and after (post-) treatment of a patient's tumor
 (n=67). For patients undergoing treatment for their tumor the length of
@@
 acetate, butyrate, and propionate followed a previously published
-protocol (28). The following changes to this protocol included the use
-of frozen fecal samples suspended in 1ml of PBS instead of fecal
+protocol that used a High-Performance Liquid Chromatography (HPLC)
+machine (28). The following changes to this protocol included the use of
+frozen fecal samples suspended in 1ml of PBS instead of fecal
 suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
 weight of fecal samples instead of the average weight for SCFA
-concentration normalizations. These changes did not affect the overall
-median concentrations of these SCFAs between the two studies (see Table
-1 (28) and Figure 1 in this report).
+concentration normalizations. These methodological changes did not
+affect the overall median concentrations of these SCFAs between the two
+studies (see Table 1 (28) and Figure 1 here).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-described previously (26, 29, 30). The major differences from these
+previously described (26, 29, 30). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity used the OptClust algorithm (31).
+97\% similarity using the OptClust algorithm (31).
 
@@
 formated table using the biom convert function, the subsequent biom file
-was processed with the normalize\_by\_copy\_number.py function, and
-subsequent imputed metagenomes created using the predict\_metagenomes.py
-function.
+was processed with the `normalize\_by\_copy\_number.py' function, and
+subsequent imputed metagenomes created using the
+`predict\_metagenomes.py' function.
 
-\textbf{Obtaining OPFs from metagenomes.} A subset of the
-cross-sectional group (n=490) containing a total of 85 individuals
-(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
+\textbf{Obtaining Operational Protein Families from metagenomes.} A
+subset of the cross-sectional group (n=490) containing a total of 85
+individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
@@
 
-\textbf{Random Forest Models.} The model was first trained on 80\% of
+\textbf{Random Forest models.} The model was first trained on 80\% of
 the data and then tested on the held out 20\% (80/20 split) using the
-Random Forest algorithm for both classification and regression models
-(19). This was repeated on 100 differen 80/20 splits of the data to
-generate a reasonable range for the AUC of the model. The reported AUCs,
-unless otherwise specified, are for the test sets. Classification models
-were built to group normal versus adenoma and normal versus carcinoma or
+Random Forest algorithm for classification models (19). This was
+repeated on 100 different 80/20 splits of the data to generate a
+reasonable range for the AUC of the model. The reported AUCs, unless
+otherwise specified, are for the test sets. The classification models
+were built to group normal versus adenoma, normal versus carcinoma, and
 high versus low SCFA concentrations.
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (39). Generally, differences between the
-different disease groups used a Kruskal-Walis rank sum test with a
-Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
+the statistical language R (39). Generally, a Kruskal-Walis rank sum
+test with a Dunn's post-hoc test was used to assess differences between
+the groups used. Where appropriate Benjamini-Hochberg was used to
 correct for multiple comparisons (40). First, we assessed differences in
@@
 and without tumor using Random Forest models. Next, the imputed gene
-counts of important mediators of SCFA creation were tested. Finally, the
-counts generated for OPFs that matched important genes involved with
-SCFA creation were analyzed. Finally, models to classify high or low
-SCFA concentration based on 16S rRNA gene sequencing data were created
-using the Random Forest algorithm.
+counts of important mediators of SCFA synthesis was tested.
+Additionally, the counts generated for OPFs that matched important genes
+involved with SCFA creation were analyzed. Finally, models to classify
+high or low SCFA concentration based on the median of each SCFA and 16S
+rRNA gene sequencing data was created using the Random Forest algorithm.
 

### 2018-06-04 submission/manuscript.Rmd
update figure legends

@@
 **Figure 1.
-Using HPLC no change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A).
+No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A).
 Butyrate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (B).
 Propionate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (C).
-The black links indicate the median SCFA concentration.
+The black lines indicate the median SCFA concentration.
 Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D).
@@
 **Figure 2.
-SCFAs do not improve OTU-based Random Forest models.** Difference between the area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
-Difference between the area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (B).
-The black linke represents the median AUC.
+SCFAs do not improve OTU-based Random Forest models.** The area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
+The area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (B).
+The black line represents the median AUC.
 The dotted line highlights an AUC of 0.5.
@@
 Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B).
-The other genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
+The other butyrate pathway genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
 
@@
 **Figure S1.
-OTU-based Random Forest models of SCFA concentrations.** Classification Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration (A).
+OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration using classification Random Forest (A).
 The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).
-Regression Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on correlation to actual SCFA concentration (C).
-The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (D).

### 2018-06-04 submission/manuscript.tex
update figure legends

@@
 
-\textbf{Figure 1. Using HPLC no change in SCFA measurements was observed
-between normal, adenoma, and carcinoma individuals.} Acetate
+\textbf{Figure 1. No change in SCFA measurements was observed between
+normal, adenoma, and carcinoma individuals using HPLC.} Acetate
 concentrations in fecal samples of individuals with normal colons,
@@
 Propionate concentrations in fecal samples of individuals with normal
-colons, adenomas, and carcinomas (C). The black links indicate the
+colons, adenomas, and carcinomas (C). The black lines indicate the
 median SCFA concentration. Acetate concentrations in fecal samples
@@
 \textbf{Figure 2. SCFAs do not improve OTU-based Random Forest models.}
-Difference between the area under the curve of 100 different 80/20 split
-OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
-(A). Difference between the area under the curve of 100 different 80/20
-OTU-based normal versus carcinoma 10-fold CV models with and without
-SCFAs (B). The black linke represents the median AUC. The dotted line
-highlights an AUC of 0.5.
+The area under the curve of 100 different 80/20 split OTU-based normal
+versus adenoma 10-fold CV models with and without SCFAs (A). The area
+under the curve of 100 different 80/20 OTU-based normal versus carcinoma
+10-fold CV models with and without SCFAs (B). The black line represents
+the median AUC. The dotted line highlights an AUC of 0.5.
 
@@
 (A). Counts per million (corrected for size and number of contigs in an
-OPF) for the Butyrate Kinase gene (B). The other genes from the PICRUSt
-analysis did not align to any of the OPFs in the metagenome analysis.
+OPF) for the Butyrate Kinase gene (B). The other butyrate pathway genes
+from the PICRUSt analysis did not align to any of the OPFs in the
+metagenome analysis.
 
@@
 \textbf{Figure S1. OTU-based Random Forest models of SCFA
-concentrations.} Classification Random Forest train and tests of 100
-different 80/20 OTU-based models with 10-fold CV based on higher or
-lower than the medain SCFA concentration (A). The top 10 OTUs based on
-mean decrease in accuracy (MDA) for each model, colored by their lowest
-taxonomic identification (B). Regression Random Forest train and tests
-of 100 different 80/20 OTU-based models with 10-fold CV based on
-correlation to actual SCFA concentration (C). The top 10 OTUs based on
-mean decrease in accuracy (MDA) for each model, colored by their lowest
-taxonomic identification (D).
+concentrations.} The train and test results of 100 different 80/20
+OTU-based models with 10-fold CV based on higher or lower than the
+medain SCFA concentration using classification Random Forest (A). The
+top 10 OTUs based on mean decrease in accuracy (MDA) for each model,
+colored by their lowest taxonomic identification (B).
 

### 2018-06-04 code/make_SCFA_classification_graph.R
change names of figures

@@ -113,79 +113,79 @@ imp_otus_classification <- combined_imp_class_summary %>%
 
 
 
-regression_model_data <- read_csv("data/process/tables/acetate_regression_RF_summary.csv") %>% 
-  mutate(scfa = "acetate") %>% 
-  bind_rows(
-    read_csv("data/process/tables/butyrate_regression_RF_summary.csv") %>% 
-      mutate(scfa = "butyrate"), 
-    read_csv("data/process/tables/propionate_regression_RF_summary.csv") %>% 
-      mutate(scfa = "propionate"))
-
-
-regression_summary <- regression_model_data %>% 
-  group_by(scfa) %>% 
-  summarise(median_train_r2 = median(train_r2), 
-            max_train_r2 = max(train_r2), 
-            min_train_r2 = min(train_r2), 
-            median_test_r2 = median(test_r2), 
-            max_test_r2 = max(test_r2), 
-            min_test_r2 = min(test_r2), 
-            median_sens = median(RMSE), max_sens = max(RMSE), min_sens = min(RMSE), 
-            median_spec = median(MAE), max_spec = max(MAE), min_spec = min(MAE)) %>% 
-  gather("groupings", "r2", median_train_r2:min_test_r2)
+# regression_model_data <- read_csv("data/process/tables/acetate_regression_RF_summary.csv") %>% 
+#   mutate(scfa = "acetate") %>% 
+#   bind_rows(
+#     read_csv("data/process/tables/butyrate_regression_RF_summary.csv") %>% 
+#       mutate(scfa = "butyrate"), 
+#     read_csv("data/process/tables/propionate_regression_RF_summary.csv") %>% 
+#       mutate(scfa = "propionate"))
+# 
+# 
+# regression_summary <- regression_model_data %>% 
+#   group_by(scfa) %>% 
+#   summarise(median_train_r2 = median(train_r2), 
+#             max_train_r2 = max(train_r2), 
+#             min_train_r2 = min(train_r2), 
+#             median_test_r2 = median(test_r2), 
+#             max_test_r2 = max(test_r2), 
+#             min_test_r2 = min(test_r2), 
+#             median_sens = median(RMSE), max_sens = max(RMSE), min_sens = min(RMSE), 
+#             median_spec = median(MAE), max_spec = max(MAE), min_spec = min(MAE)) %>% 
+#   gather("groupings", "r2", median_train_r2:min_test_r2)
 
[…]

### 2018-06-04 code/make_pi_opf_combined_graph.R

@@ -77,5 +77,5 @@ metagenome_fig <- opf_data %>%
 gene_plot <- grid.arrange(imputed_fig, metagenome_fig, 
                           layout_matrix = rbind(c(1, 1, 2)))
 # Write out to specific directory
-ggsave("results/figures/butyrate_specific_genes_graph.pdf", gene_plot, width = 9, height = 5)
+ggsave("results/figures/Figure3.pdf", gene_plot, width = 9, height = 5)
 

### 2018-06-04 code/make_rf_auc_graphs.R

@@ -75,4 +75,4 @@ carcinoma <- crc_summary %>%
 combined_plot <- grid.arrange(adenoma, carcinoma, ncol = 2, nrow = 1)
 
 # Write out to specific directory
-ggsave("results/figures/rf_otu_and_scfa_aucs_graph.pdf", combined_plot, width = 6, height = 6)
+ggsave("results/figures/Figure2.pdf", combined_plot, width = 6, height = 6)

### 2018-06-04 code/make_scfa_measures_hplc_figure.R

@@ -232,7 +232,7 @@ scfa_plot <- grid.arrange(cross_figs[["acetate"]], cross_figs[["butyrate"]], cro
                                 layout_matrix = rbind(c(1, 2, 3), c(4, 5, 6)))
 
 # Write out to specific directory
-ggsave("results/figures/full_scfa_graph.pdf", scfa_plot, width = 9.5, height = 9)
+ggsave("results/figures/Figure1.pdf", scfa_plot, width = 9.5, height = 9)
 
 
 

### 2018-06-05 code/download_bacterial_genomes.py
add new line character

@@ -75,7 +75,7 @@ def find_genomes(data_base_to_use, output_dir):
 	# Create a master txt file of the key value pairs
 	for bacterium in temp_link_list:
 
-		temp_path_file.write(bacterium+'\t'+temp_link_list[bacterium])
+		temp_path_file.write(bacterium+'\t'+temp_link_list[bacterium]+'\n')
 	# close the writing to file
 	temp_path_file.close()
 

### 2018-06-14 submission/manuscript.Rmd
spell check and add nick changes to abstract

@@
 
-**Background.** Colorectal cancer (CRC) is a growing health concern with a large percentage of the risk of developing disease being due to environmental factors.
-Certain bacterial community members have been associated to increase CRC while other taxa have been associated with normal colons.
-Some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
+**Background.** Colorectal cancer (CRC) is a growing health concern with the majority of the risk for developing disease being due to environmental factors.
+The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC, while other taxa are associated to colons without tumors.
+Some of the taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
 However, the data supporting the importance of SCFAs in human CRC is less certain.
-Here, we test the hypothesis that SCFA concentration changes in individuals with colorectal tumors.
+Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.
 
-**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate.
+**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate.
 Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.
 
-**Results.** No difference in SCFA concentration was found between normal individuals and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
-There was no difference in classification models with or without SCFAs in their ability to predict adenoma or carcinoma (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
-Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
+**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
+There was no difference in classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
+Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
 
-**Conclusions.** Although our data does not support the hypothesis that SCFAs decrease in individuals that have colorectal tumors, there may be specific scenarios where SCFAs may still be beneficial for treatment of CRC.
-Alternatively, there may be other metabolites that have not been thoroughly investigated that are more important to the development of human CRC.
+**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC.
+Alternatively, there may be other mechanisms that have not been thoroughly investigated that are more important to the development of human CRC.
 
@@
 Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
-Although there is a genetic component to the disease, the envrionment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000].
+Although there is a genetic component to the disease, the environment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000].
 This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
@@
 Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations.
-This  investigation provides additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against colon tumorigenesis.
+This  investigation provides additional information as to whether the taxa that are decreased in carcinoma use SCFAs to help protect against colon tumorigenesis.
 
@@
 **Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas.
-First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
+First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
 This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
-Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+Using a Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
 Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
@@
 Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A].
-However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A].
+However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure S1A].
 The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B].
@@
 Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016].
-The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.
+The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.
 
 **Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016].
-The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations.
+The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations.
 These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).
@@
 **Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used.
-Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.
+Briefly, the mothur shared file and metadata was converted into a biom formatted table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.
 
@@
 The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chian fatty acid analysis on the High-Performace Liquid Chromatography machine at the University of Michigan.
+We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan.
 Salary support for Marc A.
@@
 **Figure S1.
-OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration using classification Random Forest (A).
+OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
 The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).

### 2018-06-14 submission/manuscript.tex
spell check and add nick changes to abstract

@@
 \textbf{Background.} Colorectal cancer (CRC) is a growing health concern
-with a large percentage of the risk of developing disease being due to
-environmental factors. Certain bacterial community members have been
-associated to increase CRC while other taxa have been associated with
-normal colons. Some of the taxa associated with normal colons can use
-fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor
-growth in model systems. However, the data supporting the importance of
-SCFAs in human CRC is less certain. Here, we test the hypothesis that
-SCFA concentration changes in individuals with colorectal tumors.
+with the majority of the risk for developing disease being due to
+environmental factors. The microbiota is one of these environmental
+factors with certain bacterial community members being associated with
+CRC, while other taxa are associated to colons without tumors. Some of
+the taxa associated to colons without tumors can use fiber to produce
+short-chain fatty acids (SCFAs) that can inhibit tumor growth in model
+systems. However, the data supporting the importance of SCFAs in human
+CRC is less certain. Here, we test the hypothesis that SCFA
+concentrations are different in individuals with colorectal tumors.
 
-\textbf{Methods.} We analyzed a cross-sectional (n=490) and pre- and
-post-treatment (n=67) group for their concentrations of acetate,
-butyrate, and propionate. Analysis also included tumor classification
-models using Random Forest, imputed gene relative abundance with
-PICRUSt, and metagenomic sequencing on a subset (n=85) of the total
-cross-sectional group.
+\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
+pre- and post-treatment (n=67) group for their concentrations of
+acetate, butyrate, and propionate. Analysis also included tumor
+classification models using Random Forest, imputed gene relative
+abundance with PICRUSt, and metagenomic sequencing on a subset (n=85) of
+the total cross-sectional group.
 
-\textbf{Results.} No difference in SCFA concentration was found between
-normal individuals and patients with adenoma or carcinoma (P-value
-\textgreater{} 0.15). There was no difference in classification models
-with or without SCFAs in their ability to predict adenoma or carcinoma
+\textbf{Results.} No difference in SCFA concentrations were found
+between individuals without tumors and patients with adenomas or
+carcinomas (P-value \textgreater{} 0.15). There was no difference in
+classification models with or without SCFAs in their ability to predict
+patients with adenomas or carcinomas versus individuals without tumors
 (P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
-also no difference in genes involved with SCFA synthesis (P-value
-\textgreater{} 0.70).
+also no difference in genes involved with SCFA synthesis between
+individuals without tumors and patients with adenomas or carcinomas
+(P-value \textgreater{} 0.70).
 
 \textbf{Conclusions.} Although our data does not support the hypothesis
-that SCFAs decrease in individuals that have colorectal tumors, there
-may be specific scenarios where SCFAs may still be beneficial for
-treatment of CRC. Alternatively, there may be other metabolites that
-have not been thoroughly investigated that are more important to the
-development of human CRC.
+that SCFAs are different in individuals that have colorectal tumors,
+there may be context specific scenarios where SCFAs may still be
+beneficial for treatment of CRC. Alternatively, there may be other
+mechanisms that have not been thoroughly investigated that are more
+important to the development of human CRC.
 
@@
 cause of death within the US (1, 2). Although there is a genetic
-component to the disease, the envrionment has been attributed as an
+component to the disease, the environment has been attributed as an
 important and large risk factor for CRC (3). This has lead to the
@@
 investigation provides additional information as to whether the taxa
-that are decreased in carcionma use SCFAs to help protect against colon
+that are decreased in carcinoma use SCFAs to help protect against colon
 tumorigenesis.
@@
 based on 16S rRNA gene sequencing we found no difference in any of the
-genes inolved with acetate, butyrate, or propionate synthesis (P-value
+genes involved with acetate, butyrate, or propionate synthesis (P-value
 \textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
 highlighted by visualizing genes important in butyrate synthesis
-{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
+{[}Figure 3A{]}. Using a Wilcoxon rank-sum test, there also was no
 difference in imputed gene relative abundance between pre- and
@@
 high and low SCFA concentrations {[}Figure S1A{]}. However, these models
-tended to be overfit, suggesting that rarer taxa may be important for
+tended to be over fit, suggesting that rarer taxa may be important for
 this classification {[}Figure S1A{]}. The most important OTUs to these
@@
 consent was obtained from all participants in accordance to the
-guidelines set out by the Helsinki Decleration.
+guidelines set out by the Helsinki Declaration.
 
@@
 frozen fecal samples suspended in 1ml of PBS instead of fecal
-suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
+suspensions in DNA Genotek OmniGut tubes, and the use of the actual
 weight of fecal samples instead of the average weight for SCFA
@@
 Briefly, the mothur shared file and metadata was converted into a biom
-formated table using the biom convert function, the subsequent biom file
-was processed with the `normalize\_by\_copy\_number.py' function, and
-subsequent imputed metagenomes created using the
+formatted table using the biom convert function, the subsequent biom
+file was processed with the `normalize\_by\_copy\_number.py' function,
+and subsequent imputed metagenomes created using the
 `predict\_metagenomes.py' function.
@@
 would also like to thank Kwi Kim and Thomas M Schmidt for their help in
-running the short-chian fatty acid analysis on the High-Performace
+running the short-chain fatty acid analysis on the High-Performance
 Liquid Chromatography machine at the University of Michigan. Salary
@@
 OTU-based models with 10-fold CV based on higher or lower than the
-medain SCFA concentration using classification Random Forest (A). The
+median SCFA concentration using classification Random Forest (A). The
 top 10 OTUs based on mean decrease in accuracy (MDA) for each model,

### 2018-06-14 submission/manuscript.Rmd
change title

@@
 ---
-title: "Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors"
+title: "Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors"
 bibliography: references.bib

### 2018-06-14 submission/manuscript.tex
change title

@@
 \hypersetup{unicode=true,
-            pdftitle={Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors},
+            pdftitle={Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal
-Tumors}
+  \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
+Microbiota, and Colorectal Tumors}
   \pretitle{\vspace{\droptitle}\centering\huge}

### 2018-06-14 submission/manuscript.Rmd
add intro fixes suggested by Nikc

@@
 
-Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
-Although there is a genetic component to the disease, the environment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000].
-This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
-Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011].
-Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017].
+Colorectal cancer (CRC) is currently the third highest cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
+Although there is a genetic component to the disease, the environment is attributed to being a larger risk factor for CRC [@Lichtenstein2000].
+These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011].
+Many of these environmental risk factors are capable of being modified, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
+Multiple reports in case/control studies have identified bacterial taxa commonly associated with individuals without tumors to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018].
+Many of these taxa within individuals wihtout tumors actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017].
 The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016].
-Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.
+Overall, the specific bacterial taxa of the microbiota that create SCFAs are an attractive target to modulate the risk of CRC.
 
@@
 There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
-The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals with normal colons [@Ohigashi2013].
+The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals without colon tumors [@Ohigashi2013].
 Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
@@
 
-Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations.
-Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors.
-Additionally, by increasing the adenoma tumor cohort it we will identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases.
-It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors (adenoma or carcinoma).
-This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.
+Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations.
+Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors.
+We also test previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity increases by increasing the number of patients with adenomas in our study.
+Additionally, we build upon these observations and assesses the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
+Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.
 
-To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups.
+To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups.
 The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
-Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations.
-This  investigation provides additional information as to whether the taxa that are decreased in carcinoma use SCFAs to help protect against colon tumorigenesis.
+Additionally, we (i) assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], (ii) used PICRUSt [@Langille2013] and metagenomic sequencing to assess the presence of genes involved in SCFA synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
+This investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.
 

### 2018-06-14 submission/manuscript.tex
add intro fixes suggested by Nikc

@@
 
-Colorectal cancer (CRC) is currently the third leading cancer-related
+Colorectal cancer (CRC) is currently the third highest cancer-related
 cause of death within the US (1, 2). Although there is a genetic
-component to the disease, the environment has been attributed as an
-important and large risk factor for CRC (3). This has lead to the
-investigation of how the microbiota may exacerbate or cause tumorigensis
-(4--6) and whether the bacterial community is altered (7, 8). Multiple
-reports in case/control studies on the microbiota and CRC have
-identified taxa commonly associated with the normal colons to be
-decreased in patients with carcinoma tumors (9--11). Many of these taxa
-actively produce short-chain fatty acids (SCFAs) from fiber that are a
-part of our general diet (12). The most extensively studied of these
-SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
-growing health problem and the microbiota through SCFAs have shown
-promise as a potential treatment option.
+component to the disease, the environment is attributed to being a
+larger risk factor for CRC (3). These environmental risk factors include
+but are not limited to smoking cigarettes, diet, and the microbiota
+(4--6). Many of these environmental risk factors are capable of being
+modified, and this has lead to the investigation of how the microbiota
+may exacerbate or cause tumorigensis (7--9) and whether the bacterial
+community is altered (10, 11). Multiple reports in case/control studies
+have identified bacterial taxa commonly associated with individuals
+without tumors to be decreased in patients with carcinoma tumors
+(11--13). Many of these taxa within individuals wihtout tumors actively
+produce short-chain fatty acids (SCFAs) from fiber that are a part of
+our general diet (14). The most extensively studied of these SCFAs are
+acetate, butyrate, and propionate (15). Overall, the specific bacterial
+taxa of the microbiota that create SCFAs are an attractive target to
+modulate the risk of CRC.
 
 Specific SCFAs, like butyrate, have shown positive results for CRC
-treatment within model systems (14). Butyrate has been shown to inhibit
-cancer cell growth in \emph{in vitro} systems (15). Additionally,
+treatment within model systems (16). Butyrate has been shown to inhibit
+cancer cell growth in \emph{in vitro} systems (17). Additionally,
 supplementation with food sources that bacteria use to create these
@@
 supplementation in mouse models of CRC caused an overall reduction in
-tumor burden while also increasing SCFA concentrations (16). Although
+tumor burden while also increasing SCFA concentrations (18). Although
 these model systems provide important preliminary evidence towards the
@@
 butyrate, and propionate versus either patients with adenomas or
-individuals with normal colons (17). Although this would argue that
+individuals without colon tumors (19). Although this would argue that
 increasing SCFA concentrations could be protective against
 tumorigenesis, fiber supplementation in randomized controlled trials
-have consistently failed to protect against tumor recurrence (18). These
+have consistently failed to protect against tumor recurrence (20). These
 findings would argue against the utility of treatments that aim to use
@@
 
-Our current study will fill some of the current gaps in the literature
-that relate to the study of SCFAs and CRC in human populations.
-Specifically, it will attempt to replicate previous case/control
-findings on SCFA concentrations in individuals with and without tumors.
-Additionally, by increasing the adenoma tumor cohort it we will identify
-if the observation of a continuous gradual reduction in SCFA
-concentrations occurs as tumor severity increases. It also will build
-upon these observations and assess the utility of using SCFAs and
-Operational Taxonomic Units (OTUs) as a potential risk stratification
-tool of colorectal tumors (adenoma or carcinoma). This study will
-provide important information on the replicability of previous findings
-in humans by extensively studying how SCFAs could be associated with
-colorectal tumors.
+Our study fills some of the current gaps in the literature that relate
+to the study of SCFAs and CRC in human populations. Specifically, it
+tests previous case/control findings on SCFA concentrations in
+individuals with and without tumors. We also test previous suggestions
+that there is a continuous reduction in SCFA concentrations as tumor
+severity increases by increasing the number of patients with adenomas in
+our study. Additionally, we build upon these observations and assesses
+the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a
+risk stratification tool of colorectal tumors (adenoma or carcinoma).
+Collectively, this study provides important information on the
+replicability of previous findings in humans by extensively studying how
+SCFAs are associated with colorectal tumors.
 
-To accomplish this task we will directly measure the concentration of
+To accomplish this task we directly measured the concentration of
 acetate, butyrate, and propionate within fecal samples for two different
@@
 before (pre-) and after (post-)treatment for colorectal tumors.
-Additionally, we also assessed whether adding SCFA concentrations to OTU
-data improves the classification of adenoma or carcinoma using the
-Random Forest algorithm (19), used PICRUSt (20) and metagenomic
-sequencing to assess the genes involved in SCFA synthesis, and analyzed
-how well 16S rRNA gene sequencing can predict SCFA concentrations. This
-investigation provides additional information as to whether the taxa
-that are decreased in carcinoma use SCFAs to help protect against colon
+Additionally, we (i) assessed the affect adding SCFA concentrations to
+OTU data had on classification of patients with adenoma or carcinoma
+using the Random Forest algorithm (21), (ii) used PICRUSt (22) and
+metagenomic sequencing to assess the presence of genes involved in SCFA
+synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts
+SCFA concentrations. This investigation provides additional information
+as to whether SCFAs are decreased in patients with colorectal tumors and
+provides context as to whether targeting taxa to increase SCFA
+concentrations is a viable option to protect against colon
 tumorigenesis.
@@
 Much of the previous research on SCFA benefit to human CRC has been
-illustrated in model systems (14). Many SCFAs are produced through the
-breakdown of fiber (12) and a recent study in mice found that fiber
+illustrated in model systems (16). Many SCFAs are produced through the
+breakdown of fiber (14) and a recent study in mice found that fiber
 supplementation increased SCFA concentrations and decreased tumor burden
-(16). Additionally, SCFAs such as butyrate can inhibit tumor growth in
-\emph{in vitro} experiments (15). Yet, observations in humans has been
+(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
+\emph{in vitro} experiments (17). Yet, observations in humans has been
 mixed. Previous case/control studies found associations with lower SCFA
-concentrations in individuals with carcinoma tumors (17). However,
+concentrations in individuals with carcinoma tumors (19). However,
 individual randomized controlled trials and a recent meta-analysis on
 fiber supplementation to prevent tumor recurrence has found no benefit
-(18, 21). Our results align with what has been reported in
+(20, 23). Our results align with what has been reported in
 randomized-controlled trials, that SCFAs do not provide general
@@
 CRC has been that all tumors are treated as the same type. However,
-there are known differences in the types of mutations that occur (22)
+there are known differences in the types of mutations that occur (24)
 and treating all tumors as equal may actually hide any benefit that
 could be found in specific subsets of individuals. Similar to the idea
-of specific immunotherapy for specific tumors (23), SCFAs may have
+of specific immunotherapy for specific tumors (25), SCFAs may have
 beneficial effects for specific types of colorectal tumors. Future
@@
 protective. Yet, taxa that are associated with SCFA production are
-consistently higher in normal colons than those with carcinomas (7, 8,
-24).
+consistently higher in normal colons than those with carcinomas (10, 11,
+26).
 
@@
 SCFAs even though taxa associated with their production are higher in
-individuals with normal colons versus patients with carcinomas (7, 8,
-24). Protection could be via a different pathway and by extension other
+individuals with normal colons versus patients with carcinomas (10, 11,
+26). Protection could be via a different pathway and by extension other
 metabolites that have not been extensively studied. Alternatively,
@@
 exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
-\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
-10)). The idea of niche exclusion is similar to how the community
-protects against \emph{Clostridium difficile} infection (25) with
+\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (12,
+13)). The idea of niche exclusion is similar to how the community
+protects against \emph{Clostridium difficile} infection (27) with
 chronic inflammation replacing the role of antibiotics. Although we did
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (26, 27). In brief, this study used fecal
+described in detail previously (28, 29). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (26).
+removal of a tumor with or without chemotherapy and radiation (28).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (26, 27). The University
+histopathological review of biopsies obtained (28, 29). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (28). The following changes to this protocol included the use of
+machine (30). The following changes to this protocol included the use of
 frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (28) and Figure 1 here).
+studies (see Table 1 (30) and Figure 1 here).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-previously described (26, 29, 30). The major differences from these
+previously described (28, 31, 32). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptClust algorithm (31).
+97\% similarity using the OptClust algorithm (33).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (20) was used.
+1.1.2 with the recommended standard operating protocol (22) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (32). Briefly, the sequences were quality
+previously described method (34). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (33). Open Reading Frames (ORFs) were
-identified using Prodigal (34), counts generated using Diamond (35),
+to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
+identified using Prodigal (36), counts generated using Diamond (37),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (36), and OPF alignment used the KEGG database (37).
+mmseq2 (38), and OPF alignment used the KEGG database (39).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (37,
-38) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (39,
+40) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 the data and then tested on the held out 20\% (80/20 split) using the
-Random Forest algorithm for classification models (19). This was
+Random Forest algorithm for classification models (21). This was
 repeated on 100 different 80/20 splits of the data to generate a
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (39). Generally, a Kruskal-Walis rank sum
+the statistical language R (41). Generally, a Kruskal-Walis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
 the groups used. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (40). First, we assessed differences in
+correct for multiple comparisons (42). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 
+\hypertarget{ref-FlissIsakov2017}{}
+4. \textbf{Fliss-Isakov N}, \textbf{Zelber-Sagi S}, \textbf{Webb M},
+\textbf{Halpern Z}, \textbf{Kariv R}. 2017. Smoking habits are strongly
+associated with colorectal polyps in a population-based case-control
+study. Journal of Clinical Gastroenterology 1.
+doi:\href{https://doi.org/10.1097/mcg.0000000000000935}{10.1097/mcg.0000000000000935}.
+
+\hypertarget{ref-Lee2015}{}
+5. \textbf{Lee J}, \textbf{Jeon JY}, \textbf{Meyerhardt JA}. 2015. Diet
+and lifestyle in survivors of colorectal cancer. Hematology/Oncology
+Clinics of North America \textbf{29}:1--27.
+doi:\href{https://doi.org/10.1016/j.hoc.2014.09.005}{10.1016/j.hoc.2014.09.005}.
+
+\hypertarget{ref-Kostic2011}{}
+6. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
+\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
+AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
+\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
+\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
+\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
+identifies association of fusobacterium with colorectal carcinoma.
+Genome Research \textbf{22}:292--298.
+doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.
+
 \hypertarget{ref-Zackular2013}{}
-4. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
 \textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
@@
 \hypertarget{ref-Baxter2014}{}
-5. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+8. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-Zackular2015}{}
-6. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
+9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
@@
 \hypertarget{ref-Shah2017}{}
-7. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
+10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
 \textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
@@
 \hypertarget{ref-meta_analysis_crc_Sze2018}{}
-8. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+11. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-Zeller2014}{}
-9. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+12. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 \hypertarget{ref-Baxter2016}{}
-10. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 
-\hypertarget{ref-Kostic2011}{}
-11. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
-\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
-AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
-\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
-\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
-\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
-identifies association of fusobacterium with colorectal carcinoma.
-Genome Research \textbf{22}:292--298.
-doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.
-
 \hypertarget{ref-Holscher2017}{}
-12. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
+14. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
 gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
@@
 \hypertarget{ref-Louis2016}{}
-13. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
+15. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
 and butyrate by the human colonic microbiota. Environmental Microbiology
@@
 \hypertarget{ref-test_OKeefe2016}{}
-14. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+16. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
 metabolites and colon cancer. Nature Reviews Gastroenterology \&
@@
 \hypertarget{ref-Encarnao2018}{}
-15. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
+17. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
 \textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
@@
 \hypertarget{ref-Bishehsari2018}{}
-16. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+18. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
 \textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
@@
 \hypertarget{ref-Ohigashi2013}{}
-17. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
+19. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
 \textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
@@
 \hypertarget{ref-Yao2017}{}
-18. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
+20. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
 Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-19. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+21. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-Langille2013}{}
-20. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+22. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 \hypertarget{ref-Schatzkin2000}{}
-21. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+23. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
 \textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
@@
 \hypertarget{ref-Fearon1990}{}
-22. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \hypertarget{ref-Thomas2016}{}
-23. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
 agents targeting the BCR-ABL tyrosine kinase as philadelphia
@@
 \hypertarget{ref-normalization_Sze2017}{}
-24. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+26. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-Theriot2015}{}
-25. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
+27. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
 the gastrointestinal microbiome and clostridium difficile. Annual Review
@@
 \hypertarget{ref-normalization_sze2017}{}
-26. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+28. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-27. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+29. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-28. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+30. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \hypertarget{ref-Schloss2009}{}
-29. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-30. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+32. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-31. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Hannigan2017}{}
-32. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-33. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-34. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-35. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-36. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-37. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-38. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-r_citation_2017}{}
-39. \textbf{R Core Team}. 2017. R: A language and environment for
+41. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-40. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-06-14 submission/manuscript.Rmd
result changes and nick result suggestions

@@
 **Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time.
-There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
+There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
 We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples.
 Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
-Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.
+Even though there was no difference in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.
 
-**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models with OTUs only or OTUs and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
+**Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
+Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors.
+Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
 For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
-Although including SCFA concentrations did not add extra information for classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of microbes are responsible for the observed SCFA concentrations in patients with carcinoma.
+Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of taxa are responsible for the observed SCFA concentrations in patients with colorectal tumors.
 
-**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas.
+**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
 First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
 This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
-Using a Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
 Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
 This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B].
-These observations provide evidence that the gene content does not change due to colorectal tumors.
-Although these results suggest that SCFA concentrations do not change due to colorectal tumor, there may be some errors present within our results that may account for the lack of associations.
+These observations provide evidence that gene prevalence does not change due to colorectal tumors.
 
@@
 However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure S1A].
-The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B].
-These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate.
+The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were taxa that are normally associated with SCFA production [Figure S1B].
+These results highlight that within our data, SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate.
 Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations.
-Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal cancer.
+Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal tumors.
 

### 2018-06-14 submission/manuscript.tex
result changes and nick result suggestions

@@
 from 490 individuals at a cross-sectional point in time. There was no
-difference between individuals with normal colons (n=172) and patients
+difference between individuals without colon tumors (n=172) and patients
 with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
@@
 or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even
-though there was no change in SCFA concentrations between individuals
-with normal colons and those with tumors, this information could still
-be important to help classify disease.
+though there was no difference in SCFA concentrations between
+individuals with normal colons and those with tumors, this information
+could still be important to help classify disease.
 
 \textbf{Random Forest models with SCFA concentrations do not classify
-tumors better.} Using the Random Forest algorithm we built models with
-OTUs only or OTUs and SCFA concentrations to classify normal versus
-adenoma and normal versus carcinoma fecal samples. For adenoma and
-carcinoma models, there was no difference between the median AUC of
-models with or without SCFA concentrations (P-value \textgreater{} 0.05)
-{[}Figure 2{]}. Although including SCFA concentrations did not add extra
-information for classification of colorectal tumors using Random Forest
-models, it is possible that the genes for enzymes involved in SCFA
-synthesis may vary based on the type of colorectal tumor; where a
-smaller number of microbes are responsible for the observed SCFA
-concentrations in patients with carcinoma.
+tumors better.} SCFA concentrations could improve prediction of tumors
+based on specific bacterial community structures. Our OTU data can be
+used in combination with SCFAs to assess whether there is a community
+dependent context to SCFA classification of tumors. Using the Random
+Forest algorithm we built models with OTU abundance data only or OTU
+abundances and SCFA concentrations to classify normal versus adenoma and
+normal versus carcinoma fecal samples. For adenoma and carcinoma models,
+there was no difference between the median AUC of models with or without
+SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
+Although including SCFA concentrations did not improve classification of
+colorectal tumors using Random Forest models, it is possible that the
+genes for enzymes involved in SCFA synthesis may vary based on the type
+of colorectal tumor; where a smaller number of taxa are responsible for
+the observed SCFA concentrations in patients with colorectal tumors.
 
@@
 important for the synthesis of SCFAs {[}Table S1{]}, we looked for
-differences in gene abundance based on individuals having normal colons,
-adenomas, or carcinomas. First, using imputed gene relative abundance
-based on 16S rRNA gene sequencing we found no difference in any of the
-genes involved with acetate, butyrate, or propionate synthesis (P-value
-\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
-highlighted by visualizing genes important in butyrate synthesis
-{[}Figure 3A{]}. Using a Wilcoxon rank-sum test, there also was no
-difference in imputed gene relative abundance between pre- and
-post-treatment samples for any genes involved with SCFA synthesis
-(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
-these 490 fecal samples (n=85) and used metagenomic sequencing to
-confirm these results. Like the imputed gene results, metagenomic
+differences in gene abundance between individuals without colon tumors
+and patients with adenomas or carcinomas. First, using imputed gene
+relative abundance based on 16S rRNA gene sequencing we found no
+difference in any of the genes involved with acetate, butyrate, or
+propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
+similarity between groups is highlighted by visualizing genes important
+in butyrate synthesis {[}Figure 3A{]}. Using a paired Wilcoxon rank-sum
+test, there also was no difference in imputed gene relative abundance
+between pre- and post-treatment samples for any genes involved with SCFA
+synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
+subset of these 490 fecal samples (n=85) and used metagenomic sequencing
+to confirm these results. Like the imputed gene results, metagenomic
 analysis found that there was no difference in any of the genes involved
-in SCFA synthesis between individuals with normal colons (n=29) and
+in SCFA synthesis between individuals without colon tumors (n=29) and
 patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
@@
 visualize the results for butyrate kinase {[}Figure 3B{]}. These
-observations provide evidence that the gene content does not change due
-to colorectal tumors. Although these results suggest that SCFA
-concentrations do not change due to colorectal tumor, there may be some
-errors present within our results that may account for the lack of
-associations.
+observations provide evidence that gene prevalence does not change due
+to colorectal tumors.
 
@@
 this classification {[}Figure S1A{]}. The most important OTUs to these
-models (assessed with mean decrease in accuracy (MDA)) were to taxa that
+models (assessed with mean decrease in accuracy (MDA)) were taxa that
 are normally associated with SCFA production {[}Figure S1B{]}. These
-results highlight that there are no immediately apparent errors in our
-data because SCFA concentrations are associated with taxa known to
-produce acetate, butyrate, and propionate. Additionally, OTUs associated
-with these taxa are the most important to models that can classify high
-and low SCFA concentrations. Overall, our results are robust and do not
-support the hypothesis that differences in SCFA concentrations are
-associated with colorectal cancer.
+results highlight that within our data, SCFA concentrations are
+associated with taxa known to produce acetate, butyrate, and propionate.
+Additionally, OTUs associated with these taxa are the most important to
+models that can classify high and low SCFA concentrations. Overall, our
+results are robust and do not support the hypothesis that differences in
+SCFA concentrations are associated with colorectal tumors.
 

### 2018-06-15 submission/manuscript.Rmd
make discussion changes and remaining text changes suggested by Nick

@@
 For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
-Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of taxa are responsible for the observed SCFA concentrations in patients with colorectal tumors.
+Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor.
+This would be consistent with previous 16S rRNA gene sequencing results where many taxa associated with SCFA production are decreased [@Shah2017; @meta_analysis_crc_Sze2018].
 
@@
 
-The observations from this study do not support the hypothesis that SCFA concentrations are lower in individuals with tumors.
+The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors.
 Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3].
-Further, no benefit could be found in using the concentrations to help classify individuals with and without tumors [Figure 2].
-These observations are in stark contrast to some of the previous literature.
+There is an intriguing reason why taxa associated with SCFA production are decreased in CRC but the genes involoved with its' production are not.
+Mouth-associated microbes such as *Fusobacterium nucleatum* and *Porphyromonas asaccharolytica* have been found to be increased in patients with carcinomas versus individuals without tumors [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
+Both of these bacterial species are known to have strains that can produce SCFAs such as butyrate [@Vital2014].
+Thus the reason we may be observing no change in genes involved with SCFA synthesis, as well as no change to SCFAs themselves, is because the production is being supported by more inflammatory microbes associated with CRC.
+Additionally, our observations that no benefit could be found in using the concentrations to help classify individuals with and without tumors would be consistent with this reason [Figure 2].
+However, our observations are in stark contrast to some of the previous literature.
 
@@
 
-One limitation of ours and others research into the effect of SCFAs in CRC has been that all tumors are treated as the same type.
+One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type.
 However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
@@
 Future research will need to test if this is a valid hypothesis.
-Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective.
-Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
+Another limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
+However, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
+Additionally, studies that measure SCFA changes after fiber supplementation use fecal material to track these responses [@scfa_measures_venkataraman2016].
+Although there are limitations with the current research on SCFAs and colorectal tumors, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective.
+Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 
-The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
+The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 Protection could be via a different pathway and by extension other metabolites that have not been extensively studied.
 Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g.
-*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016]).
+*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]).
 The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics.
-Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results
+Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.
 
@@
 
-Our observations found no difference in SCFA concentration, their utility as a classification tool, and for genes for enzymes involved in SCFA synthesis between individuals with normal colons and patients with either adenoma or carcinoma tumors.
+Our observations found no difference in SCFA concentration, their utility as a classification tool, or for genes of enzymes involved in SCFA synthesis between individuals without colon tumors and patients with either adenoma or carcinoma tumors.
 Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.
-Overall, these results suggest that the SCFAs produced by resident microbes do not protect against tumor.
-By focusing on other types of metabolites, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
+Overall, these results suggest that the SCFAs typically produced by resident microbes do not protect against tumor.
+By focusing on other types mechanisms, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
 
@@
 **Figure 1.
-No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A).
-Butyrate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (B).
-Propionate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (C).
+No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (A).
+Butyrate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (B).
+Propionate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (C).
 The black lines indicate the median SCFA concentration.

### 2018-06-15 submission/manuscript.tex
make discussion changes and remaining text changes suggested by Nick

@@
 genes for enzymes involved in SCFA synthesis may vary based on the type
-of colorectal tumor; where a smaller number of taxa are responsible for
-the observed SCFA concentrations in patients with colorectal tumors.
+of colorectal tumor. This would be consistent with previous 16S rRNA
+gene sequencing results where many taxa associated with SCFA production
+are decreased (10, 11).
 
@@
 The observations from this study do not support the hypothesis that SCFA
-concentrations are lower in individuals with tumors. Whether we directly
-measured the SCFA concentration or investigated genes associated with
-their production, no difference could be identified {[}Figure 1 \& 3{]}.
-Further, no benefit could be found in using the concentrations to help
-classify individuals with and without tumors {[}Figure 2{]}. These
-observations are in stark contrast to some of the previous literature.
+concentrations are different in individuals with tumors. Whether we
+directly measured the SCFA concentration or investigated genes
+associated with their production, no difference could be identified
+{[}Figure 1 \& 3{]}. There is an intriguing reason why taxa associated
+with SCFA production are decreased in CRC but the genes involoved with
+its' production are not. Mouth-associated microbes such as
+\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asaccharolytica}
+have been found to be increased in patients with carcinomas versus
+individuals without tumors (10, 11, 23). Both of these bacterial species
+are known to have strains that can produce SCFAs such as butyrate (24).
+Thus the reason we may be observing no change in genes involved with
+SCFA synthesis, as well as no change to SCFAs themselves, is because the
+production is being supported by more inflammatory microbes associated
+with CRC. Additionally, our observations that no benefit could be found
+in using the concentrations to help classify individuals with and
+without tumors would be consistent with this reason {[}Figure 2{]}.
+However, our observations are in stark contrast to some of the previous
+literature.
 
@@
 fiber supplementation to prevent tumor recurrence has found no benefit
-(20, 23). Our results align with what has been reported in
+(20, 25). Our results align with what has been reported in
 randomized-controlled trials, that SCFAs do not provide general
@@
 
-One limitation of ours and others research into the effect of SCFAs in
-CRC has been that all tumors are treated as the same type. However,
-there are known differences in the types of mutations that occur (24)
-and treating all tumors as equal may actually hide any benefit that
-could be found in specific subsets of individuals. Similar to the idea
-of specific immunotherapy for specific tumors (25), SCFAs may have
-beneficial effects for specific types of colorectal tumors. Future
-research will need to test if this is a valid hypothesis. Although there
-are limitations with the current research on SCFAs and colorectal
-tumors, evidence is building that these specific metabolites may not be
-protective. Yet, taxa that are associated with SCFA production are
-consistently higher in normal colons than those with carcinomas (10, 11,
-26).
+One limitation of current research into the effect of SCFAs in CRC has
+been that all tumors are treated as the same type. However, there are
+known differences in the types of mutations that occur (26) and treating
+all tumors as equal may actually hide any benefit that could be found in
+specific subsets of individuals. Similar to the idea of specific
+immunotherapy for specific tumors (27), SCFAs may have beneficial
+effects for specific types of colorectal tumors. Future research will
+need to test if this is a valid hypothesis. Another limitation is that a
+fecal sample may not be an ideal type of biospecimen and that the effect
+SCFAs have on tumorigenesis is only detected in the colon. However, most
+\emph{in vivo} studies as well as human studies have used fecal material
+in their analysis (18, 19). Additionally, studies that measure SCFA
+changes after fiber supplementation use fecal material to track these
+responses (28). Although there are limitations with the current research
+on SCFAs and colorectal tumors, our observations along with the
+randomized controlled trials on fiber supplementation in tumor
+recurrence (20) provide evidence that these specific metabolites may not
+be protective. Yet, taxa that are associated with SCFA production are
+consistently higher in indivdiuals without colon tumors than patients
+with carcinomas (10, 11, 23).
 
@@
 SCFAs even though taxa associated with their production are higher in
-individuals with normal colons versus patients with carcinomas (10, 11,
-26). Protection could be via a different pathway and by extension other
+individuals without tumors versus patients with carcinomas (10, 11, 23).
+Protection could be via a different pathway and by extension other
 metabolites that have not been extensively studied. Alternatively,
@@
 exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
-\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (12,
-13)). The idea of niche exclusion is similar to how the community
-protects against \emph{Clostridium difficile} infection (27) with
+\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
+12, 13)). The idea of niche exclusion is similar to how the community
+protects against \emph{Clostridium difficile} infection (29) with
 chronic inflammation replacing the role of antibiotics. Although we did
@@
 tumors, we think that there are many exciting new avenues to explore
-because of these results
+because of these results.
 
@@
 Our observations found no difference in SCFA concentration, their
-utility as a classification tool, and for genes for enzymes involved in
-SCFA synthesis between individuals with normal colons and patients with
-either adenoma or carcinoma tumors. Although these results are different
-than other reports in the literature, they do align with the randomized
-controlled trials that have tested fiber use in preventing colorectal
-tumor recurrence. Overall, these results suggest that the SCFAs produced
-by resident microbes do not protect against tumor. By focusing on other
-types of metabolites, the identification of more promising therapeutic
-options for use in treating colorectal cancer may be found.
+utility as a classification tool, or for genes of enzymes involved in
+SCFA synthesis between individuals without colon tumors and patients
+with either adenoma or carcinoma tumors. Although these results are
+different than other reports in the literature, they do align with the
+randomized controlled trials that have tested fiber use in preventing
+colorectal tumor recurrence. Overall, these results suggest that the
+SCFAs typically produced by resident microbes do not protect against
+tumor. By focusing on other types mechanisms, the identification of more
+promising therapeutic options for use in treating colorectal cancer may
+be found.
 
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (28, 29). In brief, this study used fecal
+described in detail previously (30, 31). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (28).
+removal of a tumor with or without chemotherapy and radiation (30).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (28, 29). The University
+histopathological review of biopsies obtained (30, 31). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (30). The following changes to this protocol included the use of
+machine (28). The following changes to this protocol included the use of
 frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (30) and Figure 1 here).
+studies (see Table 1 (28) and Figure 1 here).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-previously described (28, 31, 32). The major differences from these
+previously described (30, 32, 33). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptClust algorithm (33).
+97\% similarity using the OptClust algorithm (34).
 
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (34). Briefly, the sequences were quality
+previously described method (35). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
-identified using Prodigal (36), counts generated using Diamond (37),
+to contig assembly with MEGAHIT (36). Open Reading Frames (ORFs) were
+identified using Prodigal (37), counts generated using Diamond (38),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (38), and OPF alignment used the KEGG database (39).
+mmseq2 (39), and OPF alignment used the KEGG database (40).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (39,
-40) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (40,
+41) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (41). Generally, a Kruskal-Walis rank sum
+the statistical language R (42). Generally, a Kruskal-Walis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
 the groups used. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (42). First, we assessed differences in
+correct for multiple comparisons (43). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 
+\hypertarget{ref-normalization_Sze2017}{}
+23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
+\hypertarget{ref-Vital2014}{}
+24. \textbf{Vital M}, \textbf{Howe AC}, \textbf{Tiedje JM}. 2014.
+Revealing the bacterial butyrate synthesis pathways by analyzing
+(meta)genomic data. mBio \textbf{5}:e00889--14--e00889--14.
+doi:\href{https://doi.org/10.1128/mbio.00889-14}{10.1128/mbio.00889-14}.
+
 \hypertarget{ref-Schatzkin2000}{}
-23. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+25. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
 \textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
@@
 \hypertarget{ref-Fearon1990}{}
-24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+26. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \hypertarget{ref-Thomas2016}{}
-25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+27. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
 agents targeting the BCR-ABL tyrosine kinase as philadelphia
@@
 
-\hypertarget{ref-normalization_Sze2017}{}
-26. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
-\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
-microbiota in patients after treatment for colonic lesions. Microbiome
-\textbf{5}.
-doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+\hypertarget{ref-scfa_measures_venkataraman2016}{}
+28. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
+Variable responses of human microbiomes to dietary supplementation with
+resistant starch. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
 
 \hypertarget{ref-Theriot2015}{}
-27. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
+29. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
 the gastrointestinal microbiome and clostridium difficile. Annual Review
@@
 \hypertarget{ref-normalization_sze2017}{}
-28. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+30. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-29. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+31. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 
-\hypertarget{ref-scfa_measures_venkataraman2016}{}
-30. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
-\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
-Variable responses of human microbiomes to dietary supplementation with
-resistant starch. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
-
 \hypertarget{ref-Schloss2009}{}
-31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-32. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+33. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+34. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Hannigan2017}{}
-34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+35. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+36. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+37. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+38. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+39. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+40. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+41. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-r_citation_2017}{}
-41. \textbf{R Core Team}. 2017. R: A language and environment for
+42. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+43. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple
@@
 normal, adenoma, and carcinoma individuals using HPLC.} Acetate
-concentrations in fecal samples of individuals with normal colons,
+concentrations in fecal samples of individuals without colon tumors,
 adenomas, and carcinomas (A). Butyrate concentrations in fecal samples
-of individuals with normal colons, adenomas, and carcinomas (B).
-Propionate concentrations in fecal samples of individuals with normal
-colons, adenomas, and carcinomas (C). The black lines indicate the
+of individuals without colon tumors, adenomas, and carcinomas (B).
+Propionate concentrations in fecal samples of individuals without colon
+tumors, adenomas, and carcinomas (C). The black lines indicate the
 median SCFA concentration. Acetate concentrations in fecal samples

### 2018-06-15 code/make_SCFA_classification_graph.R
incorporate nick fig suggestions

@@ -77,7 +77,7 @@ auc_classification <- classification_model_data %>%
   geom_boxplot(position = position_dodge(width = 1)) + 
   geom_vline(xintercept=seq(1.5, length(unique(classification_summary$scfa))-0.5, 1), 
              lwd=1, colour="gray", alpha = 0.6) + 
-  theme_bw() + coord_cartesian(ylim = c(0, 0.8)) + 
+  theme_bw() + coord_cartesian(ylim = c(0, 1.0)) + 
   labs(x = "", y = "AUC") + ggtitle("A") + 
   scale_fill_manual(name = "Group", 
                     values = c("white", "darkgray")) + 

### 2018-06-15 code/make_scfa_measures_hplc_figure.R

@@ -115,12 +115,12 @@ cross_graph <- function(scfa, dataList){
     ggplot(aes(
       factor(dx, 
              levels = c("normal", "adenoma", "cancer"), 
-             labels = c("Control", "Adenoma", "Carcinoma")), mmol_kg)) + 
+             labels = c("Control", "Adenoma", "Carcinoma")), log10(mmol_kg+1))) + 
     geom_jitter(aes(color = dx), width = 0.3, size = 2.5, alpha = 0.75, show.legend = F) + 
-    stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
-                 colour = "black", geom = "crossbar", size = 0.75, width = 0.5)  + 
+    stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
+                 colour = "black", geom = "crossbar", size = 0.75, width = 0.5)  +
     scale_color_manual(values = c('#FFD700', '#DC143C', '#228B22')) + 
-    xlab("") + ylab("mmol per Kg") + theme_bw() + ggtitle(figlegend[[scfa]]) + 
+    labs(x = "", y = expression(Log["10"]~mmol~per~Kg)) + theme_bw() + ggtitle(figlegend[[scfa]]) + 
     theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
           panel.grid.major = element_blank(), 
           panel.grid.minor = element_blank(), 
@@ -129,14 +129,14 @@ cross_graph <- function(scfa, dataList){
           axis.text.x = element_text(size = 10, face = "bold"))
   
   if(scfa == "acetate"){
-    temp_graph <- temp_graph + annotate("text", label = paste("Acetate"), x = 3, y = 300, size = 4)
+    temp_graph <- temp_graph + annotate("text", label = paste("Acetate"), x = 3, y = 3, size = 4)
   }
   if(scfa == "butyrate"){
-    temp_graph <- temp_graph + annotate("text", label = paste("Butyrate"), x = 3, y = 80, size = 4)
-  } 
+    temp_graph <- temp_graph + annotate("text", label = paste("Butyrate"), x = 3, y = 3, size = 4)
+  }
   if(scfa == "propionate"){
-    temp_graph <- temp_graph + annotate("text", label = paste("Propionate"), x = 3, y = 110, size = 4)
-    
+    temp_graph <- temp_graph + annotate("text", label = paste("Propionate"), x = 3, y = 3, size = 4)
[…]

### 2018-07-20 code/run_rf_scfa_group_analysis.R
updates

@@ -25,7 +25,7 @@ imp_vars_models <- list(final_data_normal = c(), final_data_adenoma = c(), final
 
 
 # Load in the data
-for(i in names(data_groups)){
+for(i in names(data_models)){
   
   data_models[[i]] <- sapply(scfas, function(x) read_csv(paste("data/process/tables/", i, "_",
                                             x, "_classification_RF_group_summary.csv", sep = "")) %>% 

### 2018-07-20 code/run_id_sig_changes_by_group_and_scfa.R
first commit

@@ -1,0 +1,204 @@
+### ID largest changers based on SCFA group or concentrations
+### Group SCFA analysis
+# Marc Sze
+
+
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "caret"))
+
+
+#setup variables that will be used
+scfas <- c("acetate", "butyrate", "isobutyrate", "propionate")
+
+# Samples that had to little volume to actually be measured
+not_measured <- c("3367653", "2041650", "2043650", "3027650")
+
+##############################################################################################
+############### List of functions to get things to run nice ##################################
+##############################################################################################
+
+# Function to load in the respective scfa data
+upload_scfa_data <- function(scfa_name, path_to_file, ending){
+  # scfa_name is a variable that stores the scfa of interest e.g. "acetate"
+  # path_to_file stores the directory path where the files are
+  # ending is the common ending of the file of interest
+  
+  # the command that reads in the files
+  tempData <- read_csv(paste(path_to_file, scfa_name, ending, sep = "")) %>% 
+    rename(Group = study_id)
+  # sends the data to the global work environment
+  return(tempData)
+}
+
+# Function to remove specific data from sets of interest
+pare_down_data <- function(ids_to_remove, scfa_list, otu_data, meta_data){
+  
+  # Remove data from tables first
+  temp_otuData <- otu_data %>% filter(!(Group %in% ids_to_remove))
+  
+  nzv <- nearZeroVar(temp_otuData)
+  
+  nzv_rm_tempOTU <- temp_otuData[, -nzv]
+  
+  temp_meta_data <- meta_data %>% filter(!(Group %in% ids_to_remove))
+  
+  # Remove data from lists
+  tempList <- lapply(scfa_list, 
+                     function(x) filter(x, !(Group %in% ids_to_remove)) %>% 
+                       inner_join(temp_meta_data, by = "Group") %>% 
+                       inner_join(nzv_rm_tempOTU, by = "Group")) 
+  
+  return(tempList)
+}
+
+# Function to generate the high low groupings for each respective SCFA
+get_high_low <- function(all_data_list, scfa_median_list, scfas){
+  
[…]

### 2018-07-20 code/run_id_sig_changes_by_group_and_scfa.R
add sig summaries

@@ -12,7 +12,7 @@ loadLibs(c("tidyverse", "caret"))
 
 
 #setup variables that will be used
-scfas <- c("acetate", "butyrate", "isobutyrate", "propionate")
+scfas <- c("acetate", "butyrate", "propionate")
 
 # Samples that had to little volume to actually be measured
 not_measured <- c("3367653", "2041650", "2043650", "3027650")
@@ -131,16 +131,16 @@ final_sp_data <- sapply(scfas,
                         function(x) split_dataframe(x, final_data), simplify = F)
 
 # Set up the separate analysis lists
-class_data <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
-reg_data <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
+class_data <- list(acetate = c(), butyrate = c(), propionate = c())
+reg_data <- list(acetate = c(), butyrate = c(), propionate = c())
 
-norm_class <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
-adn_class <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
-crc_class <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
+norm_class <- list(acetate = c(), butyrate = c(), propionate = c())
+adn_class <- list(acetate = c(), butyrate = c(), propionate = c())
+crc_class <- list(acetate = c(), butyrate = c(), propionate = c())
 
-norm_reg <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
-adn_reg <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
-crc_reg <- list(acetate = c(), butyrate = c(), isobutyrate = c(), propionate = c())
+norm_reg <- list(acetate = c(), butyrate = c(), propionate = c())
+adn_reg <- list(acetate = c(), butyrate = c(), propionate = c())
+crc_reg <- list(acetate = c(), butyrate = c(), propionate = c())
 
 # Set up and run the analysis
 for(i in scfas){
@@ -153,7 +153,9 @@ for(i in scfas){
     mutate(wil_test = map(data, ~wilcox.test(rel_abund ~ high_low, data = .x)),
            summary_data = map(wil_test, broom::tidy)) %>%
[…]

### 2018-07-20 code/run_id_sig_changes_by_group_and_scfa.R
add data read out

@@ -232,5 +232,12 @@ sig_reg_summary <- norm_reg %>% bind_rows() %>%
               mutate(model = "crc_reg") %>% 
               select(otu, genus, dx, statistic, estimate, p.value, bh, scfa, model))
 
+# Write out the significant data
+write_csv(sig_class_summary, "data/process/tables/significant_class_otu_comp_summary.csv")
+
+write_csv(sig_reg_summary, "data/process/tables/significant_reg_otu_comp_summary.csv")
+
+
+
 
 

### 2018-07-20 code/figure4.R
reorder figs

@@ -1,57 +1,16 @@
-### Create graph of SCFA classification AUCs
+### Graph the aggregated data
+### Visualize how well do the predictions do by group (normal, adenoma, cancer)?
 ### Marc Sze
+
 
 # Load in needed functions and libraries
 source('code/functions.R')
 
 # Load needed libraries
-loadLibs(c("tidyverse", "gridExtra"))
+loadLibs(c("tidyverse"))
 
 #setup variables that will be used
 scfas <- c("acetate", "butyrate", "propionate")
-
-
-# Read in classification data
-classification_model_data <- read_csv("data/process/tables/acetate_classification_RF_summary.csv") %>% 
-  mutate(scfa = "acetate") %>% 
-  bind_rows(
-    read_csv("data/process/tables/butyrate_classification_RF_summary.csv") %>% 
-      mutate(scfa = "butyrate"), 
-    read_csv("data/process/tables/propionate_classification_RF_summary.csv") %>% 
-      mutate(scfa = "propionate"))
-
-classification_summary <- classification_model_data %>% 
-  group_by(scfa) %>% 
-  summarise(median_train_auc = median(train_AUC), 
-            max_train_auc = max(train_AUC), 
-            min_train_auc = min(train_AUC), 
-            median_test_auc = median(test_AUC), 
-            max_test_auc = max(test_AUC), 
-            min_test_auc = min(test_AUC), 
-            median_sens = median(Sens), max_sens = max(Sens), min_sens = min(Sens), 
-            median_spec = median(Spec), max_spec = max(Spec), min_spec = min(Spec)) %>% 
-  gather("groupings", "auc", median_train_auc:min_test_auc)
-
-
-test <- classification_model_data %>% 
-  gather("groupings", "auc", train_AUC:test_AUC) %>% 
-  separate(groupings, c("group", "metric"), sep = "_") %>% 
-  select(-metric)
-
-
-classification_imp_otus <- read_csv("data/process/tables/acetateimp_otus_classification_RF_summary.csv") %>% 
-  mutate(scfa = "acetate") %>% 
-  bind_rows(
-    read_csv("data/process/tables/butyrateimp_otus_classification_RF_summary.csv") %>% 
-      mutate(scfa = "butyrate"), 
-    read_csv("data/process/tables/propionateimp_otus_classification_RF_summary.csv") %>% 
-      mutate(scfa = "propionate")) %>% 
-  group_by(scfa, otu) %>% 
[…]

### 2018-07-20 code/figure5.R

@@ -1,225 +1,0 @@
-### Graph the aggregated data
-### Visualize how well do the predictions do by group (normal, adenoma, cancer)?
-### Marc Sze
-
-
-# Load in needed functions and libraries
-source('code/functions.R')
-
-# Load needed libraries
-loadLibs(c("tidyverse"))
-
-#setup variables that will be used
-scfas <- c("acetate", "butyrate", "propionate")
-
-# Read in taxnomic data
-tax <- read_tsv("data/process/final.taxonomy") %>% 
-  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
-  separate(Taxonomy, c("domain", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
-  mutate(genus = str_replace_all(genus, "_unclassified", ""), 
-         genus = str_replace_all(genus, "_", " "))
-
-# Read in the data
-# The data is obtained from the model tracking of the run_rf_scfa_predictions.R script
-# Put in a temporary run number until the analysis is re-run with it in the data frame.
-reg_data <- map(scfas, function(x) 
-  read_csv(paste("data/process/tables/", x, "_regression_RF_train_conc_summary.csv", sep = "")) %>% 
-    mutate(scfa = x, 
-           model = "reg_train")) %>% 
-  bind_rows() %>% 
-  bind_rows(map(scfas, function(x) 
-    read_csv(paste("data/process/tables/", x, "_regression_RF_test_conc_summary.csv", sep = "")) %>% 
-      mutate(run = rep(1:100, each = 85), 
-             scfa = x, 
-             model = "reg_test")) %>% 
-      bind_rows() %>% 
-      rename(obs = mmol_kg, pred = tempPredictions, sample_id = Group)) %>% 
-  mutate(difference = obs - pred) %>% 
-  group_by(model, scfa, run, dx) %>% 
-  summarise(median_diff = median(difference), 
-            min_dff = min(difference), 
-            max_diff = max(difference)) %>% 
-  ungroup() %>% 
-  group_by(model, scfa, dx) %>% 
-  summarise(median_yes = median(median_diff, na.rm = T), 
-            min_yes = min(median_diff, na.rm = T), 
-            max_yes = max(median_diff, na.rm = T))
-
-# log_reg_data <- map(scfas, function(x) 
-#   read_csv(paste("data/process/tables/log_", x, "_regression_RF_train_conc_summary.csv", sep = "")) %>% 
-#     mutate(scfa = x, 
[…]

### 2018-07-20 code/figureS1.R

@@ -1,0 +1,188 @@
+### Create graph of SCFA classification AUCs
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "gridExtra"))
+
+#setup variables that will be used
+scfas <- c("acetate", "butyrate", "propionate")
+
+
+# Read in classification data
+classification_model_data <- read_csv("data/process/tables/acetate_classification_RF_summary.csv") %>% 
+  mutate(scfa = "acetate") %>% 
+  bind_rows(
+    read_csv("data/process/tables/butyrate_classification_RF_summary.csv") %>% 
+      mutate(scfa = "butyrate"), 
+    read_csv("data/process/tables/propionate_classification_RF_summary.csv") %>% 
+      mutate(scfa = "propionate"))
+
+classification_summary <- classification_model_data %>% 
+  group_by(scfa) %>% 
+  summarise(median_train_auc = median(train_AUC), 
+            max_train_auc = max(train_AUC), 
+            min_train_auc = min(train_AUC), 
+            median_test_auc = median(test_AUC), 
+            max_test_auc = max(test_AUC), 
+            min_test_auc = min(test_AUC), 
+            median_sens = median(Sens), max_sens = max(Sens), min_sens = min(Sens), 
+            median_spec = median(Spec), max_spec = max(Spec), min_spec = min(Spec)) %>% 
+  gather("groupings", "auc", median_train_auc:min_test_auc)
+
+
+test <- classification_model_data %>% 
+  gather("groupings", "auc", train_AUC:test_AUC) %>% 
+  separate(groupings, c("group", "metric"), sep = "_") %>% 
+  select(-metric)
+
+
+classification_imp_otus <- read_csv("data/process/tables/acetateimp_otus_classification_RF_summary.csv") %>% 
+  mutate(scfa = "acetate") %>% 
+  bind_rows(
+    read_csv("data/process/tables/butyrateimp_otus_classification_RF_summary.csv") %>% 
+      mutate(scfa = "butyrate"), 
+    read_csv("data/process/tables/propionateimp_otus_classification_RF_summary.csv") %>% 
+      mutate(scfa = "propionate")) %>% 
+  group_by(scfa, otu) %>% 
+  summarise(median_mda = median(Overall), 
+            max_mda = max(Overall), 
+            min_mda = min(Overall)) %>% 
+  arrange(desc(median_mda), .by_group = T) %>% 
[…]

### 2018-07-20 code/make_reg_group_scfa_sig_graph.R
first commit

@@ -1,0 +1,34 @@
+### Graph the significant correlations and differences 
+### Group SCFA analysis
+# Marc Sze
+
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "gridExtra", "viridis"))
+
+
+reg_data <- read_csv("data/process/tables/significant_reg_otu_comp_summary.csv") %>% 
+  mutate(genus = str_replace_all(genus, "_unclassified", ""), 
+         genus = str_replace_all(genus, "_", " "))
+
+
+graph_data <- reg_data %>% 
+  group_by(scfa, dx, genus) %>% 
+  count()
+
+graph_data %>% 
+  ungroup() %>% 
+  mutate(dx = factor(dx, 
+                     levels = c("normal", "adenoma", "cancer"), 
+                     labels = c("Normal", "Adenoma", "Carcinoma"))) %>% 
+  ggplot(aes(dx, n, group = genus, fill = genus)) + 
+  geom_bar(stat = "identity") + 
+ # geom_bar(stat = "identity", position = position_dodge(1), show.legend = F) + 
+  facet_grid(~scfa) + 
+  labs(x = "", y = "Number of Significant OTUs") + 
+  scale_fill_viridis(discrete = T, option = "magma") + 
+  theme(legend.position = "bottom")
+

### 2018-07-20 code/make_reg_group_scfa_sig_graph.R
update coding for good version

@@ -9,26 +9,54 @@ source('code/functions.R')
 # Load needed libraries
 loadLibs(c("tidyverse", "gridExtra", "viridis"))
 
+taxonomy <- read_tsv("data/process/final.taxonomy") %>% 
+  select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";")
 
 reg_data <- read_csv("data/process/tables/significant_reg_otu_comp_summary.csv") %>% 
-  mutate(genus = str_replace_all(genus, "_unclassified", ""), 
-         genus = str_replace_all(genus, "_", " "))
+  left_join(select(taxonomy, OTU, family), by = c("otu" = "OTU")) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         direction = case_when(
+           estimate < 0 ~ "neg", 
+           estimate > 0 ~ "pos", 
+           TRUE ~ "uh oh"))
+  
 
 
 graph_data <- reg_data %>% 
-  group_by(scfa, dx, genus) %>% 
+  mutate(family = case_when(
+    family %in% c("Acidaminococcaceae", "Actinomycetaceae", "Bacteria", 
+                 "Pseudomonadaceae", "Streptococcaceae", "Sutterellaceae", 
+                 "Proteobacteria", "Firmicutes", "Bacteroidaceae", 
+                 "Clostridia", "Erysipelotrichaceae", "Oxalobacteraceae", 
+                 "Pasteurellaceae") ~ "Other", 
+    TRUE ~ family)) %>% 
+  group_by(scfa, dx, direction, family) %>% 
   count()
 
 graph_data %>% 
   ungroup() %>% 
   mutate(dx = factor(dx, 
                      levels = c("normal", "adenoma", "cancer"), 
-                     labels = c("Normal", "Adenoma", "Carcinoma"))) %>% 
-  ggplot(aes(dx, n, group = genus, fill = genus)) + 
-  geom_bar(stat = "identity") + 
+                     labels = c("Normal", "Adenoma", "Carcinoma")), 
+         family = factor(family, 
+                         levels = c('Clostridiales', 'Lachnospiraceae', 'Porphyromonadaceae', 
+                                    'Rikenellaceae', 'Ruminococcaceae', 'Other'), 
[…]

### 2018-07-23 code/make_reg_group_scfa_sig_graph.R
add number of correlation fig

@@ -36,7 +36,7 @@ graph_data <- reg_data %>%
   group_by(scfa, dx, direction, family) %>% 
   count()
 
-graph_data %>% 
+number_corr_plot <- graph_data %>% 
   ungroup() %>% 
   mutate(dx = factor(dx, 
                      levels = c("normal", "adenoma", "cancer"), 
@@ -58,5 +58,16 @@ graph_data %>%
   facet_grid(scfa~direction) +
   labs(x = "", y = "Number of Significant OTUs") + 
   scale_fill_manual(name = "", values = c('#9B30FF', '#63B8FF', '#008080', '#8FBC8F', '#FF7F00', '#6C7B8B')) + 
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom") + 
+  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        legend.position = "bottom", 
+        legend.text = element_text(size = 10), 
+        axis.text.y = element_text(size = 10), 
+        axis.title = element_text(size = 12), 
+        axis.text.x = element_text(size = 10, face = "bold"))
+
+ggsave("results/figures/Figure5.pdf", number_corr_plot, width = 7, height = 8)
+
 

### 2018-07-23 code/make_class_group_scfa_sig_graph.R
first commit

@@ -1,0 +1,66 @@
+### Graph the significant correlations and differences -- classification
+### Group SCFA analysis
+# Marc Sze
+
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "gridExtra", "viridis"))
+
+taxonomy <- read_tsv("data/process/final.taxonomy") %>% 
+  select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";")
+
+class_data <- read_csv("data/process/tables/significant_class_otu_comp_summary.csv") %>% 
+  left_join(select(taxonomy, OTU, family), by = c("otu" = "OTU")) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "))
+
+graph_data <- class_data %>% 
+  mutate(family = case_when(
+    family %in% c("Acidaminococcaceae", "Actinomycetaceae", "Bacteria",
+                  "Pseudomonadaceae", "Streptococcaceae", "Sutterellaceae",
+                  "Proteobacteria", "Firmicutes", "Bacteroidaceae",
+                  "Clostridia", "Erysipelotrichaceae", "Oxalobacteraceae",
+                  "Pasteurellaceae") ~ "Other",
+    TRUE ~ family)) %>%
+  group_by(scfa, dx, family) %>% 
+  count()
+
+
+
+number_class_plot <- graph_data %>% 
+  ungroup() %>% 
+  mutate(dx = factor(dx, 
+                     levels = c("normal", "adenoma", "cancer"), 
+                     labels = c("Normal", "Adenoma", "Carcinoma")), 
+         family = factor(family, 
+                         levels = c('Clostridiales', 'Lachnospiraceae', 'Porphyromonadaceae', 
+                                    'Rikenellaceae', 'Ruminococcaceae', 'Other'), 
+                         labels = c('Clostridiales', 'Lachnospiraceae', 'Porphyromonadaceae', 
+                                    'Rikenellaceae', 'Ruminococcaceae', 'Other')), 
+         scfa = factor(scfa, 
+                       levels = c("acetate", "butyrate", "propionate"), 
[…]

### 2018-07-23 submission/supplemental.Rmd
add supplemental tables of correlations and sig diffs

@@
 ```
+
+\newpage
+
+**Table S5: Spearman tests between OTU relative abundance and SCFA concentrations.**
+
+\scriptsize
+
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+
+table_s5 <- read_csv("../data/process/tables/significant_reg_otu_comp_summary.csv") %>% 
+  left_join(read_tsv("../data/process/final.taxonomy") %>% 
+              select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
+              select(OTU, family), by = c("otu" = "OTU")) %>% 
+  select(otu, family, genus, dx, scfa, estimate, p.value, bh) %>% 
+  mutate_at(vars(p.value:bh), function(x) format(x, scientific = T, digits = 3)) %>% 
+  mutate(estimate = round(estimate, digits = 3), 
+         family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         genus = str_replace_all(family, "_unclassified", ""), 
+         genus = str_replace_all(family, "_", " "))
+
+colnames(table_s5) <-  c("OTU", "Family", "Genus", "Group", "SCFA", "Rho", "P-value", "BH")
+
+kable(table_s5, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\normalsize
+
+\newpage
+
+**Table S6: Wilcoxson tests between OTU relative abundance and high or low SCFA concentrations based on the overall median.**
+
+\scriptsize
+
+```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+
+table_s6 <- read_csv("../data/process/tables/significant_class_otu_comp_summary.csv") %>% 
+  left_join(read_tsv("../data/process/final.taxonomy") %>% 
+              select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
+              select(OTU, family), by = c("otu" = "OTU")) %>% 
+  select(otu, family, genus, dx, scfa, p.value, bh) %>% 
+  mutate_at(vars(p.value:bh), function(x) format(x, scientific = T, digits = 3)) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         genus = str_replace_all(family, "_unclassified", ""), 
+         genus = str_replace_all(family, "_", " "))
+
+colnames(table_s6) <-  c("OTU", "Family", "Genus", "Group", "SCFA", "P-value", "BH")
+
+kable(table_s6, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\normalsize

### 2018-07-23 submission/manuscript.Rmd
update fig legends and add needed analysis tables

@@
 opf_select <- read_csv("../data/process/tables/select_scfa_opf_kruskal_summary.csv")
+
+# Load in the group comparison from SCFA models
+class_group_rf_data <- read_csv("../data/process/tables/full_model_class_group_results.csv")
+reg_group_rf_data <- read_csv("../data/process/tables/full_model_reg_group_results.csv")
 
@@
 The other butyrate pathway genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
+**Figure 4.
+OTU-based regression Random Forest models of SCFA concentrations.** The train and test correlation between actual and predicted values from 100 different 80/20 split OTU-based models with 10-fold CV using regression Random Forest (A).
+The model accuracy of predicted SCFA concentrations differed between individuals without tumors, patients with adenomas, and patients with carcinomas.
+Generally, patients with carcinomas had predicted concentrations closest to their actual measured concentration (B).
+The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
+
+**Figure 5.
+Patients with adenomas had the higest number of significant negative correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
+Fewer significant positive correlations were observed overall.
+Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.
 
@@
 **Figure S1.
-OTU-based Random Forest models of SCFA concentrations.** The train and test results of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
-The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B).
+OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
+The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas.
+Patients with adenomas consistently had the best classification accuracy (B).
+The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
+
+**Figure S2.
+Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
+Fewer significant OTUs were observed in individuals without tumors (normal) and patients with carcinomas versus patients with adenomas.

### 2018-07-23 submission/manuscript.tex
update fig legends and add needed analysis tables

@@
 from the PICRUSt analysis did not align to any of the OPFs in the
-metagenome analysis.
+metagenome analysis.\\
+\textbf{Figure 4. OTU-based regression Random Forest models of SCFA
+concentrations.} The train and test correlation between actual and
+predicted values from 100 different 80/20 split OTU-based models with
+10-fold CV using regression Random Forest (A). The model accuracy of
+predicted SCFA concentrations differed between individuals without
+tumors, patients with adenomas, and patients with carcinomas. Generally,
+patients with carcinomas had predicted concentrations closest to their
+actual measured concentration (B). The top 10 OTUs based on mean
+decrease in accuracy (MDA) for each SCFA model, colored by their lowest
+taxonomic identification (C).
+
+\textbf{Figure 5. Patients with adenomas had the higest number of
+significant negative correlations between OTU relative abundance and
+SCFA concentration.} Colors denote the family or lowest taxonomic ID
+that an OTU classified to. Fewer significant positive correlations were
+observed overall. Additionally, the differences in the number of
+significant positive correlations between patients with adenomas versus
+individuals without tumors (normal) and patients with carcinomas was not
+as pronounced as the number of significant negative correlations.
 
@@
 
-\textbf{Figure S1. OTU-based Random Forest models of SCFA
-concentrations.} The train and test results of 100 different 80/20
-OTU-based models with 10-fold CV based on higher or lower than the
-median SCFA concentration using classification Random Forest (A). The
-top 10 OTUs based on mean decrease in accuracy (MDA) for each model,
-colored by their lowest taxonomic identification (B).
+\textbf{Figure S1. OTU-based classification Random Forest models of
+high/low SCFA groups based on overall SCFA median concentration.} The
+train and test results of 100 different 80/20 split OTU-based models
+with 10-fold CV based on higher or lower than the median SCFA
+concentration using classification Random Forest (A). The model accuracy
+of predicted high/low SCFA groups differed between individuals without
+tumors, patients with adenomas, and patients with carcinomas. Patients
+with adenomas consistently had the best classification accuracy (B). The
+top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA
+model, colored by their lowest taxonomic identification (C).
+
+\textbf{Figure S2. Patients with adenomas had the higest number of
+significant differences in OTU relative abundance between high/low SCFA
+groups.} Colors denote the family or lowest taxonomic ID that an OTU
+classified to. Fewer significant OTUs were observed in individuals
+without tumors (normal) and patients with carcinomas versus patients
+with adenomas.
 

### 2018-07-23 code/figure4.R
update fig with direction

@@ -175,6 +175,8 @@ test_data_graph <- reg_data %>%
   coord_cartesian(ylim = c(-20, 20)) + 
   labs(x = "", y = "Difference from Actual Concentration") + 
   scale_color_manual(name = "", values = c('#228B22', '#FFD700', '#DC143C')) + 
+  annotate("text", label = paste("Higher in Observed"), x = 2.9, y = 18, size = 3.5) + 
+  annotate("text", label = paste("Higher in Predicted"), x = 2.9, y = -18, size = 3.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 

### 2018-07-23 code/Run_full_adn_RF_data_setup.R
remove isobutyrate from models

@@ -22,7 +22,7 @@ metaI <- filter(metaI, !(sample %in% good_metaf$initial)) %>%
   mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin))
 
 #setup variables that will be used
-scfas <- c("acetate", "butyrate", "isobutyrate", "propionate")
+scfas <- c("acetate", "butyrate", "propionate")
 
 # Samples that had to little volume to actually be measured
 not_measured <- c("3367653", "2041650", "2043650", "3027650")

### 2018-07-24 code/make_disease_group_RF_imp_graph.R
update colors

@@ -30,18 +30,36 @@ all_data <- read_csv("data/process/tables/adn_full_MDA_Summary.csv") %>%
     Variable %in% c("acetate", "butyrate", "propionate") ~ str_to_title(Variable), 
     TRUE ~ family), 
     family = str_replace(family, "_unclassified", ""), 
-    family = str_replace_all(family, "_", " "))
+    family = str_replace_all(family, "_", " "), 
+    Variable = case_when(
+      Variable %in% c("acetate", "butyrate", "propionate") ~ str_to_title(Variable), 
+      TRUE ~ Variable))
 
 all_adn <- all_data %>% 
   filter(model == "adn_full") %>% 
   mutate(Variable = factor(Variable, 
                            levels = rev(Variable), 
-                           labels = rev(Variable))) %>% 
+                           labels = rev(Variable)), 
+         family = str_replace(family, "Enterobacteriaceae", "Other"), 
+         family = factor(family, 
+                         levels = c("Acetate", "Butyrate", "Clostridiales", "Lachnospiraceae", 
+                                    "Porphyromonadaceae", "Ruminococcaceae", "Other"), 
+                         labels = c("Acetate", "Butyrate", "Clostridiales", "Lachnospiraceae", 
+                                    "Porphyromonadaceae", "Ruminococcaceae", "Other"))) %>% 
   ggplot(aes(Variable, median_mda, color = family)) + 
   geom_pointrange(aes(ymin = iqr25, ymax = iqr75), position = position_dodge(width = 1), size = 1) + 
-  coord_flip(ylim = c(0, 6)) + theme_bw() + 
-  scale_color_manual(name = "", values = c('#FF3E96', '#8B475D', '#8B008B', '#63B8FF', '#008080', '#FF7F00')) + 
-  theme(legend.position = "bottom")
+  coord_flip(ylim = c(0, 3)) + theme_bw() + 
+  labs(x = "", y = "Mean Decrease in Accuracy") + 
+  scale_color_manual(name = "", values = c('#FF3E96', '#8B475D', '#9B30FF', 
+                                           '#63B8FF', '#008080', '#FF7F00', '#6C7B8B'), 
+                     guide = guide_legend(nrow = 4)) + 
+  theme(legend.position = c(0.7, 0.3), 
+        legend.background = element_rect(color = "black"), 
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
[…]

### 2018-07-24 code/figure4.R
update colors and fig panels

@@ -106,7 +106,8 @@ regression_imp_otus <- read_csv("data/process/tables/acetateimp_otus_regression_
   slice(1:10)
 
 combined_imp_reg_summary <- regression_imp_otus %>% 
-  left_join(select(tax, OTU, genus), by = c("otu" = "OTU"))
+  left_join(select(tax, OTU, family), by = c("otu" = "OTU")) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""))
 
 ### Plot the regression data ###
 r2_classification <- regression_model_data %>%
@@ -133,28 +134,108 @@ r2_classification <- regression_model_data %>%
         axis.title = element_text(size = 12),
         axis.text.x = element_text(size = 10, face = "bold"))
 
-
-imp_otus_classification <- combined_imp_reg_summary %>% 
+acetate_imp_graph <- combined_imp_reg_summary %>% 
   ungroup() %>% 
-  mutate(scfa = factor(scfa, 
-                       levels = c("acetate", "butyrate", "propionate"), 
-                       labels = c("Acetate", "Butyrate", "Propionate"))) %>% 
-  ggplot(aes(scfa, median_mda, color = genus, group = otu)) + 
+    filter(scfa == "acetate") %>% 
+  mutate(otu = factor(otu, 
+                      levels = rev(otu), 
+                      labels = rev(otu)), 
+    family = case_when(
+           family %in% c("Firmicutes", "Pasteurellaceae") ~ "Other", 
+           TRUE ~ family), 
+         family = factor(family, 
+                         levels = c("Clostridiales", "Lachnospiraceae", "Ruminococcaceae", "Other"), 
+                         labels = c("Clostridiales", "Lachnospiraceae", "Ruminococcaceae", "Other"))) %>% 
+    ggplot(aes(otu, median_mda, color = family)) + 
+    geom_pointrange(aes(ymin = min_mda, ymax = max_mda), position = position_dodge(width = 1), size = 1) + 
+    coord_flip(ylim = c(-5, 25)) + theme_bw() + ggtitle("C") + 
+    annotate("text", label = paste("Acetate"), x = 4.7, y = 16.5, size = 3.5) +
+    labs(x = "", y = "Mean Decrease in Accuracy") + 
+    scale_color_manual(name = "", values = c('#9B30FF', '#63B8FF', '#FF7F00', '#6C7B8B')) + 
+    theme(plot.title = element_text(face="bold", hjust = -0.35, size = 20), 
[…]

### 2018-07-24 code/figureS1.R

@@ -61,7 +61,8 @@ tax <- read_tsv("data/process/final.taxonomy") %>%
          genus = str_replace_all(genus, "_", " "))
 
 combined_imp_class_summary <- classification_imp_otus %>% 
-  left_join(select(tax, OTU, genus), by = c("otu" = "OTU"))
+  left_join(select(tax, OTU, family), by = c("otu" = "OTU")) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""))
 
 
 
@@ -121,27 +122,115 @@ auc_classification <- classification_model_data %>%
         axis.text.x = element_text(size = 10, face = "bold"))
 
 # Plot the top 10 OTUs in the model graph
-imp_otus_classification <- combined_imp_class_summary %>% 
+acetate_imp_graph <- combined_imp_class_summary %>% 
   ungroup() %>% 
-  mutate(scfa = factor(scfa, 
-                       levels = c("acetate", "butyrate", "propionate"), 
-                       labels = c("Acetate", "Butyrate", "Propionate"))) %>% 
-  ggplot(aes(scfa, median_mda, color = genus, group = otu)) + 
+  filter(scfa == "acetate") %>% 
+  mutate(otu = factor(otu, 
+                      levels = rev(otu), 
+                      labels = rev(otu)), 
+         family = case_when(
+           family %in% c("Firmicutes", "Pasteurellaceae", "Verrucomicrobiaceae") ~ "Other", 
+           TRUE ~ family), 
+         family = factor(family, 
+                         levels = c("Clostridiales", "Lachnospiraceae", "Ruminococcaceae", "Other"), 
+                         labels = c("Clostridiales", "Lachnospiraceae", "Ruminococcaceae", "Other"))) %>% 
+  ggplot(aes(otu, median_mda, color = family)) + 
   geom_pointrange(aes(ymin = min_mda, ymax = max_mda), position = position_dodge(width = 1), size = 1) + 
-  geom_vline(xintercept=seq(1.5, length(unique(classification_imp_otus$scfa))-0.5, 1), 
-             lwd=1, colour="gray", alpha = 0.6) + 
-  theme_bw() + labs(x = "", y = "Median MDA") + ggtitle("C") + 
-  scale_color_manual(name = "", values = c("#B0171F", "#FF6EB4", "#9B30FF", "#4169E1", 
-                                           "#63B8FF", "#FFD700", "#00FF00", "#FF7F00")) + 
[…]

### 2018-07-24 code/make_disease_group_RF_imp_graph.R
add imp otus to figure

@@ -61,11 +61,7 @@ all_adn <- all_data %>%
         axis.title = element_text(size = 12),
         axis.text.x = element_text(size = 10, face = "bold"))
 
-#values = c('#9B30FF', '#63B8FF', '#008080', '#8FBC8F', '#FF7F00', '#6C7B8B')
-#
-#values = c("#FF6EB4", "#9B30FF", "#4B0082", "#4169E1", 
-#           "#63B8FF", "#FFD700", "#00FF00", "#FF7F00")
-  
+
 only_otu_adn <- all_data %>% 
   filter(model == "adn_otu") %>% 
   mutate(Variable = factor(Variable, 
@@ -121,12 +117,6 @@ all_crc <- all_data %>%
         axis.title = element_text(size = 12),
         axis.text.x = element_text(size = 10, face = "bold"))
 
-
-
-#values = c('#9B30FF', '#63B8FF', '#008080', '#8FBC8F', '#FF7F00', '#6C7B8B')
-#
-#values = c("#FF6EB4", "#9B30FF", "#4B0082", "#4169E1", 
-#           "#63B8FF", "#FFD700", "#00FF00", "#FF7F00")
 
 
 only_otu_crc <- all_data %>% 

### 2018-07-24 code/make_rf_auc_graphs.R

@@ -21,6 +21,37 @@ crc_summary <- read_csv("data/process/tables/crc_full_AUC_model_summary.csv") %>
   bind_rows(read_csv("data/process/tables/crc_otu_only_AUC_model_summary.csv") %>% 
               mutate(model = "otu"))
 
+# Load needed libraries
+loadLibs(c("tidyverse", "gridExtra"))
+
+taxonomy <- read_tsv("data/process/final.taxonomy") %>% 
+  select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";")
+
+all_data <- read_csv("data/process/tables/adn_full_MDA_Summary.csv") %>% 
+  mutate(model = "adn_full") %>% 
+  filter(Variable != "isobutyrate") %>% 
+  slice(1:10) %>% 
+  bind_rows(read_csv("data/process/tables/crc_full_MDA_Summary.csv") %>% 
+              mutate(model = "crc_full") %>% 
+              filter(Variable != "isobutyrate") %>% 
+              slice(1:10), 
+            read_csv("data/process/tables/adn_otu_only_MDA_Summary.csv") %>% 
+              mutate(model = "adn_otu") %>% 
+              slice(1:10), 
+            read_csv("data/process/tables/crc_otu_only_MDA_Summary.csv") %>% 
+              mutate(model = "crc_otu") %>% 
+              slice(1:10)) %>% 
+  left_join(select(taxonomy, OTU, family), by = c("Variable" = "OTU")) %>% 
+  mutate(family = case_when(
+    Variable %in% c("acetate", "butyrate", "propionate") ~ str_to_title(Variable), 
+    TRUE ~ family), 
+    family = str_replace(family, "_unclassified", ""), 
+    family = str_replace_all(family, "_", " "), 
+    Variable = case_when(
+      Variable %in% c("acetate", "butyrate", "propionate") ~ str_to_title(Variable), 
+      TRUE ~ Variable))
 
 
 ##############################################################################################
@@ -60,7 +91,7 @@ carcinoma <- crc_summary %>%
   coord_cartesian(ylim = c(0, 1)) + 
   geom_hline(yintercept = 0.5, linetype = "dashed") + 
   labs(x = "", y = "Area Under the Curve") + 
[…]

### 2018-07-24 code/figure5.R
reorganize fig orders

@@ -306,4 +306,4 @@ prediction_plot <- grid.arrange(r2_classification, test_data_graph,
                                 layout_matrix = rbind(c(1, 2, 2), c(3, 4, 5)))
 
 # Write out to specific directory
-ggsave("results/figures/Figure4.pdf", prediction_plot, width = 11, height = 8)
+ggsave("results/figures/Figure5.pdf", prediction_plot, width = 11, height = 8)

### 2018-07-24 code/figureS2.R

@@ -268,7 +268,7 @@ prediction_plot <- grid.arrange(auc_classification, test_data_graph,
                                 layout_matrix = rbind(c(1, 2, 2), c(3, 4, 5)))
 
 # Write out to specific directory
-ggsave("results/figures/FigureS1.pdf", prediction_plot, width = 11, height = 8)
+ggsave("results/figures/FigureS2.pdf", prediction_plot, width = 11, height = 8)
 
 
 

### 2018-07-24 code/make_class_group_scfa_sig_graph.R

@@ -62,5 +62,5 @@ number_class_plot <- graph_data %>%
         axis.text.x = element_text(size = 10, face = "bold"))
 
 
-ggsave("results/figures/FigureS2.pdf", number_class_plot, width = 8, height = 8)
+ggsave("results/figures/FigureS1.pdf", number_class_plot, width = 8, height = 8)
 

### 2018-07-24 code/make_pi_opf_combined_graph.R

@@ -77,5 +77,5 @@ metagenome_fig <- opf_data %>%
 gene_plot <- grid.arrange(imputed_fig, metagenome_fig, 
                           layout_matrix = rbind(c(1, 1, 2)))
 # Write out to specific directory
-ggsave("results/figures/Figure3.pdf", gene_plot, width = 9, height = 5)
+ggsave("results/figures/Figure2.pdf", gene_plot, width = 9, height = 5)
 

### 2018-07-24 code/make_reg_group_scfa_sig_graph.R

@@ -68,6 +68,6 @@ number_corr_plot <- graph_data %>%
         axis.title = element_text(size = 12), 
         axis.text.x = element_text(size = 10, face = "bold"))
 
-ggsave("results/figures/Figure5.pdf", number_corr_plot, width = 7, height = 8)
+ggsave("results/figures/Figure3.pdf", number_corr_plot, width = 7, height = 8)
 
 

### 2018-07-24 code/make_rf_auc_graphs.R

@@ -232,4 +232,4 @@ prediction_plot <- grid.arrange(adenoma, all_adn, only_otu_adn,
                                 layout_matrix = rbind(c(1, 2, 3), c(4, 5, 6)))
 
 # Write out to specific directory
-ggsave("results/figures/Figure2.pdf", prediction_plot, width = 15, height = 8)
+ggsave("results/figures/Figure4.pdf", prediction_plot, width = 15, height = 8)

### 2018-07-24 submission/manuscript.Rmd
change results order and add new section

@@
 
+**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
+First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
+This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 2A].
+Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
+This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 2B].
+These observations provide evidence that gene prevalence does not change due to colorectal tumors.
+
+**The number of significant positive correlations between OTU relative abundance and SCFA concentration are similar for individuals without tumors and patients with adenoma or carcinoma tumors.** Having found no difference in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
+Using Spearmans rho we found that the majority of OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5].
+A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6].
+There was a noticeablely higher number of significant negative correlations associated with patients with adenoma versus the other groups for all SCFAs tested [Figure 3].
+In particular, OTUs from the *Ruminococcaceae* family seemed to have the largest share of these negative correlations within patients with adenoma tumors [Figure 3].
+The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3].
+Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3].
+Collectively, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
+
 **Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
@@
 Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
-For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2].
+For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
 Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor.
 This would be consistent with previous 16S rRNA gene sequencing results where many taxa associated with SCFA production are decreased [@Shah2017; @meta_analysis_crc_Sze2018].
-
-**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
-First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
-This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A].
-Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
-Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
-This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B].
-These observations provide evidence that gene prevalence does not change due to colorectal tumors.
 

### 2018-07-25 submission/manuscript.Rmd
put down new results section

@@
 Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
-Even though there was no difference in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.
 
@@
 Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3].
-Collectively, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
+Overall, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
 
-**Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
+**SCFA concentrations do not replace all important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
 Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors.
 Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
-For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
-Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor.
-This would be consistent with previous 16S rRNA gene sequencing results where many taxa associated with SCFA production are decreased [@Shah2017; @meta_analysis_crc_Sze2018].
+For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D].
+There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F].
+The reason for this seems to be because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C].
+The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining [Figure 4B & 4C].
+In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.
 
-**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations.
-Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A].
-However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure S1A].
-The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were taxa that are normally associated with SCFA production [Figure S1B].
-These results highlight that within our data, SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate.
-Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations.
-Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal tumors.
+**Random Forest models for SCFA concentrations have different top 10 most important OTUs than disease models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher than median and lower than median SCFA concentrations.
+Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high and low SCFA concentrations [Figure 5A & S2A].
+However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [[Figure 5A & S2A].
+Additionally, there were differences classification for both models based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B].
+There was minimal overlap between these model's most important OTUs and those used to classify patients with either adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
+The only OTUs that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
+OTU00167 was in the top 10 most important OTUs for the OTU adenoma model and not in the SCFA and OTU adenoma model, however, acetate and butyrate concentrations were [Figure 4B-C].
+These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
+Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with an adenoma or carcinoma tumor.
 

### 2018-07-25 submission/manuscript.tex
put down new results section

@@
 post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
-or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even
-though there was no difference in SCFA concentrations between
-individuals with normal colons and those with tumors, this information
-could still be important to help classify disease.
-
-\textbf{Random Forest models with SCFA concentrations do not classify
-tumors better.} SCFA concentrations could improve prediction of tumors
-based on specific bacterial community structures. Our OTU data can be
-used in combination with SCFAs to assess whether there is a community
-dependent context to SCFA classification of tumors. Using the Random
-Forest algorithm we built models with OTU abundance data only or OTU
-abundances and SCFA concentrations to classify normal versus adenoma and
-normal versus carcinoma fecal samples. For adenoma and carcinoma models,
-there was no difference between the median AUC of models with or without
-SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
-Although including SCFA concentrations did not improve classification of
-colorectal tumors using Random Forest models, it is possible that the
-genes for enzymes involved in SCFA synthesis may vary based on the type
-of colorectal tumor. This would be consistent with previous 16S rRNA
-gene sequencing results where many taxa associated with SCFA production
-are decreased (10, 11).
+or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
 
@@
 similarity between groups is highlighted by visualizing genes important
-in butyrate synthesis {[}Figure 3A{]}. Using a paired Wilcoxon rank-sum
+in butyrate synthesis {[}Figure 2A{]}. Using a paired Wilcoxon rank-sum
 test, there also was no difference in imputed gene relative abundance
@@
 0.70) {[}Table S4{]}. This lack of difference is highlighted when we
-visualize the results for butyrate kinase {[}Figure 3B{]}. These
+visualize the results for butyrate kinase {[}Figure 2B{]}. These
 observations provide evidence that gene prevalence does not change due
@@
 
-\textbf{Expected taxa are associated with higher SCFA concentrations
-regardless of tumor status.} Using OTU data we built Random Forest
-models to classify higher than median and lower than median SCFA
-concentrations. Overall, OTU data had a reasonable ability to classify
-high and low SCFA concentrations {[}Figure S1A{]}. However, these models
-tended to be over fit, suggesting that rarer taxa may be important for
-this classification {[}Figure S1A{]}. The most important OTUs to these
-models (assessed with mean decrease in accuracy (MDA)) were taxa that
-are normally associated with SCFA production {[}Figure S1B{]}. These
-results highlight that within our data, SCFA concentrations are
-associated with taxa known to produce acetate, butyrate, and propionate.
-Additionally, OTUs associated with these taxa are the most important to
-models that can classify high and low SCFA concentrations. Overall, our
-results are robust and do not support the hypothesis that differences in
-SCFA concentrations are associated with colorectal tumors.
+\textbf{The number of significant positive correlations between OTU
+relative abundance and SCFA concentration are similar for individuals
+without tumors and patients with adenoma or carcinoma tumors.} Having
+found no difference in SCFA concentrations or genes coding for enzymes
+involved with SCFA synthesis, we next investigated if specific OTUs
+correlated with SCFA concentrations. Using Spearmans rho we found that
+the majority of OTU correlations were to taxa from \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} {[}Figure 3 \& Table
+S5{]}. A similar pattern was observed when using high/low SCFA groups
+based on the overall median concentration for that specific SCFA
+{[}Figure S1 \& Table S6{]}. There was a noticeablely higher number of
+significant negative correlations associated with patients with adenoma
+versus the other groups for all SCFAs tested {[}Figure 3{]}. In
+particular, OTUs from the \emph{Ruminococcaceae} family seemed to have
+the largest share of these negative correlations within patients with
+adenoma tumors {[}Figure 3{]}. The number of positive correlations
+between OTUs and SCFA concentrations was similar between individuals
+without tumors and patients with a carcinoma {[}Figure 3{]}. Although
+patients with adenomas had more positive correlations between OTUs and
+SCFA concentrations, their total number was more similar to the other
+groups than the number of negative correlations {[}Figure 3{]}. Overall,
+these results suggest that the resident taxa that may change the most
+due to carcinoma tumors may not be ones that are responsible for the
+production of acetate, butyrate, or propionate.
+
+\textbf{SCFA concentrations do not replace all important
+\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
+OTUs in Random Forest models built to classify tumors.} SCFA
+concentrations could improve prediction of tumors based on specific
+bacterial community structures. Our OTU data can be used in combination
+with SCFAs to assess whether there is a community dependent context to
+SCFA classification of tumors. Using the Random Forest algorithm we
+built models with OTU abundance data only or OTU abundances and SCFA
+concentrations to classify normal versus adenoma and normal versus
+carcinoma fecal samples. For adenoma and carcinoma models, there was no
+difference between the median AUC of models with or without SCFA
+concentrations (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
+There was little difference between the top 10 most important OTUs, as
+measured by mean decrease in accuracy, in models using SCFA and OTUs
+versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The reason for this seems to
+be because SCFA concentrations replaced the information gained by
+specific OTUs rather than add new information to the model {[}Figure
+4B-C{]}. The SCFA concentrations also only replaced select OTUs with
+taxonomic classification to \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} with many OTUs within
+these taxa remaining {[}Figure 4B \& 4C{]}. In combination with the
+previous results on taxa correlations, these observations provide
+additional evidence that the resident taxa within models that classify
+tumor are not ones associated with acetate, butyrate, or propionate
+production.
+
+\textbf{Random Forest models for SCFA concentrations have different top
+10 most important OTUs than disease models.} Using OTU data we built
+Random Forest models to classify either SCFA concentration or higher
+than median and lower than median SCFA concentrations. Overall, OTU data
+had a reasonable ability to classify both SCFA concentrations and high
+and low SCFA concentrations {[}Figure 5A \& S2A{]}. However, these
+models tended to be over fit, suggesting that rarer taxa may be
+important for this classification {[}{[}Figure 5A \& S2A{]}.
+Additionally, there were differences classification for both models
+based on whether the individual had no tumors, an adenoma, or a
+carcinoma {[}Figure 5B \& S2B{]}. There was minimal overlap between
+these model's most important OTUs and those used to classify patients
+with either adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and
+S2C-E{]}. The only OTUs that did overlap between the models was OTU00167
+(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
+OTU00167 was in the top 10 most important OTUs for the OTU adenoma model
+and not in the SCFA and OTU adenoma model, however, acetate and butyrate
+concentrations were {[}Figure 4B-C{]}. These observations provide
+further evidence that it is possible to identify specific OTUs
+associated with higher SCFA concentrations and that these OTUs belong to
+taxa known to produce acetate, butyrate, and propionate. Although it is
+possible to identify OTUs associated with SCFA production, our results
+do not support the hypothesis that SCFA concentration or OTUs associated
+with their production are different between individuals with no tumors
+and patients with an adenoma or carcinoma tumor.
 

### 2018-07-25 submission/manuscript.Rmd
update fig legends

@@
 **Figure 2.
-SCFAs do not improve OTU-based Random Forest models.** The area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
-The area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (B).
-The black line represents the median AUC.
-The dotted line highlights an AUC of 0.5.
-
-**Figure 3.
 No change in butyrate producing genes identified between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A).
@@
 The other butyrate pathway genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
+
+**Figure 3.
+Patients with adenomas had the higest number of significant negative correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
+Fewer significant positive correlations were observed overall.
+Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.
+
 **Figure 4.
+SCFA concentrations do not improve OTU-based Random Forest models.** The area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
+The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma model (B).
+The top 10 most important OTUs in the OTU adenoma model (C).
+The area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (D).
+The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma model (E).
+The top 10 most important OTUs in the OTU carcinoma model (F).
+For (A) and (D) the black line represents the median AUC.
+The dotted line highlights an AUC of 0.5.
+
+**Figure 5.
 OTU-based regression Random Forest models of SCFA concentrations.** The train and test correlation between actual and predicted values from 100 different 80/20 split OTU-based models with 10-fold CV using regression Random Forest (A).
@@
 
-**Figure 5.
-Patients with adenomas had the higest number of significant negative correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
-Fewer significant positive correlations were observed overall.
-Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.
-
 \newpage
@@
 **Figure S1.
+Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
+Fewer significant OTUs were observed in individuals without tumors (normal) and patients with carcinomas versus patients with adenomas.
+
+**Figure S2.
 OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
@@
 The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
-
-**Figure S2.
-Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
-Fewer significant OTUs were observed in individuals without tumors (normal) and patients with carcinomas versus patients with adenomas.

### 2018-07-25 submission/manuscript.tex
update fig legends

@@
 
-\textbf{Figure 2. SCFAs do not improve OTU-based Random Forest models.}
-The area under the curve of 100 different 80/20 split OTU-based normal
-versus adenoma 10-fold CV models with and without SCFAs (A). The area
-under the curve of 100 different 80/20 OTU-based normal versus carcinoma
-10-fold CV models with and without SCFAs (B). The black line represents
-the median AUC. The dotted line highlights an AUC of 0.5.
-
-\textbf{Figure 3. No change in butyrate producing genes identified
+\textbf{Figure 2. No change in butyrate producing genes identified
 between normal, adenoma, and carcinoma individuals.} Imputed gene
@@
 from the PICRUSt analysis did not align to any of the OPFs in the
-metagenome analysis.\\
-\textbf{Figure 4. OTU-based regression Random Forest models of SCFA
+metagenome analysis.
+
+\textbf{Figure 3. Patients with adenomas had the higest number of
+significant negative correlations between OTU relative abundance and
+SCFA concentration.} Colors denote the family or lowest taxonomic ID
+that an OTU classified to. Fewer significant positive correlations were
+observed overall. Additionally, the differences in the number of
+significant positive correlations between patients with adenomas versus
+individuals without tumors (normal) and patients with carcinomas was not
+as pronounced as the number of significant negative correlations.
+
+\textbf{Figure 4. SCFA concentrations do not improve OTU-based Random
+Forest models.} The area under the curve of 100 different 80/20 split
+OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
+(A). The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma
+model (B). The top 10 most important OTUs in the OTU adenoma model (C).
+The area under the curve of 100 different 80/20 OTU-based normal versus
+carcinoma 10-fold CV models with and without SCFAs (D). The top 10 most
+important OTUs or SCFAs in the SCFA and OTU carcinoma model (E). The top
+10 most important OTUs in the OTU carcinoma model (F). For (A) and (D)
+the black line represents the median AUC. The dotted line highlights an
+AUC of 0.5.
+
+\textbf{Figure 5. OTU-based regression Random Forest models of SCFA
 concentrations.} The train and test correlation between actual and
@@
 
-\textbf{Figure 5. Patients with adenomas had the higest number of
-significant negative correlations between OTU relative abundance and
-SCFA concentration.} Colors denote the family or lowest taxonomic ID
-that an OTU classified to. Fewer significant positive correlations were
-observed overall. Additionally, the differences in the number of
-significant positive correlations between patients with adenomas versus
-individuals without tumors (normal) and patients with carcinomas was not
-as pronounced as the number of significant negative correlations.
-
 \newpage
 
-\textbf{Figure S1. OTU-based classification Random Forest models of
+\textbf{Figure S1. Patients with adenomas had the higest number of
+significant differences in OTU relative abundance between high/low SCFA
+groups.} Colors denote the family or lowest taxonomic ID that an OTU
+classified to. Fewer significant OTUs were observed in individuals
+without tumors (normal) and patients with carcinomas versus patients
+with adenomas.
+
+\textbf{Figure S2. OTU-based classification Random Forest models of
 high/low SCFA groups based on overall SCFA median concentration.} The
@@
 
-\textbf{Figure S2. Patients with adenomas had the higest number of
-significant differences in OTU relative abundance between high/low SCFA
-groups.} Colors denote the family or lowest taxonomic ID that an OTU
-classified to. Fewer significant OTUs were observed in individuals
-without tumors (normal) and patients with carcinomas versus patients
-with adenomas.
-
 \end{document}

### 2018-07-25 submission/manuscript.Rmd
make new edits and streamline results section

@@
 We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples.
-Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
+Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
 
@@
 Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
-This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 2B].
-These observations provide evidence that gene prevalence does not change due to colorectal tumors.
+This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 2B].
+These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors.
 
-**The number of significant positive correlations between OTU relative abundance and SCFA concentration are similar for individuals without tumors and patients with adenoma or carcinoma tumors.** Having found no difference in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
-Using Spearmans rho we found that the majority of OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5].
+**Total significant positive correlations between OTU relative abundance and SCFA concentration were similar for individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
+Using Spearmans rho we found that the majority of significant OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5].
 A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6].
-There was a noticeablely higher number of significant negative correlations associated with patients with adenoma versus the other groups for all SCFAs tested [Figure 3].
-In particular, OTUs from the *Ruminococcaceae* family seemed to have the largest share of these negative correlations within patients with adenoma tumors [Figure 3].
+There was a noticeablely higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3].
+In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas  [Figure 3].
+Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
 The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3].
-Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3].
-Overall, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
+Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
 
-**SCFA concentrations do not replace all important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
+**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
 Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors.
-Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
+Using the Random Forest algorithm we built models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
 For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D].
 There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F].
-The reason for this seems to be because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C].
-The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining [Figure 4B & 4C].
+The main reason for this was because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C].
+The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining in the top 10 [Figure 4B & 4C].
 In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.
 
-**Random Forest models for SCFA concentrations have different top 10 most important OTUs than disease models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher than median and lower than median SCFA concentrations.
-Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high and low SCFA concentrations [Figure 5A & S2A].
-However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [[Figure 5A & S2A].
-Additionally, there were differences classification for both models based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B].
-There was minimal overlap between these model's most important OTUs and those used to classify patients with either adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
-The only OTUs that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
-OTU00167 was in the top 10 most important OTUs for the OTU adenoma model and not in the SCFA and OTU adenoma model, however, acetate and butyrate concentrations were [Figure 4B-C].
+**Random Forest models for SCFA concentrations have different top 10 most important OTUs than tumor models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher/lower than median SCFA concentrations.
+Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high/low SCFA groups [Figure 5A & S2A].
+However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure 5A & S2A].
+Additionally, there were differences in accuracy for both model types based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B].
+There also was minimal overlap between these SCFA classification model's most important OTUs and those used to classify patients with adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
+The only OTU that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
+Additionally, OTU00167 was in the top 10 most important OTUs for the OTU adenoma model but not in the SCFA and OTU adenoma model while acetate and butyrate concentrations were [Figure 4B-C].
 These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
-Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with an adenoma or carcinoma tumor.
+Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with adenomas or carcinomas.
 

### 2018-07-25 submission/manuscript.tex
make new edits and streamline results section

@@
 after treatment for tumors, there was no significant difference pre- and
-post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
-or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
+post-treatment for patients with adenomas (P-value \textgreater{} 0.20)
+or carcinomas (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
 
@@
 patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
-0.70) {[}Table S4{]}. This lack of difference is highlighted when we
+0.70) {[}Table S4{]}. This lack of difference can be highlighted when we
 visualize the results for butyrate kinase {[}Figure 2B{]}. These
-observations provide evidence that gene prevalence does not change due
-to colorectal tumors.
+observations provide evidence that gene prevalence for enzymes involved
+in SCFA synthesis does not change due to colorectal tumors.
 
-\textbf{The number of significant positive correlations between OTU
-relative abundance and SCFA concentration are similar for individuals
-without tumors and patients with adenoma or carcinoma tumors.} Having
-found no difference in SCFA concentrations or genes coding for enzymes
+\textbf{Total significant positive correlations between OTU relative
+abundance and SCFA concentration were similar for individuals without
+tumors and patients with adenomas or carcinomas.} Having found no
+difference between individuals without tumors and patients with adenomas
+or carcinomas in SCFA concentrations or genes coding for enzymes
 involved with SCFA synthesis, we next investigated if specific OTUs
 correlated with SCFA concentrations. Using Spearmans rho we found that
-the majority of OTU correlations were to taxa from \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} {[}Figure 3 \& Table
-S5{]}. A similar pattern was observed when using high/low SCFA groups
-based on the overall median concentration for that specific SCFA
-{[}Figure S1 \& Table S6{]}. There was a noticeablely higher number of
-significant negative correlations associated with patients with adenoma
-versus the other groups for all SCFAs tested {[}Figure 3{]}. In
-particular, OTUs from the \emph{Ruminococcaceae} family seemed to have
-the largest share of these negative correlations within patients with
-adenoma tumors {[}Figure 3{]}. The number of positive correlations
-between OTUs and SCFA concentrations was similar between individuals
-without tumors and patients with a carcinoma {[}Figure 3{]}. Although
-patients with adenomas had more positive correlations between OTUs and
-SCFA concentrations, their total number was more similar to the other
-groups than the number of negative correlations {[}Figure 3{]}. Overall,
-these results suggest that the resident taxa that may change the most
-due to carcinoma tumors may not be ones that are responsible for the
-production of acetate, butyrate, or propionate.
+the majority of significant OTU correlations were to taxa from
+\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
+{[}Figure 3 \& Table S5{]}. A similar pattern was observed when using
+high/low SCFA groups based on the overall median concentration for that
+specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeablely
+higher number of significant negative correlations associated with
+patients with adenomas for all SCFAs tested {[}Figure 3{]}. In
+particular, OTUs from the \emph{Ruminococcaceae} family had the largest
+share of these negative correlations within patients with adenomas
+{[}Figure 3{]}. Although patients with adenomas had more positive
+correlations between OTUs and SCFA concentrations, their total number
+was more similar to individuals without tumors or patients with
+carcinomas versus the analogous comparison for the number of negative
+correlations {[}Figure 3{]}. The number of positive correlations between
+OTUs and SCFA concentrations was similar between individuals without
+tumors and patients with a carcinoma {[}Figure 3{]}. Overall, these
+results suggest that the resident taxa that may change the most due to
+colon tumors may not be ones that are responsible for the production of
+acetate, butyrate, or propionate.
 
-\textbf{SCFA concentrations do not replace all important
+\textbf{SCFA concentrations do not replace important
 \emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
@@
 SCFA classification of tumors. Using the Random Forest algorithm we
-built models with OTU abundance data only or OTU abundances and SCFA
+built models with OTU abundance data or OTU abundances and SCFA
 concentrations to classify normal versus adenoma and normal versus
@@
 measured by mean decrease in accuracy, in models using SCFA and OTUs
-versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The reason for this seems to
-be because SCFA concentrations replaced the information gained by
-specific OTUs rather than add new information to the model {[}Figure
-4B-C{]}. The SCFA concentrations also only replaced select OTUs with
-taxonomic classification to \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} with many OTUs within
-these taxa remaining {[}Figure 4B \& 4C{]}. In combination with the
-previous results on taxa correlations, these observations provide
-additional evidence that the resident taxa within models that classify
-tumor are not ones associated with acetate, butyrate, or propionate
-production.
+versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The main reason for this was
+because SCFA concentrations replaced the information gained by specific
+OTUs rather than add new information to the model {[}Figure 4B-C{]}. The
+SCFA concentrations also only replaced select OTUs with taxonomic
+classification to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} with many OTUs within these taxa remaining in the
+top 10 {[}Figure 4B \& 4C{]}. In combination with the previous results
+on taxa correlations, these observations provide additional evidence
+that the resident taxa within models that classify tumor are not ones
+associated with acetate, butyrate, or propionate production.
 
 \textbf{Random Forest models for SCFA concentrations have different top
-10 most important OTUs than disease models.} Using OTU data we built
-Random Forest models to classify either SCFA concentration or higher
-than median and lower than median SCFA concentrations. Overall, OTU data
-had a reasonable ability to classify both SCFA concentrations and high
-and low SCFA concentrations {[}Figure 5A \& S2A{]}. However, these
-models tended to be over fit, suggesting that rarer taxa may be
-important for this classification {[}{[}Figure 5A \& S2A{]}.
-Additionally, there were differences classification for both models
-based on whether the individual had no tumors, an adenoma, or a
-carcinoma {[}Figure 5B \& S2B{]}. There was minimal overlap between
-these model's most important OTUs and those used to classify patients
-with either adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and
-S2C-E{]}. The only OTUs that did overlap between the models was OTU00167
+10 most important OTUs than tumor models.} Using OTU data we built
+Random Forest models to classify either SCFA concentration or
+higher/lower than median SCFA concentrations. Overall, OTU data had a
+reasonable ability to classify both SCFA concentrations and high/low
+SCFA groups {[}Figure 5A \& S2A{]}. However, these models tended to be
+over fit, suggesting that rarer taxa may be important for this
+classification {[}Figure 5A \& S2A{]}. Additionally, there were
+differences in accuracy for both model types based on whether the
+individual had no tumors, an adenoma, or a carcinoma {[}Figure 5B \&
+S2B{]}. There also was minimal overlap between these SCFA classification
+model's most important OTUs and those used to classify patients with
+adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
+The only OTU that did overlap between the models was OTU00167
 (\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
-OTU00167 was in the top 10 most important OTUs for the OTU adenoma model
-and not in the SCFA and OTU adenoma model, however, acetate and butyrate
-concentrations were {[}Figure 4B-C{]}. These observations provide
-further evidence that it is possible to identify specific OTUs
+Additionally, OTU00167 was in the top 10 most important OTUs for the OTU
+adenoma model but not in the SCFA and OTU adenoma model while acetate
+and butyrate concentrations were {[}Figure 4B-C{]}. These observations
+provide further evidence that it is possible to identify specific OTUs
 associated with higher SCFA concentrations and that these OTUs belong to
@@
 with their production are different between individuals with no tumors
-and patients with an adenoma or carcinoma tumor.
+and patients with adenomas or carcinomas.
 

### 2018-07-26 submission/manuscript.Rmd
update first paragraph of discussion

@@
 The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors.
-Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3].
-There is an intriguing reason why taxa associated with SCFA production are decreased in CRC but the genes involoved with its' production are not.
-Mouth-associated microbes such as *Fusobacterium nucleatum* and *Porphyromonas asaccharolytica* have been found to be increased in patients with carcinomas versus individuals without tumors [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Both of these bacterial species are known to have strains that can produce SCFAs such as butyrate [@Vital2014].
-Thus the reason we may be observing no change in genes involved with SCFA synthesis, as well as no change to SCFAs themselves, is because the production is being supported by more inflammatory microbes associated with CRC.
-Additionally, our observations that no benefit could be found in using the concentrations to help classify individuals with and without tumors would be consistent with this reason [Figure 2].
-However, our observations are in stark contrast to some of the previous literature.
+Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2].
+Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1].
+Instead, SCFA concentrations provided similar information to what specific OTUs were already providing to the tumor classification models [Figure 4].
+Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to  these taxa in the tumor models [Figure 4-5].
+Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carinomas, are not that same as those involved with SCFA production.
 

### 2018-07-26 submission/manuscript.tex
update first paragraph of discussion

@@
 associated with their production, no difference could be identified
-{[}Figure 1 \& 3{]}. There is an intriguing reason why taxa associated
-with SCFA production are decreased in CRC but the genes involoved with
-its' production are not. Mouth-associated microbes such as
-\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asaccharolytica}
-have been found to be increased in patients with carcinomas versus
-individuals without tumors (10, 11, 23). Both of these bacterial species
-are known to have strains that can produce SCFAs such as butyrate (24).
-Thus the reason we may be observing no change in genes involved with
-SCFA synthesis, as well as no change to SCFAs themselves, is because the
-production is being supported by more inflammatory microbes associated
-with CRC. Additionally, our observations that no benefit could be found
-in using the concentrations to help classify individuals with and
-without tumors would be consistent with this reason {[}Figure 2{]}.
-However, our observations are in stark contrast to some of the previous
-literature.
+between individuals without tumors and patients with adenomas or
+carcinomas {[}Figure 1 \& 2{]}. Although there were differences in the
+number of significant correlations between SCFA concentration and OTU
+relative abundance based on whether individuals did not have tumors, had
+an adenoma, or had a carcinoma, SCFA concentrations did not provide
+increased model accuracy for tumor classification {[}Figure 3-4 \&
+S1{]}. Instead, SCFA concentrations provided similar information to what
+specific OTUs were already providing to the tumor classification models
+{[}Figure 4{]}. Additionally, when models using OTU relative abundance
+to classify SCFA concentrations were assessed, the OTUs that classified
+to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} were not the same as the OTUs that classified to
+these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
+observations suggest that resident taxa from \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae}, that are different
+between individuals without tumors and patients with adenomas or
+carinomas, are not that same as those involved with SCFA production.
 
@@
 fiber supplementation to prevent tumor recurrence has found no benefit
-(20, 25). Our results align with what has been reported in
+(20, 23). Our results align with what has been reported in
 randomized-controlled trials, that SCFAs do not provide general
@@
 been that all tumors are treated as the same type. However, there are
-known differences in the types of mutations that occur (26) and treating
+known differences in the types of mutations that occur (24) and treating
 all tumors as equal may actually hide any benefit that could be found in
 specific subsets of individuals. Similar to the idea of specific
-immunotherapy for specific tumors (27), SCFAs may have beneficial
+immunotherapy for specific tumors (25), SCFAs may have beneficial
 effects for specific types of colorectal tumors. Future research will
@@
 changes after fiber supplementation use fecal material to track these
-responses (28). Although there are limitations with the current research
+responses (26). Although there are limitations with the current research
 on SCFAs and colorectal tumors, our observations along with the
@@
 consistently higher in indivdiuals without colon tumors than patients
-with carcinomas (10, 11, 23).
+with carcinomas (10, 11, 27).
 
@@
 SCFAs even though taxa associated with their production are higher in
-individuals without tumors versus patients with carcinomas (10, 11, 23).
+individuals without tumors versus patients with carcinomas (10, 11, 27).
 Protection could be via a different pathway and by extension other
@@
 12, 13)). The idea of niche exclusion is similar to how the community
-protects against \emph{Clostridium difficile} infection (29) with
+protects against \emph{Clostridium difficile} infection (28) with
 chronic inflammation replacing the role of antibiotics. Although we did
@@
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (30, 31). In brief, this study used fecal
+described in detail previously (29, 30). In brief, this study used fecal
 samples obtained at either a single cross-sectional time point (n=490)
@@
 days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (30).
+removal of a tumor with or without chemotherapy and radiation (29).
 Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (30, 31). The University
+histopathological review of biopsies obtained (29, 30). The University
 of Michigan Institutional Review Board approved the study and informed
@@
 protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (28). The following changes to this protocol included the use of
+machine (26). The following changes to this protocol included the use of
 frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (28) and Figure 1 here).
+studies (see Table 1 (26) and Figure 1 here).
 
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-previously described (30, 32, 33). The major differences from these
+previously described (29, 31, 32). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptClust algorithm (34).
+97\% similarity using the OptClust algorithm (33).
 
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (35). Briefly, the sequences were quality
+previously described method (34). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (36). Open Reading Frames (ORFs) were
-identified using Prodigal (37), counts generated using Diamond (38),
+to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
+identified using Prodigal (36), counts generated using Diamond (37),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (39), and OPF alignment used the KEGG database (40).
+mmseq2 (38), and OPF alignment used the KEGG database (39).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (40,
-41) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (39,
+40) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (42). Generally, a Kruskal-Walis rank sum
+the statistical language R (41). Generally, a Kruskal-Walis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
 the groups used. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (43). First, we assessed differences in
+correct for multiple comparisons (42). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 
-\hypertarget{ref-normalization_Sze2017}{}
-23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
-\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
-microbiota in patients after treatment for colonic lesions. Microbiome
-\textbf{5}.
-doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
-
-\hypertarget{ref-Vital2014}{}
-24. \textbf{Vital M}, \textbf{Howe AC}, \textbf{Tiedje JM}. 2014.
-Revealing the bacterial butyrate synthesis pathways by analyzing
-(meta)genomic data. mBio \textbf{5}:e00889--14--e00889--14.
-doi:\href{https://doi.org/10.1128/mbio.00889-14}{10.1128/mbio.00889-14}.
-
 \hypertarget{ref-Schatzkin2000}{}
-25. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+23. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
 \textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
@@
 \hypertarget{ref-Fearon1990}{}
-26. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \hypertarget{ref-Thomas2016}{}
-27. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
 agents targeting the BCR-ABL tyrosine kinase as philadelphia
@@
 \hypertarget{ref-scfa_measures_venkataraman2016}{}
-28. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+26. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 
+\hypertarget{ref-normalization_Sze2017}{}
+27. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
 \hypertarget{ref-Theriot2015}{}
-29. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
+28. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
 the gastrointestinal microbiome and clostridium difficile. Annual Review
@@
 \hypertarget{ref-normalization_sze2017}{}
-30. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+29. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 \hypertarget{ref-crc_model_baxter2016}{}
-31. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+30. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-Schloss2009}{}
-32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-Kozich2013}{}
-33. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+32. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-opticlust_westcott2017}{}
-34. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Hannigan2017}{}
-35. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-36. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-37. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-38. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-39. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-40. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-41. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 \hypertarget{ref-r_citation_2017}{}
-42. \textbf{R Core Team}. 2017. R: A language and environment for
+41. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-43. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-07-27 submission/manuscript.Rmd
second paragraph half changed
update discussion and conclusion sections

@@
 
-Much of the previous research on SCFA benefit to human CRC has been illustrated in model systems [@test_OKeefe2016].
-Many SCFAs are produced through the breakdown of fiber [@Holscher2017] and a recent study in mice found that fiber supplementation increased SCFA concentrations and decreased tumor burden [@Bishehsari2018].
-Additionally, SCFAs such as butyrate can inhibit tumor growth in *in vitro* experiments [@Encarnao2018].
-Yet, observations in humans has been mixed.
-Previous case/control studies found associations with lower SCFA concentrations in individuals with carcinoma tumors [@Ohigashi2013].
-However, individual randomized controlled trials and a recent meta-analysis on fiber supplementation to prevent tumor recurrence has found no benefit [@Schatzkin2000; @Yao2017].
-Our results align with what has been reported in randomized-controlled trials, that SCFAs do not provide general protection against colorectal tumors.
-It is possible though that there are specific instances where SCFAs may be beneficial.
+Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018].
+Additionally, many of the *in vivo* studies investigate proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017].
+Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems.
+In addition to using fiber supplementation as a proxy for SCFA treatment, observations on the benefit of SCFAs in preventing tumorigenesis has been mixed.
+In previous case/control studies lower SCFA concentrations between patients with carcinomas and those without carcinomas was observed [@Ohigashi2013].
+Yet, this is in contrast to multiple randomized-controlled trials that have found no difference between patients who do and do not get fiber supplementation to try and prevent tumor recurrence [@Schatzkin2000; @Yao2017].
+This could be because fiber has other actions in humans besides providing an energy source for bacteria to breakdown to make SCFAs.
+It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
+This information taken together with our observations would suggest that either indviduals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.
 
+In addition to not having a response to fiber supplementation due to lack of the required microbes, it is also possible that there are specific instances of colorectal cancer where SCFAs may be beneficial.
 One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type.
@@
 Future research will need to test if this is a valid hypothesis.
-Another limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
-However, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
-Additionally, studies that measure SCFA changes after fiber supplementation use fecal material to track these responses [@scfa_measures_venkataraman2016].
-Although there are limitations with the current research on SCFAs and colorectal tumors, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective.
+Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.
+
+One possible technical limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
+However, this is unlikely to be a major confounder.
+First, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
+Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016].
+Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be cause of this.
+Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer.
 Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 
-The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Protection could be via a different pathway and by extension other metabolites that have not been extensively studied.
+The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
+Indeed our data would support the contention that the taxa are similar to those associated with SCFA production but that these specific microbes or OTUs themselves are not associated with SCFAs.
+In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis.
+This leads to the possibilty that protection may be through two other routes.
+First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis.
 Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g.
@@
 The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics.
-Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.
+Even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.
 
@@
 Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.
-Overall, these results suggest that the SCFAs typically produced by resident microbes do not protect against tumor.
-By focusing on other types mechanisms, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
+Additionally, these observations suggest that resident microbes that are not involved in SCFA production may be the important resident community members involved with preventing tumorigenesis.
+By focusing on the alternative mechanisms that are associated with these non-SCFA producing resident microbes, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
 

### 2018-07-27 submission/manuscript.tex
update discussion and conclusion sections

@@
 
-Much of the previous research on SCFA benefit to human CRC has been
-illustrated in model systems (16). Many SCFAs are produced through the
-breakdown of fiber (14) and a recent study in mice found that fiber
-supplementation increased SCFA concentrations and decreased tumor burden
-(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
-\emph{in vitro} experiments (17). Yet, observations in humans has been
-mixed. Previous case/control studies found associations with lower SCFA
-concentrations in individuals with carcinoma tumors (19). However,
-individual randomized controlled trials and a recent meta-analysis on
-fiber supplementation to prevent tumor recurrence has found no benefit
-(20, 23). Our results align with what has been reported in
-randomized-controlled trials, that SCFAs do not provide general
-protection against colorectal tumors. It is possible though that there
-are specific instances where SCFAs may be beneficial.
+Although SCFAs have been shown to be anti-tumorigenic, most of these
+studies have been performed in model systems (16, 17). Additionally,
+many of the \emph{in vivo} studies investigate proxies such as fiber
+supplementation rather than SCFAs directly (14). Although it is well
+known that breakdown products from gut bacteria results in SCFA
+production, fiber effects on tumorigenesis may be through other
+mechanisms in these \emph{in vivo} model systems. In addition to using
+fiber supplementation as a proxy for SCFA treatment, observations on the
+benefit of SCFAs in preventing tumorigenesis has been mixed. In previous
+case/control studies lower SCFA concentrations between patients with
+carcinomas and those without carcinomas was observed (19). Yet, this is
+in contrast to multiple randomized-controlled trials that have found no
+difference between patients who do and do not get fiber supplementation
+to try and prevent tumor recurrence (20, 23). This could be because
+fiber has other actions in humans besides providing an energy source for
+bacteria to breakdown to make SCFAs. It could also be due to the fact
+that SCFA concentrations and responses to fiber can vary quite a bit
+between healthy individuals (24). This information taken together with
+our observations would suggest that either indviduals who do not respond
+to fiber supplementation would need to acquire these bacteria to achieve
+a benefit or that SCFAs provide little to no benefit as an
+anti-tumorigenic compound in colorectal cancer.
 
+In addition to not having a response to fiber supplementation due to
+lack of the required microbes, it is also possible that there are
+specific instances of colorectal cancer where SCFAs may be beneficial.
 One limitation of current research into the effect of SCFAs in CRC has
 been that all tumors are treated as the same type. However, there are
-known differences in the types of mutations that occur (24) and treating
+known differences in the types of mutations that occur (25) and treating
 all tumors as equal may actually hide any benefit that could be found in
 specific subsets of individuals. Similar to the idea of specific
-immunotherapy for specific tumors (25), SCFAs may have beneficial
+immunotherapy for specific tumors (26), SCFAs may have beneficial
 effects for specific types of colorectal tumors. Future research will
-need to test if this is a valid hypothesis. Another limitation is that a
-fecal sample may not be an ideal type of biospecimen and that the effect
-SCFAs have on tumorigenesis is only detected in the colon. However, most
-\emph{in vivo} studies as well as human studies have used fecal material
-in their analysis (18, 19). Additionally, studies that measure SCFA
-changes after fiber supplementation use fecal material to track these
-responses (26). Although there are limitations with the current research
-on SCFAs and colorectal tumors, our observations along with the
-randomized controlled trials on fiber supplementation in tumor
-recurrence (20) provide evidence that these specific metabolites may not
-be protective. Yet, taxa that are associated with SCFA production are
-consistently higher in indivdiuals without colon tumors than patients
-with carcinomas (10, 11, 27).
+need to test if this is a valid hypothesis. Regardless of this
+limitation, our results in combination to previous randomized controlled
+trials on fiber supplementation suggests that using SCFAs as a general
+treatment for colorectal cancer is unlikely to provide a reduction in
+tumorigenesis.
+
+One possible technical limitation is that a fecal sample may not be an
+ideal type of biospecimen and that the effect SCFAs have on
+tumorigenesis is only detected in the colon. However, this is unlikely
+to be a major confounder. First, most \emph{in vivo} studies as well as
+human studies have used fecal material in their analysis (18, 19).
+Second, previous studies that measure SCFA changes after fiber
+supplementation use fecal material to track these responses with a great
+deal of success (24). Although there are limitations with the current
+research on SCFAs and colorectal tumors, technical limitations are less
+likely to be cause of this. Additionally, as mentioned earlier, our
+observations along with the randomized controlled trials on fiber
+supplementation in tumor recurrence (20) provide evidence that these
+specific metabolites may not be protective or used as a general
+treatment option in colorectal cancer. Yet, taxa that are associated
+with SCFA production are consistently higher in indivdiuals without
+colon tumors than patients with carcinomas (10, 11, 27).
 
 The potential protection against colorectal cancers may not be from
-SCFAs even though taxa associated with their production are higher in
+SCFAs, even though taxa associated with their production are higher in
 individuals without tumors versus patients with carcinomas (10, 11, 27).
-Protection could be via a different pathway and by extension other
-metabolites that have not been extensively studied. Alternatively,
-protection may not occur via a metabolite but instead through niche
-exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
+Indeed our data would support the contention that the taxa are similar
+to those associated with SCFA production but that these specific
+microbes or OTUs themselves are not associated with SCFAs. In
+particular, our results showing that different OTUs from the same
+taxonomic classification are in tumor and SCFA Random Forest models
+supports this hypothesis. This leads to the possibilty that protection
+may be through two other routes. First, there could be a different
+pathway or other less extensively studied metabolites that provides the
+necessary protection against tumorigenesis. Alternatively, protection
+may not occur via a metabolite but instead through niche exclusion of
+mouth-associated microbes (e.g. \emph{Fusobacterium},
 \emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
@@
 protects against \emph{Clostridium difficile} infection (28) with
-chronic inflammation replacing the role of antibiotics. Although we did
-not find lower concentrations of SCFAs associated with colorectal
+chronic inflammation replacing the role of antibiotics. Even though we
+did not find lower concentrations of SCFAs associated with colorectal
 tumors, we think that there are many exciting new avenues to explore
@@
 randomized controlled trials that have tested fiber use in preventing
-colorectal tumor recurrence. Overall, these results suggest that the
-SCFAs typically produced by resident microbes do not protect against
-tumor. By focusing on other types mechanisms, the identification of more
-promising therapeutic options for use in treating colorectal cancer may
-be found.
+colorectal tumor recurrence. Additionally, these observations suggest
+that resident microbes that are not involved in SCFA production may be
+the important resident community members involved with preventing
+tumorigenesis. By focusing on the alternative mechanisms that are
+associated with these non-SCFA producing resident microbes, the
+identification of more promising therapeutic options for use in treating
+colorectal cancer may be found.
 
@@
 protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (26). The following changes to this protocol included the use of
+machine (24). The following changes to this protocol included the use of
 frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (26) and Figure 1 here).
+studies (see Table 1 (24) and Figure 1 here).
 
@@
 
+\hypertarget{ref-scfa_measures_venkataraman2016}{}
+24. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
+Variable responses of human microbiomes to dietary supplementation with
+resistant starch. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
+
 \hypertarget{ref-Fearon1990}{}
-24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+25. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \hypertarget{ref-Thomas2016}{}
-25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
+26. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
 agents targeting the BCR-ABL tyrosine kinase as philadelphia
@@
 doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
-
-\hypertarget{ref-scfa_measures_venkataraman2016}{}
-26. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
-\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
-Variable responses of human microbiomes to dietary supplementation with
-resistant starch. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
 

### 2018-07-30 submission/manuscript.Rmd
update methods section

@@
 **Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016].
-In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`).
+In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`).
 For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
@@
 **16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017].
+In brief, sequences were quality filtered and contigs created from the paired end reads.
+Any sequences with ambiguous base calls were discarded.
+Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012].
+Chimeras were identifed and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
 The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].
@@
 
-**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification models [@randomforest_citation_2002].
+**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@randomforest_citation_2002; @caret_citation].
 This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model.
@@
 The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.
+The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.
 
@@
 We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment.
-Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models.
 Next, the imputed gene counts of important mediators of SCFA synthesis was tested.
 Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
-Finally, models to classify high or low SCFA concentration based on the median of each SCFA and 16S rRNA gene sequencing data was created using the Random Forest algorithm.
+From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho.
+Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models.
+Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm.
+For all Random Forest models, the assessment of the most important variables was based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in accuracy.
 

### 2018-07-30 submission/manuscript.tex
update methods section

@@
 or from before (pre-) and after (post-) treatment of a patient's tumor
-(n=67). For patients undergoing treatment for their tumor the length of
-time between their initial and follow up sample ranged from 188 - 546
-days. Our use of treatment has been previously defined as encompassing
-removal of a tumor with or without chemotherapy and radiation (29).
-Diagnosis of tumor was made by colonoscopic examination and
-histopathological review of biopsies obtained (29, 30). The University
-of Michigan Institutional Review Board approved the study and informed
-consent was obtained from all participants in accordance to the
-guidelines set out by the Helsinki Declaration.
+(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
+for their tumor the length of time between their initial and follow up
+sample ranged from 188 - 546 days. Our use of treatment has been
+previously defined as encompassing removal of a tumor with or without
+chemotherapy and radiation (29). Diagnosis of tumor was made by
+colonoscopic examination and histopathological review of biopsies
+obtained (29, 30). The University of Michigan Institutional Review Board
+approved the study and informed consent was obtained from all
+participants in accordance to the guidelines set out by the Helsinki
+Declaration.
 
@@
 \textbf{16s rRNA gene sequencing.} The workflow and processing have been
-previously described (29, 31, 32). The major differences from these
+previously described (29, 31, 32). In brief, sequences were quality
+filtered and contigs created from the paired end reads. Any sequences
+with ambiguous base calls were discarded. Contigs were then checked for
+matches to the V4 region of the 16S rRNA gene using the SILVA database
+(33). Chimeras were identifed and removed using UCHIME and OTUs
+clustered at 97\% similarity (34). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptClust algorithm (33).
+97\% similarity using the OptClust algorithm (35).
 
@@
 sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (34). Briefly, the sequences were quality
+previously described method (36). Briefly, the sequences were quality
 filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
-identified using Prodigal (36), counts generated using Diamond (37),
+to contig assembly with MEGAHIT (37). Open Reading Frames (ORFs) were
+identified using Prodigal (38), counts generated using Diamond (39),
 subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (38), and OPF alignment used the KEGG database (39).
+mmseq2 (40), and OPF alignment used the KEGG database (41).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (39,
-40) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (41,
+42) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 the data and then tested on the held out 20\% (80/20 split) using the
-Random Forest algorithm for classification models (21). This was
-repeated on 100 different 80/20 splits of the data to generate a
-reasonable range for the AUC of the model. The reported AUCs, unless
-otherwise specified, are for the test sets. The classification models
-were built to group normal versus adenoma, normal versus carcinoma, and
-high versus low SCFA concentrations.
+Random Forest algorithm for classification and regression models via the
+caret package (21, 43). This was repeated on 100 different 80/20 splits
+of the data to generate a reasonable range for the AUC of the model. The
+reported AUCs, unless otherwise specified, are for the test sets. The
+classification models were built to group normal versus adenoma, normal
+versus carcinoma, and high versus low SCFA concentrations. The
+regression models were built to classify the SCFA concentrations of
+acetate, butyrate, and propionate regardless of disease status.
 
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (41). Generally, a Kruskal-Walis rank sum
+the statistical language R (44). Generally, a Kruskal-Walis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
 the groups used. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (42). First, we assessed differences in
+correct for multiple comparisons (45). First, we assessed differences in
 SCFA concentrations measured by HPLC between individuals with normal
@@
 whether SCFA concentrations changed in patients with an adenoma or
-carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
-alone or OTUs and SCFAs were better able to classify individuals with
-and without tumor using Random Forest models. Next, the imputed gene
-counts of important mediators of SCFA synthesis was tested.
-Additionally, the counts generated for OPFs that matched important genes
-involved with SCFA creation were analyzed. Finally, models to classify
-high or low SCFA concentration based on the median of each SCFA and 16S
-rRNA gene sequencing data was created using the Random Forest algorithm.
+carcinoma pre- versus post-treatment. Next, the imputed gene counts of
+important mediators of SCFA synthesis was tested. Additionally, the
+counts generated for OPFs that matched important genes involved with
+SCFA creation were analyzed. From here we analyzed the number of
+significant positive and negative correlations between OTU relative
+abundance and SCFA concentrations in individuals without tumors and
+patients with adenomas or carcinomas using Spearman's rho. Next, we
+assessed whether OTUs alone or OTUs and SCFAs were better able to
+classify individuals with and without tumors using Random Forest models.
+Finally, models to classify high or low SCFA concentration based on the
+median of each SCFA or the actual concentration using 16S rRNA gene
+sequencing data was created using the Random Forest algorithm. For all
+Random Forest models, the assessment of the most important variables was
+based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in
+accuracy.
 
@@
 
+\hypertarget{ref-Quast2012}{}
+33. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
+J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
+project: Improved data processing and web-based tools. Nucleic Acids
+Research \textbf{41}:D590--D596.
+doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
+
+\hypertarget{ref-Edgar2011}{}
+34. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
+and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+
 \hypertarget{ref-opticlust_westcott2017}{}
-33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+35. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \hypertarget{ref-Hannigan2017}{}
-34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+36. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
 \textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
@@
 \hypertarget{ref-Li2015}{}
-35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+37. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \hypertarget{ref-Hyatt2010}{}
-36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+38. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \hypertarget{ref-Buchfink2014}{}
-37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+39. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \hypertarget{ref-Steinegger2017}{}
-38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+40. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 \hypertarget{ref-Kanehisa2015}{}
-39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+41. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 \hypertarget{ref-scfa_baxter2014}{}
-40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+42. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
 \textbf{Schloss PD}. 2014. Structure of the gut microbiome following
@@
 
+\hypertarget{ref-caret_citation}{}
+43. \textbf{Jed Wing MKC from}, \textbf{Weston S}, \textbf{Williams A},
+\textbf{Keefer C}, \textbf{Engelhardt A}, \textbf{Cooper T},
+\textbf{Mayer Z}, \textbf{Kenkel B}, \textbf{R Core Team},
+\textbf{Benesty M}, \textbf{Lescarbeau R}, \textbf{Ziem A},
+\textbf{Scrucca L}, \textbf{Tang Y}, \textbf{Candan C}, \textbf{Hunt.
+T}. 2017. Caret: Classification and regression training.
+
 \hypertarget{ref-r_citation_2017}{}
-41. \textbf{R Core Team}. 2017. R: A language and environment for
+44. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+45. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple

### 2018-07-30 submission/manuscript.Rmd
update introduction

@@
 
-Colorectal cancer (CRC) is currently the third highest cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
-Although there is a genetic component to the disease, the environment is attributed to being a larger risk factor for CRC [@Lichtenstein2000].
+Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
+Although there is a genetic component to the disease, the environment has been attributed to being a larger risk factor for CRC [@Lichtenstein2000].
 These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011].
-Many of these environmental risk factors are capable of being modified, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018].
-Multiple reports in case/control studies have identified bacterial taxa commonly associated with individuals without tumors to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018].
-Many of these taxa within individuals wihtout tumors actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017].
+Many of these environmental risk factors, including the microbiota, could be easily modifiable, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018].
+Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018].
+Many of the bacterial species from these resident taxa actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017].
 The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016].
-Overall, the specific bacterial taxa of the microbiota that create SCFAs are an attractive target to modulate the risk of CRC.
+Ultimately, these SCFAs could be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.
 
-Specific SCFAs, like butyrate, have shown positive results for CRC treatment within model systems [@test_OKeefe2016].
-Butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
-Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects.
-For example, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
-Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.
+Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent.
+Specific SCFAs, like butyrate, have shown positive results within model systems [@test_OKeefe2016].
+For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
+Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
+These exciting results in model systems suggests that supplementation with food sources that bacteria use to create these SCFAs may be able to confer beneficial effects against CRC.
+However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.
 
-There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
-The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals without colon tumors [@Ohigashi2013].
+Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
+The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013].
 Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
@@
 Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors.
-We also test previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity increases by increasing the number of patients with adenomas in our study.
-Additionally, we build upon these observations and assesses the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
+It also tests previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity worsens by increasing the number of patients with adenomas within the study.
+Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
+We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations.
 Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.
@@
 The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
-Additionally, we (i) assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], (ii) used PICRUSt [@Langille2013] and metagenomic sequencing to assess the presence of genes involved in SCFA synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
-This investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.
+We also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate if there were any differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas.
+Next, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas.
+Additionally, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002].
+We also analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
+Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.
 

### 2018-07-30 submission/manuscript.tex
update introduction

@@
 
-Colorectal cancer (CRC) is currently the third highest cancer-related
+Colorectal cancer (CRC) is currently the third leading cancer-related
 cause of death within the US (1, 2). Although there is a genetic
-component to the disease, the environment is attributed to being a
+component to the disease, the environment has been attributed to being a
 larger risk factor for CRC (3). These environmental risk factors include
 but are not limited to smoking cigarettes, diet, and the microbiota
-(4--6). Many of these environmental risk factors are capable of being
-modified, and this has lead to the investigation of how the microbiota
-may exacerbate or cause tumorigensis (7--9) and whether the bacterial
-community is altered (10, 11). Multiple reports in case/control studies
-have identified bacterial taxa commonly associated with individuals
-without tumors to be decreased in patients with carcinoma tumors
-(11--13). Many of these taxa within individuals wihtout tumors actively
-produce short-chain fatty acids (SCFAs) from fiber that are a part of
-our general diet (14). The most extensively studied of these SCFAs are
-acetate, butyrate, and propionate (15). Overall, the specific bacterial
-taxa of the microbiota that create SCFAs are an attractive target to
-modulate the risk of CRC.
-
-Specific SCFAs, like butyrate, have shown positive results for CRC
-treatment within model systems (16). Butyrate has been shown to inhibit
-cancer cell growth in \emph{in vitro} systems (17). Additionally,
-supplementation with food sources that bacteria use to create these
-SCFAs may also be able to confer beneficial effects. For example, fiber
-supplementation in mouse models of CRC caused an overall reduction in
-tumor burden while also increasing SCFA concentrations (18). Although
-these model systems provide important preliminary evidence towards the
-ability of SCFAs to reduce and treat tumors, the studies reporting
-benefit in humans has been less convincing.
-
-There is a lack of evidence on the benefit of increasing SCFA
+(4--6). Many of these environmental risk factors, including the
+microbiota, could be easily modifiable, and this has lead to the
+investigation of how the microbiota may exacerbate or cause tumorigensis
+(7--9) and whether the bacterial community is altered in CRC (10, 11).
+Many of these previous case/control studies have identified resident
+bacterial taxa to be decreased in patients with carcinoma tumors
+(11--13). Many of the bacterial species from these resident taxa
+actively produce short-chain fatty acids (SCFAs) from fiber that are
+part of our general diet (14). The most extensively studied of these
+SCFAs are acetate, butyrate, and propionate (15). Ultimately, these
+SCFAs could be the main metabolites involved with protection against
+tumorigenesis and could help to reduce the risk of CRC.
+
+Prior research suggests that SCFAs have promise in acting as an
+anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown
+positive results within model systems (16). For example, butyrate has
+been shown to inhibit cancer cell growth in \emph{in vitro} systems
+(17). Additionally, fiber supplementation in mouse models of CRC caused
+an overall reduction in tumor burden while also increasing SCFA
+concentrations (18). These exciting results in model systems suggests
+that supplementation with food sources that bacteria use to create these
+SCFAs may be able to confer beneficial effects against CRC. However, it
+is important to note that these model systems provide only preliminary
+evidence towards the ability of SCFAs to reduce and treat tumors and the
+studies reporting benefit in humans has been less convincing.
+
+Overall, there is a lack of evidence on the benefit of increasing SCFA
 concentrations to protect against CRC in human populations. The initial
@@
 that patients with carcinomas had lower concentrations of acetate,
-butyrate, and propionate versus either patients with adenomas or
-individuals without colon tumors (19). Although this would argue that
-increasing SCFA concentrations could be protective against
-tumorigenesis, fiber supplementation in randomized controlled trials
-have consistently failed to protect against tumor recurrence (20). These
-findings would argue against the utility of treatments that aim to use
-SCFAs to reduce or protect against tumorigenesis. Given the lack of
-clear evidence in human studies of the benefit of SCFAs in CRC, there is
-a need for more investigation into this area.
+butyrate, and propionate versus patients with adenomas or individuals
+without colon tumors (19). Although this would argue that increasing
+SCFA concentrations could be protective against tumorigenesis, fiber
+supplementation in randomized controlled trials have consistently failed
+to protect against tumor recurrence (20). These findings would argue
+against the utility of treatments that aim to use SCFAs to reduce or
+protect against tumorigenesis. Given the lack of clear evidence in human
+studies of the benefit of SCFAs in CRC, there is a need for more
+investigation into this area.
 
@@
 tests previous case/control findings on SCFA concentrations in
-individuals with and without tumors. We also test previous suggestions
+individuals with and without tumors. It also tests previous suggestions
 that there is a continuous reduction in SCFA concentrations as tumor
-severity increases by increasing the number of patients with adenomas in
-our study. Additionally, we build upon these observations and assesses
-the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a
-risk stratification tool of colorectal tumors (adenoma or carcinoma).
-Collectively, this study provides important information on the
-replicability of previous findings in humans by extensively studying how
-SCFAs are associated with colorectal tumors.
+severity worsens by increasing the number of patients with adenomas
+within the study. Additionally, we build upon these observations by
+assessing the utility of using SCFAs and Operational Taxonomic Units
+(OTUs) as a risk stratification tool of colorectal tumors (adenoma or
+carcinoma). We also investigate whether OTUs that are most important to
+these models are closely associated with the classification of SCFA
+concentrations. Collectively, this study provides important information
+on the replicability of previous findings in humans by extensively
+studying how SCFAs are associated with colorectal tumors.
 
@@
 sectional point in time while the second group had samples obtained
-before (pre-) and after (post-)treatment for colorectal tumors.
-Additionally, we (i) assessed the affect adding SCFA concentrations to
-OTU data had on classification of patients with adenoma or carcinoma
-using the Random Forest algorithm (21), (ii) used PICRUSt (22) and
-metagenomic sequencing to assess the presence of genes involved in SCFA
-synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts
-SCFA concentrations. This investigation provides additional information
-as to whether SCFAs are decreased in patients with colorectal tumors and
+before (pre-) and after (post-)treatment for colorectal tumors. We also
+used PICRUSt (21) and metagenomic sequencing to investigate if there
+were any differences in genes involved with SCFA synthesis between
+individuals without tumors, patients with adenomas, and patients with
+carcinomas. Next, using the cross-sectional data, we analyzed the number
+of correlations between OTU relative abundance and SCFA concentrations
+across individuals without tumors and patients with adenomas or
+carcinomas. Additionally, we assessed the affect adding SCFA
+concentrations to OTU data had on classification of patients with
+adenomas or carcinomas using the Random Forest algorithm (22). We also
+analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
+Collectively, this investigation provides additional information as to
+whether SCFAs are decreased in patients with colorectal tumors and
 provides context as to whether targeting taxa to increase SCFA
@@
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (22) was used.
+1.1.2 with the recommended standard operating protocol (21) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 Random Forest algorithm for classification and regression models via the
-caret package (21, 43). This was repeated on 100 different 80/20 splits
+caret package (22, 43). This was repeated on 100 different 80/20 splits
 of the data to generate a reasonable range for the AUC of the model. The
@@
 
-\hypertarget{ref-randomforest_citation_2002}{}
-21. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News \textbf{2}:18--22.
-
 \hypertarget{ref-Langille2013}{}
-22. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
+
+\hypertarget{ref-randomforest_citation_2002}{}
+22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
 

### 2018-07-30 submission/manuscript.Rmd
update abstract

@@
 
-**Background.** Colorectal cancer (CRC) is a growing health concern with the majority of the risk for developing disease being due to environmental factors.
-The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC, while other taxa are associated to colons without tumors.
-Some of the taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
+**Background.** Colorectal cancer (CRC) is a continuing health concern with the majority of the risk for developing disease being due to environmental factors.
+The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC and other taxa being associated to colons without tumors.
+Some of the bacterial species in taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
 However, the data supporting the importance of SCFAs in human CRC is less certain.
@@
 **Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate.
-Analysis also included tumor classification models using Random Forest, imputed gene relative abundance with PICRUSt, and metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group.
+Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.
 
 **Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
-There was no difference in classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
 Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
+Finally, there was no difference between classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
 
 **Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC.
-Alternatively, there may be other mechanisms that have not been thoroughly investigated that are more important to the development of human CRC.
+Alternatively, our observations also support the hypothesis that there may be other mechanisms that have not been thoroughly investigated that are more involved with the development of human CRC.
 

### 2018-07-30 submission/manuscript.tex
update abstract

@@
 
-\textbf{Background.} Colorectal cancer (CRC) is a growing health concern
-with the majority of the risk for developing disease being due to
-environmental factors. The microbiota is one of these environmental
+\textbf{Background.} Colorectal cancer (CRC) is a continuing health
+concern with the majority of the risk for developing disease being due
+to environmental factors. The microbiota is one of these environmental
 factors with certain bacterial community members being associated with
-CRC, while other taxa are associated to colons without tumors. Some of
-the taxa associated to colons without tumors can use fiber to produce
-short-chain fatty acids (SCFAs) that can inhibit tumor growth in model
-systems. However, the data supporting the importance of SCFAs in human
-CRC is less certain. Here, we test the hypothesis that SCFA
-concentrations are different in individuals with colorectal tumors.
+CRC and other taxa being associated to colons without tumors. Some of
+the bacterial species in taxa associated to colons without tumors can
+use fiber to produce short-chain fatty acids (SCFAs) that can inhibit
+tumor growth in model systems. However, the data supporting the
+importance of SCFAs in human CRC is less certain. Here, we test the
+hypothesis that SCFA concentrations are different in individuals with
+colorectal tumors.
 
@@
 pre- and post-treatment (n=67) group for their concentrations of
-acetate, butyrate, and propionate. Analysis also included tumor
-classification models using Random Forest, imputed gene relative
-abundance with PICRUSt, and metagenomic sequencing on a subset (n=85) of
-the total cross-sectional group.
+acetate, butyrate, and propionate. Analysis also included imputed gene
+relative abundance with PICRUSt, metagenomic sequencing on a subset
+(n=85) of the total cross-sectional group, and tumor classification and
+SCFA prediction models using Random Forest.
 
@@
 between individuals without tumors and patients with adenomas or
-carcinomas (P-value \textgreater{} 0.15). There was no difference in
-classification models with or without SCFAs in their ability to predict
-patients with adenomas or carcinomas versus individuals without tumors
-(P-value \textgreater{} 0.05). Using metagenomic sequencing, there was
-also no difference in genes involved with SCFA synthesis between
-individuals without tumors and patients with adenomas or carcinomas
-(P-value \textgreater{} 0.70).
+carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
+there was also no difference in genes involved with SCFA synthesis
+between individuals without tumors and patients with adenomas or
+carcinomas (P-value \textgreater{} 0.70). Finally, there was no
+difference between classification models with or without SCFAs in their
+ability to predict patients with adenomas or carcinomas versus
+individuals without tumors (P-value \textgreater{} 0.05).
 
@@
 there may be context specific scenarios where SCFAs may still be
-beneficial for treatment of CRC. Alternatively, there may be other
-mechanisms that have not been thoroughly investigated that are more
-important to the development of human CRC.
+beneficial for treatment of CRC. Alternatively, our observations also
+support the hypothesis that there may be other mechanisms that have not
+been thoroughly investigated that are more involved with the development
+of human CRC.
 

### 2018-08-03 submission/manuscript.Rmd
update abstract with Ada suggestions

@@
 
-Marc A Sze${^1}$, Nicholas A Lesniak${^1}$, Mack T Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
+Marc A. Sze${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
 
@@
 
-**Background.** Colorectal cancer (CRC) is a continuing health concern with the majority of the risk for developing disease being due to environmental factors.
-The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC and other taxa being associated to colons without tumors.
-Some of the bacterial species in taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems.
+**Background.** Colorectal cancer (CRC) is increasing in prevalence in individuals under 50 and because of this will be a continuing health concern for the foreseeable future.
+The majority of the risk for developing CRC is attributable to environmental factors.
+One of these environmental factors is the microbiota, with certain bacterial community members being associated with CRC and other taxa being associated to individuals without tumors.
+Some of the bacterial species in taxa associated to individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems.
 However, the data supporting the importance of SCFAs in human CRC is less certain.
-Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.
+Here, we test the hypothesis that SCFA concentrations and taxa associated with their production are different in individuals with colorectal tumors.
 
-**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate.
+**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their fecal concentrations of acetate, butyrate, and propionate.
 Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.
@@
 Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
-Finally, there was no difference between classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
+Finally, there was no difference between the ability of Random Forest models to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
 
-**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC.
-Alternatively, our observations also support the hypothesis that there may be other mechanisms that have not been thoroughly investigated that are more involved with the development of human CRC.
+**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations in patients in the general CRC population are different, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC.
+Alternatively, our observations also support the hypothesis that there may be other metabolites or mechanisms (e.g.
+bacterial niche exclusion) that may be more protective against tumorigenesis and have not been thoroughly investigated in the context of human CRC.
 
@@
 A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6].
-There was a noticeablely higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3].
+There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3].
 In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas  [Figure 3].
@@
 Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to  these taxa in the tumor models [Figure 4-5].
-Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carinomas, are not that same as those involved with SCFA production.
+Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carcinomas, are not that same as those involved with SCFA production.
 
@@
 It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
-This information taken together with our observations would suggest that either indviduals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.
+This information taken together with our observations would suggest that either individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.
 
@@
 
-One possible technical limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
+One possible technical limitation is that a fecal sample may not be an ideal type of bio-specimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
 However, this is unlikely to be a major confounder.
@@
 Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer.
-Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
+Yet, taxa that are associated with SCFA production are consistently higher in individuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
 
@@
 In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis.
-This leads to the possibilty that protection may be through two other routes.
+This leads to the possibility that protection may be through two other routes.
 First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis.
@@
 Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012].
-Chimeras were identifed and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
+Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
 The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].
@@
 Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm.
-For all Random Forest models, the assessment of the most important variables was based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in accuracy.
+For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.
 

### 2018-08-03 submission/manuscript.tex
update abstract with Ada suggestions

@@
 
-Marc A Sze\({^1}\), Nicholas A Lesniak\({^1}\), Mack T Ruffin
+Marc A. Sze\({^1}\), Nicholas A. Lesniak\({^1}\), Mack T. Ruffin
 IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)
@@
 
-\textbf{Background.} Colorectal cancer (CRC) is a continuing health
-concern with the majority of the risk for developing disease being due
-to environmental factors. The microbiota is one of these environmental
-factors with certain bacterial community members being associated with
-CRC and other taxa being associated to colons without tumors. Some of
-the bacterial species in taxa associated to colons without tumors can
-use fiber to produce short-chain fatty acids (SCFAs) that can inhibit
-tumor growth in model systems. However, the data supporting the
-importance of SCFAs in human CRC is less certain. Here, we test the
-hypothesis that SCFA concentrations are different in individuals with
-colorectal tumors.
+\textbf{Background.} Colorectal cancer (CRC) is increasing in prevalence
+in individuals under 50 and because of this will be a continuing health
+concern for the foreseeable future. The majority of the risk for
+developing CRC is attributable to environmental factors. One of these
+environmental factors is the microbiota, with certain bacterial
+community members being associated with CRC and other taxa being
+associated to individuals without tumors. Some of the bacterial species
+in taxa associated to individuals without tumors can use fiber to
+produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
+model systems. However, the data supporting the importance of SCFAs in
+human CRC is less certain. Here, we test the hypothesis that SCFA
+concentrations and taxa associated with their production are different
+in individuals with colorectal tumors.
 
 \textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
-pre- and post-treatment (n=67) group for their concentrations of
+pre- and post-treatment (n=67) group for their fecal concentrations of
 acetate, butyrate, and propionate. Analysis also included imputed gene
@@
 carcinomas (P-value \textgreater{} 0.70). Finally, there was no
-difference between classification models with or without SCFAs in their
-ability to predict patients with adenomas or carcinomas versus
-individuals without tumors (P-value \textgreater{} 0.05).
+difference between the ability of Random Forest models to predict
+patients with adenomas or carcinomas versus individuals without tumors
+(P-value \textgreater{} 0.05).
 
 \textbf{Conclusions.} Although our data does not support the hypothesis
-that SCFAs are different in individuals that have colorectal tumors,
-there may be context specific scenarios where SCFAs may still be
-beneficial for treatment of CRC. Alternatively, our observations also
-support the hypothesis that there may be other mechanisms that have not
-been thoroughly investigated that are more involved with the development
-of human CRC.
+that fecal SCFA concentrations in patients in the general CRC population
+are different, there still may be specific types of colorectal tumors
+where SCFAs may be beneficial for treatment of CRC. Alternatively, our
+observations also support the hypothesis that there may be other
+metabolites or mechanisms (e.g.~bacterial niche exclusion) that may be
+more protective against tumorigenesis and have not been thoroughly
+investigated in the context of human CRC.
 
@@
 high/low SCFA groups based on the overall median concentration for that
-specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeablely
-higher number of significant negative correlations associated with
-patients with adenomas for all SCFAs tested {[}Figure 3{]}. In
-particular, OTUs from the \emph{Ruminococcaceae} family had the largest
-share of these negative correlations within patients with adenomas
-{[}Figure 3{]}. Although patients with adenomas had more positive
-correlations between OTUs and SCFA concentrations, their total number
-was more similar to individuals without tumors or patients with
-carcinomas versus the analogous comparison for the number of negative
-correlations {[}Figure 3{]}. The number of positive correlations between
-OTUs and SCFA concentrations was similar between individuals without
-tumors and patients with a carcinoma {[}Figure 3{]}. Overall, these
-results suggest that the resident taxa that may change the most due to
-colon tumors may not be ones that are responsible for the production of
-acetate, butyrate, or propionate.
+specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeably higher
+number of significant negative correlations associated with patients
+with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
+from the \emph{Ruminococcaceae} family had the largest share of these
+negative correlations within patients with adenomas {[}Figure 3{]}.
+Although patients with adenomas had more positive correlations between
+OTUs and SCFA concentrations, their total number was more similar to
+individuals without tumors or patients with carcinomas versus the
+analogous comparison for the number of negative correlations {[}Figure
+3{]}. The number of positive correlations between OTUs and SCFA
+concentrations was similar between individuals without tumors and
+patients with a carcinoma {[}Figure 3{]}. Overall, these results suggest
+that the resident taxa that may change the most due to colon tumors may
+not be ones that are responsible for the production of acetate,
+butyrate, or propionate.
 
@@
 between individuals without tumors and patients with adenomas or
-carinomas, are not that same as those involved with SCFA production.
+carcinomas, are not that same as those involved with SCFA production.
 
@@
 between healthy individuals (24). This information taken together with
-our observations would suggest that either indviduals who do not respond
-to fiber supplementation would need to acquire these bacteria to achieve
-a benefit or that SCFAs provide little to no benefit as an
+our observations would suggest that either individuals who do not
+respond to fiber supplementation would need to acquire these bacteria to
+achieve a benefit or that SCFAs provide little to no benefit as an
 anti-tumorigenic compound in colorectal cancer.
@@
 One possible technical limitation is that a fecal sample may not be an
-ideal type of biospecimen and that the effect SCFAs have on
+ideal type of bio-specimen and that the effect SCFAs have on
 tumorigenesis is only detected in the colon. However, this is unlikely
@@
 treatment option in colorectal cancer. Yet, taxa that are associated
-with SCFA production are consistently higher in indivdiuals without
+with SCFA production are consistently higher in individuals without
 colon tumors than patients with carcinomas (10, 11, 27).
@@
 taxonomic classification are in tumor and SCFA Random Forest models
-supports this hypothesis. This leads to the possibilty that protection
+supports this hypothesis. This leads to the possibility that protection
 may be through two other routes. First, there could be a different
@@
 matches to the V4 region of the 16S rRNA gene using the SILVA database
-(33). Chimeras were identifed and removed using UCHIME and OTUs
+(33). Chimeras were identified and removed using UCHIME and OTUs
 clustered at 97\% similarity (34). The major differences from these
@@
 Random Forest models, the assessment of the most important variables was
-based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in
+based on the top 10 features (OTUs or SCFAs) using the mean decrease in
 accuracy.

### 2018-08-03 submission/manuscript.Rmd
update intro with Ada suggestions

@@
 
-Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016].
-Although there is a genetic component to the disease, the environment has been attributed to being a larger risk factor for CRC [@Lichtenstein2000].
+Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US and the prevelance is increasing in invidiauls under 50 years of age [@Haggar2009; @Siegel2016].
+Although there is a genetic component to the disease, the environment is considered a larger risk factor for CRC [@Lichtenstein2000].
 These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011].
-Many of these environmental risk factors, including the microbiota, could be easily modifiable, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018].
+Many of these environmental risk factors, including the microbiota, are modifiable.
+This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018].
 Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018].
-Many of the bacterial species from these resident taxa actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017].
-The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016].
-Ultimately, these SCFAs could be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.
+Many of the bacterial species from these resident taxa identified in these case/control studies actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017].
+The most extensively studied SCFAs are acetate, butyrate, and propionate [@Louis2016].
+These SCFAs are hypothesized to be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.
 
 Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent.
-Specific SCFAs, like butyrate, have shown positive results within model systems [@test_OKeefe2016].
+Specific SCFAs have shown positive results within model systems [@test_OKeefe2016].
 For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
 Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
-These exciting results in model systems suggests that supplementation with food sources that bacteria use to create these SCFAs may be able to confer beneficial effects against CRC.
+These exciting results in model systems suggest that supplementation with food sources that bacteria use to create these SCFAs may confer beneficial effects against CRC.
 However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.
@@
 Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
-These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.
+These findings argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.
 Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.
@@
 Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations.
-Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors.
-It also tests previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity worsens by increasing the number of patients with adenomas within the study.
+Specifically, by using a seperate cohort it tests previous case/control findings on SCFA concentrations in individuals with and without tumors.
+Additionally, prior investigations grouped patients with adenoma and individuals without tumors into a single group to compare against patients with carcinomas.
+Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without tumors.
+We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without tumors into a single group.
 Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
@@
 The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
-We also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate if there were any differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas.
-Next, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas.
-Additionally, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002].
-We also analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
+To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas.
+Next, we investigated whether taxa associated with SCFA production were important to disease classification models.
+First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas.
+Second, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002].
+Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
+Finally, we compared whether SCFA concentrations replace important taxa in disease models and whether the same OTUs are the most important variables used to classify disease and SCFA concentration.
 Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

### 2018-08-03 submission/manuscript.tex
update intro with Ada suggestions

@@
 Colorectal cancer (CRC) is currently the third leading cancer-related
-cause of death within the US (1, 2). Although there is a genetic
-component to the disease, the environment has been attributed to being a
-larger risk factor for CRC (3). These environmental risk factors include
-but are not limited to smoking cigarettes, diet, and the microbiota
-(4--6). Many of these environmental risk factors, including the
-microbiota, could be easily modifiable, and this has lead to the
-investigation of how the microbiota may exacerbate or cause tumorigensis
-(7--9) and whether the bacterial community is altered in CRC (10, 11).
-Many of these previous case/control studies have identified resident
-bacterial taxa to be decreased in patients with carcinoma tumors
-(11--13). Many of the bacterial species from these resident taxa
-actively produce short-chain fatty acids (SCFAs) from fiber that are
-part of our general diet (14). The most extensively studied of these
-SCFAs are acetate, butyrate, and propionate (15). Ultimately, these
-SCFAs could be the main metabolites involved with protection against
-tumorigenesis and could help to reduce the risk of CRC.
+cause of death within the US and the prevelance is increasing in
+invidiauls under 50 years of age (1, 2). Although there is a genetic
+component to the disease, the environment is considered a larger risk
+factor for CRC (3). These environmental risk factors include but are not
+limited to smoking cigarettes, diet, and the microbiota (4--6). Many of
+these environmental risk factors, including the microbiota, are
+modifiable. This has lead to the investigation of how the microbiota may
+exacerbate or cause tumorigensis (7--9) and whether the bacterial
+community is altered in CRC (10, 11). Many of these previous
+case/control studies have identified resident bacterial taxa to be
+decreased in patients with carcinoma tumors (11--13). Many of the
+bacterial species from these resident taxa identified in these
+case/control studies actively produce short-chain fatty acids (SCFAs)
+from fiber that are part of our general diet (14). The most extensively
+studied SCFAs are acetate, butyrate, and propionate (15). These SCFAs
+are hypothesized to be the main metabolites involved with protection
+against tumorigenesis and could help to reduce the risk of CRC.
 
 Prior research suggests that SCFAs have promise in acting as an
-anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown
-positive results within model systems (16). For example, butyrate has
-been shown to inhibit cancer cell growth in \emph{in vitro} systems
-(17). Additionally, fiber supplementation in mouse models of CRC caused
-an overall reduction in tumor burden while also increasing SCFA
-concentrations (18). These exciting results in model systems suggests
+anti-tumorigenic agent. Specific SCFAs have shown positive results
+within model systems (16). For example, butyrate has been shown to
+inhibit cancer cell growth in \emph{in vitro} systems (17).
+Additionally, fiber supplementation in mouse models of CRC caused an
+overall reduction in tumor burden while also increasing SCFA
+concentrations (18). These exciting results in model systems suggest
 that supplementation with food sources that bacteria use to create these
-SCFAs may be able to confer beneficial effects against CRC. However, it
-is important to note that these model systems provide only preliminary
+SCFAs may confer beneficial effects against CRC. However, it is
+important to note that these model systems provide only preliminary
 evidence towards the ability of SCFAs to reduce and treat tumors and the
@@
 supplementation in randomized controlled trials have consistently failed
-to protect against tumor recurrence (20). These findings would argue
-against the utility of treatments that aim to use SCFAs to reduce or
-protect against tumorigenesis. Given the lack of clear evidence in human
-studies of the benefit of SCFAs in CRC, there is a need for more
-investigation into this area.
+to protect against tumor recurrence (20). These findings argue against
+the utility of treatments that aim to use SCFAs to reduce or protect
+against tumorigenesis. Given the lack of clear evidence in human studies
+of the benefit of SCFAs in CRC, there is a need for more investigation
+into this area.
 
 Our study fills some of the current gaps in the literature that relate
-to the study of SCFAs and CRC in human populations. Specifically, it
-tests previous case/control findings on SCFA concentrations in
-individuals with and without tumors. It also tests previous suggestions
-that there is a continuous reduction in SCFA concentrations as tumor
-severity worsens by increasing the number of patients with adenomas
-within the study. Additionally, we build upon these observations by
-assessing the utility of using SCFAs and Operational Taxonomic Units
-(OTUs) as a risk stratification tool of colorectal tumors (adenoma or
-carcinoma). We also investigate whether OTUs that are most important to
-these models are closely associated with the classification of SCFA
-concentrations. Collectively, this study provides important information
-on the replicability of previous findings in humans by extensively
-studying how SCFAs are associated with colorectal tumors.
+to the study of SCFAs and CRC in human populations. Specifically, by
+using a seperate cohort it tests previous case/control findings on SCFA
+concentrations in individuals with and without tumors. Additionally,
+prior investigations grouped patients with adenoma and individuals
+without tumors into a single group to compare against patients with
+carcinomas. Despite doing this, the suggestion was made that a reduction
+in SCFA concentration would also be observed between patients with
+adenomas and indivdiuals without tumors. We also can test this
+suggestion because we have a larger number of patients with adenomas
+within our study and do not need to group patients with adenomas and
+individuals without tumors into a single group. Additionally, we build
+upon these observations by assessing the utility of using SCFAs and
+Operational Taxonomic Units (OTUs) as a risk stratification tool of
+colorectal tumors (adenoma or carcinoma). We also investigate whether
+OTUs that are most important to these models are closely associated with
+the classification of SCFA concentrations. Collectively, this study
+provides important information on the replicability of previous findings
+in humans by extensively studying how SCFAs are associated with
+colorectal tumors.
 
@@
 sectional point in time while the second group had samples obtained
-before (pre-) and after (post-)treatment for colorectal tumors. We also
-used PICRUSt (21) and metagenomic sequencing to investigate if there
-were any differences in genes involved with SCFA synthesis between
-individuals without tumors, patients with adenomas, and patients with
-carcinomas. Next, using the cross-sectional data, we analyzed the number
-of correlations between OTU relative abundance and SCFA concentrations
-across individuals without tumors and patients with adenomas or
-carcinomas. Additionally, we assessed the affect adding SCFA
+before (pre-) and after (post-)treatment for colorectal tumors. To
+provide further support for our SCFA findings we also used PICRUSt (21)
+and metagenomic sequencing to investigate differences in genes involved
+with SCFA synthesis between individuals without tumors, patients with
+adenomas, and patients with carcinomas. Next, we investigated whether
+taxa associated with SCFA production were important to disease
+classification models. First, using the cross-sectional data, we
+analyzed the number of correlations between OTU relative abundance and
+SCFA concentrations across individuals without tumors and patients with
+adenomas or carcinomas. Second, we assessed the affect adding SCFA
 concentrations to OTU data had on classification of patients with
-adenomas or carcinomas using the Random Forest algorithm (22). We also
+adenomas or carcinomas using the Random Forest algorithm (22). Third, we
 analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
-Collectively, this investigation provides additional information as to
-whether SCFAs are decreased in patients with colorectal tumors and
-provides context as to whether targeting taxa to increase SCFA
-concentrations is a viable option to protect against colon
-tumorigenesis.
+Finally, we compared whether SCFA concentrations replace important taxa
+in disease models and whether the same OTUs are the most important
+variables used to classify disease and SCFA concentration. Collectively,
+this investigation provides additional information as to whether SCFAs
+are decreased in patients with colorectal tumors and provides context as
+to whether targeting taxa to increase SCFA concentrations is a viable
+option to protect against colon tumorigenesis.
 

### 2018-08-03 code/make_pi_opf_combined_graph.R
update fig with only butyrate kinase

@@ -29,20 +29,17 @@ opf_data <- read_csv("data/process/tables/select_scfa_opf_data.csv") %>%
   filter(kegg_id %in% butyrate_genes)
 
 imputed_fig <- picrust_data %>% 
+  filter(kegg_ortholog == "K00929") %>% 
   mutate(dx = factor(dx, 
                 levels = c("normal", "adenoma", "cancer"), 
-                labels = c("Control", "Adenoma", "Carcinoma")), 
-         kegg_ortholog = factor(kegg_ortholog, 
-                                levels = c("K00929", "K01034", "K01035"), 
-                                labels = c("Butyrate Kinase", "Acetate CoA\n(alpha subunit)", 
-                                           "Acetate CoA\n(beta subunit)"))) %>% 
+                labels = c("Control", "Adenoma", "Carcinoma"))) %>% 
   ggplot(aes(kegg_ortholog, log10(rel_abund), fill = dx)) + 
   geom_boxplot(position = position_dodge(width = 1)) + 
   geom_vline(xintercept=seq(1.5, length(unique(picrust_data$kegg_ortholog))-0.5, 1), 
              lwd=1, colour="gray") + 
-  theme_bw() + ggtitle("A") + 
+  theme_bw() + ggtitle("A") + coord_cartesian(ylim = c(0, 8)) + 
   scale_fill_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
-  labs(x = "", y = expression(Log["10"]~Imputed~Relative~Abundance)) + 
+  labs(x = "Butyrate Kinase", y = expression(Log["10"]~Imputed~Relative~Abundance)) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
@@ -50,7 +47,8 @@ imputed_fig <- picrust_data %>%
         legend.title = element_blank(),
         axis.text.y = element_text(size = 10), 
         axis.title = element_text(size = 12), 
-        axis.text.x = element_text(size = 10, face = "bold"))
+        axis.text.x = element_blank(), 
+        axis.ticks.x = element_blank())
 
 
 metagenome_fig <- opf_data %>% 
@@ -63,7 +61,7 @@ metagenome_fig <- opf_data %>%
   geom_boxplot(position = position_dodge(width = 1)) + 
   theme_bw() + ggtitle("B") + 
   scale_fill_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
[…]

### 2018-08-03 submission/manuscript.Rmd
include Ada suggestions results 1st 2 paragraphs

@@
 
-**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time.
+**Decreased SCFA concentrations are not associated with adenoma or carcinoma tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time.
 There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
 We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples.
-Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
+Although there was a general trend for increasing acetate, butyrate, and propionate concentrations following treatment for tumors, there was no significant difference between pre- and post-treatment for either patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
 
-**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
-First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
-This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 2A].
-Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
+**Gene abundance for enzymes involved in SCFA synthesis are the same for individuals without tumors and patients with adenomas or carcinomas.** In order to provide further confirmation and support of our SCFA concentration results we investigated the genes encoding specific enzymes involved with SCFA synthesis [Table S1].
+Using this list of specific genes [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
+Although we intended to analyze all the genes in the list [Table S1], not all of the KEGG genes were identified during our analysis.
+We first analyzed imputed gene relative abundance, calculated from our OTU data that was generated from 16S rRNA gene sequencing.
+We found no difference in any of the imputed gene relative abundances for enzymes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
+Since butyrate is one of the most studied SCFAs in the context of CRC, we visualized the observed lack of difference between indviduals without tumors and patients with adenomas or carcinomas in gene abundance for enzymes involved with SCFA synthesis using the butyrate kinase gene [Figure 2A].
+Additionally, using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
 Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
-This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 2B].
-These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors.
+Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenomas (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinomas (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
+This similarity between indviduals without tumors and patients with adenomas or carcinomas is highlighted again by visualizing the gene abundance for the butyrate kinase gene [Figure 2B].
+These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors and provides further support for our original SCFA concentration observations.
 

### 2018-08-03 submission/manuscript.tex
include Ada suggestions results 1st 2 paragraphs

@@
 
-\textbf{Decreased SCFA concentrations are not associated with tumors.}
-We used high-performance liquid chromatography (HPLC) to measure
-acetate, butyrate, and propionate concentrations of frozen fecal samples
-from 490 individuals at a cross-sectional point in time. There was no
-difference between individuals without colon tumors (n=172) and patients
-with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
-measured after multiple comparison correction (P-value \textgreater{}
-0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
-in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
-had pre- and post-treatment fecal samples. Although there was a general
-trend for increasing acetate, butyrate, and propionate concentrations
-after treatment for tumors, there was no significant difference pre- and
-post-treatment for patients with adenomas (P-value \textgreater{} 0.20)
-or carcinomas (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
-
-\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
-associated with tumors.} Using a list of specific genes that are
-important for the synthesis of SCFAs {[}Table S1{]}, we looked for
-differences in gene abundance between individuals without colon tumors
-and patients with adenomas or carcinomas. First, using imputed gene
-relative abundance based on 16S rRNA gene sequencing we found no
-difference in any of the genes involved with acetate, butyrate, or
-propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
-similarity between groups is highlighted by visualizing genes important
-in butyrate synthesis {[}Figure 2A{]}. Using a paired Wilcoxon rank-sum
-test, there also was no difference in imputed gene relative abundance
-between pre- and post-treatment samples for any genes involved with SCFA
-synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
-subset of these 490 fecal samples (n=85) and used metagenomic sequencing
-to confirm these results. Like the imputed gene results, metagenomic
+\textbf{Decreased SCFA concentrations are not associated with adenoma or
+carcinoma tumors.} We used high-performance liquid chromatography (HPLC)
+to measure acetate, butyrate, and propionate concentrations of frozen
+fecal samples from 490 individuals at a cross-sectional point in time.
+There was no difference between individuals without colon tumors (n=172)
+and patients with either an adenoma (n=198) or carcinoma (n=120) for any
+of the SCFAs measured after multiple comparison correction (P-value
+\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
+concentration of SCFAs in 67 patients with an adenoma (n=41) or
+carcinoma (n=26) in which we had pre- and post-treatment fecal samples.
+Although there was a general trend for increasing acetate, butyrate, and
+propionate concentrations following treatment for tumors, there was no
+significant difference between pre- and post-treatment for either
+patients with adenomas (P-value \textgreater{} 0.20) or carcinomas
+(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
+
+\textbf{Gene abundance for enzymes involved in SCFA synthesis are the
+same for individuals without tumors and patients with adenomas or
+carcinomas.} In order to provide further confirmation and support of our
+SCFA concentration results we investigated the genes encoding specific
+enzymes involved with SCFA synthesis {[}Table S1{]}. Using this list of
+specific genes {[}Table S1{]}, we looked for differences in gene
+abundance between individuals without colon tumors and patients with
+adenomas or carcinomas. Although we intended to analyze all the genes in
+the list {[}Table S1{]}, not all of the KEGG genes were identified
+during our analysis. We first analyzed imputed gene relative abundance,
+calculated from our OTU data that was generated from 16S rRNA gene
+sequencing. We found no difference in any of the imputed gene relative
+abundances for enzymes involved with acetate, butyrate, or propionate
+synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. Since butyrate
+is one of the most studied SCFAs in the context of CRC, we visualized
+the observed lack of difference between indviduals without tumors and
+patients with adenomas or carcinomas in gene abundance for enzymes
+involved with SCFA synthesis using the butyrate kinase gene {[}Figure
+2A{]}. Additionally, using a paired Wilcoxon rank-sum test, there also
+was no difference in imputed gene relative abundance between pre- and
+post-treatment samples for any genes involved with SCFA synthesis
+(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
+these 490 fecal samples (n=85) and used metagenomic sequencing to
+confirm these results. Like the imputed gene results, metagenomic
 analysis found that there was no difference in any of the genes involved
 in SCFA synthesis between individuals without colon tumors (n=29) and
-patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
-0.70) {[}Table S4{]}. This lack of difference can be highlighted when we
-visualize the results for butyrate kinase {[}Figure 2B{]}. These
-observations provide evidence that gene prevalence for enzymes involved
-in SCFA synthesis does not change due to colorectal tumors.
+patients with adenomas (n=28) or carcinomas (n=28) (P-value
+\textgreater{} 0.70) {[}Table S4{]}. This similarity between indviduals
+without tumors and patients with adenomas or carcinomas is highlighted
+again by visualizing the gene abundance for the butyrate kinase gene
+{[}Figure 2B{]}. These observations provide evidence that gene
+prevalence for enzymes involved in SCFA synthesis does not change due to
+colorectal tumors and provides further support for our original SCFA
+concentration observations.
 

### 2018-08-06 submission/manuscript.Rmd
3rd paragraph results - ada suggestions

@@
 
-**Total significant positive correlations between OTU relative abundance and SCFA concentration were similar for individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
-Using Spearmans rho we found that the majority of significant OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5].
-A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6].
+**The number OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
+The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have tumors, had an adenoma, or had a carcinoma.
+To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero.
+We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5].
 There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3].
 In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3].
-Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
-The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3].
+Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
+Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3].
+Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6].
 Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

### 2018-08-06 submission/manuscript.tex
3rd paragraph results - ada suggestions

@@
 
-\textbf{Total significant positive correlations between OTU relative
-abundance and SCFA concentration were similar for individuals without
-tumors and patients with adenomas or carcinomas.} Having found no
-difference between individuals without tumors and patients with adenomas
-or carcinomas in SCFA concentrations or genes coding for enzymes
-involved with SCFA synthesis, we next investigated if specific OTUs
-correlated with SCFA concentrations. Using Spearmans rho we found that
-the majority of significant OTU correlations were to taxa from
-\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
-{[}Figure 3 \& Table S5{]}. A similar pattern was observed when using
-high/low SCFA groups based on the overall median concentration for that
-specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeably higher
+\textbf{The number OTUs positively associated with SCFA concentration
+were similar between individuals without tumors and patients with
+adenomas or carcinomas.} Having found no difference between individuals
+without tumors and patients with adenomas or carcinomas in SCFA
+concentrations or genes encoding enzymes involved with SCFA synthesis,
+we next investigated if specific OTUs correlated with SCFA
+concentrations. The main goal of this analysis was to identify if there
+were OTUs that were significantly associated with SCFA concentrations
+and if this was different between individuals that did not have tumors,
+had an adenoma, or had a carcinoma. To accomplish this we used Spearmans
+rho, a non-parametric measure of association, and tested if there was a
+correlation that was signficantly greater than zero. We found that taxa
+from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} dominated statistically significant OTU
+correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
 number of significant negative correlations associated with patients
@@
 negative correlations within patients with adenomas {[}Figure 3{]}.
-Although patients with adenomas had more positive correlations between
-OTUs and SCFA concentrations, their total number was more similar to
+Patients with adenomas also had more positive correlations between OTUs
+and SCFA concentrations, but their total number was more similar to
 individuals without tumors or patients with carcinomas versus the
 analogous comparison for the number of negative correlations {[}Figure
-3{]}. The number of positive correlations between OTUs and SCFA
-concentrations was similar between individuals without tumors and
-patients with a carcinoma {[}Figure 3{]}. Overall, these results suggest
-that the resident taxa that may change the most due to colon tumors may
-not be ones that are responsible for the production of acetate,
-butyrate, or propionate.
+3{]}. Additionally, the number of positive correlations between OTUs and
+SCFA concentrations was similar between individuals without tumors and
+patients with a carcinoma {[}Figure 3{]}. Finally, when we used high/low
+SCFA groups based on the overall median concentration for each SCFA
+instead of SCFA concentrations a similar pattern was still observed
+{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
+resident taxa that may change the most due to colon tumors may not be
+ones that are responsible for the production of acetate, butyrate, or
+propionate.
 

### 2018-08-06 submission/manuscript.Rmd
update with ada suggestions paragraph 4

@@
 reg_group_rf_data <- read_csv("../data/process/tables/full_model_reg_group_results.csv")
+
+# Load in sig associations and top 10 RF model presence
+sig_assoc_top10_model <- read.csv("../data/process/tables/sig_association_and_RF_top10.csv")
 
@@
 
-**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures.
-Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors.
-Using the Random Forest algorithm we built models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
-For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D].
-There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F].
-The main reason for this was because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C].
-The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining in the top 10 [Figure 4B & 4C].
-In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.
+**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** Despite the lack of difference in positive correlations between OTUs and SCFA concentrations between individuals with and without tumors, OTUs associated with SCFA concentrations could still be the most important variables to Random Forest models built to classify patients with adenomas or carcinomas.
+We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
+With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained in the OTU and SCFA model.
+Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations.
+Both our adenoma and carcinoma models classified patients with a similar degree of success, as measured by the area under the curve (AUC) (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D].
+After the addition of SCFA concentrations to the adenoma or carcinoma models, many OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables to the model [Figure Figure 4B-C & 4E-F].
+After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations.
+When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration remained as part of the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations.
+In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.
 

### 2018-08-06 submission/manuscript.tex
update with ada suggestions paragraph 4

@@
 \emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
-OTUs in Random Forest models built to classify tumors.} SCFA
-concentrations could improve prediction of tumors based on specific
-bacterial community structures. Our OTU data can be used in combination
-with SCFAs to assess whether there is a community dependent context to
-SCFA classification of tumors. Using the Random Forest algorithm we
-built models with OTU abundance data or OTU abundances and SCFA
-concentrations to classify normal versus adenoma and normal versus
-carcinoma fecal samples. For adenoma and carcinoma models, there was no
-difference between the median AUC of models with or without SCFA
-concentrations (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
-There was little difference between the top 10 most important OTUs, as
-measured by mean decrease in accuracy, in models using SCFA and OTUs
-versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The main reason for this was
-because SCFA concentrations replaced the information gained by specific
-OTUs rather than add new information to the model {[}Figure 4B-C{]}. The
-SCFA concentrations also only replaced select OTUs with taxonomic
-classification to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} with many OTUs within these taxa remaining in the
-top 10 {[}Figure 4B \& 4C{]}. In combination with the previous results
-on taxa correlations, these observations provide additional evidence
-that the resident taxa within models that classify tumor are not ones
-associated with acetate, butyrate, or propionate production.
+OTUs in Random Forest models built to classify tumors.} Despite the lack
+of difference in positive correlations between OTUs and SCFA
+concentrations between individuals with and without tumors, OTUs
+associated with SCFA concentrations could still be the most important
+variables to Random Forest models built to classify patients with
+adenomas or carcinomas. We tested this by using the Random Forest
+algorithm to build models with OTU abundance data or OTU abundances and
+SCFA concentrations to classify normal versus adenoma and normal versus
+carcinoma fecal samples. With these models we compared whether OTUs with
+taxonomic classification to \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained in the OTU
+and SCFA model. Additionally, we also compared whether any of the
+important OTUs that remained also had significant correlations with SCFA
+concentrations. Both our adenoma and carcinoma models classified
+patients with a similar degree of success, as measured by the area under
+the curve (AUC) (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
+After the addition of SCFA concentrations to the adenoma or carcinoma
+models, many OTUs with taxonomic classification to \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained as important
+variables to the model {[}Figure Figure 4B-C \& 4E-F{]}. After adding
+SCFA concentrations to the adenoma model, there were only 2 OTUs
+significantly associated with SCFA concentration that were part of the
+top 10 most important variables and both were postively associated with
+acetate, butyrate, or propionate concentrations. When SCFA
+concentrations were added to the carcinoma model, only 1 OTU that was
+associated with SCFA concentration remained as part of the top 10 most
+important variables and it was negatively associated with acetate and
+butyrate concentrations. In combination with the previous results on OTU
+correlations, these observations provide additional evidence that the
+resident taxa that are associated with protection against tumorigeneis
+are not ones associated with acetate, butyrate, or propionate
+production.
 

### 2018-08-06 submission/manuscript.Rmd
complete Ada comments on results

@@
 
-**Random Forest models for SCFA concentrations have different top 10 most important OTUs than tumor models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher/lower than median SCFA concentrations.
-Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high/low SCFA groups [Figure 5A & S2A].
-However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure 5A & S2A].
-Additionally, there were differences in accuracy for both model types based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B].
-There also was minimal overlap between these SCFA classification model's most important OTUs and those used to classify patients with adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
-The only OTU that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E].
-Additionally, OTU00167 was in the top 10 most important OTUs for the OTU adenoma model but not in the SCFA and OTU adenoma model while acetate and butyrate concentrations were [Figure 4B-C].
-These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
-Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with adenomas or carcinomas.
+**The most important OTUs in Random Forest models built to classify SCFA concentrations or tumors are not the same.** It is possible that due to the way the Random Forest algorithm works, OTUs associated with SCFA concentrations could be downweighted in importance within the adenoma or carcinoma models when SCFA concentrations are included.
+To test if this is the case we used OTU data and built Random Forest models to classify SCFA concentrations.
+Although the correlation between the predicted and actual SCFA concentrations were moderately associated with each other [Figure 5A], there was a difference in accuracy based on whether the fecal sample was from an individual without tumors or from patients with adenomas or carcinomas [Figure 5B].
+Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa are important for these classifications [Figure 5A].
+When comparing the adenoma or carcinoma model to the SCFA concentration models there was minimal overlap between these model's most important OTUs [Figure 4B-C, 4E-F and 5C-E].
+The only OTU that was in the top 10 most important variables that had overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E].
+Similar observations were made when using high/low SCFA groups based on higher/lower than median SCFA concentrations [Figure S2].
+Collectively, these observations provide evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and accordingly these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
+Although it is possible to identify OTUs associated with SCFA production, overall, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with adenomas or carcinomas.
 

### 2018-08-06 submission/manuscript.tex
complete Ada comments on results

@@
 
-\textbf{Random Forest models for SCFA concentrations have different top
-10 most important OTUs than tumor models.} Using OTU data we built
-Random Forest models to classify either SCFA concentration or
-higher/lower than median SCFA concentrations. Overall, OTU data had a
-reasonable ability to classify both SCFA concentrations and high/low
-SCFA groups {[}Figure 5A \& S2A{]}. However, these models tended to be
-over fit, suggesting that rarer taxa may be important for this
-classification {[}Figure 5A \& S2A{]}. Additionally, there were
-differences in accuracy for both model types based on whether the
-individual had no tumors, an adenoma, or a carcinoma {[}Figure 5B \&
-S2B{]}. There also was minimal overlap between these SCFA classification
-model's most important OTUs and those used to classify patients with
-adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
-The only OTU that did overlap between the models was OTU00167
-(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
-Additionally, OTU00167 was in the top 10 most important OTUs for the OTU
-adenoma model but not in the SCFA and OTU adenoma model while acetate
-and butyrate concentrations were {[}Figure 4B-C{]}. These observations
-provide further evidence that it is possible to identify specific OTUs
-associated with higher SCFA concentrations and that these OTUs belong to
-taxa known to produce acetate, butyrate, and propionate. Although it is
-possible to identify OTUs associated with SCFA production, our results
-do not support the hypothesis that SCFA concentration or OTUs associated
+\textbf{The most important OTUs in Random Forest models built to
+classify SCFA concentrations or tumors are not the same.} It is possible
+that due to the way the Random Forest algorithm works, OTUs associated
+with SCFA concentrations could be downweighted in importance within the
+adenoma or carcinoma models when SCFA concentrations are included. To
+test if this is the case we used OTU data and built Random Forest models
+to classify SCFA concentrations. Although the correlation between the
+predicted and actual SCFA concentrations were moderately associated with
+each other {[}Figure 5A{]}, there was a difference in accuracy based on
+whether the fecal sample was from an individual without tumors or from
+patients with adenomas or carcinomas {[}Figure 5B{]}. Additionally,
+because the training set R\textsuperscript{2} was always higher than the
+test set R\textsuperscript{2}, all SCFA concentration models tended to
+be over fit, suggesting that rarer taxa are important for these
+classifications {[}Figure 5A{]}. When comparing the adenoma or carcinoma
+model to the SCFA concentration models there was minimal overlap between
+these model's most important OTUs {[}Figure 4B-C, 4E-F and 5C-E{]}. The
+only OTU that was in the top 10 most important variables that had
+overlap between the models was OTU00167 (\emph{Clostridiales}) {[}Figure
+4B-C, 4E-F, 5C-E{]}. Similar observations were made when using high/low
+SCFA groups based on higher/lower than median SCFA concentrations
+{[}Figure S2{]}. Collectively, these observations provide evidence that
+it is possible to identify specific OTUs associated with higher SCFA
+concentrations and accordingly these OTUs belong to taxa known to
+produce acetate, butyrate, and propionate. Although it is possible to
+identify OTUs associated with SCFA production, overall, our results do
+not support the hypothesis that SCFA concentration or OTUs associated
 with their production are different between individuals with no tumors

### 2018-08-06 submission/manuscript.Rmd
final results update based on Ada suggesionts

@@
 To test if this is the case we used OTU data and built Random Forest models to classify SCFA concentrations.
-Although the correlation between the predicted and actual SCFA concentrations were moderately associated with each other [Figure 5A], there was a difference in accuracy based on whether the fecal sample was from an individual without tumors or from patients with adenomas or carcinomas [Figure 5B].
-Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa are important for these classifications [Figure 5A].
+Overall, the correlation between the predicted and actual SCFA concentrations were moderately associated with each other [Figure 5A].
+Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa were important for these classifications [Figure 5A].
+There also was a difference in accuracy based on whether the fecal sample was from an individual without tumors or from patients with adenomas or carcinomas [Figure 5B].
 When comparing the adenoma or carcinoma model to the SCFA concentration models there was minimal overlap between these model's most important OTUs [Figure 4B-C, 4E-F and 5C-E].
-The only OTU that was in the top 10 most important variables that had overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E].
-Similar observations were made when using high/low SCFA groups based on higher/lower than median SCFA concentrations [Figure S2].
+The only OTU that was in the top 10 most important variables and had overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E].
+Similar observations were made when using high/low SCFA groups based on the median SCFA concentration [Figure S2].
 Collectively, these observations provide evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and accordingly these OTUs belong to taxa known to produce acetate, butyrate, and propionate.

### 2018-08-06 submission/manuscript.tex
final results update based on Ada suggesionts

@@
 test if this is the case we used OTU data and built Random Forest models
-to classify SCFA concentrations. Although the correlation between the
+to classify SCFA concentrations. Overall, the correlation between the
 predicted and actual SCFA concentrations were moderately associated with
-each other {[}Figure 5A{]}, there was a difference in accuracy based on
+each other {[}Figure 5A{]}. Additionally, because the training set
+R\textsuperscript{2} was always higher than the test set
+R\textsuperscript{2}, all SCFA concentration models tended to be over
+fit, suggesting that rarer taxa were important for these classifications
+{[}Figure 5A{]}. There also was a difference in accuracy based on
 whether the fecal sample was from an individual without tumors or from
-patients with adenomas or carcinomas {[}Figure 5B{]}. Additionally,
-because the training set R\textsuperscript{2} was always higher than the
-test set R\textsuperscript{2}, all SCFA concentration models tended to
-be over fit, suggesting that rarer taxa are important for these
-classifications {[}Figure 5A{]}. When comparing the adenoma or carcinoma
-model to the SCFA concentration models there was minimal overlap between
-these model's most important OTUs {[}Figure 4B-C, 4E-F and 5C-E{]}. The
-only OTU that was in the top 10 most important variables that had
-overlap between the models was OTU00167 (\emph{Clostridiales}) {[}Figure
-4B-C, 4E-F, 5C-E{]}. Similar observations were made when using high/low
-SCFA groups based on higher/lower than median SCFA concentrations
-{[}Figure S2{]}. Collectively, these observations provide evidence that
-it is possible to identify specific OTUs associated with higher SCFA
-concentrations and accordingly these OTUs belong to taxa known to
-produce acetate, butyrate, and propionate. Although it is possible to
-identify OTUs associated with SCFA production, overall, our results do
-not support the hypothesis that SCFA concentration or OTUs associated
-with their production are different between individuals with no tumors
-and patients with adenomas or carcinomas.
+patients with adenomas or carcinomas {[}Figure 5B{]}. When comparing the
+adenoma or carcinoma model to the SCFA concentration models there was
+minimal overlap between these model's most important OTUs {[}Figure
+4B-C, 4E-F and 5C-E{]}. The only OTU that was in the top 10 most
+important variables and had overlap between the models was OTU00167
+(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E{]}. Similar
+observations were made when using high/low SCFA groups based on the
+median SCFA concentration {[}Figure S2{]}. Collectively, these
+observations provide evidence that it is possible to identify specific
+OTUs associated with higher SCFA concentrations and accordingly these
+OTUs belong to taxa known to produce acetate, butyrate, and propionate.
+Although it is possible to identify OTUs associated with SCFA
+production, overall, our results do not support the hypothesis that SCFA
+concentration or OTUs associated with their production are different
+between individuals with no tumors and patients with adenomas or
+carcinomas.
 

### 2018-08-07 code/run_rf_sig_reg_comparison.R
add comparison code

@@ -1,0 +1,76 @@
+### Compare the number of significant correlations that are in the top 10 variables 
+### Get a count of whether SCFAs replace OTUs associated with them
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "gridExtra"))
+
+#setup variables that will be used
+scfas <- c("acetate", "butyrate", "propionate")
+
+# Read in first data table
+sig_reg <- read_csv("data/process/tables/significant_reg_otu_comp_summary.csv") %>% 
+  left_join(read_tsv("data/process/final.taxonomy") %>% 
+              select(-Size) %>% 
+              mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+              separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
+              select(OTU, family), by = c("otu" = "OTU")) %>% 
+  select(otu, family, genus, dx, scfa, estimate, p.value, bh) %>% 
+  mutate_at(vars(p.value:bh), function(x) format(x, scientific = T, digits = 3)) %>% 
+  mutate(estimate = round(estimate, digits = 3), 
+         family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         genus = str_replace_all(family, "_unclassified", ""), 
+         genus = str_replace_all(family, "_", " "))
+
+# Read in second data table
+taxonomy <- read_tsv("data/process/final.taxonomy") %>% 
+  select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";")
+
+all_data <- read_csv("data/process/tables/adn_full_MDA_Summary.csv") %>% 
+  mutate(model = "adn_full") %>% 
+  filter(Variable != "isobutyrate") %>% 
+  slice(1:10) %>% 
+  bind_rows(read_csv("data/process/tables/crc_full_MDA_Summary.csv") %>% 
+              mutate(model = "crc_full") %>% 
+              filter(Variable != "isobutyrate") %>% 
+              slice(1:10), 
[…]

### 2018-08-07 submission/manuscript.Rmd
paragraph 1 discussion Ada suggestions

@@
 The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors.
-Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2].
+Whether we directly measured the SCFA concentration or investigated genes that encoded enzymes used for their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2].
 Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1].
-Instead, SCFA concentrations provided similar information to what specific OTUs were already providing to the tumor classification models [Figure 4].
+In models with SCFA concentrations included, many OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables for the model [Figure 4].
 Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to these taxa in the tumor models [Figure 4-5].
-Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carcinomas, are not that same as those involved with SCFA production.
+Collectively, our observations suggest that the resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* that are different between individuals without tumors and patients with adenomas or carcinomas, are not the same as those involved with SCFA production.
 

### 2018-08-07 submission/manuscript.tex
paragraph 1 discussion Ada suggestions

@@
 concentrations are different in individuals with tumors. Whether we
-directly measured the SCFA concentration or investigated genes
-associated with their production, no difference could be identified
-between individuals without tumors and patients with adenomas or
-carcinomas {[}Figure 1 \& 2{]}. Although there were differences in the
-number of significant correlations between SCFA concentration and OTU
-relative abundance based on whether individuals did not have tumors, had
-an adenoma, or had a carcinoma, SCFA concentrations did not provide
+directly measured the SCFA concentration or investigated genes that
+encoded enzymes used for their production, no difference could be
+identified between individuals without tumors and patients with adenomas
+or carcinomas {[}Figure 1 \& 2{]}. Although there were differences in
+the number of significant correlations between SCFA concentration and
+OTU relative abundance based on whether individuals did not have tumors,
+had an adenoma, or had a carcinoma, SCFA concentrations did not provide
 increased model accuracy for tumor classification {[}Figure 3-4 \&
-S1{]}. Instead, SCFA concentrations provided similar information to what
-specific OTUs were already providing to the tumor classification models
+S1{]}. In models with SCFA concentrations included, many OTUs that
+classified to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} remained as important variables for the model
 {[}Figure 4{]}. Additionally, when models using OTU relative abundance
@@
 these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
-observations suggest that resident taxa from \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae}, that are different
+observations suggest that the resident taxa from \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} that are different
 between individuals without tumors and patients with adenomas or
-carcinomas, are not that same as those involved with SCFA production.
+carcinomas, are not the same as those involved with SCFA production.
 

### 2018-08-07 submission/manuscript.Rmd
P2 Ada suggestions discussion

@@
 Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018].
-Additionally, many of the *in vivo* studies investigate proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017].
+Many of the *in vivo* studies use proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017].
 Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems.
-In addition to using fiber supplementation as a proxy for SCFA treatment, observations on the benefit of SCFAs in preventing tumorigenesis has been mixed.
-In previous case/control studies lower SCFA concentrations between patients with carcinomas and those without carcinomas was observed [@Ohigashi2013].
-Yet, this is in contrast to multiple randomized-controlled trials that have found no difference between patients who do and do not get fiber supplementation to try and prevent tumor recurrence [@Schatzkin2000; @Yao2017].
-This could be because fiber has other actions in humans besides providing an energy source for bacteria to breakdown to make SCFAs.
-It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
+Additionally, the observations in humans on the benefit of SCFAs in preventing tumorigenesis have been mixed.
+In previous case/control studies lower SCFA concentrations were observed in patients with carcinomas versus those without carcinomas [@Ohigashi2013].
+Yet, this is in contrast to multiple randomized-controlled trials that have found no difference in tumor recurrence between patients who do and do not get fiber supplementation [@Schatzkin2000; @Yao2017].
+In contrast to the *in vivo* model findings, the observations made in these randomized-controlled trials would suggest that SCFAs do not prevent or slow tumorigenesis.
+One reason for these results is that SCFA concentrations and responses to fiber vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
 This information taken together with our observations would suggest that either individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.

### 2018-08-07 submission/manuscript.tex
P2 Ada suggestions discussion

@@
 Although SCFAs have been shown to be anti-tumorigenic, most of these
-studies have been performed in model systems (16, 17). Additionally,
-many of the \emph{in vivo} studies investigate proxies such as fiber
-supplementation rather than SCFAs directly (14). Although it is well
-known that breakdown products from gut bacteria results in SCFA
-production, fiber effects on tumorigenesis may be through other
-mechanisms in these \emph{in vivo} model systems. In addition to using
-fiber supplementation as a proxy for SCFA treatment, observations on the
-benefit of SCFAs in preventing tumorigenesis has been mixed. In previous
-case/control studies lower SCFA concentrations between patients with
-carcinomas and those without carcinomas was observed (19). Yet, this is
-in contrast to multiple randomized-controlled trials that have found no
-difference between patients who do and do not get fiber supplementation
-to try and prevent tumor recurrence (20, 23). This could be because
-fiber has other actions in humans besides providing an energy source for
-bacteria to breakdown to make SCFAs. It could also be due to the fact
-that SCFA concentrations and responses to fiber can vary quite a bit
-between healthy individuals (24). This information taken together with
-our observations would suggest that either individuals who do not
-respond to fiber supplementation would need to acquire these bacteria to
-achieve a benefit or that SCFAs provide little to no benefit as an
-anti-tumorigenic compound in colorectal cancer.
+studies have been performed in model systems (16, 17). Many of the
+\emph{in vivo} studies use proxies such as fiber supplementation rather
+than SCFAs directly (14). Although it is well known that breakdown
+products from gut bacteria results in SCFA production, fiber effects on
+tumorigenesis may be through other mechanisms in these \emph{in vivo}
+model systems. Additionally, the observations in humans on the benefit
+of SCFAs in preventing tumorigenesis have been mixed. In previous
+case/control studies lower SCFA concentrations were observed in patients
+with carcinomas versus those without carcinomas (19). Yet, this is in
+contrast to multiple randomized-controlled trials that have found no
+difference in tumor recurrence between patients who do and do not get
+fiber supplementation (20, 23). In contrast to the \emph{in vivo} model
+findings, the observations made in these randomized-controlled trials
+would suggest that SCFAs do not prevent or slow tumorigenesis. One
+reason for these results is that SCFA concentrations and responses to
+fiber vary quite a bit between healthy individuals (24). This
+information taken together with our observations would suggest that
+either individuals who do not respond to fiber supplementation would
+need to acquire these bacteria to achieve a benefit or that SCFAs
+provide little to no benefit as an anti-tumorigenic compound in
+colorectal cancer.
 

### 2018-08-07 submission/manuscript.Rmd
P3 Ada discussion edits

@@
 
-**The number OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
+**The number of OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
 The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have tumors, had an adenoma, or had a carcinoma.
@@
 We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
-With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained in the OTU and SCFA model.
+With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained when SCFA concentrations were added to the model.
 Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations.
@@
 
-In addition to not having a response to fiber supplementation due to lack of the required microbes, it is also possible that there are specific instances of colorectal cancer where SCFAs may be beneficial.
-One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type.
-However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals.
-Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors.
+Another possible alternative explanation as to why no difference in SCFA concentration between individuals without tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs.
+One limitation of current research into the effect of SCFAs and the microbiota in CRC has been that all tumors are treated as the same type.
+However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in certain subsets of individuals.
+Similar to the idea of using immunotherapy as a targeted treatment option for specific tumors [@Thomas2016], SCFAs may have beneficial effects for distinct types of colorectal tumors.
 Future research will need to test if this is a valid hypothesis.
-Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.
+Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation [@Schatzkin2000; @Yao2017] suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.
 

### 2018-08-07 submission/manuscript.tex
P3 Ada discussion edits

@@
 
-\textbf{The number OTUs positively associated with SCFA concentration
+\textbf{The number of OTUs positively associated with SCFA concentration
 were similar between individuals without tumors and patients with
@@
 taxonomic classification to \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained in the OTU
-and SCFA model. Additionally, we also compared whether any of the
-important OTUs that remained also had significant correlations with SCFA
-concentrations. Both our adenoma and carcinoma models classified
-patients with a similar degree of success, as measured by the area under
-the curve (AUC) (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
-After the addition of SCFA concentrations to the adenoma or carcinoma
-models, many OTUs with taxonomic classification to \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained as important
-variables to the model {[}Figure Figure 4B-C \& 4E-F{]}. After adding
-SCFA concentrations to the adenoma model, there were only 2 OTUs
-significantly associated with SCFA concentration that were part of the
-top 10 most important variables and both were postively associated with
-acetate, butyrate, or propionate concentrations. When SCFA
-concentrations were added to the carcinoma model, only 1 OTU that was
-associated with SCFA concentration remained as part of the top 10 most
-important variables and it was negatively associated with acetate and
-butyrate concentrations. In combination with the previous results on OTU
-correlations, these observations provide additional evidence that the
-resident taxa that are associated with protection against tumorigeneis
-are not ones associated with acetate, butyrate, or propionate
-production.
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained when SCFA
+concentrations were added to the model. Additionally, we also compared
+whether any of the important OTUs that remained also had significant
+correlations with SCFA concentrations. Both our adenoma and carcinoma
+models classified patients with a similar degree of success, as measured
+by the area under the curve (AUC) (P-value \textgreater{} 0.05)
+{[}Figure 4A \& 4D{]}. After the addition of SCFA concentrations to the
+adenoma or carcinoma models, many OTUs with taxonomic classification to
+\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
+remained as important variables to the model {[}Figure Figure 4B-C \&
+4E-F{]}. After adding SCFA concentrations to the adenoma model, there
+were only 2 OTUs significantly associated with SCFA concentration that
+were part of the top 10 most important variables and both were postively
+associated with acetate, butyrate, or propionate concentrations. When
+SCFA concentrations were added to the carcinoma model, only 1 OTU that
+was associated with SCFA concentration remained as part of the top 10
+most important variables and it was negatively associated with acetate
+and butyrate concentrations. In combination with the previous results on
+OTU correlations, these observations provide additional evidence that
+the resident taxa that are associated with protection against
+tumorigeneis are not ones associated with acetate, butyrate, or
+propionate production.
 
@@
 
-In addition to not having a response to fiber supplementation due to
-lack of the required microbes, it is also possible that there are
-specific instances of colorectal cancer where SCFAs may be beneficial.
-One limitation of current research into the effect of SCFAs in CRC has
-been that all tumors are treated as the same type. However, there are
-known differences in the types of mutations that occur (25) and treating
-all tumors as equal may actually hide any benefit that could be found in
-specific subsets of individuals. Similar to the idea of specific
-immunotherapy for specific tumors (26), SCFAs may have beneficial
-effects for specific types of colorectal tumors. Future research will
-need to test if this is a valid hypothesis. Regardless of this
-limitation, our results in combination to previous randomized controlled
-trials on fiber supplementation suggests that using SCFAs as a general
-treatment for colorectal cancer is unlikely to provide a reduction in
-tumorigenesis.
+Another possible alternative explanation as to why no difference in SCFA
+concentration between individuals without tumors and patients with
+adenomas or carcinomas was observed, could be because only certain types
+of colorectal cancers are affected by SCFAs. One limitation of current
+research into the effect of SCFAs and the microbiota in CRC has been
+that all tumors are treated as the same type. However, there are known
+differences in the types of mutations that occur (25) and treating all
+tumors as equal may actually hide any benefit that could be found in
+certain subsets of individuals. Similar to the idea of using
+immunotherapy as a targeted treatment option for specific tumors (26),
+SCFAs may have beneficial effects for distinct types of colorectal
+tumors. Future research will need to test if this is a valid hypothesis.
+Regardless of this limitation, our results in combination to previous
+randomized controlled trials on fiber supplementation (20, 23) suggests
+that using SCFAs as a general treatment for colorectal cancer is
+unlikely to provide a reduction in tumorigenesis.
 

### 2018-08-07 submission/manuscript.Rmd
P3 ada suggestions

@@
 However, this is unlikely to be a major confounder.
-First, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
+First, most *in vivo* studies, as well as human studies, have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
 Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016].
-Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be cause of this.
-Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer.
+Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be the cause of this.
+Clearly, our observations provide robust evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer.
 Yet, taxa that are associated with SCFA production are consistently higher in individuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

### 2018-08-07 submission/manuscript.tex
P3 ada suggestions

@@
 tumorigenesis is only detected in the colon. However, this is unlikely
-to be a major confounder. First, most \emph{in vivo} studies as well as
-human studies have used fecal material in their analysis (18, 19).
+to be a major confounder. First, most \emph{in vivo} studies, as well as
+human studies, have used fecal material in their analysis (18, 19).
 Second, previous studies that measure SCFA changes after fiber
@@
 research on SCFAs and colorectal tumors, technical limitations are less
-likely to be cause of this. Additionally, as mentioned earlier, our
-observations along with the randomized controlled trials on fiber
-supplementation in tumor recurrence (20) provide evidence that these
-specific metabolites may not be protective or used as a general
-treatment option in colorectal cancer. Yet, taxa that are associated
-with SCFA production are consistently higher in individuals without
-colon tumors than patients with carcinomas (10, 11, 27).
+likely to be the cause of this. Clearly, our observations provide robust
+evidence that these specific metabolites may not be protective or used
+as a general treatment option in colorectal cancer. Yet, taxa that are
+associated with SCFA production are consistently higher in individuals
+without colon tumors than patients with carcinomas (10, 11, 27).
 

### 2018-08-07 submission/manuscript.Rmd
last paragraph ada comments discussion

@@
 The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Indeed our data would support the contention that the taxa are similar to those associated with SCFA production but that these specific microbes or OTUs themselves are not associated with SCFAs.
-In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis.
-This leads to the possibility that protection may be through two other routes.
-First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis.
-Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g.
-*Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]).
-The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics.
-Even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.
+Our data is consistent with the literature despite finding no difference in SCFA concentrations between individuals without tumors and patients with adenomas or carcinomas.
+In particular, we observed that the important OTUs in our tumor models were to taxa that are associated with SCFA production but that these OTUs were not themselves associated with acetate, butyrate, or propionate production.
+One of the clearest results to highlight this within our study were the results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models [Figure 4 & 5].
+This leads to the possibility that protection may be through other routes.
+First, there could be a different pathway or other less extensively studied metabolites that provide the necessary protection against tumorigenesis.
+Alternatively, previous research has shown that non-resident oral microbes, such as *Fusobacterium*, *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* are associated with carcinomas [@Zeller2014; @Baxter2016; @Kostic2011] and can increase inflammation [@Flynn2016].
+When these non-resident oral microbes increase inflammation they change the existing environment to one in which is more hospitable for themselves [@Flynn2016].
+It is possible that these resident microbes provide protection not through a metabolite but instead through active exclusion of these non-resident oral microbes which in turn prevents them from establishing themselves within the community.
+Overall, even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that our results open up many new and exciting avenues for investigation into how metabolites and the microbiota can prevent tumorigenesis.
 

### 2018-08-07 submission/manuscript.tex
last paragraph ada comments discussion

@@
 individuals without tumors versus patients with carcinomas (10, 11, 27).
-Indeed our data would support the contention that the taxa are similar
-to those associated with SCFA production but that these specific
-microbes or OTUs themselves are not associated with SCFAs. In
-particular, our results showing that different OTUs from the same
-taxonomic classification are in tumor and SCFA Random Forest models
-supports this hypothesis. This leads to the possibility that protection
-may be through two other routes. First, there could be a different
-pathway or other less extensively studied metabolites that provides the
-necessary protection against tumorigenesis. Alternatively, protection
-may not occur via a metabolite but instead through niche exclusion of
-mouth-associated microbes (e.g. \emph{Fusobacterium},
-\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
-12, 13)). The idea of niche exclusion is similar to how the community
-protects against \emph{Clostridium difficile} infection (28) with
-chronic inflammation replacing the role of antibiotics. Even though we
-did not find lower concentrations of SCFAs associated with colorectal
-tumors, we think that there are many exciting new avenues to explore
-because of these results.
+Our data is consistent with the literature despite finding no difference
+in SCFA concentrations between individuals without tumors and patients
+with adenomas or carcinomas. In particular, we observed that the
+important OTUs in our tumor models were to taxa that are associated with
+SCFA production but that these OTUs were not themselves associated with
+acetate, butyrate, or propionate production. One of the clearest results
+to highlight this within our study were the results showing that
+different OTUs from the same taxonomic classification are in tumor and
+SCFA Random Forest models {[}Figure 4 \& 5{]}. This leads to the
+possibility that protection may be through other routes. First, there
+could be a different pathway or other less extensively studied
+metabolites that provide the necessary protection against tumorigenesis.
+Alternatively, previous research has shown that non-resident oral
+microbes, such as \emph{Fusobacterium}, \emph{Porphyromonas},
+\emph{Parvimonas}, and \emph{Peptostreptococcus} are associated with
+carcinomas (6, 12, 13) and can increase inflammation (28). When these
+non-resident oral microbes increase inflammation they change the
+existing environment to one in which is more hospitable for themselves
+(28). It is possible that these resident microbes provide protection not
+through a metabolite but instead through active exclusion of these
+non-resident oral microbes which in turn prevents them from establishing
+themselves within the community. Overall, even though we did not find
+lower concentrations of SCFAs associated with colorectal tumors, we
+think that our results open up many new and exciting avenues for
+investigation into how metabolites and the microbiota can prevent
+tumorigenesis.
 
@@
 
-\hypertarget{ref-Theriot2015}{}
-28. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
-the gastrointestinal microbiome and clostridium difficile. Annual Review
-of Microbiology \textbf{69}:445--461.
-doi:\href{https://doi.org/10.1146/annurev-micro-091014-104115}{10.1146/annurev-micro-091014-104115}.
+\hypertarget{ref-Flynn2016}{}
+28. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+Metabolic and community synergy of oral bacteria in colorectal cancer.
+mSphere \textbf{1}.
+doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
 

### 2018-08-08 submission/manuscript.Rmd
update conclusion

@@
 
-Our observations found no difference in SCFA concentration, their utility as a classification tool, or for genes of enzymes involved in SCFA synthesis between individuals without colon tumors and patients with either adenoma or carcinoma tumors.
+Our observations found no difference in SCFA concentration or in genes encoding enzymes involved in their production between individuals without colon tumors and patients with either adenoma or carcinoma tumors.
+The SCFA concentration also did not improve classification of adenoma or carcinoma tumors.
+The most important OTUs to these models classifying patients with adenomas or carcinomas were mostly not associated with SCFA production.
 Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.
 Additionally, these observations suggest that resident microbes that are not involved in SCFA production may be the important resident community members involved with preventing tumorigenesis.
-By focusing on the alternative mechanisms that are associated with these non-SCFA producing resident microbes, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
+By focusing on alternative mechanisms that are associated with these non-SCFA producing resident microbes or SCFA effects on distinct colorectal cancer groups the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
 

### 2018-08-08 submission/manuscript.tex
update conclusion

@@
 
-Our observations found no difference in SCFA concentration, their
-utility as a classification tool, or for genes of enzymes involved in
-SCFA synthesis between individuals without colon tumors and patients
-with either adenoma or carcinoma tumors. Although these results are
-different than other reports in the literature, they do align with the
-randomized controlled trials that have tested fiber use in preventing
-colorectal tumor recurrence. Additionally, these observations suggest
-that resident microbes that are not involved in SCFA production may be
-the important resident community members involved with preventing
-tumorigenesis. By focusing on the alternative mechanisms that are
-associated with these non-SCFA producing resident microbes, the
-identification of more promising therapeutic options for use in treating
-colorectal cancer may be found.
+Our observations found no difference in SCFA concentration or in genes
+encoding enzymes involved in their production between individuals
+without colon tumors and patients with either adenoma or carcinoma
+tumors. The SCFA concentration also did not improve classification of
+adenoma or carcinoma tumors. The most important OTUs to these models
+classifying patients with adenomas or carcinomas were mostly not
+associated with SCFA production. Although these results are different
+than other reports in the literature, they do align with the randomized
+controlled trials that have tested fiber use in preventing colorectal
+tumor recurrence. Additionally, these observations suggest that resident
+microbes that are not involved in SCFA production may be the important
+resident community members involved with preventing tumorigenesis. By
+focusing on alternative mechanisms that are associated with these
+non-SCFA producing resident microbes or SCFA effects on distinct
+colorectal cancer groups the identification of more promising
+therapeutic options for use in treating colorectal cancer may be found.
 

### 2018-08-08 submission/manuscript.Rmd
fix figure legend captions

@@
 We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan.
+We would also like to thank Ada Hagan for providing valuable feedback on earlier drafts of the manuscript.
 Salary support for Marc A.
@@
 **Figure 1.
-No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (A).
-Butyrate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (B).
-Propionate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (C).
+No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate (A), butyrate (B), and propionate (C) concentrations in fecal samples of individuals without colon tumors (Control), adenomas, and carcinomas (C).
 The black lines indicate the median SCFA concentration.
-Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D).
-Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E).
-Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F).
+Acetate (D), butyrate (E), and propionate (F) concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red).
 The black dots and lines represent the median change in SCFA concentration.
@@
 **Figure 2.
-No change in butyrate producing genes identified between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A).
+No change in the butyrate kinase gene abudnance was observed between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A).
 Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B).
-The other butyrate pathway genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.
 
 **Figure 3.
-Patients with adenomas had the higest number of significant negative correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
+Patients with adenomas had the higest number of statistically significant correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
 Fewer significant positive correlations were observed overall.
@@
 **Figure 4.
-SCFA concentrations do not improve OTU-based Random Forest models.** The area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A).
+SCFA concentrations do not improve OTU-based Random Forest models.** Individuals without tumors versus patients with adenomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (A).
 The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma model (B).
 The top 10 most important OTUs in the OTU adenoma model (C).
-The area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (D).
+Individuals without tumors versus patients with carcinomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D).
 The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma model (E).
@@
 The model accuracy of predicted SCFA concentrations differed between individuals without tumors, patients with adenomas, and patients with carcinomas.
-Generally, patients with carcinomas had predicted concentrations closest to their actual measured concentration (B).
+Differences between the actual and predicted SCFA concentrations based on whether individuals did not have tumors (Normal), had an adenoma, or had a carcinoma (B).
 The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
@@
 Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
-Fewer significant OTUs were observed in individuals without tumors (normal) and patients with carcinomas versus patients with adenomas.
 
@@
 OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
-The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas.
-Patients with adenomas consistently had the best classification accuracy (B).
-The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
+The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas (B).
+The top 10 OTUs based on mean decrease in accuracy for each SCFA model, colored by their lowest taxonomic identification (C).

### 2018-08-08 submission/manuscript.tex
fix figure legend captions

@@
 running the short-chain fatty acid analysis on the High-Performance
-Liquid Chromatography machine at the University of Michigan. Salary
-support for Marc A. Sze came from the Canadian Institute of Health
-Research and NIH grant UL1TR002240. Salary support for Patrick D.
-Schloss came from NIH grants P30DK034933 and 1R01CA215574.
+Liquid Chromatography machine at the University of Michigan. We would
+also like to thank Ada Hagan for providing valuable feedback on earlier
+drafts of the manuscript. Salary support for Marc A. Sze came from the
+Canadian Institute of Health Research and NIH grant UL1TR002240. Salary
+support for Patrick D. Schloss came from NIH grants P30DK034933 and
+1R01CA215574.
 
@@
 \textbf{Figure 1. No change in SCFA measurements was observed between
-normal, adenoma, and carcinoma individuals using HPLC.} Acetate
-concentrations in fecal samples of individuals without colon tumors,
-adenomas, and carcinomas (A). Butyrate concentrations in fecal samples
-of individuals without colon tumors, adenomas, and carcinomas (B).
-Propionate concentrations in fecal samples of individuals without colon
-tumors, adenomas, and carcinomas (C). The black lines indicate the
-median SCFA concentration. Acetate concentrations in fecal samples
-before and after treatment for adenoma (yellow) and carcinoma (red) (D).
-Butyrate concentrations in fecal samples before and after treatment for
-adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in
-fecal samples before and after treatment for adenoma (yellow) and
-carcinoma (red) (F). The black dots and lines represent the median
-change in SCFA concentration.
-
-\textbf{Figure 2. No change in butyrate producing genes identified
-between normal, adenoma, and carcinoma individuals.} Imputed gene
-relative abundance of important butyrate pathway genes using PICRUSt
-(A). Counts per million (corrected for size and number of contigs in an
-OPF) for the Butyrate Kinase gene (B). The other butyrate pathway genes
-from the PICRUSt analysis did not align to any of the OPFs in the
-metagenome analysis.
+normal, adenoma, and carcinoma individuals using HPLC.} Acetate (A),
+butyrate (B), and propionate (C) concentrations in fecal samples of
+individuals without colon tumors (Control), adenomas, and carcinomas
+(C). The black lines indicate the median SCFA concentration. Acetate
+(D), butyrate (E), and propionate (F) concentrations in fecal samples
+before and after treatment for adenoma (yellow) and carcinoma (red). The
+black dots and lines represent the median change in SCFA concentration.
+
+\textbf{Figure 2. No change in the butyrate kinase gene abudnance was
+observed between normal, adenoma, and carcinoma individuals.} Imputed
+gene relative abundance of important butyrate pathway genes using
+PICRUSt (A). Counts per million (corrected for size and number of
+contigs in an OPF) for the Butyrate Kinase gene (B).
 
 \textbf{Figure 3. Patients with adenomas had the higest number of
-significant negative correlations between OTU relative abundance and
-SCFA concentration.} Colors denote the family or lowest taxonomic ID
+statistically significant correlations between OTU relative abundance
+and SCFA concentration.} Colors denote the family or lowest taxonomic ID
 that an OTU classified to. Fewer significant positive correlations were
@@
 \textbf{Figure 4. SCFA concentrations do not improve OTU-based Random
-Forest models.} The area under the curve of 100 different 80/20 split
-OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
-(A). The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma
-model (B). The top 10 most important OTUs in the OTU adenoma model (C).
-The area under the curve of 100 different 80/20 OTU-based normal versus
-carcinoma 10-fold CV models with and without SCFAs (D). The top 10 most
-important OTUs or SCFAs in the SCFA and OTU carcinoma model (E). The top
-10 most important OTUs in the OTU carcinoma model (F). For (A) and (D)
-the black line represents the median AUC. The dotted line highlights an
-AUC of 0.5.
+Forest models.} Individuals without tumors versus patients with adenomas
+model with and without SCFAs area under the curve (AUC) from 100 80/20
+splits with 10-fold cross-validation (A). The top 10 most important OTUs
+or SCFAs in the SCFA and OTU adenoma model (B). The top 10 most
+important OTUs in the OTU adenoma model (C). Individuals without tumors
+versus patients with carcinomas model with and without SCFAs area under
+the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D).
+The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma
+model (E). The top 10 most important OTUs in the OTU carcinoma model
+(F). For (A) and (D) the black line represents the median AUC. The
+dotted line highlights an AUC of 0.5.
 
@@
 predicted SCFA concentrations differed between individuals without
-tumors, patients with adenomas, and patients with carcinomas. Generally,
-patients with carcinomas had predicted concentrations closest to their
-actual measured concentration (B). The top 10 OTUs based on mean
-decrease in accuracy (MDA) for each SCFA model, colored by their lowest
-taxonomic identification (C).
+tumors, patients with adenomas, and patients with carcinomas.
+Differences between the actual and predicted SCFA concentrations based
+on whether individuals did not have tumors (Normal), had an adenoma, or
+had a carcinoma (B). The top 10 OTUs based on mean decrease in accuracy
+(MDA) for each SCFA model, colored by their lowest taxonomic
+identification (C).
 
@@
 groups.} Colors denote the family or lowest taxonomic ID that an OTU
-classified to. Fewer significant OTUs were observed in individuals
-without tumors (normal) and patients with carcinomas versus patients
-with adenomas.
+classified to.
 
@@
 of predicted high/low SCFA groups differed between individuals without
-tumors, patients with adenomas, and patients with carcinomas. Patients
-with adenomas consistently had the best classification accuracy (B). The
-top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA
-model, colored by their lowest taxonomic identification (C).
+tumors, patients with adenomas, and patients with carcinomas (B). The
+top 10 OTUs based on mean decrease in accuracy for each SCFA model,
+colored by their lowest taxonomic identification (C).
 

### 2018-08-08 submission/manuscript.Rmd
change to normal referencing

@@
 **Figure 1.
-No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate (A), butyrate (B), and propionate (C) concentrations in fecal samples of individuals without colon tumors (Control), adenomas, and carcinomas (C).
+No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate (A), butyrate (B), and propionate (C) concentrations in fecal samples of individuals without colon tumors (Normal), adenomas, and carcinomas (C).
 The black lines indicate the median SCFA concentration.

### 2018-08-08 submission/manuscript.tex
change to normal referencing

@@
 butyrate (B), and propionate (C) concentrations in fecal samples of
-individuals without colon tumors (Control), adenomas, and carcinomas
-(C). The black lines indicate the median SCFA concentration. Acetate
-(D), butyrate (E), and propionate (F) concentrations in fecal samples
-before and after treatment for adenoma (yellow) and carcinoma (red). The
-black dots and lines represent the median change in SCFA concentration.
+individuals without colon tumors (Normal), adenomas, and carcinomas (C).
+The black lines indicate the median SCFA concentration. Acetate (D),
+butyrate (E), and propionate (F) concentrations in fecal samples before
+and after treatment for adenoma (yellow) and carcinoma (red). The black
+dots and lines represent the median change in SCFA concentration.
 

### 2018-08-08 code/make_pi_opf_combined_graph.R
convert to normal instead of control

@@ -32,7 +32,7 @@ imputed_fig <- picrust_data %>%
   filter(kegg_ortholog == "K00929") %>% 
   mutate(dx = factor(dx, 
                 levels = c("normal", "adenoma", "cancer"), 
-                labels = c("Control", "Adenoma", "Carcinoma"))) %>% 
+                labels = c("Normal", "Adenoma", "Carcinoma"))) %>% 
   ggplot(aes(kegg_ortholog, log10(rel_abund), fill = dx)) + 
   geom_boxplot(position = position_dodge(width = 1)) + 
   geom_vline(xintercept=seq(1.5, length(unique(picrust_data$kegg_ortholog))-0.5, 1), 
@@ -54,7 +54,7 @@ imputed_fig <- picrust_data %>%
 metagenome_fig <- opf_data %>% 
   mutate(disease = factor(disease, 
                 levels = c("Healthy", "Adenoma", "Cancer"), 
-                labels = c("Control", "Adenoma", "Carcinoma")), 
+                labels = c("Normal", "Adenoma", "Carcinoma")), 
          kegg_id = factor(kegg_id, 
                           levels = c("K00929"), labels = c("Butyrate Kinase"))) %>% 
   ggplot(aes(kegg_id, combined_count, fill = disease)) + 

### 2018-08-08 code/make_scfa_measures_hplc_figure.R

@@ -115,7 +115,7 @@ cross_graph <- function(scfa, dataList){
     ggplot(aes(
       factor(dx, 
              levels = c("normal", "adenoma", "cancer"), 
-             labels = c("Control", "Adenoma", "Carcinoma")), log10(mmol_kg+1))) + 
+             labels = c("Normal", "Adenoma", "Carcinoma")), log10(mmol_kg+1))) + 
     geom_jitter(aes(color = dx), width = 0.3, size = 2.5, alpha = 0.75, show.legend = F) + 
     stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                  colour = "black", geom = "crossbar", size = 0.75, width = 0.5)  +

### 2018-08-08 code/make_class_group_scfa_sig_graph.R
add italics

@@ -56,7 +56,7 @@ number_class_plot <- graph_data %>%
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         legend.position = "bottom", 
-        legend.text = element_text(size = 10), 
+        legend.text = element_text(size = 10, face = "italic"), 
         axis.text.y = element_text(size = 10), 
         axis.title = element_text(size = 12), 
         axis.text.x = element_text(size = 10, face = "bold"))

### 2018-08-08 code/make_reg_group_scfa_sig_graph.R

@@ -63,7 +63,7 @@ number_corr_plot <- graph_data %>%
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         legend.position = "bottom", 
-        legend.text = element_text(size = 10), 
+        legend.text = element_text(size = 10, face = "italic"), 
         axis.text.y = element_text(size = 10), 
         axis.title = element_text(size = 12), 
         axis.text.x = element_text(size = 10, face = "bold"))

### 2018-08-08 submission/supplemental.Rmd
update with ada suggestions

--- 2018-05-24 submission/supplemental.Rmd
@@
 
-**Table S1: Breakdown of select genes and corresponding KEGG ID analyzed.**
+**Table S1: SCFA enzymes and corresponding KEGG ID analyzed.**
 ```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-colnames(sup_table1) <-  c("KEGG IDs", "Gene Name")
+colnames(sup_table1) <-  c("KEGG IDs", "Enzyme Name")
 
@@
   select(kegg_ortholog, group, statistic, p.value, bh) %>% 
-   mutate_at(vars(p.value:bh), function(x) round(x, digits = 3))
+   mutate_at(vars(p.value:bh), function(x) round(x, digits = 3)) %>% 
+  arrange(kegg_ortholog, group)
 
@@
 ```
+
+\newpage
+
+**Table S5: Spearman tests between OTU relative abundance and SCFA concentrations.**
+
+\scriptsize
+
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+
+table_s5 <- read_csv("../data/process/tables/significant_reg_otu_comp_summary.csv") %>% 
+  left_join(read_tsv("../data/process/final.taxonomy") %>% 
+              select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
+              select(OTU, family), by = c("otu" = "OTU")) %>% 
+  select(otu, family, genus, dx, scfa, estimate, p.value, bh) %>% 
+  mutate_at(vars(p.value:bh), function(x) format(x, scientific = T, digits = 3)) %>% 
+  mutate(estimate = round(estimate, digits = 3), 
+         family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         genus = str_replace_all(family, "_unclassified", ""), 
+         genus = str_replace_all(family, "_", " ")) %>% 
+  arrange(scfa, dx, bh)
+
+colnames(table_s5) <-  c("OTU", "Family", "Genus", "Group", "SCFA", "Rho", "P-value", "BH")
+
+kable(table_s5, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\normalsize
+
+\newpage
+
+**Table S6: Wilcoxson tests between OTU relative abundance and high or low SCFA concentrations based on the overall median.**
+
+\scriptsize
+
+```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+
+table_s6 <- read_csv("../data/process/tables/significant_class_otu_comp_summary.csv") %>% 
+  left_join(read_tsv("../data/process/final.taxonomy") %>% 
+              select(-Size) %>% 
+  mutate(Taxonomy = str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>% 
+  separate(Taxonomy, c("kingdom", "phyla", "class", "order", "family", "genus", "species"), sep = ";") %>% 
+              select(OTU, family), by = c("otu" = "OTU")) %>% 
+  select(otu, family, genus, dx, scfa, p.value, bh) %>% 
+  mutate_at(vars(p.value:bh), function(x) format(x, scientific = T, digits = 3)) %>% 
+  mutate(family = str_replace_all(family, "_unclassified", ""), 
+         family = str_replace_all(family, "_", " "), 
+         genus = str_replace_all(family, "_unclassified", ""), 
+         genus = str_replace_all(family, "_", " ")) %>% 
+  arrange(scfa, dx, bh)
+
+colnames(table_s6) <-  c("OTU", "Family", "Genus", "Group", "SCFA", "P-value", "BH")
+
+kable(table_s6, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\normalsize

### 2018-08-09 submission/manuscript.Rmd
final 1st draft read through

@@
 The majority of the risk for developing CRC is attributable to environmental factors.
-One of these environmental factors is the microbiota, with certain bacterial community members being associated with CRC and other taxa being associated to individuals without tumors.
-Some of the bacterial species in taxa associated to individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems.
+One of these environmental factors is the microbiota, with certain bacterial community members being more prevalent in CRC while other taxa are higher in individuals without tumors.
+Some of the bacterial species in taxa that are more abundant in individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems.
 However, the data supporting the importance of SCFAs in human CRC is less certain.
@@
 
-**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations in patients in the general CRC population are different, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC.
+**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations are different in the general CRC population, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC.
 Alternatively, our observations also support the hypothesis that there may be other metabolites or mechanisms (e.g.
@@
 Additionally, prior investigations grouped patients with adenoma and individuals without tumors into a single group to compare against patients with carcinomas.
-Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without tumors.
-We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without tumors into a single group.
+Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without colon tumors.
+We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without colon tumors into a single group.
 Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
@@
 The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
-To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas.
+To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes encoding enzymes involved with SCFA synthesis between individuals without colon tumors, patients with adenomas, and patients with carcinomas.
 Next, we investigated whether taxa associated with SCFA production were important to disease classification models.
-First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas.
+First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without colon tumors and patients with adenomas or carcinomas.
 Second, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002].
@@
 
-**Gene abundance for enzymes involved in SCFA synthesis are the same for individuals without tumors and patients with adenomas or carcinomas.** In order to provide further confirmation and support of our SCFA concentration results we investigated the genes encoding specific enzymes involved with SCFA synthesis [Table S1].
+**Gene abundance for enzymes involved in SCFA synthesis are the same for individuals without colon tumors and patients with adenomas or carcinomas.** In order to provide further confirmation and support of our SCFA concentration results we investigated the genes encoding specific enzymes involved with SCFA synthesis [Table S1].
 Using this list of specific genes [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
@@
 We found no difference in any of the imputed gene relative abundances for enzymes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
-Since butyrate is one of the most studied SCFAs in the context of CRC, we visualized the observed lack of difference between indviduals without tumors and patients with adenomas or carcinomas in gene abundance for enzymes involved with SCFA synthesis using the butyrate kinase gene [Figure 2A].
+Since butyrate is one of the most studied SCFAs in the context of CRC, we visualized the observed lack of difference between indviduals without colon tumors and patients with adenomas or carcinomas in gene abundance for enzymes involved with SCFA synthesis using the butyrate kinase gene [Figure 2A].
 Additionally, using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
@@
 Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenomas (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinomas (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
-This similarity between indviduals without tumors and patients with adenomas or carcinomas is highlighted again by visualizing the gene abundance for the butyrate kinase gene [Figure 2B].
+This similarity between indviduals without colon tumors and patients with adenomas or carcinomas is highlighted again by visualizing the gene abundance for the butyrate kinase gene [Figure 2B].
 These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors and provides further support for our original SCFA concentration observations.
 
-**The number of OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
-The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have tumors, had an adenoma, or had a carcinoma.
+**The number of OTUs positively associated with SCFA concentration were similar between individuals without colon tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without colon tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
+The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma.
 To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero.
@@
 In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3].
-Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
-Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3].
+Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without colon tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
+Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without colon tumors and patients with a carcinoma [Figure 3].
 Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6].
@@
 After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations.
-When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration remained as part of the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations.
+When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration was in the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations.
 In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.
 
-**The most important OTUs in Random Forest models built to classify SCFA concentrations or tumors are not the same.** It is possible that due to the way the Random Forest algorithm works, OTUs associated with SCFA concentrations could be downweighted in importance within the adenoma or carcinoma models when SCFA concentrations are included.
+**The most important OTUs for SCFA concentration and tumor Random Forest models are not the same.** It is possible that due to the way the Random Forest algorithm works, OTUs associated with SCFA concentrations could be downweighted in importance within the adenoma or carcinoma models when SCFA concentrations are included.
 To test if this is the case we used OTU data and built Random Forest models to classify SCFA concentrations.
@@
 Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa were important for these classifications [Figure 5A].
-There also was a difference in accuracy based on whether the fecal sample was from an individual without tumors or from patients with adenomas or carcinomas [Figure 5B].
+There also was a difference in accuracy based on whether the fecal sample was from an individual without colon tumors or from patients with adenomas or carcinomas [Figure 5B].
 When comparing the adenoma or carcinoma model to the SCFA concentration models there was minimal overlap between these model's most important OTUs [Figure 4B-C, 4E-F and 5C-E].
@@
 Collectively, these observations provide evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and accordingly these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
-Although it is possible to identify OTUs associated with SCFA production, overall, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with adenomas or carcinomas.
+Although it is possible to identify OTUs associated with SCFA production, overall, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no colon tumors and patients with adenomas or carcinomas.
 
@@
 Whether we directly measured the SCFA concentration or investigated genes that encoded enzymes used for their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2].
-Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1].
+There were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have colon tumors, had an adenoma, or had a carcinoma.
+However, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1].
 In models with SCFA concentrations included, many OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables for the model [Figure 4].
 Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to these taxa in the tumor models [Figure 4-5].
-Collectively, our observations suggest that the resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* that are different between individuals without tumors and patients with adenomas or carcinomas, are not the same as those involved with SCFA production.
+Collectively, our observations suggest that the resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* that are different between individuals without colon tumors and patients with adenomas or carcinomas, are not the same as those involved with SCFA production.
 
@@
 One reason for these results is that SCFA concentrations and responses to fiber vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
-This information taken together with our observations would suggest that either individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.
+This information taken together with our observations would suggest that either SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer or that individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit.
 
-Another possible alternative explanation as to why no difference in SCFA concentration between individuals without tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs.
+Another possible alternative explanation as to why no difference in SCFA concentration between individuals without colon tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs.
 One limitation of current research into the effect of SCFAs and the microbiota in CRC has been that all tumors are treated as the same type.
@@
 The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Our data is consistent with the literature despite finding no difference in SCFA concentrations between individuals without tumors and patients with adenomas or carcinomas.
+Our data is consistent with the literature despite finding no difference in SCFA concentrations between individuals without colon tumors and patients with adenomas or carcinomas.
 In particular, we observed that the important OTUs in our tumor models were to taxa that are associated with SCFA production but that these OTUs were not themselves associated with acetate, butyrate, or propionate production.
@@
 This leads to the possibility that protection may be through other routes.
-First, there could be a different pathway or other less extensively studied metabolites that provide the necessary protection against tumorigenesis.
+For example, there could be a different pathway or other less extensively studied metabolites that provide the necessary protection against tumorigenesis.
 Alternatively, previous research has shown that non-resident oral microbes, such as *Fusobacterium*, *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* are associated with carcinomas [@Zeller2014; @Baxter2016; @Kostic2011] and can increase inflammation [@Flynn2016].

### 2018-08-09 submission/manuscript.tex
final 1st draft read through

@@
 environmental factors is the microbiota, with certain bacterial
-community members being associated with CRC and other taxa being
-associated to individuals without tumors. Some of the bacterial species
-in taxa associated to individuals without tumors can use fiber to
-produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
+community members being more prevalent in CRC while other taxa are
+higher in individuals without tumors. Some of the bacterial species in
+taxa that are more abundant in individuals without tumors can use fiber
+to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
 model systems. However, the data supporting the importance of SCFAs in
@@
 \textbf{Conclusions.} Although our data does not support the hypothesis
-that fecal SCFA concentrations in patients in the general CRC population
-are different, there still may be specific types of colorectal tumors
-where SCFAs may be beneficial for treatment of CRC. Alternatively, our
+that fecal SCFA concentrations are different in the general CRC
+population, there still may be specific types of colorectal tumors where
+SCFAs may be beneficial for treatment of CRC. Alternatively, our
 observations also support the hypothesis that there may be other
@@
 in SCFA concentration would also be observed between patients with
-adenomas and indivdiuals without tumors. We also can test this
+adenomas and indivdiuals without colon tumors. We also can test this
 suggestion because we have a larger number of patients with adenomas
 within our study and do not need to group patients with adenomas and
-individuals without tumors into a single group. Additionally, we build
-upon these observations by assessing the utility of using SCFAs and
-Operational Taxonomic Units (OTUs) as a risk stratification tool of
+individuals without colon tumors into a single group. Additionally, we
+build upon these observations by assessing the utility of using SCFAs
+and Operational Taxonomic Units (OTUs) as a risk stratification tool of
 colorectal tumors (adenoma or carcinoma). We also investigate whether
@@
 provide further support for our SCFA findings we also used PICRUSt (21)
-and metagenomic sequencing to investigate differences in genes involved
-with SCFA synthesis between individuals without tumors, patients with
-adenomas, and patients with carcinomas. Next, we investigated whether
-taxa associated with SCFA production were important to disease
-classification models. First, using the cross-sectional data, we
-analyzed the number of correlations between OTU relative abundance and
-SCFA concentrations across individuals without tumors and patients with
-adenomas or carcinomas. Second, we assessed the affect adding SCFA
-concentrations to OTU data had on classification of patients with
-adenomas or carcinomas using the Random Forest algorithm (22). Third, we
-analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
-Finally, we compared whether SCFA concentrations replace important taxa
-in disease models and whether the same OTUs are the most important
-variables used to classify disease and SCFA concentration. Collectively,
-this investigation provides additional information as to whether SCFAs
-are decreased in patients with colorectal tumors and provides context as
-to whether targeting taxa to increase SCFA concentrations is a viable
-option to protect against colon tumorigenesis.
+and metagenomic sequencing to investigate differences in genes encoding
+enzymes involved with SCFA synthesis between individuals without colon
+tumors, patients with adenomas, and patients with carcinomas. Next, we
+investigated whether taxa associated with SCFA production were important
+to disease classification models. First, using the cross-sectional data,
+we analyzed the number of correlations between OTU relative abundance
+and SCFA concentrations across individuals without colon tumors and
+patients with adenomas or carcinomas. Second, we assessed the affect
+adding SCFA concentrations to OTU data had on classification of patients
+with adenomas or carcinomas using the Random Forest algorithm (22).
+Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA
+concentrations. Finally, we compared whether SCFA concentrations replace
+important taxa in disease models and whether the same OTUs are the most
+important variables used to classify disease and SCFA concentration.
+Collectively, this investigation provides additional information as to
+whether SCFAs are decreased in patients with colorectal tumors and
+provides context as to whether targeting taxa to increase SCFA
+concentrations is a viable option to protect against colon
+tumorigenesis.
 
@@
 \textbf{Gene abundance for enzymes involved in SCFA synthesis are the
-same for individuals without tumors and patients with adenomas or
+same for individuals without colon tumors and patients with adenomas or
 carcinomas.} In order to provide further confirmation and support of our
@@
 is one of the most studied SCFAs in the context of CRC, we visualized
-the observed lack of difference between indviduals without tumors and
-patients with adenomas or carcinomas in gene abundance for enzymes
+the observed lack of difference between indviduals without colon tumors
+and patients with adenomas or carcinomas in gene abundance for enzymes
 involved with SCFA synthesis using the butyrate kinase gene {[}Figure
@@
 \textgreater{} 0.70) {[}Table S4{]}. This similarity between indviduals
-without tumors and patients with adenomas or carcinomas is highlighted
-again by visualizing the gene abundance for the butyrate kinase gene
-{[}Figure 2B{]}. These observations provide evidence that gene
-prevalence for enzymes involved in SCFA synthesis does not change due to
-colorectal tumors and provides further support for our original SCFA
-concentration observations.
+without colon tumors and patients with adenomas or carcinomas is
+highlighted again by visualizing the gene abundance for the butyrate
+kinase gene {[}Figure 2B{]}. These observations provide evidence that
+gene prevalence for enzymes involved in SCFA synthesis does not change
+due to colorectal tumors and provides further support for our original
+SCFA concentration observations.
 
 \textbf{The number of OTUs positively associated with SCFA concentration
-were similar between individuals without tumors and patients with
+were similar between individuals without colon tumors and patients with
 adenomas or carcinomas.} Having found no difference between individuals
-without tumors and patients with adenomas or carcinomas in SCFA
+without colon tumors and patients with adenomas or carcinomas in SCFA
 concentrations or genes encoding enzymes involved with SCFA synthesis,
@@
 were OTUs that were significantly associated with SCFA concentrations
-and if this was different between individuals that did not have tumors,
-had an adenoma, or had a carcinoma. To accomplish this we used Spearmans
-rho, a non-parametric measure of association, and tested if there was a
-correlation that was signficantly greater than zero. We found that taxa
-from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
+and if this was different between individuals that did not have colon
+tumors, had an adenoma, or had a carcinoma. To accomplish this we used
+Spearmans rho, a non-parametric measure of association, and tested if
+there was a correlation that was signficantly greater than zero. We
+found that taxa from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
 \emph{Ruminococcaceae} dominated statistically significant OTU
@@
 and SCFA concentrations, but their total number was more similar to
-individuals without tumors or patients with carcinomas versus the
+individuals without colon tumors or patients with carcinomas versus the
 analogous comparison for the number of negative correlations {[}Figure
 3{]}. Additionally, the number of positive correlations between OTUs and
-SCFA concentrations was similar between individuals without tumors and
-patients with a carcinoma {[}Figure 3{]}. Finally, when we used high/low
-SCFA groups based on the overall median concentration for each SCFA
-instead of SCFA concentrations a similar pattern was still observed
+SCFA concentrations was similar between individuals without colon tumors
+and patients with a carcinoma {[}Figure 3{]}. Finally, when we used
+high/low SCFA groups based on the overall median concentration for each
+SCFA instead of SCFA concentrations a similar pattern was still observed
 {[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
@@
 SCFA concentrations were added to the carcinoma model, only 1 OTU that
-was associated with SCFA concentration remained as part of the top 10
-most important variables and it was negatively associated with acetate
-and butyrate concentrations. In combination with the previous results on
-OTU correlations, these observations provide additional evidence that
-the resident taxa that are associated with protection against
-tumorigeneis are not ones associated with acetate, butyrate, or
-propionate production.
-
-\textbf{The most important OTUs in Random Forest models built to
-classify SCFA concentrations or tumors are not the same.} It is possible
-that due to the way the Random Forest algorithm works, OTUs associated
-with SCFA concentrations could be downweighted in importance within the
-adenoma or carcinoma models when SCFA concentrations are included. To
-test if this is the case we used OTU data and built Random Forest models
-to classify SCFA concentrations. Overall, the correlation between the
-predicted and actual SCFA concentrations were moderately associated with
-each other {[}Figure 5A{]}. Additionally, because the training set
+was associated with SCFA concentration was in the top 10 most important
+variables and it was negatively associated with acetate and butyrate
+concentrations. In combination with the previous results on OTU
+correlations, these observations provide additional evidence that the
+resident taxa that are associated with protection against tumorigeneis
+are not ones associated with acetate, butyrate, or propionate
+production.
+
+\textbf{The most important OTUs for SCFA concentration and tumor Random
+Forest models are not the same.} It is possible that due to the way the
+Random Forest algorithm works, OTUs associated with SCFA concentrations
+could be downweighted in importance within the adenoma or carcinoma
+models when SCFA concentrations are included. To test if this is the
+case we used OTU data and built Random Forest models to classify SCFA
+concentrations. Overall, the correlation between the predicted and
+actual SCFA concentrations were moderately associated with each other
+{[}Figure 5A{]}. Additionally, because the training set
 R\textsuperscript{2} was always higher than the test set
@@
 {[}Figure 5A{]}. There also was a difference in accuracy based on
-whether the fecal sample was from an individual without tumors or from
-patients with adenomas or carcinomas {[}Figure 5B{]}. When comparing the
-adenoma or carcinoma model to the SCFA concentration models there was
-minimal overlap between these model's most important OTUs {[}Figure
-4B-C, 4E-F and 5C-E{]}. The only OTU that was in the top 10 most
-important variables and had overlap between the models was OTU00167
-(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E{]}. Similar
+whether the fecal sample was from an individual without colon tumors or
+from patients with adenomas or carcinomas {[}Figure 5B{]}. When
+comparing the adenoma or carcinoma model to the SCFA concentration
+models there was minimal overlap between these model's most important
+OTUs {[}Figure 4B-C, 4E-F and 5C-E{]}. The only OTU that was in the top
+10 most important variables and had overlap between the models was
+OTU00167 (\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E{]}. Similar
 observations were made when using high/low SCFA groups based on the
@@
 concentration or OTUs associated with their production are different
-between individuals with no tumors and patients with adenomas or
+between individuals with no colon tumors and patients with adenomas or
 carcinomas.
@@
 identified between individuals without tumors and patients with adenomas
-or carcinomas {[}Figure 1 \& 2{]}. Although there were differences in
-the number of significant correlations between SCFA concentration and
-OTU relative abundance based on whether individuals did not have tumors,
-had an adenoma, or had a carcinoma, SCFA concentrations did not provide
-increased model accuracy for tumor classification {[}Figure 3-4 \&
-S1{]}. In models with SCFA concentrations included, many OTUs that
+or carcinomas {[}Figure 1 \& 2{]}. There were differences in the number
+of significant correlations between SCFA concentration and OTU relative
+abundance based on whether individuals did not have colon tumors, had an
+adenoma, or had a carcinoma. However, SCFA concentrations did not
+provide increased model accuracy for tumor classification {[}Figure 3-4
+\& S1{]}. In models with SCFA concentrations included, many OTUs that
 classified to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
@@
 \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} that are different
-between individuals without tumors and patients with adenomas or
+between individuals without colon tumors and patients with adenomas or
 carcinomas, are not the same as those involved with SCFA production.
@@
 information taken together with our observations would suggest that
-either individuals who do not respond to fiber supplementation would
-need to acquire these bacteria to achieve a benefit or that SCFAs
-provide little to no benefit as an anti-tumorigenic compound in
-colorectal cancer.
+either SCFAs provide little to no benefit as an anti-tumorigenic
+compound in colorectal cancer or that individuals who do not respond to
+fiber supplementation would need to acquire these bacteria to achieve a
+benefit.
 
 Another possible alternative explanation as to why no difference in SCFA
-concentration between individuals without tumors and patients with
+concentration between individuals without colon tumors and patients with
 adenomas or carcinomas was observed, could be because only certain types
@@
 Our data is consistent with the literature despite finding no difference
-in SCFA concentrations between individuals without tumors and patients
-with adenomas or carcinomas. In particular, we observed that the
-important OTUs in our tumor models were to taxa that are associated with
-SCFA production but that these OTUs were not themselves associated with
-acetate, butyrate, or propionate production. One of the clearest results
-to highlight this within our study were the results showing that
+in SCFA concentrations between individuals without colon tumors and
+patients with adenomas or carcinomas. In particular, we observed that
+the important OTUs in our tumor models were to taxa that are associated
+with SCFA production but that these OTUs were not themselves associated
+with acetate, butyrate, or propionate production. One of the clearest
+results to highlight this within our study were the results showing that
 different OTUs from the same taxonomic classification are in tumor and
 SCFA Random Forest models {[}Figure 4 \& 5{]}. This leads to the
-possibility that protection may be through other routes. First, there
-could be a different pathway or other less extensively studied
+possibility that protection may be through other routes. For example,
+there could be a different pathway or other less extensively studied
 metabolites that provide the necessary protection against tumorigenesis.

### 2018-08-09 submission/manuscript.Rmd
methods final draft 1 read through

@@
 The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma.
-To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero.
+To accomplish this we used Spearman's rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero.
 We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5].
@@
 For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
-Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017].
+Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017].
 Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016].
@@
 Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
-The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].
+The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].
 
@@
 **Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017].
-Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used.
+Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas.
 Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995].
@@
 We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment.
-Next, the imputed gene counts of important mediators of SCFA synthesis was tested.
-Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.
-From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho.
+Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested.
+Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed.
+From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho.
 Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models.
-Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm.
+Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm.
 For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

### 2018-08-09 submission/manuscript.tex
methods final draft 1 read through

@@
 tumors, had an adenoma, or had a carcinoma. To accomplish this we used
-Spearmans rho, a non-parametric measure of association, and tested if
+Spearman's rho, a non-parametric measure of association, and tested if
 there was a correlation that was signficantly greater than zero. We
@@
 sample ranged from 188 - 546 days. Our use of treatment has been
-previously defined as encompassing removal of a tumor with or without
-chemotherapy and radiation (29). Diagnosis of tumor was made by
-colonoscopic examination and histopathological review of biopsies
-obtained (29, 30). The University of Michigan Institutional Review Board
-approved the study and informed consent was obtained from all
-participants in accordance to the guidelines set out by the Helsinki
-Declaration.
+previously defined as encompassing removal of a tumor (surgery or
+colonoscopy) with or without chemotherapy and radiation (29). Diagnosis
+of tumor was made by colonoscopic examination and histopathological
+review of biopsies obtained (29, 30). The University of Michigan
+Institutional Review Board approved the study and informed consent was
+obtained from all participants in accordance to the guidelines set out
+by the Helsinki Declaration.
 
@@
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptClust algorithm (35).
+97\% similarity using the OptiClust algorithm (35).
 
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (44). Generally, a Kruskal-Walis rank sum
+the statistical language R (44). Generally, a Kruskal-Wallis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
-the groups used. Where appropriate Benjamini-Hochberg was used to
+individuals without colon tumors, patients with adenomas, and patients
+with carcinomas. Where appropriate Benjamini-Hochberg was used to
 correct for multiple comparisons (45). First, we assessed differences in
@@
 carcinoma pre- versus post-treatment. Next, the imputed gene counts of
-important mediators of SCFA synthesis was tested. Additionally, the
-counts generated for OPFs that matched important genes involved with
-SCFA creation were analyzed. From here we analyzed the number of
+genes encoding enzymes involved in SCFA synthesis was tested.
+Additionally, metagenomic sequencing counts for important genes involved
+with SCFA production were analyzed. From here we analyzed the number of
 significant positive and negative correlations between OTU relative
-abundance and SCFA concentrations in individuals without tumors and
-patients with adenomas or carcinomas using Spearman's rho. Next, we
+abundance and SCFA concentrations in individuals without colon tumors
+and patients with adenomas or carcinomas using Spearman's rho. Next, we
 assessed whether OTUs alone or OTUs and SCFAs were better able to
 classify individuals with and without tumors using Random Forest models.
-Finally, models to classify high or low SCFA concentration based on the
-median of each SCFA or the actual concentration using 16S rRNA gene
+Finally, models to classify the actual SCFA concentration or high/low
+SCFA concentration based on the median of each SCFA using 16S rRNA gene
 sequencing data was created using the Random Forest algorithm. For all

### 2019-03-11 submission/manuscript.Rmd
Rewrite introduction

@@
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
-source('../code/functions.R')
+\vspace{35mm}
 
-loadLibs(c("tidyverse", "knitr", "rmarkdown"))
-```
-
-```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-# Read in meta data
-metaf <- read_csv("../data/raw/metadata/good_metaf_final.csv")
-metai <- read_csv("../data/raw/metadata/metaI_final.csv")
-geof_samples <- read_tsv("../data/raw/metadata/303030_subset/NexteraXT003Map.tsv", 
-                         col_names = F) %>% filter(X22 != "Negative")
-
-# Read in analysis of SCFA by HPLC
-hplc_groups_scfa_kruskal <- read_csv("../data/process/tables/acetate_kruskal_crc_groups.csv") %>% 
-  mutate(scfa = "acetate") %>% slice(1) %>% 
-  bind_rows(
-    read_csv("../data/process/tables/butyrate_kruskal_crc_groups.csv") %>% 
-      mutate(scfa = "butyrate") %>% slice(1), 
-    read_csv("../data/process/tables/propionate_kruskal_crc_groups.csv") %>% 
-      mutate(scfa = "propionate") %>% slice(1)) %>% 
-  mutate(bh = p.adjust(pvalue, method = "BH"))
-
-hplc_treat_scfa_kruskal <- read_csv("../data/process/tables/scfa_treatment_comparison_pvalues.csv") %>% 
-  filter(scfa != "isobutyrate")
-
-# Load in Random Forest with and without SCFA summary data
-disease_rf <- read_csv("../data/process/tables/rf_model_ttest_comparisons.csv")
-
-# Load in data for select genes in SCFA pathways
-picrust_select <- read_csv("../data/process/tables/specific_scfa_kruskal_picrust_summary.csv")
-picrust_select_treat <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv")
-opf_select <- read_csv("../data/process/tables/select_scfa_opf_kruskal_summary.csv")
-
-# Load in the group comparison from SCFA models
-class_group_rf_data <- read_csv("../data/process/tables/full_model_class_group_results.csv")
-reg_group_rf_data <- read_csv("../data/process/tables/full_model_reg_group_results.csv")
-
-# Load in sig associations and top 10 RF model presence
-sig_assoc_top10_model <- read.csv("../data/process/tables/sig_association_and_RF_top10.csv")
-
-```
+Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure
 
@@
 
-Running title: SCFAs and colorectal tumors
-
-\vspace{35mm}
-
-Marc A. Sze${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
+Marc A. Sze${^1}$, Begüm Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
 
@@
 
-**Background.** Colorectal cancer (CRC) is increasing in prevalence in individuals under 50 and because of this will be a continuing health concern for the foreseeable future.
-The majority of the risk for developing CRC is attributable to environmental factors.
-One of these environmental factors is the microbiota, with certain bacterial community members being more prevalent in CRC while other taxa are higher in individuals without tumors.
-Some of the bacterial species in taxa that are more abundant in individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems.
-However, the data supporting the importance of SCFAs in human CRC is less certain.
-Here, we test the hypothesis that SCFA concentrations and taxa associated with their production are different in individuals with colorectal tumors.
+\newpage
 
-**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their fecal concentrations of acetate, butyrate, and propionate.
-Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.
+## Introduction (400 words)
 
-**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15).
-Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70).
-Finally, there was no difference between the ability of Random Forest models to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).
+Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016].
+Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990].
+This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015].
+Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
+Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016].
+Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017].
+Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012].
+These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.
 
-**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations are different in the general CRC population, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC.
-Alternatively, our observations also support the hypothesis that there may be other metabolites or mechanisms (e.g.
-bacterial niche exclusion) that may be more protective against tumorigenesis and have not been thoroughly investigated in the context of human CRC.
+SCFAs have have anti-inflammatory and anti-proliferative activities [@Encarnao2018; @Verma2018; @Zheng2017; @OKeefe2016].
+Furthermore, manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018; @Tian2018].
+These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
+Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
+Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Vital2014; @Sanna2019; @Liu2019; @Meisel2016].
+Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013].
+Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Schatzkin2000; @Yao2017; @Kunzmann2015; @Murphy2012; @Gianfredi2018].
+These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.
+
+To better understand the connection between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of SCFAs in feces of previously characterized individuals with normal colons, adenomas, and carcinomas and subset of those individuals after they underwent treatment for their lesions.
+Our goals included: (1) testing whether there was an association between SCFA concentration and tumor status, (2) determining whether SCFA concentrations could be used as biomarkers to improve the detection of colonic lesions, and (3) predicting SCFA concentrations based on the relative abundance of fecal bacteria and their genes.
 
@@
 
-## Introduction
+## Results (700 words)
 
-Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US and the prevelance is increasing in invidiauls under 50 years of age [@Haggar2009; @Siegel2016].
-Although there is a genetic component to the disease, the environment is considered a larger risk factor for CRC [@Lichtenstein2000].
-These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011].
-Many of these environmental risk factors, including the microbiota, are modifiable.
-This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018].
-Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018].
-Many of the bacterial species from these resident taxa identified in these case/control studies actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017].
-The most extensively studied SCFAs are acetate, butyrate, and propionate [@Louis2016].
-These SCFAs are hypothesized to be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.
-
-Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent.
-Specific SCFAs have shown positive results within model systems [@test_OKeefe2016].
-For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018].
-Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018].
-These exciting results in model systems suggest that supplementation with food sources that bacteria use to create these SCFAs may confer beneficial effects against CRC.
-However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.
-
-Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations.
-The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013].
-Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017].
-These findings argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis.
-Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.
-
-Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations.
-Specifically, by using a seperate cohort it tests previous case/control findings on SCFA concentrations in individuals with and without tumors.
-Additionally, prior investigations grouped patients with adenoma and individuals without tumors into a single group to compare against patients with carcinomas.
-Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without colon tumors.
-We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without colon tumors into a single group.
-Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma).
-We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations.
-Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.
-
-To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups.
-The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors.
-To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes encoding enzymes involved with SCFA synthesis between individuals without colon tumors, patients with adenomas, and patients with carcinomas.
-Next, we investigated whether taxa associated with SCFA production were important to disease classification models.
-First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without colon tumors and patients with adenomas or carcinomas.
-Second, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002].
-Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
-Finally, we compared whether SCFA concentrations replace important taxa in disease models and whether the same OTUs are the most important variables used to classify disease and SCFA concentration.
-Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.
+* There is no association between presence of tumors and SCFA concentration
+	- Wilcoxon/T-test
+	- Cross sectional
+	- Aggregating SCFAs does not improve association with CRC
+	- Combining 16S and SCFA does not improve prediction of CRC
+	- Lack of change pre and post treatment
+* Community structure is not predictive of SCFA production
+	- Random forest regression
+	- 16S
+	- picrust
+	- shotgun metagenomic data
 
@@
 
-## Results
+## Discussion (200 words)
 
-**Decreased SCFA concentrations are not associated with adenoma or carcinoma tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time.
-There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C].
-We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples.
-Although there was a general trend for increasing acetate, butyrate, and propionate concentrations following treatment for tumors, there was no significant difference between pre- and post-treatment for either patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].
-
-**Gene abundance for enzymes involved in SCFA synthesis are the same for individuals without colon tumors and patients with adenomas or carcinomas.** In order to provide further confirmation and support of our SCFA concentration results we investigated the genes encoding specific enzymes involved with SCFA synthesis [Table S1].
-Using this list of specific genes [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas.
-Although we intended to analyze all the genes in the list [Table S1], not all of the KEGG genes were identified during our analysis.
-We first analyzed imputed gene relative abundance, calculated from our OTU data that was generated from 16S rRNA gene sequencing.
-We found no difference in any of the imputed gene relative abundances for enzymes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2].
-Since butyrate is one of the most studied SCFAs in the context of CRC, we visualized the observed lack of difference between indviduals without colon tumors and patients with adenomas or carcinomas in gene abundance for enzymes involved with SCFA synthesis using the butyrate kinase gene [Figure 2A].
-Additionally, using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3].
-Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results.
-Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenomas (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinomas (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4].
-This similarity between indviduals without colon tumors and patients with adenomas or carcinomas is highlighted again by visualizing the gene abundance for the butyrate kinase gene [Figure 2B].
-These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors and provides further support for our original SCFA concentration observations.
-
-**The number of OTUs positively associated with SCFA concentration were similar between individuals without colon tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without colon tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations.
-The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma.
-To accomplish this we used Spearman's rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero.
-We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5].
-There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3].
-In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3].
-Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without colon tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3].
-Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without colon tumors and patients with a carcinoma [Figure 3].
-Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6].
-Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.
-
-**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** Despite the lack of difference in positive correlations between OTUs and SCFA concentrations between individuals with and without tumors, OTUs associated with SCFA concentrations could still be the most important variables to Random Forest models built to classify patients with adenomas or carcinomas.
-We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples.
-With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained when SCFA concentrations were added to the model.
-Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations.
-Both our adenoma and carcinoma models classified patients with a similar degree of success, as measured by the area under the curve (AUC) (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D].
-After the addition of SCFA concentrations to the adenoma or carcinoma models, many OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables to the model [Figure Figure 4B-C & 4E-F].
-After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations.
-When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration was in the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations.
-In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.
-
-**The most important OTUs for SCFA concentration and tumor Random Forest models are not the same.** It is possible that due to the way the Random Forest algorithm works, OTUs associated with SCFA concentrations could be downweighted in importance within the adenoma or carcinoma models when SCFA concentrations are included.
-To test if this is the case we used OTU data and built Random Forest models to classify SCFA concentrations.
-Overall, the correlation between the predicted and actual SCFA concentrations were moderately associated with each other [Figure 5A].
-Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa were important for these classifications [Figure 5A].
-There also was a difference in accuracy based on whether the fecal sample was from an individual without colon tumors or from patients with adenomas or carcinomas [Figure 5B].
-When comparing the adenoma or carcinoma model to the SCFA concentration models there was minimal overlap between these model's most important OTUs [Figure 4B-C, 4E-F and 5C-E].
-The only OTU that was in the top 10 most important variables and had overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E].
-Similar observations were made when using high/low SCFA groups based on the median SCFA concentration [Figure S2].
-Collectively, these observations provide evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and accordingly these OTUs belong to taxa known to produce acetate, butyrate, and propionate.
-Although it is possible to identify OTUs associated with SCFA production, overall, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no colon tumors and patients with adenomas or carcinomas.
-
-\newpage
-
-## Discussion
-
-The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors.
-Whether we directly measured the SCFA concentration or investigated genes that encoded enzymes used for their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2].
-There were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have colon tumors, had an adenoma, or had a carcinoma.
-However, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1].
-In models with SCFA concentrations included, many OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables for the model [Figure 4].
-Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to these taxa in the tumor models [Figure 4-5].
-Collectively, our observations suggest that the resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* that are different between individuals without colon tumors and patients with adenomas or carcinomas, are not the same as those involved with SCFA production.
-
-Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018].
-Many of the *in vivo* studies use proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017].
-Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems.
-Additionally, the observations in humans on the benefit of SCFAs in preventing tumorigenesis have been mixed.
-In previous case/control studies lower SCFA concentrations were observed in patients with carcinomas versus those without carcinomas [@Ohigashi2013].
-Yet, this is in contrast to multiple randomized-controlled trials that have found no difference in tumor recurrence between patients who do and do not get fiber supplementation [@Schatzkin2000; @Yao2017].
-In contrast to the *in vivo* model findings, the observations made in these randomized-controlled trials would suggest that SCFAs do not prevent or slow tumorigenesis.
-One reason for these results is that SCFA concentrations and responses to fiber vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016].
-This information taken together with our observations would suggest that either SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer or that individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit.
-
-Another possible alternative explanation as to why no difference in SCFA concentration between individuals without colon tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs.
-One limitation of current research into the effect of SCFAs and the microbiota in CRC has been that all tumors are treated as the same type.
-However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in certain subsets of individuals.
-Similar to the idea of using immunotherapy as a targeted treatment option for specific tumors [@Thomas2016], SCFAs may have beneficial effects for distinct types of colorectal tumors.
-Future research will need to test if this is a valid hypothesis.
-Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation [@Schatzkin2000; @Yao2017] suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.
-
-One possible technical limitation is that a fecal sample may not be an ideal type of bio-specimen and that the effect SCFAs have on tumorigenesis is only detected in the colon.
-However, this is unlikely to be a major confounder.
-First, most *in vivo* studies, as well as human studies, have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018].
-Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016].
-Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be the cause of this.
-Clearly, our observations provide robust evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer.
-Yet, taxa that are associated with SCFA production are consistently higher in individuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-
-The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].
-Our data is consistent with the literature despite finding no difference in SCFA concentrations between individuals without colon tumors and patients with adenomas or carcinomas.
-In particular, we observed that the important OTUs in our tumor models were to taxa that are associated with SCFA production but that these OTUs were not themselves associated with acetate, butyrate, or propionate production.
-One of the clearest results to highlight this within our study were the results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models [Figure 4 & 5].
-This leads to the possibility that protection may be through other routes.
-For example, there could be a different pathway or other less extensively studied metabolites that provide the necessary protection against tumorigenesis.
-Alternatively, previous research has shown that non-resident oral microbes, such as *Fusobacterium*, *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* are associated with carcinomas [@Zeller2014; @Baxter2016; @Kostic2011] and can increase inflammation [@Flynn2016].
-When these non-resident oral microbes increase inflammation they change the existing environment to one in which is more hospitable for themselves [@Flynn2016].
-It is possible that these resident microbes provide protection not through a metabolite but instead through active exclusion of these non-resident oral microbes which in turn prevents them from establishing themselves within the community.
-Overall, even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that our results open up many new and exciting avenues for investigation into how metabolites and the microbiota can prevent tumorigenesis.
-
-\newpage
-
-## Conclusions
-
-Our observations found no difference in SCFA concentration or in genes encoding enzymes involved in their production between individuals without colon tumors and patients with either adenoma or carcinoma tumors.
-The SCFA concentration also did not improve classification of adenoma or carcinoma tumors.
-The most important OTUs to these models classifying patients with adenomas or carcinomas were mostly not associated with SCFA production.
-Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence.
-Additionally, these observations suggest that resident microbes that are not involved in SCFA production may be the important resident community members involved with preventing tumorigenesis.
-By focusing on alternative mechanisms that are associated with these non-SCFA producing resident microbes or SCFA effects on distinct colorectal cancer groups the identification of more promising therapeutic options for use in treating colorectal cancer may be found.
+* Unlikely that SCFAs are the primary mechanism that limits tumorigenesis
+* Limitations
+	- Observing the community once the tumor is developed
+	- Not looking at mucosal communities
+* Need to identify other possible mechanisms that drive and prevent tumorigenesis
+* Supports evidence that fiber consumption has a null effect on tumorigenesis
 
@@
 
-**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016].
+**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @colorectal cancer_model_baxter2016].
 In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`).
@@
 Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017].
-Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016].
+Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @colorectal cancer_model_baxter2016].
 The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.
@@
 
-**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017].
+**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017].
 In brief, sequences were quality filtered and contigs created from the paired end reads.

### 2019-03-13 code/rf_classification.R
Correct internal cross validation steps

@@ -231,7 +231,7 @@ get_data <- function(path) {
 		mg_tag <- feature_sources[which(feature_sources %in% metagenomics)]
 		mg_file_name <- paste0("data/metagenome/metag.", mg_tag, ".shared")
 
-		data <- read_shared(mg_file_name, min_abundance=30) %>%
+		data <- read_shared(mg_file_name, min_abundance=5*78) %>%
 			inner_join(data, ., by=c("sample"="Group"))
 
 	}

### 2019-03-13 code/rf_regression.R

@@ -42,7 +42,7 @@ testTransformed  <- dataTransformed[-inTraining,]
 # cv index to make sure the internal 5-folds are stratified for diagnosis classes and also resampled 100 times.
 # 100 repeat internally is necessary to get robust readings of hyperparameter setting performance
   folds <- 5
-	cvIndex <- createMultiFolds(factor(trainTransformed$classes), folds, times=100) #returnTrain = T default for multifolds
+	cvIndex <- createMultiFolds(factor(trainTransformed$regress), folds, times=100) #returnTrain = T default for multifolds
   cv <- trainControl(method="repeatedcv",
                      number= folds,
                      index = cvIndex,
@@ -200,7 +200,7 @@ get_data <- function(path){
 		mg_tag <- feature_sources[which(feature_sources %in% metagenomics)]
 		mg_file_name <- paste0("data/metagenome/metag.", mg_tag, ".shared")
 
-		data <- read_shared(mg_file_name, min_abundance=30) %>%
+		data <- read_shared(mg_file_name, min_abundance=5*78) %>%
 			inner_join(data, ., by=c("sample"="Group"))
 
 	}

### 2019-03-13 submission/manuscript.Rmd
Build Figure 1

@@
 **Figure 1.
-No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate (A), butyrate (B), and propionate (C) concentrations in fecal samples of individuals without colon tumors (Normal), adenomas, and carcinomas (C).
-The black lines indicate the median SCFA concentration.
-Acetate (D), butyrate (E), and propionate (F) concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red).
-The black dots and lines represent the median change in SCFA concentration.
+No differences in SCFA measurements were observed between individuals with normal colons or those with adenoma or carcinomas before treatment (A) or due to treatment (B).** Concentrations of SCFAs did not differ among individuals in the different diagnosis categories (normal=XXX, adenoma=XXX, carcinoma=XXX) except for isobutyrate, which was significantly higher among individuals with cancer than those with adenomas (P=0.XXXX), but no different from those with normal colons (all P>0.XXX) (A).
+The change in SCFA concentrations among individuals undergoing treatment who initially had adenoma (N=XX) and carcinoma (N=XX) was only significant for the isobutyrate concentration among individuals with carcinoma (P=0.XXX, all others P>0.XXX).
 
-**Figure 2.
-No change in the butyrate kinase gene abudnance was observed between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A).
-Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B).
+**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions.**
 
-**Figure 3.
-Patients with adenomas had the higest number of statistically significant correlations between OTU relative abundance and SCFA concentration.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
-Fewer significant positive correlations were observed overall.
-Additionally, the differences in the number of significant positive correlations between patients with adenomas versus individuals without tumors (normal) and patients with carcinomas was not as pronounced as the number of significant negative correlations.
+**Figure 3. 16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**
 
-**Figure 4.
-SCFA concentrations do not improve OTU-based Random Forest models.** Individuals without tumors versus patients with adenomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (A).
-The top 10 most important OTUs or SCFAs in the SCFA and OTU adenoma model (B).
-The top 10 most important OTUs in the OTU adenoma model (C).
-Individuals without tumors versus patients with carcinomas model with and without SCFAs area under the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D).
-The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma model (E).
-The top 10 most important OTUs in the OTU carcinoma model (F).
-For (A) and (D) the black line represents the median AUC.
-The dotted line highlights an AUC of 0.5.
-
-**Figure 5.
-OTU-based regression Random Forest models of SCFA concentrations.** The train and test correlation between actual and predicted values from 100 different 80/20 split OTU-based models with 10-fold CV using regression Random Forest (A).
-The model accuracy of predicted SCFA concentrations differed between individuals without tumors, patients with adenomas, and patients with carcinomas.
-Differences between the actual and predicted SCFA concentrations based on whether individuals did not have tumors (Normal), had an adenoma, or had a carcinoma (B).
-The top 10 OTUs based on mean decrease in accuracy (MDA) for each SCFA model, colored by their lowest taxonomic identification (C).
-
-\newpage
-
-**Figure S1.
-Patients with adenomas had the higest number of significant differences in OTU relative abundance between high/low SCFA groups.** Colors denote the family or lowest taxonomic ID that an OTU classified to.
-
-**Figure S2.
-OTU-based classification Random Forest models of high/low SCFA groups based on overall SCFA median concentration.** The train and test results of 100 different 80/20 split OTU-based models with 10-fold CV based on higher or lower than the median SCFA concentration using classification Random Forest (A).
-The model accuracy of predicted high/low SCFA groups differed between individuals without tumors, patients with adenomas, and patients with carcinomas (B).
-The top 10 OTUs based on mean decrease in accuracy for each SCFA model, colored by their lowest taxonomic identification (C).
+**Figure S1. Comparison of training and testing results.**

### 2019-03-13 code/figure_1.R

@@ -1,0 +1,82 @@
+library(tidyverse)
+library(RColorBrewer)
+library(cowplot)
+
+### Helper function
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+### Read in scfa data and metadata
+scfa_data <- read_tsv('data/scfa/scfa_composite.tsv') %>%
+	mutate(scfa=factor(scfa, levels=c("acetate", "propionate", "isobutyrate", "butyrate"))) %>%
+	filter(scfa != "pooled")
+
+cross_section_metadata <- read_csv("data/metadata/cross_section.csv",
+																col_type=cols(sample=col_character())) %>%
+														mutate(dx=factor(dx, levels=c("normal", "adenoma", "cancer")))
+follow_up_metadata <- read_csv("data/metadata/follow_up.csv",
+														col_type=cols(initial=col_character(), followUp=col_character())) %>%
+												mutate(dx=factor(dx, levels=c("normal", "adenoma", "cancer")))
+
+
+### Plot SCFAs for each SCFA and diagnosis
+cross_section <- inner_join(scfa_data, cross_section_metadata, by=c("study_id"="sample")) %>%
+	select(scfa, mmol_kg, dx) %>%
+	ggplot(aes(x=dx, y=mmol_kg, fill=dx)) +
+		geom_violin(show.legend=FALSE, draw_quantiles=0.5) +
+		facet_wrap(~scfa, scales="free_y", nrow=1, labeller=labeller(.default=capitalize)) +
+		labs(x=NULL, y="mmol of SCFA/kg of feces") +
+		scale_x_discrete(labels = NULL) +
+		scale_fill_manual(
+			breaks=c("normal", "adenoma", "cancer"),
+			labels=c("Normal", "Adenoma", "Cancer"),
+			values=brewer.pal(n=3, "YlOrRd"),
+			name=NULL
+		) +
+		theme_classic() +
+		theme(
+			strip.background = element_blank(),
+			axis.ticks.x = element_blank(),
+			strip.text = element_text(size=10)
+		)
+
+
+### Plot SCFAs for each SCFA and diagnosis
+longitudinal <- inner_join(scfa_data, follow_up_metadata, by=c("study_id"="initial")) %>%
+	rename(initial = mmol_kg) %>%
[…]

### 2019-03-13 submission/manuscript.tex
Clean up manuscript and get it to render

@@
 \setlength{\droptitle}{-2em}
+
   \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
@@
 
-Running title: SCFAs and colorectal tumors
+Fecal short chain fatty acids are not predictive of colorectal cancer
+status and cannot be predicted based on bacterial community structure
 
@@
 
-Marc A. Sze\({^1}\), Nicholas A. Lesniak\({^1}\), Mack T. Ruffin
-IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)
+Marc A. Sze\({^1}\), Begüm Topçuoğlu\({^1}\), Nicholas A.
+Lesniak\({^1}\), Mack T. Ruffin IV\({^2}\), Patrick D.
+Schloss\({^1}\)\({^\dagger}\)
 
@@
 \newpage
-
 \linenumbers
 
-\subsection{Abstract}\label{abstract}
-
-\textbf{Background.} Colorectal cancer (CRC) is increasing in prevalence
-in individuals under 50 and because of this will be a continuing health
-concern for the foreseeable future. The majority of the risk for
-developing CRC is attributable to environmental factors. One of these
-environmental factors is the microbiota, with certain bacterial
-community members being more prevalent in CRC while other taxa are
-higher in individuals without tumors. Some of the bacterial species in
-taxa that are more abundant in individuals without tumors can use fiber
-to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
-model systems. However, the data supporting the importance of SCFAs in
-human CRC is less certain. Here, we test the hypothesis that SCFA
-concentrations and taxa associated with their production are different
-in individuals with colorectal tumors.
-
-\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
-pre- and post-treatment (n=67) group for their fecal concentrations of
-acetate, butyrate, and propionate. Analysis also included imputed gene
-relative abundance with PICRUSt, metagenomic sequencing on a subset
-(n=85) of the total cross-sectional group, and tumor classification and
-SCFA prediction models using Random Forest.
-
-\textbf{Results.} No difference in SCFA concentrations were found
-between individuals without tumors and patients with adenomas or
-carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
-there was also no difference in genes involved with SCFA synthesis
-between individuals without tumors and patients with adenomas or
-carcinomas (P-value \textgreater{} 0.70). Finally, there was no
-difference between the ability of Random Forest models to predict
-patients with adenomas or carcinomas versus individuals without tumors
-(P-value \textgreater{} 0.05).
-
-\textbf{Conclusions.} Although our data does not support the hypothesis
-that fecal SCFA concentrations are different in the general CRC
-population, there still may be specific types of colorectal tumors where
-SCFAs may be beneficial for treatment of CRC. Alternatively, our
-observations also support the hypothesis that there may be other
-metabolites or mechanisms (e.g.~bacterial niche exclusion) that may be
-more protective against tumorigenesis and have not been thoroughly
-investigated in the context of human CRC.
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
@@
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction-400-words}{%
+\subsection{Introduction (400 words)}\label{introduction-400-words}}
 
-Colorectal cancer (CRC) is currently the third leading cancer-related
-cause of death within the US and the prevelance is increasing in
-invidiauls under 50 years of age (1, 2). Although there is a genetic
-component to the disease, the environment is considered a larger risk
-factor for CRC (3). These environmental risk factors include but are not
-limited to smoking cigarettes, diet, and the microbiota (4--6). Many of
-these environmental risk factors, including the microbiota, are
-modifiable. This has lead to the investigation of how the microbiota may
-exacerbate or cause tumorigensis (7--9) and whether the bacterial
-community is altered in CRC (10, 11). Many of these previous
-case/control studies have identified resident bacterial taxa to be
-decreased in patients with carcinoma tumors (11--13). Many of the
-bacterial species from these resident taxa identified in these
-case/control studies actively produce short-chain fatty acids (SCFAs)
-from fiber that are part of our general diet (14). The most extensively
-studied SCFAs are acetate, butyrate, and propionate (15). These SCFAs
-are hypothesized to be the main metabolites involved with protection
-against tumorigenesis and could help to reduce the risk of CRC.
+Colorectal cancer is the third leading cancer-related cause of death
+within the United States (1, 2). Less than 10\% of cases can be
+attributed to genetic risk factors (3). This leaves a significant role
+for environmental, behavioral, and other factors such as smoking and
+diet (4, 5). Colorectal cancer is thought to be initiated by a series of
+mutations that accumulate as mutated cells proliferate leading to
+adenomatous lesions, which are succeeded by carcinomas (3). Throughout
+this progression, there are ample opportunities for bacterial
+populations to create mutations, induce inflammation, and accelerate
+tumorigenesis (6). Numerous studies in murine models have supported this
+model (7--11). Additional cross sectional studies in humans have
+identified microbiome-based biomarkers of disease (12--20). These
+studies suggest that in some cases, it is the loss of bacterial
+populations that produce short-chain fatty acids (SCFAs) that results in
+increased inflammation and tumorigenesis.
 
-Prior research suggests that SCFAs have promise in acting as an
-anti-tumorigenic agent. Specific SCFAs have shown positive results
-within model systems (16). For example, butyrate has been shown to
-inhibit cancer cell growth in \emph{in vitro} systems (17).
-Additionally, fiber supplementation in mouse models of CRC caused an
-overall reduction in tumor burden while also increasing SCFA
-concentrations (18). These exciting results in model systems suggest
-that supplementation with food sources that bacteria use to create these
-SCFAs may confer beneficial effects against CRC. However, it is
-important to note that these model systems provide only preliminary
-evidence towards the ability of SCFAs to reduce and treat tumors and the
-studies reporting benefit in humans has been less convincing.
+SCFAs have have anti-inflammatory and anti-proliferative activities
+(21--24). Furthermore, manipulation of SCFAs in mouse models of
+colorectal cancer by direct supplementation or feeding of fiber caused
+an overall reduction in tumor burden (25, 26). These results suggest
+that supplementation with substrates that bacteria can ferment to
+produce SCFAs may confer beneficial effects against colorectal cancer.
+Regardless, there is a lack of evidence that increasing SCFA
+concentrations can protect against colorectal cancer in humans. Based on
+similar observations, many microbiome studies use the concentrations of
+SCFAs and the presence of 16S rRNA gene sequences from organisms and the
+genes involved in producing them as a biomarker of a healthy microbiota
+(27--30). Case-control studies that have investigated SCFA
+concentrations in colorectal cancer found that patients with carcinomas
+had lower concentrations of SCFAs versus patients with adenomas or
+individuals without colon tumors (31). Although this would argue that
+increasing SCFA concentrations could be protective against
+tumorigenesis, in randomized controlled trials fiber supplementation has
+been inconsistently associated with protection against tumor formation
+and recurrence (32--36). These findings temper enthusiasm for treatments
+that aim to use SCFAs as biomarkers or protection against tumorigenesis.
 
-Overall, there is a lack of evidence on the benefit of increasing SCFA
-concentrations to protect against CRC in human populations. The initial
-case/control studies that investigated SCFA concentrations in CRC found
-that patients with carcinomas had lower concentrations of acetate,
-butyrate, and propionate versus patients with adenomas or individuals
-without colon tumors (19). Although this would argue that increasing
-SCFA concentrations could be protective against tumorigenesis, fiber
-supplementation in randomized controlled trials have consistently failed
-to protect against tumor recurrence (20). These findings argue against
-the utility of treatments that aim to use SCFAs to reduce or protect
-against tumorigenesis. Given the lack of clear evidence in human studies
-of the benefit of SCFAs in CRC, there is a need for more investigation
-into this area.
-
-Our study fills some of the current gaps in the literature that relate
-to the study of SCFAs and CRC in human populations. Specifically, by
-using a seperate cohort it tests previous case/control findings on SCFA
-concentrations in individuals with and without tumors. Additionally,
-prior investigations grouped patients with adenoma and individuals
-without tumors into a single group to compare against patients with
-carcinomas. Despite doing this, the suggestion was made that a reduction
-in SCFA concentration would also be observed between patients with
-adenomas and indivdiuals without colon tumors. We also can test this
-suggestion because we have a larger number of patients with adenomas
-within our study and do not need to group patients with adenomas and
-individuals without colon tumors into a single group. Additionally, we
-build upon these observations by assessing the utility of using SCFAs
-and Operational Taxonomic Units (OTUs) as a risk stratification tool of
-colorectal tumors (adenoma or carcinoma). We also investigate whether
-OTUs that are most important to these models are closely associated with
-the classification of SCFA concentrations. Collectively, this study
-provides important information on the replicability of previous findings
-in humans by extensively studying how SCFAs are associated with
-colorectal tumors.
-
-To accomplish this task we directly measured the concentration of
-acetate, butyrate, and propionate within fecal samples for two different
-groups. The first group had a sample obtained at a single cross
-sectional point in time while the second group had samples obtained
-before (pre-) and after (post-)treatment for colorectal tumors. To
-provide further support for our SCFA findings we also used PICRUSt (21)
-and metagenomic sequencing to investigate differences in genes encoding
-enzymes involved with SCFA synthesis between individuals without colon
-tumors, patients with adenomas, and patients with carcinomas. Next, we
-investigated whether taxa associated with SCFA production were important
-to disease classification models. First, using the cross-sectional data,
-we analyzed the number of correlations between OTU relative abundance
-and SCFA concentrations across individuals without colon tumors and
-patients with adenomas or carcinomas. Second, we assessed the affect
-adding SCFA concentrations to OTU data had on classification of patients
-with adenomas or carcinomas using the Random Forest algorithm (22).
-Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA
-concentrations. Finally, we compared whether SCFA concentrations replace
-important taxa in disease models and whether the same OTUs are the most
-important variables used to classify disease and SCFA concentration.
-Collectively, this investigation provides additional information as to
-whether SCFAs are decreased in patients with colorectal tumors and
-provides context as to whether targeting taxa to increase SCFA
-concentrations is a viable option to protect against colon
-tumorigenesis.
+To better understand the connection between colorectal cancer, the
+microbiome, and SCFAs, we quantified the concentration of SCFAs in feces
+of previously characterized individuals with normal colons, adenomas,
+and carcinomas and subset of those individuals after they underwent
+treatment for their lesions (14, 16). Our goals included: (1) testing
+whether there was an association between SCFA concentration and tumor
+status, (2) determining whether SCFA concentrations could be used as
+biomarkers to improve the detection of colonic lesions, and (3)
+predicting SCFA concentrations based on the relative abundance of fecal
+bacteria and their genes.
 
@@
 
-\subsection{Results}\label{results}
+\hypertarget{results-700-words}{%
+\subsection{Results (700 words)}\label{results-700-words}}
 
-\textbf{Decreased SCFA concentrations are not associated with adenoma or
-carcinoma tumors.} We used high-performance liquid chromatography (HPLC)
-to measure acetate, butyrate, and propionate concentrations of frozen
-fecal samples from 490 individuals at a cross-sectional point in time.
-There was no difference between individuals without colon tumors (n=172)
-and patients with either an adenoma (n=198) or carcinoma (n=120) for any
-of the SCFAs measured after multiple comparison correction (P-value
-\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
-concentration of SCFAs in 67 patients with an adenoma (n=41) or
-carcinoma (n=26) in which we had pre- and post-treatment fecal samples.
-Although there was a general trend for increasing acetate, butyrate, and
-propionate concentrations following treatment for tumors, there was no
-significant difference between pre- and post-treatment for either
-patients with adenomas (P-value \textgreater{} 0.20) or carcinomas
-(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.
+\begin{itemize}
+\tightlist
+\item
+  There is no association between presence of tumors and SCFA
+  concentration
 
-\textbf{Gene abundance for enzymes involved in SCFA synthesis are the
-same for individuals without colon tumors and patients with adenomas or
-carcinomas.} In order to provide further confirmation and support of our
-SCFA concentration results we investigated the genes encoding specific
-enzymes involved with SCFA synthesis {[}Table S1{]}. Using this list of
-specific genes {[}Table S1{]}, we looked for differences in gene
-abundance between individuals without colon tumors and patients with
-adenomas or carcinomas. Although we intended to analyze all the genes in
-the list {[}Table S1{]}, not all of the KEGG genes were identified
-during our analysis. We first analyzed imputed gene relative abundance,
-calculated from our OTU data that was generated from 16S rRNA gene
-sequencing. We found no difference in any of the imputed gene relative
-abundances for enzymes involved with acetate, butyrate, or propionate
-synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. Since butyrate
-is one of the most studied SCFAs in the context of CRC, we visualized
-the observed lack of difference between indviduals without colon tumors
-and patients with adenomas or carcinomas in gene abundance for enzymes
-involved with SCFA synthesis using the butyrate kinase gene {[}Figure
-2A{]}. Additionally, using a paired Wilcoxon rank-sum test, there also
-was no difference in imputed gene relative abundance between pre- and
-post-treatment samples for any genes involved with SCFA synthesis
-(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
-these 490 fecal samples (n=85) and used metagenomic sequencing to
-confirm these results. Like the imputed gene results, metagenomic
-analysis found that there was no difference in any of the genes involved
-in SCFA synthesis between individuals without colon tumors (n=29) and
-patients with adenomas (n=28) or carcinomas (n=28) (P-value
-\textgreater{} 0.70) {[}Table S4{]}. This similarity between indviduals
-without colon tumors and patients with adenomas or carcinomas is
-highlighted again by visualizing the gene abundance for the butyrate
-kinase gene {[}Figure 2B{]}. These observations provide evidence that
-gene prevalence for enzymes involved in SCFA synthesis does not change
-due to colorectal tumors and provides further support for our original
-SCFA concentration observations.
+  \begin{itemize}
+  \tightlist
+  \item
+    Cross sectional
+  \item
+    Aggregating SCFAs does not improve association with CRC
+  \item
+    Lack of change pre and post treatment
+  \item
+    Aggregating SCFAs does not improve association with CRC
+  \end{itemize}
+\item
+  Adding SCFA data to diagnostic models does not improve performance
 
-\textbf{The number of OTUs positively associated with SCFA concentration
-were similar between individuals without colon tumors and patients with
-adenomas or carcinomas.} Having found no difference between individuals
-without colon tumors and patients with adenomas or carcinomas in SCFA
-concentrations or genes encoding enzymes involved with SCFA synthesis,
-we next investigated if specific OTUs correlated with SCFA
-concentrations. The main goal of this analysis was to identify if there
-were OTUs that were significantly associated with SCFA concentrations
-and if this was different between individuals that did not have colon
-tumors, had an adenoma, or had a carcinoma. To accomplish this we used
-Spearman's rho, a non-parametric measure of association, and tested if
-there was a correlation that was signficantly greater than zero. We
-found that taxa from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} dominated statistically significant OTU
-correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
-number of significant negative correlations associated with patients
-with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
-from the \emph{Ruminococcaceae} family had the largest share of these
-negative correlations within patients with adenomas {[}Figure 3{]}.
-Patients with adenomas also had more positive correlations between OTUs
-and SCFA concentrations, but their total number was more similar to
-individuals without colon tumors or patients with carcinomas versus the
-analogous comparison for the number of negative correlations {[}Figure
-3{]}. Additionally, the number of positive correlations between OTUs and
-SCFA concentrations was similar between individuals without colon tumors
-and patients with a carcinoma {[}Figure 3{]}. Finally, when we used
-high/low SCFA groups based on the overall median concentration for each
-SCFA instead of SCFA concentrations a similar pattern was still observed
-{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
-resident taxa that may change the most due to colon tumors may not be
-ones that are responsible for the production of acetate, butyrate, or
-propionate.
+  \begin{itemize}
+  \tightlist
+  \item
+    Combining 16S and SCFA does not improve prediction of CRC
+  \item
+    OTU, genus, OPF, kegg
+  \end{itemize}
+\item
+  Community structure is not predictive of SCFA production
 
-\textbf{SCFA concentrations do not replace important
-\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
-OTUs in Random Forest models built to classify tumors.} Despite the lack
-of difference in positive correlations between OTUs and SCFA
-concentrations between individuals with and without tumors, OTUs
-associated with SCFA concentrations could still be the most important
-variables to Random Forest models built to classify patients with
-adenomas or carcinomas. We tested this by using the Random Forest
-algorithm to build models with OTU abundance data or OTU abundances and
-SCFA concentrations to classify normal versus adenoma and normal versus
-carcinoma fecal samples. With these models we compared whether OTUs with
-taxonomic classification to \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained when SCFA
-concentrations were added to the model. Additionally, we also compared
-whether any of the important OTUs that remained also had significant
-correlations with SCFA concentrations. Both our adenoma and carcinoma
-models classified patients with a similar degree of success, as measured
-by the area under the curve (AUC) (P-value \textgreater{} 0.05)
-{[}Figure 4A \& 4D{]}. After the addition of SCFA concentrations to the
-adenoma or carcinoma models, many OTUs with taxonomic classification to
-\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
-remained as important variables to the model {[}Figure Figure 4B-C \&
-4E-F{]}. After adding SCFA concentrations to the adenoma model, there
-were only 2 OTUs significantly associated with SCFA concentration that
-were part of the top 10 most important variables and both were postively
-associated with acetate, butyrate, or propionate concentrations. When
-SCFA concentrations were added to the carcinoma model, only 1 OTU that
-was associated with SCFA concentration was in the top 10 most important
-variables and it was negatively associated with acetate and butyrate
-concentrations. In combination with the previous results on OTU
-correlations, these observations provide additional evidence that the
-resident taxa that are associated with protection against tumorigeneis
-are not ones associated with acetate, butyrate, or propionate
-production.
-
-\textbf{The most important OTUs for SCFA concentration and tumor Random
-Forest models are not the same.} It is possible that due to the way the
-Random Forest algorithm works, OTUs associated with SCFA concentrations
-could be downweighted in importance within the adenoma or carcinoma
-models when SCFA concentrations are included. To test if this is the
-case we used OTU data and built Random Forest models to classify SCFA
-concentrations. Overall, the correlation between the predicted and
-actual SCFA concentrations were moderately associated with each other
-{[}Figure 5A{]}. Additionally, because the training set
-R\textsuperscript{2} was always higher than the test set
-R\textsuperscript{2}, all SCFA concentration models tended to be over
-fit, suggesting that rarer taxa were important for these classifications
-{[}Figure 5A{]}. There also was a difference in accuracy based on
-whether the fecal sample was from an individual without colon tumors or
-from patients with adenomas or carcinomas {[}Figure 5B{]}. When
-comparing the adenoma or carcinoma model to the SCFA concentration
-models there was minimal overlap between these model's most important
-OTUs {[}Figure 4B-C, 4E-F and 5C-E{]}. The only OTU that was in the top
-10 most important variables and had overlap between the models was
-OTU00167 (\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E{]}. Similar
-observations were made when using high/low SCFA groups based on the
-median SCFA concentration {[}Figure S2{]}. Collectively, these
-observations provide evidence that it is possible to identify specific
-OTUs associated with higher SCFA concentrations and accordingly these
-OTUs belong to taxa known to produce acetate, butyrate, and propionate.
-Although it is possible to identify OTUs associated with SCFA
-production, overall, our results do not support the hypothesis that SCFA
-concentration or OTUs associated with their production are different
-between individuals with no colon tumors and patients with adenomas or
-carcinomas.
+  \begin{itemize}
+  \tightlist
+  \item
+    Random forest regression
+  \item
+    16S
+  \item
+    picrust
+  \item
+    shotgun metagenomic data
+  \end{itemize}
+\end{itemize}
 
@@
 
-\subsection{Discussion}\label{discussion}
+\hypertarget{discussion-200-words}{%
+\subsection{Discussion (200 words)}\label{discussion-200-words}}
 
-The observations from this study do not support the hypothesis that SCFA
-concentrations are different in individuals with tumors. Whether we
-directly measured the SCFA concentration or investigated genes that
-encoded enzymes used for their production, no difference could be
-identified between individuals without tumors and patients with adenomas
-or carcinomas {[}Figure 1 \& 2{]}. There were differences in the number
-of significant correlations between SCFA concentration and OTU relative
-abundance based on whether individuals did not have colon tumors, had an
-adenoma, or had a carcinoma. However, SCFA concentrations did not
-provide increased model accuracy for tumor classification {[}Figure 3-4
-\& S1{]}. In models with SCFA concentrations included, many OTUs that
-classified to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} remained as important variables for the model
-{[}Figure 4{]}. Additionally, when models using OTU relative abundance
-to classify SCFA concentrations were assessed, the OTUs that classified
-to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} were not the same as the OTUs that classified to
-these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
-observations suggest that the resident taxa from \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} that are different
-between individuals without colon tumors and patients with adenomas or
-carcinomas, are not the same as those involved with SCFA production.
+\begin{itemize}
+\tightlist
+\item
+  Unlikely that SCFAs are the primary mechanism that limits
+  tumorigenesis
+\item
+  Limitations
 
-Although SCFAs have been shown to be anti-tumorigenic, most of these
-studies have been performed in model systems (16, 17). Many of the
-\emph{in vivo} studies use proxies such as fiber supplementation rather
-than SCFAs directly (14). Although it is well known that breakdown
-products from gut bacteria results in SCFA production, fiber effects on
-tumorigenesis may be through other mechanisms in these \emph{in vivo}
-model systems. Additionally, the observations in humans on the benefit
-of SCFAs in preventing tumorigenesis have been mixed. In previous
-case/control studies lower SCFA concentrations were observed in patients
-with carcinomas versus those without carcinomas (19). Yet, this is in
-contrast to multiple randomized-controlled trials that have found no
-difference in tumor recurrence between patients who do and do not get
-fiber supplementation (20, 23). In contrast to the \emph{in vivo} model
-findings, the observations made in these randomized-controlled trials
-would suggest that SCFAs do not prevent or slow tumorigenesis. One
-reason for these results is that SCFA concentrations and responses to
-fiber vary quite a bit between healthy individuals (24). This
-information taken together with our observations would suggest that
-either SCFAs provide little to no benefit as an anti-tumorigenic
-compound in colorectal cancer or that individuals who do not respond to
-fiber supplementation would need to acquire these bacteria to achieve a
-benefit.
-
-Another possible alternative explanation as to why no difference in SCFA
-concentration between individuals without colon tumors and patients with
-adenomas or carcinomas was observed, could be because only certain types
-of colorectal cancers are affected by SCFAs. One limitation of current
-research into the effect of SCFAs and the microbiota in CRC has been
-that all tumors are treated as the same type. However, there are known
-differences in the types of mutations that occur (25) and treating all
-tumors as equal may actually hide any benefit that could be found in
-certain subsets of individuals. Similar to the idea of using
-immunotherapy as a targeted treatment option for specific tumors (26),
-SCFAs may have beneficial effects for distinct types of colorectal
-tumors. Future research will need to test if this is a valid hypothesis.
-Regardless of this limitation, our results in combination to previous
-randomized controlled trials on fiber supplementation (20, 23) suggests
-that using SCFAs as a general treatment for colorectal cancer is
-unlikely to provide a reduction in tumorigenesis.
-
-One possible technical limitation is that a fecal sample may not be an
-ideal type of bio-specimen and that the effect SCFAs have on
-tumorigenesis is only detected in the colon. However, this is unlikely
-to be a major confounder. First, most \emph{in vivo} studies, as well as
-human studies, have used fecal material in their analysis (18, 19).
-Second, previous studies that measure SCFA changes after fiber
-supplementation use fecal material to track these responses with a great
-deal of success (24). Although there are limitations with the current
-research on SCFAs and colorectal tumors, technical limitations are less
-likely to be the cause of this. Clearly, our observations provide robust
-evidence that these specific metabolites may not be protective or used
-as a general treatment option in colorectal cancer. Yet, taxa that are
-associated with SCFA production are consistently higher in individuals
-without colon tumors than patients with carcinomas (10, 11, 27).
-
-The potential protection against colorectal cancers may not be from
-SCFAs, even though taxa associated with their production are higher in
-individuals without tumors versus patients with carcinomas (10, 11, 27).
-Our data is consistent with the literature despite finding no difference
-in SCFA concentrations between individuals without colon tumors and
-patients with adenomas or carcinomas. In particular, we observed that
-the important OTUs in our tumor models were to taxa that are associated
-with SCFA production but that these OTUs were not themselves associated
-with acetate, butyrate, or propionate production. One of the clearest
-results to highlight this within our study were the results showing that
-different OTUs from the same taxonomic classification are in tumor and
-SCFA Random Forest models {[}Figure 4 \& 5{]}. This leads to the
-possibility that protection may be through other routes. For example,
-there could be a different pathway or other less extensively studied
-metabolites that provide the necessary protection against tumorigenesis.
-Alternatively, previous research has shown that non-resident oral
-microbes, such as \emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}, and \emph{Peptostreptococcus} are associated with
-carcinomas (6, 12, 13) and can increase inflammation (28). When these
-non-resident oral microbes increase inflammation they change the
-existing environment to one in which is more hospitable for themselves
-(28). It is possible that these resident microbes provide protection not
-through a metabolite but instead through active exclusion of these
-non-resident oral microbes which in turn prevents them from establishing
-themselves within the community. Overall, even though we did not find
-lower concentrations of SCFAs associated with colorectal tumors, we
-think that our results open up many new and exciting avenues for
-investigation into how metabolites and the microbiota can prevent
-tumorigenesis.
+  \begin{itemize}
+  \tightlist
+  \item
+    Observing the community once the tumor is developed
+  \item
+    Not looking at mucosal communities
+  \end{itemize}
+\item
+  Need to identify other possible mechanisms that drive and prevent
+  tumorigenesis
+\item
+  Supports evidence that fiber consumption has a null effect on
+  tumorigenesis
+\end{itemize}
 
@@
 
-\subsection{Conclusions}\label{conclusions}
-
-Our observations found no difference in SCFA concentration or in genes
-encoding enzymes involved in their production between individuals
-without colon tumors and patients with either adenoma or carcinoma
-tumors. The SCFA concentration also did not improve classification of
-adenoma or carcinoma tumors. The most important OTUs to these models
-classifying patients with adenomas or carcinomas were mostly not
-associated with SCFA production. Although these results are different
-than other reports in the literature, they do align with the randomized
-controlled trials that have tested fiber use in preventing colorectal
-tumor recurrence. Additionally, these observations suggest that resident
-microbes that are not involved in SCFA production may be the important
-resident community members involved with preventing tumorigenesis. By
-focusing on alternative mechanisms that are associated with these
-non-SCFA producing resident microbes or SCFA effects on distinct
-colorectal cancer groups the identification of more promising
-therapeutic options for use in treating colorectal cancer may be found.
-
-\newpage
-
-\subsection{Materials and Methods}\label{materials-and-methods}
+\hypertarget{materials-and-methods}{%
+\subsection{Materials and Methods}\label{materials-and-methods}}
 
 \textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (29, 30). In brief, this study used fecal
-samples obtained at either a single cross-sectional time point (n=490)
-or from before (pre-) and after (post-) treatment of a patient's tumor
-(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
-for their tumor the length of time between their initial and follow up
-sample ranged from 188 - 546 days. Our use of treatment has been
-previously defined as encompassing removal of a tumor (surgery or
-colonoscopy) with or without chemotherapy and radiation (29). Diagnosis
+described in detail previously (14, 37). In brief, this study used fecal
+samples obtained at either a single cross-sectional time point
+(n=\textbf{r
+length(metai\(sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**).
+For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf
+\)time)} - \textbf{r max(metaf\$time)} days. Our use of treatment has
+been previously defined as encompassing removal of a tumor (surgery or
+colonoscopy) with or without chemotherapy and radiation (37). Diagnosis
 of tumor was made by colonoscopic examination and histopathological
-review of biopsies obtained (29, 30). The University of Michigan
+review of biopsies obtained (14, 37). The University of Michigan
 Institutional Review Board approved the study and informed consent was
@@
 protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (24). The following changes to this protocol included the use of
+machine (38). The following changes to this protocol included the use of
 frozen fecal samples suspended in 1ml of PBS instead of fecal
@@
 affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (24) and Figure 1 here).
+studies (see Table 1 (38) and Figure 1 here).
 
-\textbf{16s rRNA gene sequencing.} The workflow and processing have been
-previously described (29, 31, 32). In brief, sequences were quality
+\textbf{16S rRNA gene sequencing.} The workflow and processing have been
+previously described (37, 39, 40). In brief, sequences were quality
 filtered and contigs created from the paired end reads. Any sequences
@@
 matches to the V4 region of the 16S rRNA gene using the SILVA database
-(33). Chimeras were identified and removed using UCHIME and OTUs
-clustered at 97\% similarity (34). The major differences from these
+(41). Chimeras were identified and removed using UCHIME and OTUs
+clustered at 97\% similarity (42). The major differences from these
 previous reports include: the use of version 1.39.5 of the mothur
 software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptiClust algorithm (35).
+97\% similarity using the OptiClust algorithm (43).
 
 \textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (21) was used.
+1.1.2 with the recommended standard operating protocol (44) was used.
 Briefly, the mothur shared file and metadata was converted into a biom
@@
 \textbf{Obtaining Operational Protein Families from metagenomes.} A
-subset of the cross-sectional group (n=490) containing a total of 85
-individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
-sequenced on an Illumina HiSeq using 125 bp paired end reads and a
-previously described method (36). Briefly, the sequences were quality
-filtered and sequences aligning to the human genome were removed prior
-to contig assembly with MEGAHIT (37). Open Reading Frames (ORFs) were
-identified using Prodigal (38), counts generated using Diamond (39),
-subsequent clustering into Operational Protein Families (OPFs) used
-mmseq2 (40), and OPF alignment used the KEGG database (41).
+subset of the cross-sectional group (n=\textbf{r
+length(metai\(sample)**) containing a total of **r length(geof_samples\)X1)}
+individuals (normal n=\textbf{r filter(geof\_samples, X30 ==
+``Healthy'') \%\textgreater{}\% pull(X30) \%\textgreater{}\% length()},
+adenoma n=\textbf{r filter(geof\_samples, X30 == ``Adenoma'')
+\%\textgreater{}\% pull(X30) \%\textgreater{}\% length()}, and carcinoma
+n=\textbf{r filter(geof\_samples, X30 == ``Cancer'') \%\textgreater{}\%
+pull(X30) \%\textgreater{}\% length()}) was shotgun sequenced on an
+Illumina HiSeq using 125 bp paired end reads and a previously described
+method (16). Briefly, the sequences were quality filtered and sequences
+aligning to the human genome were removed prior to contig assembly with
+MEGAHIT (45). Open Reading Frames (ORFs) were identified using Prodigal
+(46), counts generated using Diamond (47), subsequent clustering into
+Operational Protein Families (OPFs) used mmseq2 (48), and OPF alignment
+used the KEGG database (49).
 
@@
 between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (41,
-42) and a list can be found in the supplemental material {[}Table S1{]}.
+genes were based on pathways from KEGG as well as previous research (8,
+49) and a list can be found in the supplemental material {[}Table S1{]}.
 
@@
 Random Forest algorithm for classification and regression models via the
-caret package (22, 43). This was repeated on 100 different 80/20 splits
+caret package (50, 51). This was repeated on 100 different 80/20 splits
 of the data to generate a reasonable range for the AUC of the model. The
@@
 \textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (44). Generally, a Kruskal-Wallis rank sum
+the statistical language R (52). Generally, a Kruskal-Wallis rank sum
 test with a Dunn's post-hoc test was used to assess differences between
@@
 with carcinomas. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons (45). First, we assessed differences in
-SCFA concentrations measured by HPLC between individuals with normal
-colons and patients with tumors (adenoma or carcinoma). We then analyzed
-whether SCFA concentrations changed in patients with an adenoma or
-carcinoma pre- versus post-treatment. Next, the imputed gene counts of
-genes encoding enzymes involved in SCFA synthesis was tested.
-Additionally, metagenomic sequencing counts for important genes involved
-with SCFA production were analyzed. From here we analyzed the number of
-significant positive and negative correlations between OTU relative
-abundance and SCFA concentrations in individuals without colon tumors
-and patients with adenomas or carcinomas using Spearman's rho. Next, we
-assessed whether OTUs alone or OTUs and SCFAs were better able to
-classify individuals with and without tumors using Random Forest models.
-Finally, models to classify the actual SCFA concentration or high/low
-SCFA concentration based on the median of each SCFA using 16S rRNA gene
-sequencing data was created using the Random Forest algorithm. For all
-Random Forest models, the assessment of the most important variables was
-based on the top 10 features (OTUs or SCFAs) using the mean decrease in
-accuracy.
+correct for multiple comparisons {[}\textbf{XXXXXX}{]}. First, we
+assessed differences in SCFA concentrations measured by HPLC between
+individuals with normal colons and patients with tumors (adenoma or
+carcinoma). We then analyzed whether SCFA concentrations changed in
+patients with an adenoma or carcinoma pre- versus post-treatment. Next,
+the imputed gene counts of genes encoding enzymes involved in SCFA
+synthesis was tested. Additionally, metagenomic sequencing counts for
+important genes involved with SCFA production were analyzed. From here
+we analyzed the number of significant positive and negative correlations
+between OTU relative abundance and SCFA concentrations in individuals
+without colon tumors and patients with adenomas or carcinomas using
+Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs
+were better able to classify individuals with and without tumors using
+Random Forest models. Finally, models to classify the actual SCFA
+concentration or high/low SCFA concentration based on the median of each
+SCFA using 16S rRNA gene sequencing data was created using the Random
+Forest algorithm. For all Random Forest models, the assessment of the
+most important variables was based on the top 10 features (OTUs or
+SCFAs) using the mean decrease in accuracy.
 
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-Haggar2009}{}
+\leavevmode\hypertarget{ref-Haggar2009}{}%
 1. \textbf{Haggar F}, \textbf{Boushey R}. 2009. Colorectal cancer
@@
 
-\hypertarget{ref-Siegel2016}{}
+\leavevmode\hypertarget{ref-Siegel2016}{}%
 2. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
@@
 
-\hypertarget{ref-Lichtenstein2000}{}
-3. \textbf{Lichtenstein P}, \textbf{Holm NV}, \textbf{Verkasalo PK},
-\textbf{Iliadou A}, \textbf{Kaprio J}, \textbf{Koskenvuo M},
-\textbf{Pukkala E}, \textbf{Skytthe A}, \textbf{Hemminki K}. 2000.
-Environmental and heritable factors in the causation of cancer analyses
-of cohorts of twins from sweden, denmark, and finland. New England
-Journal of Medicine \textbf{343}:78--85.
-doi:\href{https://doi.org/10.1056/nejm200007133430201}{10.1056/nejm200007133430201}.
+\leavevmode\hypertarget{ref-Fearon1990}{}%
+3. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+colorectal tumorigenesis. Cell \textbf{61}:759--767.
+doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.
 
-\hypertarget{ref-FlissIsakov2017}{}
+\leavevmode\hypertarget{ref-FlissIsakov2017}{}%
 4. \textbf{Fliss-Isakov N}, \textbf{Zelber-Sagi S}, \textbf{Webb M},
@@
 
-\hypertarget{ref-Lee2015}{}
+\leavevmode\hypertarget{ref-Lee2015}{}%
 5. \textbf{Lee J}, \textbf{Jeon JY}, \textbf{Meyerhardt JA}. 2015. Diet
@@
 
-\hypertarget{ref-Kostic2011}{}
-6. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
+\leavevmode\hypertarget{ref-Flynn2016}{}%
+6. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+Metabolic and community synergy of oral bacteria in colorectal cancer.
+mSphere \textbf{1}.
+doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
+
+\leavevmode\hypertarget{ref-Zackular2013}{}%
+7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
+tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
+doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.
+
+\leavevmode\hypertarget{ref-Baxter2014}{}%
+8. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
+colonization with human feces determines colonic tumor burden.
+Microbiome \textbf{2}:20.
+doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
+
+\leavevmode\hypertarget{ref-Zackular2015}{}%
+9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
+role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
+doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.
+
+\leavevmode\hypertarget{ref-DeStefanoShields2016}{}%
+10. \textbf{Shields CED}, \textbf{Meerbeke SWV}, \textbf{Housseau F},
+\textbf{Wang H}, \textbf{Huso DL}, \textbf{Casero RA}, \textbf{O'Hagan
+HM}, \textbf{Sears CL}. 2016. Reduction of murine colon tumorigenesis
+driven by EnterotoxigenicBacteroides fragilisUsing cefoxitin treatment.
+Journal of Infectious Diseases \textbf{214}:122--129.
+doi:\href{https://doi.org/10.1093/infdis/jiw069}{10.1093/infdis/jiw069}.
+
+\leavevmode\hypertarget{ref-Tomkovich2017}{}%
+11. \textbf{Tomkovich S}, \textbf{Yang Y}, \textbf{Winglee K},
+\textbf{Gauthier J}, \textbf{Mühlbauer M}, \textbf{Sun X},
+\textbf{Mohamadzadeh M}, \textbf{Liu X}, \textbf{Martin P}, \textbf{Wang
+GP}, \textbf{Oswald E}, \textbf{Fodor AA}, \textbf{Jobin C}. 2017.
+Locoregional effects of microbiota in a preclinical model of colon
+carcinogenesis. Cancer Research \textbf{77}:2620--2632.
+doi:\href{https://doi.org/10.1158/0008-5472.can-16-3472}{10.1158/0008-5472.can-16-3472}.
+
+\leavevmode\hypertarget{ref-Kostic2011}{}%
+12. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
 \textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
@@
 
-\hypertarget{ref-Zackular2013}{}
-7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
-\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
-tumorigenesis. mBio \textbf{4}:e00692--13--e00692--13.
-doi:\href{https://doi.org/10.1128/mbio.00692-13}{10.1128/mbio.00692-13}.
-
-\hypertarget{ref-Baxter2014}{}
-8. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
-colonization with human feces determines colonic tumor burden.
-Microbiome \textbf{2}:20.
-doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
-
-\hypertarget{ref-Zackular2015}{}
-9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
-role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
-doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.
-
-\hypertarget{ref-Shah2017}{}
-10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
-\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
-\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
-sequence-based faecal microbial community survey data to identify a
-composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.
-
-\hypertarget{ref-meta_analysis_crc_Sze2018}{}
-11. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+\leavevmode\hypertarget{ref-Sze2018}{}%
+13. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 
-\hypertarget{ref-Zeller2014}{}
-12. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\leavevmode\hypertarget{ref-Baxter2016}{}%
+14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\leavevmode\hypertarget{ref-Zackular2014}{}%
+15. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prevention Research \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.capr-14-0129}{10.1158/1940-6207.capr-14-0129}.
+
+\leavevmode\hypertarget{ref-Hannigan2017}{}%
+16. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
+\& the interactive dynamics of the colorectal cancer virome.
+doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
+
+\leavevmode\hypertarget{ref-Zeller2014}{}%
+17. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 
-\hypertarget{ref-Baxter2016}{}
-13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+\leavevmode\hypertarget{ref-Shah2017}{}%
+18. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
+\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
+\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
+sequence-based faecal microbial community survey data to identify a
+composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.
 
-\hypertarget{ref-Holscher2017}{}
-14. \textbf{Holscher HD}. 2017. Dietary fiber and prebiotics and the
-gastrointestinal microbiota. Gut Microbes \textbf{8}:172--184.
-doi:\href{https://doi.org/10.1080/19490976.2017.1290756}{10.1080/19490976.2017.1290756}.
+\leavevmode\hypertarget{ref-Dejea2018}{}%
+19. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
+\textbf{Boleij A}, \textbf{Taddese R}, \textbf{Geis AL}, \textbf{Wu X},
+\textbf{Shields CED}, \textbf{Hechenbleikner EM}, \textbf{Huso DL},
+\textbf{Anders RA}, \textbf{Giardiello FM}, \textbf{Wick EC},
+\textbf{Wang H}, \textbf{Wu S}, \textbf{Pardoll DM}, \textbf{Housseau
+F}, \textbf{Sears CL}. 2018. Patients with familial adenomatous
+polyposis harbor colonic biofilms containing tumorigenic bacteria.
+Science \textbf{359}:592--597.
+doi:\href{https://doi.org/10.1126/science.aah3648}{10.1126/science.aah3648}.
 
-\hypertarget{ref-Louis2016}{}
-15. \textbf{Louis P}, \textbf{Flint HJ}. 2016. Formation of propionate
-and butyrate by the human colonic microbiota. Environmental Microbiology
-\textbf{19}:29--41.
-doi:\href{https://doi.org/10.1111/1462-2920.13589}{10.1111/1462-2920.13589}.
+\leavevmode\hypertarget{ref-Arthur2012}{}%
+20. \textbf{Arthur JC}, \textbf{Perez-Chanona E}, \textbf{Muhlbauer M},
+\textbf{Tomkovich S}, \textbf{Uronis JM}, \textbf{Fan T-J},
+\textbf{Campbell BJ}, \textbf{Abujamel T}, \textbf{Dogan B},
+\textbf{Rogers AB}, \textbf{Rhodes JM}, \textbf{Stintzi A},
+\textbf{Simpson KW}, \textbf{Hansen JJ}, \textbf{Keku TO}, \textbf{Fodor
+AA}, \textbf{Jobin C}. 2012. Intestinal inflammation targets
+cancer-inducing activity of the microbiota. Science
+\textbf{338}:120--123.
+doi:\href{https://doi.org/10.1126/science.1224820}{10.1126/science.1224820}.
 
-\hypertarget{ref-test_OKeefe2016}{}
-16. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
-metabolites and colon cancer. Nature Reviews Gastroenterology \&
-Hepatology \textbf{13}:691--706.
-doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
-
-\hypertarget{ref-Encarnao2018}{}
-17. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
+\leavevmode\hypertarget{ref-Encarnao2018}{}%
+21. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
 \textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
@@
 
-\hypertarget{ref-Bishehsari2018}{}
-18. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+\leavevmode\hypertarget{ref-Verma2018}{}%
+22. \textbf{Verma MS}, \textbf{Fink MJ}, \textbf{Salmon GL},
+\textbf{Fornelos N}, \textbf{Ohara TE}, \textbf{Ryu SH},
+\textbf{Vlamakis H}, \textbf{Xavier RJ}, \textbf{Stappenbeck TS},
+\textbf{Whitesides GM}. 2018. A common mechanism links activities of
+butyrate in the colon. ACS Chemical Biology \textbf{13}:1291--1298.
+doi:\href{https://doi.org/10.1021/acschembio.8b00073}{10.1021/acschembio.8b00073}.
+
+\leavevmode\hypertarget{ref-Zheng2017}{}%
+23. \textbf{Zheng L}, \textbf{Kelly CJ}, \textbf{Battista KD},
+\textbf{Schaefer R}, \textbf{Lanis JM}, \textbf{Alexeev EE},
+\textbf{Wang RX}, \textbf{Onyiah JC}, \textbf{Kominsky DJ},
+\textbf{Colgan SP}. 2017. Microbial-derived butyrate promotes epithelial
+barrier function through IL-10 receptorDependent repression of
+claudin-2. The Journal of Immunology \textbf{199}:2976--2984.
+doi:\href{https://doi.org/10.4049/jimmunol.1700105}{10.4049/jimmunol.1700105}.
+
+\leavevmode\hypertarget{ref-OKeefe2016}{}%
+24. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+metabolites and colon cancer. Nature Reviews Gastroenterology \&
+Hepatology \textbf{13}:691--706.
+doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
+
+\leavevmode\hypertarget{ref-Bishehsari2018}{}%
+25. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
 \textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
@@
 
-\hypertarget{ref-Ohigashi2013}{}
-19. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
+\leavevmode\hypertarget{ref-Tian2018}{}%
+26. \textbf{Tian Y}, \textbf{Xu Q}, \textbf{Sun L}, \textbf{Ye Y},
+\textbf{Ji G}. 2018. Short-chain fatty acids administration is
+protective in colitis-associated colorectal cancer development. The
+Journal of Nutritional Biochemistry \textbf{57}:103--109.
+doi:\href{https://doi.org/10.1016/j.jnutbio.2018.03.007}{10.1016/j.jnutbio.2018.03.007}.
+
+\leavevmode\hypertarget{ref-Vital2014}{}%
+27. \textbf{Vital M}, \textbf{Howe AC}, \textbf{Tiedje JM}. 2014.
+Revealing the bacterial butyrate synthesis pathways by analyzing
+(meta)genomic data. mBio \textbf{5}:e00889--14--e00889--14.
+doi:\href{https://doi.org/10.1128/mbio.00889-14}{10.1128/mbio.00889-14}.
+
+\leavevmode\hypertarget{ref-Sanna2019}{}%
+28. \textbf{Sanna S}, \textbf{Zuydam NR van}, \textbf{Mahajan A},
+\textbf{Kurilshikov A}, \textbf{Vila AV}, \textbf{Võsa U},
+\textbf{Mujagic Z}, \textbf{Masclee AAM}, \textbf{Jonkers DMAE},
+\textbf{Oosting M}, \textbf{Joosten LAB}, \textbf{Netea MG},
+\textbf{Franke L}, \textbf{Zhernakova A}, \textbf{Fu J},
+\textbf{Wijmenga C}, \textbf{McCarthy MI}. 2019. Causal relationships
+among the gut microbiome, short-chain fatty acids and metabolic
+diseases. Nature Genetics.
+doi:\href{https://doi.org/10.1038/s41588-019-0350-x}{10.1038/s41588-019-0350-x}.
+
+\leavevmode\hypertarget{ref-Liu2019}{}%
+29. \textbf{Liu S}, \textbf{Li E}, \textbf{Sun Z}, \textbf{Fu D},
+\textbf{Duan G}, \textbf{Jiang M}, \textbf{Yu Y}, \textbf{Mei L},
+\textbf{Yang P}, \textbf{Tang Y}, \textbf{Zheng P}. 2019. Altered gut
+microbiota and short chain fatty acids in chinese children with autism
+spectrum disorder. Scientific Reports \textbf{9}.
+doi:\href{https://doi.org/10.1038/s41598-018-36430-z}{10.1038/s41598-018-36430-z}.
+
+\leavevmode\hypertarget{ref-Meisel2016}{}%
+30. \textbf{Meisel M}, \textbf{Mayassi T}, \textbf{Fehlner-Peach H},
+\textbf{Koval JC}, \textbf{O'Brien SL}, \textbf{Hinterleitner R},
+\textbf{Lesko K}, \textbf{Kim S}, \textbf{Bouziat R}, \textbf{Chen L},
+\textbf{Weber CR}, \textbf{Mazmanian SK}, \textbf{Jabri B},
+\textbf{Antonopoulos DA}. 2016. Interleukin-15 promotes intestinal
+dysbiosis with butyrate deficiency associated with increased
+susceptibility to colitis. The ISME Journal \textbf{11}:15--30.
+doi:\href{https://doi.org/10.1038/ismej.2016.114}{10.1038/ismej.2016.114}.
+
+\leavevmode\hypertarget{ref-Ohigashi2013}{}%
+31. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
 \textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
@@
 
-\hypertarget{ref-Yao2017}{}
-20. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
-Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
-for the prevention of recurrent colorectal adenomas and carcinomas.
-Cochrane Database of Systematic Reviews.
-doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.
-
-\hypertarget{ref-Langille2013}{}
-21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
-\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
-\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
-\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
-Predictive functional profiling of microbial communities using 16S rRNA
-marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
-doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
-
-\hypertarget{ref-randomforest_citation_2002}{}
-22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News \textbf{2}:18--22.
-
-\hypertarget{ref-Schatzkin2000}{}
-23. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
+\leavevmode\hypertarget{ref-Schatzkin2000}{}%
+32. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
 \textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
@@
 
-\hypertarget{ref-scfa_measures_venkataraman2016}{}
-24. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+\leavevmode\hypertarget{ref-Yao2017}{}%
+33. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
+Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
+for the prevention of recurrent colorectal adenomas and carcinomas.
+Cochrane Database of Systematic Reviews.
+doi:\href{https://doi.org/10.1002/14651858.cd003430.pub2}{10.1002/14651858.cd003430.pub2}.
+
+\leavevmode\hypertarget{ref-Kunzmann2015}{}%
+34. \textbf{Kunzmann AT}, \textbf{Coleman HG}, \textbf{Huang W-Y},
+\textbf{Kitahara CM}, \textbf{Cantwell MM}, \textbf{Berndt SI}. 2015.
+Dietary fiber intake and risk of colorectal cancer and incident and
+recurrent adenoma in the prostate, lung, colorectal, and ovarian cancer
+screening trial. The American Journal of Clinical Nutrition
+\textbf{102}:881--890.
+doi:\href{https://doi.org/10.3945/ajcn.115.113282}{10.3945/ajcn.115.113282}.
+
+\leavevmode\hypertarget{ref-Murphy2012}{}%
+35. \textbf{Murphy N}, \textbf{Norat T}, \textbf{Ferrari P},
+\textbf{Jenab M}, \textbf{Bueno-de-Mesquita B}, \textbf{Skeie G},
+\textbf{Dahm CC}, \textbf{Overvad K}, \textbf{Olsen A},
+\textbf{Tjønneland A}, \textbf{Clavel-Chapelon F},
+\textbf{Boutron-Ruault MC}, \textbf{Racine A}, \textbf{Kaaks R},
+\textbf{Teucher B}, \textbf{Boeing H}, \textbf{Bergmann MM},
+\textbf{Trichopoulou A}, \textbf{Trichopoulos D}, \textbf{Lagiou P},
+\textbf{Palli D}, \textbf{Pala V}, \textbf{Panico S}, \textbf{Tumino R},
+\textbf{Vineis P}, \textbf{Siersema P}, \textbf{Duijnhoven F van},
+\textbf{Peeters PHM}, \textbf{Hjartaker A}, \textbf{Engeset D},
+\textbf{González CA}, \textbf{Sánchez M-J}, \textbf{Dorronsoro M},
+\textbf{Navarro C}, \textbf{Ardanaz E}, \textbf{Quirós JR},
+\textbf{Sonestedt E}, \textbf{Ericson U}, \textbf{Nilsson L},
+\textbf{Palmqvist R}, \textbf{Khaw K-T}, \textbf{Wareham N}, \textbf{Key
+TJ}, \textbf{Crowe FL}, \textbf{Fedirko V}, \textbf{Wark PA},
+\textbf{Chuang S-C}, \textbf{Riboli E}. 2012. Dietary fibre intake and
+risks of cancers of the colon and rectum in the european prospective
+investigation into cancer and nutrition (EPIC). PLoS ONE
+\textbf{7}:e39361.
+doi:\href{https://doi.org/10.1371/journal.pone.0039361}{10.1371/journal.pone.0039361}.
+
+\leavevmode\hypertarget{ref-Gianfredi2018}{}%
+36. \textbf{Gianfredi V}, \textbf{Salvatori T}, \textbf{Villarini M},
+\textbf{Moretti M}, \textbf{Nucci D}, \textbf{Realdon S}. 2018. Is
+dietary fibre truly protective against colon cancer? A systematic review
+and meta-analysis. International Journal of Food Sciences and Nutrition
+\textbf{69}:904--915.
+doi:\href{https://doi.org/10.1080/09637486.2018.1446917}{10.1080/09637486.2018.1446917}.
+
+\leavevmode\hypertarget{ref-Sze2017}{}%
+37. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
+
+\leavevmode\hypertarget{ref-Venkataraman2016}{}%
+38. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 
-\hypertarget{ref-Fearon1990}{}
-25. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
-colorectal tumorigenesis. Cell \textbf{61}:759--767.
-doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.
-
-\hypertarget{ref-Thomas2016}{}
-26. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
-agents targeting the BCR-ABL tyrosine kinase as philadelphia
-chromosome-positive acute lymphoblastic leukemia treatment. Expert
-Opinion on Drug Discovery \textbf{11}:1061--1070.
-doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
-
-\hypertarget{ref-normalization_Sze2017}{}
-27. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
-\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
-microbiota in patients after treatment for colonic lesions. Microbiome
-\textbf{5}.
-doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
-
-\hypertarget{ref-Flynn2016}{}
-28. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
-Metabolic and community synergy of oral bacteria in colorectal cancer.
-mSphere \textbf{1}.
-doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
-
-\hypertarget{ref-normalization_sze2017}{}
-29. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
-\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
-microbiota in patients after treatment for colonic lesions. Microbiome
-\textbf{5}.
-doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
-
-\hypertarget{ref-crc_model_baxter2016}{}
-30. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-Schloss2009}{}
-31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\hypertarget{ref-Kozich2013}{}
-32. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+40. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 
-\hypertarget{ref-Quast2012}{}
-33. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+\leavevmode\hypertarget{ref-Quast2012}{}%
+41. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
 \textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
@@
 
-\hypertarget{ref-Edgar2011}{}
-34. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\leavevmode\hypertarget{ref-Edgar2011}{}%
+42. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
 \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
@@
 
-\hypertarget{ref-opticlust_westcott2017}{}
-35. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 
-\hypertarget{ref-Hannigan2017}{}
-36. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
-\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
-\& the interactive dynamics of the colorectal cancer virome.
-doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
+\leavevmode\hypertarget{ref-Langille2013}{}%
+44. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
+\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
+\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
+Predictive functional profiling of microbial communities using 16S rRNA
+marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
+doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.
 
-\hypertarget{ref-Li2015}{}
-37. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+\leavevmode\hypertarget{ref-Li2015}{}%
+45. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 
-\hypertarget{ref-Hyatt2010}{}
-38. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+\leavevmode\hypertarget{ref-Hyatt2010}{}%
+46. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 
-\hypertarget{ref-Buchfink2014}{}
-39. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+\leavevmode\hypertarget{ref-Buchfink2014}{}%
+47. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 
-\hypertarget{ref-Steinegger2017}{}
-40. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+\leavevmode\hypertarget{ref-Steinegger2017}{}%
+48. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 
-\hypertarget{ref-Kanehisa2015}{}
-41. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+\leavevmode\hypertarget{ref-Kanehisa2015}{}%
+49. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 
-\hypertarget{ref-scfa_baxter2014}{}
-42. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
-colonization with human feces determines colonic tumor burden.
-Microbiome \textbf{2}:20.
-doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
+\leavevmode\hypertarget{ref-Liaw2002}{}%
+50. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
 
-\hypertarget{ref-caret_citation}{}
-43. \textbf{Jed Wing MKC from}, \textbf{Weston S}, \textbf{Williams A},
-\textbf{Keefer C}, \textbf{Engelhardt A}, \textbf{Cooper T},
-\textbf{Mayer Z}, \textbf{Kenkel B}, \textbf{R Core Team},
-\textbf{Benesty M}, \textbf{Lescarbeau R}, \textbf{Ziem A},
-\textbf{Scrucca L}, \textbf{Tang Y}, \textbf{Candan C}, \textbf{Hunt.
-T}. 2017. Caret: Classification and regression training.
+\leavevmode\hypertarget{ref-Kuhn2017}{}%
+51. \textbf{Kuhn M}. 2017. Caret: Classification and regression
+training.
 
-\hypertarget{ref-r_citation_2017}{}
-44. \textbf{R Core Team}. 2017. R: A language and environment for
+\leavevmode\hypertarget{ref-r_citation_2017}{}%
+52. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 
-\hypertarget{ref-benjamini_controlling_1995}{}
-45. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
-false discovery rate: A practical and powerful approach to multiple
-testing. Journal of the Royal Statistical Society Series B
-(Methodological) \textbf{57}:289--300.
-
 \newpage
 
-\textbf{Figure 1. No change in SCFA measurements was observed between
-normal, adenoma, and carcinoma individuals using HPLC.} Acetate (A),
-butyrate (B), and propionate (C) concentrations in fecal samples of
-individuals without colon tumors (Normal), adenomas, and carcinomas (C).
-The black lines indicate the median SCFA concentration. Acetate (D),
-butyrate (E), and propionate (F) concentrations in fecal samples before
-and after treatment for adenoma (yellow) and carcinoma (red). The black
-dots and lines represent the median change in SCFA concentration.
+\textbf{Figure 1. No differences in SCFA measurements were observed
+between individuals with normal colons or those with adenoma or
+carcinomas before treatment (A) or due to treatment (B).} Concentrations
+of SCFAs did not differ among individuals in the different diagnosis
+categories (normal=\textbf{XXX}, adenoma=\textbf{XXX},
+carcinoma=\textbf{XXX}) except for isobutyrate, which was significantly
+higher among individuals with cancer than those with adenomas
+(P=\textbf{0.XXXX}), but no different from those with normal colons (all
+P\textgreater{}\textbf{0.XXX}) (A). The change in SCFA concentrations
+among individuals undergoing treatment who initially had adenoma
+(N=\textbf{XX}) and carcinoma (N=\textbf{XX}) was only significant for
+the isobutyrate concentration among individuals with carcinoma
+(P=\textbf{0.XXX}, all others P\textgreater{}\textbf{0.XXX}).
 
-\textbf{Figure 2. No change in the butyrate kinase gene abudnance was
-observed between normal, adenoma, and carcinoma individuals.} Imputed
-gene relative abundance of important butyrate pathway genes using
-PICRUSt (A). Counts per million (corrected for size and number of
-contigs in an OPF) for the Butyrate Kinase gene (B).
+\textbf{Figure 2. SCFA concentrations do not improve models for
+diagnosing the presence of adenomas, carcinomas, or all lesions.}
 
-\textbf{Figure 3. Patients with adenomas had the higest number of
-statistically significant correlations between OTU relative abundance
-and SCFA concentration.} Colors denote the family or lowest taxonomic ID
-that an OTU classified to. Fewer significant positive correlations were
-observed overall. Additionally, the differences in the number of
-significant positive correlations between patients with adenomas versus
-individuals without tumors (normal) and patients with carcinomas was not
-as pronounced as the number of significant negative correlations.
+\textbf{Figure 3. 16S rRNA gene and metagenomic sequence data do not
+predict SCFAs concentrations.}
 
-\textbf{Figure 4. SCFA concentrations do not improve OTU-based Random
-Forest models.} Individuals without tumors versus patients with adenomas
-model with and without SCFAs area under the curve (AUC) from 100 80/20
-splits with 10-fold cross-validation (A). The top 10 most important OTUs
-or SCFAs in the SCFA and OTU adenoma model (B). The top 10 most
-important OTUs in the OTU adenoma model (C). Individuals without tumors
-versus patients with carcinomas model with and without SCFAs area under
-the curve (AUC) from 100 80/20 splits with 10-fold cross-validation (D).
-The top 10 most important OTUs or SCFAs in the SCFA and OTU carcinoma
-model (E). The top 10 most important OTUs in the OTU carcinoma model
-(F). For (A) and (D) the black line represents the median AUC. The
-dotted line highlights an AUC of 0.5.
-
-\textbf{Figure 5. OTU-based regression Random Forest models of SCFA
-concentrations.} The train and test correlation between actual and
-predicted values from 100 different 80/20 split OTU-based models with
-10-fold CV using regression Random Forest (A). The model accuracy of
-predicted SCFA concentrations differed between individuals without
-tumors, patients with adenomas, and patients with carcinomas.
-Differences between the actual and predicted SCFA concentrations based
-on whether individuals did not have tumors (Normal), had an adenoma, or
-had a carcinoma (B). The top 10 OTUs based on mean decrease in accuracy
-(MDA) for each SCFA model, colored by their lowest taxonomic
-identification (C).
-
-\newpage
-
-\textbf{Figure S1. Patients with adenomas had the higest number of
-significant differences in OTU relative abundance between high/low SCFA
-groups.} Colors denote the family or lowest taxonomic ID that an OTU
-classified to.
-
-\textbf{Figure S2. OTU-based classification Random Forest models of
-high/low SCFA groups based on overall SCFA median concentration.} The
-train and test results of 100 different 80/20 split OTU-based models
-with 10-fold CV based on higher or lower than the median SCFA
-concentration using classification Random Forest (A). The model accuracy
-of predicted high/low SCFA groups differed between individuals without
-tumors, patients with adenomas, and patients with carcinomas (B). The
-top 10 OTUs based on mean decrease in accuracy for each SCFA model,
-colored by their lowest taxonomic identification (C).
+\textbf{Figure S1. Comparison of training and testing results.}
 

### 2019-03-13 submission/manuscript.Rmd
First pass at Results

@@
 ---
+
+```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
+options(tidyverse.quiet = TRUE)
+
+library(tidyverse)
+library(knitr)
+
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("message" = FALSE)
+opts_chunk$set("cache" = FALSE)
+
+inline_hook <- function(x){
+  print(x)
+  if(is.list(x)){
+    x <- unlist(x)
+  }
+  if(is.numeric(x)){
+    if(abs(x - round(x)) < .Machine$double.eps^0.5){
+      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+    } else {
+      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
+    }
+  } else {
+      paste(x)
+  }
+}
+knitr::knit_hooks$set(inline=inline_hook)
+```
 
@@
 
-## Abstract
+## Abstract (250 words)
+
+Something clever
+
+## Importance (150 words)
+
+Something clever
 
 \newpage
-
-## Introduction (400 words)
 
@@
 Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017].
-Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012].
+Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Hannigan2017; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012].
 These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.
@@
 
-To better understand the connection between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of SCFAs in feces of previously characterized individuals with normal colons, adenomas, and carcinomas and subset of those individuals after they underwent treatment for their lesions.
-Our goals included: (1) testing whether there was an association between SCFA concentration and tumor status, (2) determining whether SCFA concentrations could be used as biomarkers to improve the detection of colonic lesions, and (3) predicting SCFA concentrations based on the relative abundance of fecal bacteria and their genes.
+```{r, concentration}
+n <- read_csv('../data/metadata/cross_section.csv') %>% count(dx)
+n_normal <- n %>% filter(dx == "normal") %>% pull(n)
+n_adenoma <- n %>% filter(dx == "adenoma") %>% pull(n)
+n_cancer <- n %>% filter(dx == "cancer") %>% pull(n)
 
-\newpage
+cross_test <- read_tsv('../data/process/scfa_cross_section_stats.tsv')
+iso_p_cross <- cross_test %>% filter(p.value < 0.05)
+stopifnot(nrow(iso_p_cross) == 1)
 
-## Results (700 words)
+n_follow <- read_csv('../data/metadata/follow_up.csv') %>% count(dx)
+n_follow_adenoma <- n_follow %>% filter(dx == "adenoma") %>% pull(n)
+n_follow_cancer <- n_follow %>% filter(dx == "cancer") %>% pull(n)
 
-* There is no association between presence of tumors and SCFA concentration
-	- Wilcoxon/T-test
-	- Cross sectional
-	- Aggregating SCFAs does not improve association with CRC
-	- Combining 16S and SCFA does not improve prediction of CRC
-	- Lack of change pre and post treatment
-* Community structure is not predictive of SCFA production
-	- Random forest regression
-	- 16S
-	- picrust
-	- shotgun metagenomic data
+follow_test <- read_tsv('../data/process/scfa_pre_post_stats.tsv')
+iso_p_follow <- follow_test %>% filter(p.value < 0.05)
+stopifnot(nrow(iso_p_follow) == 1)
 
-\newpage
+pooled_test <- min(c(cross_test %>% filter(scfa=="pooled") %>% pull(p.value), follow_test %>% filter(scfa=="pooled") %>% pull(p.value)))
 
-## Discussion (200 words)
+```
+
+**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, isobutyrate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
+The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A).
+The median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas.
+The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`).
+The median concentration of isobutyrate in people with normal colons (`r iso_p_cross$normal` mmol/kg) was between those with adenomas (`r iso_p_cross$adenoma` mmol/kg) and carcinomas (`r iso_p_cross$cancer` mmol/kg).
+Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
+The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg (P=`r iso_p_follow$p.value`; Figure 1B).
+For both the pre-treatment cross-sectional data and the pre/post treatment data, we pooled the SCFA concentrations on a per molecule of carbon basis and again failed to see any significant differences (P>`r pooled_test`).
+The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to this observation.
+
+```{r, classification}
+```
+
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models.
+We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model.
+A model trained on the individual or pool SCFAs alone had a median area under the receiver operator characteristic curve (AUROC) of **0.XX** (IQR=**0.XX**).
+When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances (AUROC~OTU+SCFA~=**0.XX**, AUROC~genus+SCFA~=**0.XX**), the AUROC values were no different from the models trained without the SCFA concentrations (AUROC~OTU~=**0.XX**, AUROC~genus~=**0.XX**; Figure 2A).
+We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=**XX**, N~adenoma~=**XX**, and N~cancer~=**XX**).
+We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories [**XXXXXX**].
+Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (Figure 2A).
+These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.
+
+```{r, regression}
+```
+
+**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations.
+We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs.
+Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B).
+Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs.
+Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
+Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using picrust2.
+Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
+The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.
+
+**Conclusion.**
 
@@
 
-**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @colorectal cancer_model_baxter2016].
-In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`).
-For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days.
-Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017].
-Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @colorectal cancer_model_baxter2016].
+**Study design and sampling.** The overall protocol has been described in detail previously [@Sze2017; @Baxter2016].
+In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=**r length(metai$sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**).
+For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf$time)** - **r max(metaf$time)** days.
+Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@Sze2017].
+Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@Sze2017; @Baxter2016].
 The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.
 
-**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016].
+**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@Venkataraman2016].
 The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations.
-These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).
+These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).
 
-**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017].
+**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @Sze2017].
 In brief, sequences were quality filtered and contigs created from the paired end reads.
@@
 Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
-The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].
+The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@Westcott2017].
 
@@
 
-**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()`, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
+**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=**r length(metai$sample)**) containing a total of **r length(geof_samples$X1)** individuals (normal n=**r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()**, adenoma n=**r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()**, and carcinoma n=**r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()**) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
 Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015].
@@
 
-**Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors.
-These genes were based on pathways from KEGG as well as previous research [@scfa_baxter2014; @Kanehisa2015] and a list can be found in the supplemental material [Table S1].
-
-**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@randomforest_citation_2002; @caret_citation].
+**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@Liaw2002; @Kuhn2017].
 This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model.
@@
 Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas.
-Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995].
+Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [**XXXXXX**].
 First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma).
@@
 The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan.
-We would also like to thank Ada Hagan for providing valuable feedback on earlier drafts of the manuscript.
-Salary support for Marc A.
-Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240.
-Salary support for Patrick D.
-Schloss came from NIH grants P30DK034933 and 1R01CA215574.
+We would thank the University of Michigan Center for Microbial Systems for enabling our short-chain fatty acid analysis.
+Support for MAS came from the Canadian Institute of Health Research and the National Institutes of Health (UL1TR002240).
+Support for PDS came from the National Institutes of Health (P30DK034933 and R01CA215574).
 
@@
 
+\includegraphics{../results/figures/scfa_abundance.pdf}
+
 **Figure 1.
-No differences in SCFA measurements were observed between individuals with normal colons or those with adenoma or carcinomas before treatment (A) or due to treatment (B).** Concentrations of SCFAs did not differ among individuals in the different diagnosis categories (normal=XXX, adenoma=XXX, carcinoma=XXX) except for isobutyrate, which was significantly higher among individuals with cancer than those with adenomas (P=0.XXXX), but no different from those with normal colons (all P>0.XXX) (A).
-The change in SCFA concentrations among individuals undergoing treatment who initially had adenoma (N=XX) and carcinoma (N=XX) was only significant for the isobutyrate concentration among individuals with carcinoma (P=0.XXX, all others P>0.XXX).
+SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) was for isobutyrate.
+(B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.
 
-**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions.**
+\newpage
 
-**Figure 3. 16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**
+**Figure 2.
+SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions.
+16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**
+
+\newpage
 

### 2019-03-13 submission/manuscript.tex
First pass at Results

@@
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
+\hypertarget{abstract-250-words}{%
+\subsection{Abstract (250 words)}\label{abstract-250-words}}
+
+Something clever
+
+\hypertarget{importance-150-words}{%
+\subsection{Importance (150 words)}\label{importance-150-words}}
+
+Something clever
 
 \newpage
-
-\hypertarget{introduction-400-words}{%
-\subsection{Introduction (400 words)}\label{introduction-400-words}}
 
@@
 
-To better understand the connection between colorectal cancer, the
-microbiome, and SCFAs, we quantified the concentration of SCFAs in feces
-of previously characterized individuals with normal colons, adenomas,
-and carcinomas and subset of those individuals after they underwent
-treatment for their lesions (14, 16). Our goals included: (1) testing
-whether there was an association between SCFA concentration and tumor
-status, (2) determining whether SCFA concentrations could be used as
-biomarkers to improve the detection of colonic lesions, and (3)
-predicting SCFA concentrations based on the relative abundance of fecal
-bacteria and their genes.
+\textbf{SCFA concentrations do not meaningfully vary with diagnosis or
+treatment.} To quantify the associations between colorectal cancer, the
+microbiome, and SCFAs, we quantified the concentration of acetate,
+propionate, isobutyrate, and butyrate in feces of previously
+characterized individuals with normal colons (N=172) and those with
+colonic adenomas (N=198) or carcinomas (N=120) (14). The only SCFA that
+had a significantly different concentration across the diagnoses was
+isobutyrate (P=0.0091; Figure 1A). The median concentration of
+isobutyrate was 3.30 mmol/kg in people with normal colons and it was
+3.00 and 3.84 mmol/kg in people with colonic adenomas or carcinomas. The
+difference in isobutyrate concentration between people with adenomas and
+carcinomas was significantly different (P=0.0065); however, the
+differences between people with normal colons and those with adenomas or
+carcinomas was not significant (P=0.19 and P=0.11). The median
+concentration of isobutyrate in people with normal colons (3.30 mmol/kg)
+was between those with adenomas (3.00 mmol/kg) and carcinomas (3.84
+mmol/kg). Among the subjects with adenomas and carcinomas, a subset
+(N\textsubscript{adenoma}=41, N\textsubscript{carcinoma}=26) were
+treated and sampled a year later (37). The only SCFA that changed
+following treatment was isobutyrate, which decreased by 0.99 mmol/kg
+(P=0.002; Figure 1B). For both the pre-treatment cross-sectional data
+and the pre/post treatment data, we pooled the SCFA concentrations on a
+per molecule of carbon basis and again failed to see any significant
+differences (P\textgreater{}0.15). The low concentration of isobutyrate
+relative to the other SCFAs, inconsistent concentrations, and unexpected
+decrease in concentration with treatment makes it difficult to ascribe
+much biological relevance to this observation.
 
-\newpage
+\textbf{Combining SCFA and microbiome data does not improve the ability
+to diagnose individual as having adenomas or carcinomas.} We previously
+found that binning 16S rRNA gene sequence data into operational
+taxonomic units based on 97\% similarity or into genera enabled us to
+classify individuals as having adenomas or carcinomas using Random
+Forest machine learning models. We repeated that analysis but added the
+concentration of the individual or pooled SCFAs as possible features to
+train a model. A model trained on the individual or pool SCFAs alone had
+a median area under the receiver operator characteristic curve (AUROC)
+of \textbf{0.XX} (IQR=\textbf{0.XX}). When we trained the models with
+the SCFAs concentrations and OTU or genus-level relative abundances
+(AUROC\textsubscript{OTU+SCFA}=\textbf{0.XX},
+AUROC\textsubscript{genus+SCFA}=\textbf{0.XX}), the AUROC values were no
+different from the models trained without the SCFA concentrations
+(AUROC\textsubscript{OTU}=\textbf{0.XX},
+AUROC\textsubscript{genus}=\textbf{0.XX}; Figure 2A). We also trained
+models using a smaller dataset that generated shotgun metagenomic
+sequencing data from a subset of our cohort
+(N\textsubscript{normal}=\textbf{XX},
+N\textsubscript{adenoma}=\textbf{XX}, and
+N\textsubscript{cancer}=\textbf{XX}). We binned genes extracted from the
+assembled metagenomes into operational protein families (OPFs) or KEGG
+categories {[}\textbf{XXXXXX}{]}. Again, the performance of the models
+trained with the meteagenomic data did not improve when the SCFA
+concentrations were added as possible features when training the model
+(Figure 2A). These data demonstrate that knowledge of the SCFA profile
+from a patient's fecal sample does not improve the ability to diagnose a
+colonic lesion.
 
-\hypertarget{results-700-words}{%
-\subsection{Results (700 words)}\label{results-700-words}}
+\textbf{Knowledge of microbial community structure does not predict SCFA
+concentrations.} Regardless of a person's diagnosis, we next asked
+whether the fecal community structure was predictive of fecal SCFA
+concentrations. We trained Random Forest regression models using 16S
+rRNA gene sequence data binned into OTUs and genera to predict the
+concentration of the individual and pooled SCFAs. Regardless of the
+binning method or SCFA, the largest amount of variation that the model
+could explain was \textbf{0.XXXX} (Figure 2B). Next, we trained Random
+Forest regression models using metagenomic sequence data binned into
+OPFs and KEGG categories to predict the concentration of the individual
+and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence
+data, the metagenomic data was not predictive of SCFA concentration; the
+largest amount of variation that the models could explain was
+\textbf{0.XXXX} (Figure 2B). Because of the limited number of samples
+that we were able to generate metagenomic sequence data from, we used
+our 16S rRNA gene sequence data to impute metagenomes that were binned
+into metabolic pathways, enzyme commission numbers, or KEGG categories
+using picrust2. Again, SCFA concentrations could not be predicted based
+on the imputed metagenomic data; the largest amount of variation that
+the models could explain was \textbf{0.XXXX} (Figure 2B). The inability
+to model SCFA concentrations from microbiome data indicates that the
+knowledge of the abundance of organisms and genes is insufficient to
+predict SCFA concentrations.
 
-\begin{itemize}
-\tightlist
-\item
-  There is no association between presence of tumors and SCFA
-  concentration
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Cross sectional
-  \item
-    Aggregating SCFAs does not improve association with CRC
-  \item
-    Lack of change pre and post treatment
-  \item
-    Aggregating SCFAs does not improve association with CRC
-  \end{itemize}
-\item
-  Adding SCFA data to diagnostic models does not improve performance
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Combining 16S and SCFA does not improve prediction of CRC
-  \item
-    OTU, genus, OPF, kegg
-  \end{itemize}
-\item
-  Community structure is not predictive of SCFA production
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Random forest regression
-  \item
-    16S
-  \item
-    picrust
-  \item
-    shotgun metagenomic data
-  \end{itemize}
-\end{itemize}
-
-\newpage
-
-\hypertarget{discussion-200-words}{%
-\subsection{Discussion (200 words)}\label{discussion-200-words}}
+\textbf{Conclusion.}
 
@@
 
-\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
-located near the end of the pathways involved in the synthesis of
-acetate, butyrate, and propionate were analyzed for any differences
-between individuals with normal colons and those with tumors. These
-genes were based on pathways from KEGG as well as previous research (8,
-49) and a list can be found in the supplemental material {[}Table S1{]}.
-
 \textbf{Random Forest models.} The model was first trained on 80\% of
@@
 Network for providing the fecal samples that were used in this study. We
-would also like to thank Kwi Kim and Thomas M Schmidt for their help in
-running the short-chain fatty acid analysis on the High-Performance
-Liquid Chromatography machine at the University of Michigan. We would
-also like to thank Ada Hagan for providing valuable feedback on earlier
-drafts of the manuscript. Salary support for Marc A. Sze came from the
-Canadian Institute of Health Research and NIH grant UL1TR002240. Salary
-support for Patrick D. Schloss came from NIH grants P30DK034933 and
-1R01CA215574.
+would thank the University of Michigan Center for Microbial Systems for
+enabling our short-chain fatty acid analysis. Support for MAS came from
+the Canadian Institute of Health Research and the National Institutes of
+Health (UL1TR002240). Support for PDS came from the National Institutes
+of Health (P30DK034933 and R01CA215574).
 
@@
 
-\textbf{Figure 1. No differences in SCFA measurements were observed
-between individuals with normal colons or those with adenoma or
-carcinomas before treatment (A) or due to treatment (B).} Concentrations
-of SCFAs did not differ among individuals in the different diagnosis
-categories (normal=\textbf{XXX}, adenoma=\textbf{XXX},
-carcinoma=\textbf{XXX}) except for isobutyrate, which was significantly
-higher among individuals with cancer than those with adenomas
-(P=\textbf{0.XXXX}), but no different from those with normal colons (all
-P\textgreater{}\textbf{0.XXX}) (A). The change in SCFA concentrations
-among individuals undergoing treatment who initially had adenoma
-(N=\textbf{XX}) and carcinoma (N=\textbf{XX}) was only significant for
-the isobutyrate concentration among individuals with carcinoma
-(P=\textbf{0.XXX}, all others P\textgreater{}\textbf{0.XXX}).
+\includegraphics{../results/figures/scfa_abundance.pdf}
+
+\textbf{Figure 1. SCFA concentrations did not vary meaningfully with
+diagnosis of colonic lesions or with treatment for adenomas or
+carcinomas.} (A) We measured the concentration of fecal SCFAs from
+individuals with normal colons (N=172) or those with adenoma (N=198) or
+carcinomas (N=120) was for isobutyrate. (B) A subset of individuals
+diagnosed with adenomas (N=41) or carcinomas (N=26) who underwent
+treatment were resampled a year after the initial sampling; one extreme
+propionate value (124.4 mmol/kg) was included in the adenoma analysis
+but censored from the visualization for clarity.
+
+\newpage
 
 \textbf{Figure 2. SCFA concentrations do not improve models for
-diagnosing the presence of adenomas, carcinomas, or all lesions.}
+diagnosing the presence of adenomas, carcinomas, or all lesions. 16S
+rRNA gene and metagenomic sequence data do not predict SCFAs
+concentrations.}
 
-\textbf{Figure 3. 16S rRNA gene and metagenomic sequence data do not
-predict SCFAs concentrations.}
+\newpage
 

### 2019-03-19 submission/manuscript.Rmd
First pass at Conclusion

@@
 
-**Conclusion.**
-
-* Unlikely that SCFAs are the primary mechanism that limits tumorigenesis
-* Limitations
-	- Observing the community once the tumor is developed
-	- Not looking at mucosal communities
-* Need to identify other possible mechanisms that drive and prevent tumorigenesis
-* Supports evidence that fiber consumption has a null effect on tumorigenesis
+**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions.
+Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of individual or pooled SCFA concentrations.
+These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors.
+It is important to note that our analysis concerned fecal SCFA concentrations and microbiome characterization and that observations along the mucosa near the site of lesions may provide a stronger association.
+Regardless, given the growing literature in this area, it is unlikely that SCFAs are the primary mechanism that limits tumorigenesis.
+This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally.
+Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis to better understand the disease process and identify biomarkers that will allow early detection of tumors.
 

### 2019-03-19 code/rf_pool_pools.R
Clean up code for processing RF results files

@@ -1,6 +1,9 @@
 library(tidyverse)
 
+options(warn = 1)
+
 cv_test_compare_files <- function(file) {
+	print(file)
 
 	read_tsv(file, col_types=cols(col_character(), col_double(), col_double(), col_double())) %>%
 		mutate(condition=str_replace(condition, "_", ".")) %>%

### 2019-03-19 code/rf_compare_classification_models.R
Analyze classification data

@@ -1,0 +1,68 @@
+library(tidyverse)
+library(broom)
+
+data_file_name <- "data/rf/classification_data_pool.tsv"
+model_data <- read_tsv(data_file_name)
+
+# compare cv and test models' AUCs
+model_data %>%
+	group_by(class, input) %>%
+	nest() %>%
+	mutate(test = map(data, ~tidy(wilcox.test(x=.x$cv_aucs, y=.x$test_aucs, exact=FALSE))),
+				cv_lci = map(data, ~quantile(.x$cv_aucs, 0.025)),
+				cv_q1 = map(data, ~quantile(.x$cv_aucs, 0.25)),
+				cv_median = map(data, ~median(.x$cv_aucs)),
+				cv_q3 = map(data, ~quantile(.x$cv_aucs, 0.75)),
+				cv_uci = map(data, ~quantile(.x$cv_aucs, 0.975)),
+				test_lci = map(data, ~quantile(.x$test_aucs, 0.025)),
+				test_q1 = map(data, ~quantile(.x$test_aucs, 0.25)),
+				test_median = map(data, ~median(.x$test_aucs)),
+				test_q3 = map(data, ~quantile(.x$test_aucs, 0.75)),
+				test_uci = map(data, ~quantile(.x$test_aucs, 0.975))
+			) %>%
+	select(class, input, test,
+				cv_lci, cv_q1, cv_median, cv_q3, cv_uci,
+				test_lci, test_q1, test_median, test_q3, test_uci
+			) %>%
+	unnest() %>%
+	write_tsv("data/rf/classification_cv_test_compare.tsv")
+
+
+# compare models constructed w/ and w/o SCFA data
+model_data %>%
+	mutate(scfa = str_detect(input, "scfa"), microbiome=str_replace(input, "scfa_", ""), microbiome=str_replace(microbiome, "_scfa", "")) %>%
+	filter(microbiome != "scfa") %>%
+	select(class, microbiome, scfa, cv_aucs, test_aucs) %>%
+	group_by(class, microbiome) %>%
+	nest() %>%
+	mutate(test = map(data, ~tidy(wilcox.test(.x$test_aucs~as.factor(.x$scfa), exact=FALSE, alternative="less"))),
+				wo_lci = map(data, ~quantile(.x$test_aucs[!.x$scfa], probs=0.025)),
[…]

### 2019-03-19 code/metagenomics_get_ns.R
Count the number of metagenomes per diagnosis group

@@ -1,0 +1,14 @@
+library(tidyverse)
+library(data.table)
+
+fread("data/metagenome/metag.kegg.shared")$Group %>%
+	tibble(group=.) %>%
+	mutate(dx = str_replace_all(group, "\\d", ""),
+				dx = case_when(
+					dx == "C" ~ "cancer",
+					dx == "A" ~ "adenoma",
+					dx == "H" ~ "normal"
+				)
+		) %>%
+	count(dx) %>%
+	write_tsv("data/metagenome/metag.sample.counts")

### 2019-03-19 submission/manuscript.Rmd
Add code to classification paragraph

@@
 ```{r, classification}
+random <- read_tsv("../data/rf/classification_SCFA_to_random.tsv")
+stopifnot(random$p.value < 0.001)
+
+w_wo_compare <- read_tsv("../data/rf/classification_w_wo_SCFA.tsv")
+
+min_p_compare_tax <- w_wo_compare %>%
+	filter(class=="adenoma" | class == "cancer") %>%
+	filter(microbiome =="otu" | microbiome == "genus") %>%
+  select(p.value) %>%
+	min()
+
+min_p_compare_metag <- w_wo_compare %>%
+	filter(class=="adenoma" | class == "cancer") %>%
+	filter(microbiome =="opf" | microbiome == "kegg") %>%
+	select(p.value) %>%
+	min()
+
+n_s <- read_tsv("../data/metagenome/metag.sample.counts")
 ```
 
-**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models.
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
 We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model.
-A model trained on the individual or pool SCFAs alone had a median area under the receiver operator characteristic curve (AUROC) of **0.XX** (IQR=**0.XX**).
-When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances (AUROC~OTU+SCFA~=**0.XX**, AUROC~genus+SCFA~=**0.XX**), the AUROC values were no different from the models trained without the SCFA concentrations (AUROC~OTU~=**0.XX**, AUROC~genus~=**0.XX**; Figure 2A).
-We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=**XX**, N~adenoma~=**XX**, and N~cancer~=**XX**).
-We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories [**XXXXXX**].
-Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (Figure 2A).
+Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
+When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
+We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
+We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories.
+Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A).
 These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

### 2019-03-19 submission/manuscript.tex
Add code to classification paragraph

@@
 classify individuals as having adenomas or carcinomas using Random
-Forest machine learning models. We repeated that analysis but added the
-concentration of the individual or pooled SCFAs as possible features to
-train a model. A model trained on the individual or pool SCFAs alone had
-a median area under the receiver operator characteristic curve (AUROC)
-of \textbf{0.XX} (IQR=\textbf{0.XX}). When we trained the models with
-the SCFAs concentrations and OTU or genus-level relative abundances
-(AUROC\textsubscript{OTU+SCFA}=\textbf{0.XX},
-AUROC\textsubscript{genus+SCFA}=\textbf{0.XX}), the AUROC values were no
-different from the models trained without the SCFA concentrations
-(AUROC\textsubscript{OTU}=\textbf{0.XX},
-AUROC\textsubscript{genus}=\textbf{0.XX}; Figure 2A). We also trained
-models using a smaller dataset that generated shotgun metagenomic
-sequencing data from a subset of our cohort
-(N\textsubscript{normal}=\textbf{XX},
-N\textsubscript{adenoma}=\textbf{XX}, and
-N\textsubscript{cancer}=\textbf{XX}). We binned genes extracted from the
+Forest machine learning models (13, 14). We repeated that analysis but
+added the concentration of the individual or pooled SCFAs as possible
+features to train a model. Models trained using SCFAs to classify
+individuals as having adenomas or carcinomas rather than normal colons
+had median areas under the receiver operator characteristic curve
+(AUROC) that were significantly greater than 0.5
+(P\textsubscript{adenoma}\textless{}0.001 and
+P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
+to detect the presence of adenomas or carcinomas were only 0.54 and
+0.55, respectively (Figure 2A). When we trained the models with the
+SCFAs concentrations and OTU or genus-level relative abundances the
+AUROC values were not significantly different from the models trained
+without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). We
+also trained models using a smaller dataset that generated shotgun
+metagenomic sequencing data from a subset of our cohort
+(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
+N\textsubscript{cancer}=26) (16). We binned genes extracted from the
 assembled metagenomes into operational protein families (OPFs) or KEGG
-categories {[}\textbf{XXXXXX}{]}. Again, the performance of the models
-trained with the meteagenomic data did not improve when the SCFA
-concentrations were added as possible features when training the model
-(Figure 2A). These data demonstrate that knowledge of the SCFA profile
+categories. Again, the performance of the models trained with the
+meteagenomic data did not improve when the SCFA concentrations were
+added as possible features when training the model (P\textgreater{}0.24;
+Figure 2A). These data demonstrate that knowledge of the SCFA profile
 from a patient's fecal sample does not improve the ability to diagnose a
@@
 
-\textbf{Conclusion.}
-
-\begin{itemize}
-\tightlist
-\item
-  Unlikely that SCFAs are the primary mechanism that limits
-  tumorigenesis
-\item
-  Limitations
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Observing the community once the tumor is developed
-  \item
-    Not looking at mucosal communities
-  \end{itemize}
-\item
-  Need to identify other possible mechanisms that drive and prevent
-  tumorigenesis
-\item
-  Supports evidence that fiber consumption has a null effect on
-  tumorigenesis
-\end{itemize}
+\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
+are not associated with the presence of adenomas or carcinomas and that
+they provide weak predictive power to improve the ability to diagnose
+someone with one of these lesions. Furthermore, knowledge of the
+taxonomic and genetic structure of gut microbiota was not predictive of
+individual or pooled SCFA concentrations. These results complement
+existing literature that suggest that fiber consumption and the
+production of SCFAs are unable to prevent the risk of developing colonic
+tumors. It is important to note that our analysis concerned fecal SCFA
+concentrations and microbiome characterization and that observations
+along the mucosa near the site of lesions may provide a stronger
+association. Regardless, given the growing literature in this area, it
+is unlikely that SCFAs are the primary mechanism that limits
+tumorigenesis. This may be a cautionary result to temper enthusiasm for
+SCFAs as a biomarker of gut health more generally. Going forward it is
+critical to develop additional hypotheses for how the microbiome and
+host interact to drive tumorigenesis to better understand the disease
+process and identify biomarkers that will allow early detection of
+tumors.
 

### 2019-03-19 code/rf_classification.R
Remove pooled column from scfa data frame

@@ -180,6 +180,7 @@ get_data <- function(path) {
 	if("scfa" %in% feature_sources){
 		data <- read_tsv('data/scfa/scfa_composite.tsv',
 										col_types=cols(study_id=col_character())) %>%
+		filter(scfa != "pooled") %>% 
 		spread(key=scfa, value=mmol_kg) %>%
 		inner_join(data, ., by=c("sample"="study_id"))
 	}

### 2019-03-20 submission/manuscript.Rmd
First pass at Materials and Methods

@@
 
+follow_up_time <- read_csv('../data/metadata/follow_up.csv') %>%
+ 	summarize(q1=quantile(time, prob=0.25), median=median(time), q3=quantile(time, prob=0.75))
+
 follow_test <- read_tsv('../data/process/scfa_pre_post_stats.tsv')
@@
 Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
-When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
+When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
 We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
@@
 
-**Study design and sampling.** The overall protocol has been described in detail previously [@Sze2017; @Baxter2016].
-In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=**r length(metai$sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**).
-For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf$time)** - **r max(metaf$time)** days.
-Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@Sze2017].
-Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@Sze2017; @Baxter2016].
-The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.
+**Study design and sampling.** The overall study design and the resulting sequence data have been previously described [@Sze2017; @Baxter2016].
+In brief, fecal samples were obtained from `r n_normal` individuals with normal colons, `r n_adenoma` individuals with colonic adenomas, and `r n_cancer` individuals with carcinomas.
+Of the individuals diagnosed as having adenomas or carcinomas, a subset (N~adenoma~=`r n_follow_adenoma` and N~carcinoma~=`r n_follow_cancer`) were sampled after treatment of the lesion (median=`r follow_up_time$median` days between sampling, IQR=`r follow_up_time$q1` to `r follow_up_time$q3` days).
+Tumor diagnosis was made by colonoscopic examination and histopathological review of the biopsies [@Baxter2016].
+The University of Michigan Institutional Review Board approved the studies that generated the samples and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.
 
-**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@Venkataraman2016].
-The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations.
-These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).
+**Measuring specific SCFAs.** The measurement of acetate, propionate, isobutyrate, and butyrate used a previously published protocol that used High-Performance Liquid Chromatography (HPLC) [@Venkataraman2016].
+Two changes were made to the protocol.
+First, instead of using fecal samples suspended in DNA Genotek OmniGut tubes, we suspended frozen fecal samples in 1 mL of PBS.
+Second, instead of using the average weight of fecal sample aliquots to normalize SCFA concentrations, we used the actual weight of the fecal samples.
+These methodological changes did not affect the range of concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).
 
-**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @Sze2017].
-In brief, sequences were quality filtered and contigs created from the paired end reads.
-Any sequences with ambiguous base calls were discarded.
-Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012].
-Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011].
-The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@Westcott2017].
+**16S rRNA gene sequence data analysis.** Sequence data from Baxter et al.
+[@Baxter2016] and Sze et al.
+[@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009].
+The sequence generation workflow and processing steps have been previously described [@Schloss2009; @Kozich2013].
+In brief, the original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer [@Kozich2013].
+The resulting sequence data were assembled into contigs and screened for quality aberrations including chimeras, which were removed based on the use of UCHIME [@Edgar2011].
+Sequences were clustered into OTUs at 97% similarity using OptiClust [@Westcott2017] and assigned to the closest possible genus using a naive Bayesian classifier with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16) [@Wang2007].
+We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].
 
-**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used.
-Briefly, the mothur shared file and metadata was converted into a biom formatted table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.
+**Metagenomic DNA sequence analysis.** A subset of the samples from the samples described by Baxter et al.
+[@Baxter2016] were used to generate metagenomic sequence data (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`).
+These data were generated by Hannigan et al.
+[@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915).
+Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads.
+The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data.
+We downloaded the sequences and assembled them into contigs using MEGAHIT [@Li2015], which were used to identify open reading frames (ORFs) using Prodigal [@Hyatt2010].
+We determined the abundance of each ORF by mapping the raw reads back to the ORFs using Diamond [@Buchfink2014].
+We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017; @Schloss2008].
+We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped [@Kanehisa2015].
 
-**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=**r length(metai$sample)**) containing a total of **r length(geof_samples$X1)** individuals (normal n=**r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()**, adenoma n=**r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()**, and carcinoma n=**r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()**) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017].
-Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015].
-Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].
+**Random Forest models.**
 
-**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@Liaw2002; @Kuhn2017].
-This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model.
-The reported AUCs, unless otherwise specified, are for the test sets.
-The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.
-The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.
+**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.1) with the tidyverse package (v.1.2.1).
+To assess differences in SCFA concentrations between individuals normal colons and those with adenomas or carcinomas, we used the Kruskal-Wallis rank sum test.
+If a test had a P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test with a Benjamini-Hochberg correction for multiple comparisons.
+To assess differences in SCFA concentrations between individuals samples before and after treatment we used paired Wilcoxon rank sum tests to test for significance.
+To compare the median AUCROC for the held out data for the model generated using only the SCFAs, we compared the distribution of the data to the expected median of 0.5 using the Wilcoxon rank sum test to test whether the model performed better than would be achieved by randomly assigning the data to each diagnosis.
+When we compared the Random Forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.
 
-**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017].
-Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas.
-Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [**XXXXXX**].
-First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma).
-We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment.
-Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested.
-Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed.
-From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho.
-Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models.
-Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm.
-For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.
+**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/.
 

### 2019-03-20 submission/manuscript.tex
First pass at Materials and Methods

@@
 0.55, respectively (Figure 2A). When we trained the models with the
-SCFAs concentrations and OTU or genus-level relative abundances the
-AUROC values were not significantly different from the models trained
-without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). We
-also trained models using a smaller dataset that generated shotgun
-metagenomic sequencing data from a subset of our cohort
-(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
-N\textsubscript{cancer}=26) (16). We binned genes extracted from the
+SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
+relative abundances the AUROC values were not significantly different
+from the models trained without the SCFA concentrations
+(P\textgreater{}0.21; Figure 2A). We also trained models using a smaller
+dataset that generated shotgun metagenomic sequencing data from a subset
+of our cohort (N\textsubscript{normal}=27, N\textsubscript{adenoma}=25,
+and N\textsubscript{cancer}=26) (16). We binned genes extracted from the
 assembled metagenomes into operational protein families (OPFs) or KEGG
@@
 
-\textbf{Study design and sampling.} The overall protocol has been
-described in detail previously (14, 37). In brief, this study used fecal
-samples obtained at either a single cross-sectional time point
-(n=\textbf{r
-length(metai\(sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**).
-For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf
-\)time)} - \textbf{r max(metaf\$time)} days. Our use of treatment has
-been previously defined as encompassing removal of a tumor (surgery or
-colonoscopy) with or without chemotherapy and radiation (37). Diagnosis
-of tumor was made by colonoscopic examination and histopathological
-review of biopsies obtained (14, 37). The University of Michigan
-Institutional Review Board approved the study and informed consent was
-obtained from all participants in accordance to the guidelines set out
-by the Helsinki Declaration.
+\textbf{Study design and sampling.} The overall study design and the
+resulting sequence data have been previously described (14, 37). In
+brief, fecal samples were obtained from 172 individuals with normal
+colons, 198 individuals with colonic adenomas, and 120 individuals with
+carcinomas. Of the individuals diagnosed as having adenomas or
+carcinomas, a subset (N\textsubscript{adenoma}=41 and
+N\textsubscript{carcinoma}=26) were sampled after treatment of the
+lesion (median=255 days between sampling, IQR=233 to 334 days). Tumor
+diagnosis was made by colonoscopic examination and histopathological
+review of the biopsies (14). The University of Michigan Institutional
+Review Board approved the studies that generated the samples and
+informed consent was obtained from all participants in accordance to the
+guidelines set out by the Helsinki Declaration.
 
-\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
-acetate, butyrate, and propionate followed a previously published
-protocol that used a High-Performance Liquid Chromatography (HPLC)
-machine (38). The following changes to this protocol included the use of
-frozen fecal samples suspended in 1ml of PBS instead of fecal
-suspensions in DNA Genotek OmniGut tubes, and the use of the actual
-weight of fecal samples instead of the average weight for SCFA
-concentration normalizations. These methodological changes did not
-affect the overall median concentrations of these SCFAs between the two
-studies (see Table 1 (38) and Figure 1 here).
+\textbf{Measuring specific SCFAs.} The measurement of acetate,
+propionate, isobutyrate, and butyrate used a previously published
+protocol that used High-Performance Liquid Chromatography (HPLC) (38).
+Two changes were made to the protocol. First, instead of using fecal
+samples suspended in DNA Genotek OmniGut tubes, we suspended frozen
+fecal samples in 1 mL of PBS. Second, instead of using the average
+weight of fecal sample aliquots to normalize SCFA concentrations, we
+used the actual weight of the fecal samples. These methodological
+changes did not affect the range of concentrations of these SCFAs
+between the two studies (see Table 1 (38) and Figure 1 here).
 
-\textbf{16S rRNA gene sequencing.} The workflow and processing have been
-previously described (37, 39, 40). In brief, sequences were quality
-filtered and contigs created from the paired end reads. Any sequences
-with ambiguous base calls were discarded. Contigs were then checked for
-matches to the V4 region of the 16S rRNA gene using the SILVA database
-(41). Chimeras were identified and removed using UCHIME and OTUs
-clustered at 97\% similarity (42). The major differences from these
-previous reports include: the use of version 1.39.5 of the mothur
-software package and clustering Operational Taxonomic Units (OTUs) at
-97\% similarity using the OptiClust algorithm (43).
+\textbf{16S rRNA gene sequence data analysis.} Sequence data from Baxter
+et al.~(14) and Sze et al.~(37) were obtained from the Sequence Read
+Archive (studies SRP062005 and SRP096978) and reprocessed using using
+mothur v.1.42 (39). The sequence generation workflow and processing
+steps have been previously described (39, 40). In brief, the original
+studies generated sequence data from V4 region of the 16S rRNA gene
+using paired 250 nt reads on an Illumina MiSeq sequencer (40). The
+resulting sequence data were assembled into contigs and screened for
+quality aberrations including chimeras, which were removed based on the
+use of UCHIME (41). Sequences were clustered into OTUs at 97\%
+similarity using OptiClust (42) and assigned to the closest possible
+genus using a naive Bayesian classifier with an 80\% confidence
+threshold trained on the reference collection from the Ribosomal
+Database Project (v.16) (43). We used PICRUSt (v.2.1.0-b) with the
+recommended standard operating protocol to generate imputed metagenomes
+based on the expected metabolic pathways, KEGG categories, and enzyme
+commission numbers (44).
 
-\textbf{Generating imputed metagenomes.} The use of PICRUSt version
-1.1.2 with the recommended standard operating protocol (44) was used.
-Briefly, the mothur shared file and metadata was converted into a biom
-formatted table using the biom convert function, the subsequent biom
-file was processed with the `normalize\_by\_copy\_number.py' function,
-and subsequent imputed metagenomes created using the
-`predict\_metagenomes.py' function.
+\textbf{Metagenomic DNA sequence analysis.} A subset of the samples from
+the samples described by Baxter et al.~(14) were used to generate
+metagenomic sequence data (N\textsubscript{normal}=27,
+N\textsubscript{adenoma}=25, and N\textsubscript{cancer}=26). These data
+were generated by Hannigan et al.~(16) and deposited into the Sequence
+Read Archive (study SRP108915). Fecal DNA was subjected to shotgun
+sequencing on an Illumina HiSeq using 125 bp paired end reads. The
+archived sequences were already quality filtered and aligned to the
+human genome to remove contaminating sequence data. We downloaded the
+sequences and assembled them into contigs using MEGAHIT (45), which were
+used to identify open reading frames (ORFs) using Prodigal (46). We
+determined the abundance of each ORF by mapping the raw reads back to
+the ORFs using Diamond (47). We clustered the ORFs into operational
+protein families (OPFs) in which the clustered ORFs were more than 40\%
+identical to each other using mmseq2 (48, 49). We also used mmseq2 to
+map the ORFs to the KEGG database and clustered the ORFs according to
+which category the ORFs mapped (50).
 
-\textbf{Obtaining Operational Protein Families from metagenomes.} A
-subset of the cross-sectional group (n=\textbf{r
-length(metai\(sample)**) containing a total of **r length(geof_samples\)X1)}
-individuals (normal n=\textbf{r filter(geof\_samples, X30 ==
-``Healthy'') \%\textgreater{}\% pull(X30) \%\textgreater{}\% length()},
-adenoma n=\textbf{r filter(geof\_samples, X30 == ``Adenoma'')
-\%\textgreater{}\% pull(X30) \%\textgreater{}\% length()}, and carcinoma
-n=\textbf{r filter(geof\_samples, X30 == ``Cancer'') \%\textgreater{}\%
-pull(X30) \%\textgreater{}\% length()}) was shotgun sequenced on an
-Illumina HiSeq using 125 bp paired end reads and a previously described
-method (16). Briefly, the sequences were quality filtered and sequences
-aligning to the human genome were removed prior to contig assembly with
-MEGAHIT (45). Open Reading Frames (ORFs) were identified using Prodigal
-(46), counts generated using Diamond (47), subsequent clustering into
-Operational Protein Families (OPFs) used mmseq2 (48), and OPF alignment
-used the KEGG database (49).
+\textbf{Random Forest models.}
 
-\textbf{Random Forest models.} The model was first trained on 80\% of
-the data and then tested on the held out 20\% (80/20 split) using the
-Random Forest algorithm for classification and regression models via the
-caret package (50, 51). This was repeated on 100 different 80/20 splits
-of the data to generate a reasonable range for the AUC of the model. The
-reported AUCs, unless otherwise specified, are for the test sets. The
-classification models were built to group normal versus adenoma, normal
-versus carcinoma, and high versus low SCFA concentrations. The
-regression models were built to classify the SCFA concentrations of
-acetate, butyrate, and propionate regardless of disease status.
+\textbf{Statistical analysis workflow.} Data summaries, statistical
+analysis, and data visualizations were performed using R (v.3.5.1) with
+the tidyverse package (v.1.2.1). To assess differences in SCFA
+concentrations between individuals normal colons and those with adenomas
+or carcinomas, we used the Kruskal-Wallis rank sum test. If a test had a
+P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test
+with a Benjamini-Hochberg correction for multiple comparisons. To assess
+differences in SCFA concentrations between individuals samples before
+and after treatment we used paired Wilcoxon rank sum tests to test for
+significance. To compare the median AUCROC for the held out data for the
+model generated using only the SCFAs, we compared the distribution of
+the data to the expected median of 0.5 using the Wilcoxon rank sum test
+to test whether the model performed better than would be achieved by
+randomly assigning the data to each diagnosis. When we compared the
+Random Forest models generated without and with SCFA data included, we
+used Wilcoxon rank sum tests to determine whether the models with the
+SCFA data included did better.
 
-\textbf{Statistical analysis workflow.} All analysis was performed using
-the statistical language R (52). Generally, a Kruskal-Wallis rank sum
-test with a Dunn's post-hoc test was used to assess differences between
-individuals without colon tumors, patients with adenomas, and patients
-with carcinomas. Where appropriate Benjamini-Hochberg was used to
-correct for multiple comparisons {[}\textbf{XXXXXX}{]}. First, we
-assessed differences in SCFA concentrations measured by HPLC between
-individuals with normal colons and patients with tumors (adenoma or
-carcinoma). We then analyzed whether SCFA concentrations changed in
-patients with an adenoma or carcinoma pre- versus post-treatment. Next,
-the imputed gene counts of genes encoding enzymes involved in SCFA
-synthesis was tested. Additionally, metagenomic sequencing counts for
-important genes involved with SCFA production were analyzed. From here
-we analyzed the number of significant positive and negative correlations
-between OTU relative abundance and SCFA concentrations in individuals
-without colon tumors and patients with adenomas or carcinomas using
-Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs
-were better able to classify individuals with and without tumors using
-Random Forest models. Finally, models to classify the actual SCFA
-concentration or high/low SCFA concentration based on the median of each
-SCFA using 16S rRNA gene sequencing data was created using the Random
-Forest algorithm. For all Random Forest models, the assessment of the
-most important variables was based on the top 10 features (OTUs or
-SCFAs) using the mean decrease in accuracy.
+\textbf{Code availability.} The code for all sequence curation and
+analysis steps including an Rmarkdown version of this manuscript is
+available at \url{https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/}.
 
@@
 
-\leavevmode\hypertarget{ref-Quast2012}{}%
-41. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
-\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
-J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
-project: Improved data processing and web-based tools. Nucleic Acids
-Research \textbf{41}:D590--D596.
-doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
-
 \leavevmode\hypertarget{ref-Edgar2011}{}%
-42. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+41. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
 \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
@@
 \leavevmode\hypertarget{ref-Westcott2017}{}%
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
+\leavevmode\hypertarget{ref-Wang2007}{}%
+43. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
 
@@
 
+\leavevmode\hypertarget{ref-Schloss2008}{}%
+49. \textbf{Schloss PD}, \textbf{Handelsman J}. 2008. A statistical
+toolbox for metagenomics: Assessing functional diversity in microbial
+communities. BMC Bioinformatics \textbf{9}.
+doi:\href{https://doi.org/10.1186/1471-2105-9-34}{10.1186/1471-2105-9-34}.
+
 \leavevmode\hypertarget{ref-Kanehisa2015}{}%
-49. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
+50. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
 \textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
@@
 doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.
-
-\leavevmode\hypertarget{ref-Liaw2002}{}%
-50. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
-regression by randomForest. R News \textbf{2}:18--22.
-
-\leavevmode\hypertarget{ref-Kuhn2017}{}%
-51. \textbf{Kuhn M}. 2017. Caret: Classification and regression
-training.
-
-\leavevmode\hypertarget{ref-r_citation_2017}{}%
-52. \textbf{R Core Team}. 2017. R: A language and environment for
-statistical computing. R Foundation for Statistical Computing, Vienna,
-Austria.
 

### 2019-03-20 code/rf_compare_classification_models.R

@@ -60,7 +60,7 @@ model_data %>%
 model_data %>%
 	filter(input == "scfa") %>%
 	group_by(class) %>%
-	summarize(test=list(tidy(wilcox.test(test_aucs, mu=0.5))),
+	summarize(test=list(tidy(wilcox.test(test_aucs, mu=0.5, alternative="great"))),
 						summary=list(tidy(summary(test_aucs)))) %>%
 	unnest() %>%
 	write_tsv("data/rf/classification_SCFA_to_random.tsv")

### 2019-03-20 submission/manuscript.Rmd
First pass at reducing the number of References

@@
 
-Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016].
+Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016].
 Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990].
@@
 Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
-Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016].
-Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017].
-Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Hannigan2017; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012].
+Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis.
+Numerous studies in murine models have supported this model [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
+Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018].
 These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.
 
-SCFAs have have anti-inflammatory and anti-proliferative activities [@Encarnao2018; @Verma2018; @Zheng2017; @OKeefe2016].
-Furthermore, manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018; @Tian2018].
+SCFAs have have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
+Manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018].
 These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
 Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
-Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Vital2014; @Sanna2019; @Liu2019; @Meisel2016].
+Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
 Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013].
-Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Schatzkin2000; @Yao2017; @Kunzmann2015; @Murphy2012; @Gianfredi2018].
+Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
 These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.
@@
 The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A).
-The median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas.
+Interestingly, the median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas, respectively.
 The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`).
-The median concentration of isobutyrate in people with normal colons (`r iso_p_cross$normal` mmol/kg) was between those with adenomas (`r iso_p_cross$adenoma` mmol/kg) and carcinomas (`r iso_p_cross$cancer` mmol/kg).
 Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
@@
 Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
-Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using picrust2.
+Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using PICRUSt.
 Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
@@
 Second, instead of using the average weight of fecal sample aliquots to normalize SCFA concentrations, we used the actual weight of the fecal samples.
-These methodological changes did not affect the range of concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).
+These methodological changes did not affect the range of concentrations of these SCFAs between the two studies.
 
@@
 [@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009].
-The sequence generation workflow and processing steps have been previously described [@Schloss2009; @Kozich2013].
-In brief, the original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer [@Kozich2013].
-The resulting sequence data were assembled into contigs and screened for quality aberrations including chimeras, which were removed based on the use of UCHIME [@Edgar2011].
-Sequences were clustered into OTUs at 97% similarity using OptiClust [@Westcott2017] and assigned to the closest possible genus using a naive Bayesian classifier with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16) [@Wang2007].
+The original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer.
+The resulting sequence data were assembled into contigs, screened to remove low quality contigs and chimeras.
+The curated sequences were then clustered into OTUs at a 97% similarity threshold and assigned to the closest possible genus with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16).
 We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].
@@
 We determined the abundance of each ORF by mapping the raw reads back to the ORFs using Diamond [@Buchfink2014].
-We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017; @Schloss2008].
-We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped [@Kanehisa2015].
+We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017].
+We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped.
 

### 2019-03-20 submission/manuscript.tex
First pass at reducing the number of References

@@
 Colorectal cancer is the third leading cancer-related cause of death
-within the United States (1, 2). Less than 10\% of cases can be
-attributed to genetic risk factors (3). This leaves a significant role
-for environmental, behavioral, and other factors such as smoking and
-diet (4, 5). Colorectal cancer is thought to be initiated by a series of
+within the United States (1). Less than 10\% of cases can be attributed
+to genetic risk factors (2). This leaves a significant role for
+environmental, behavioral, and other factors such as smoking and diet
+(3, 4). Colorectal cancer is thought to be initiated by a series of
 mutations that accumulate as mutated cells proliferate leading to
-adenomatous lesions, which are succeeded by carcinomas (3). Throughout
+adenomatous lesions, which are succeeded by carcinomas (2). Throughout
 this progression, there are ample opportunities for bacterial
 populations to create mutations, induce inflammation, and accelerate
-tumorigenesis (6). Numerous studies in murine models have supported this
-model (7--11). Additional cross sectional studies in humans have
-identified microbiome-based biomarkers of disease (12--20). These
-studies suggest that in some cases, it is the loss of bacterial
-populations that produce short-chain fatty acids (SCFAs) that results in
-increased inflammation and tumorigenesis.
+tumorigenesis. Numerous studies in murine models have supported this
+model (5--7). Additional cross sectional studies in humans have
+identified microbiome-based biomarkers of disease (8). These studies
+suggest that in some cases, it is the loss of bacterial populations that
+produce short-chain fatty acids (SCFAs) that results in increased
+inflammation and tumorigenesis.
 
-SCFAs have have anti-inflammatory and anti-proliferative activities
-(21--24). Furthermore, manipulation of SCFAs in mouse models of
-colorectal cancer by direct supplementation or feeding of fiber caused
-an overall reduction in tumor burden (25, 26). These results suggest
-that supplementation with substrates that bacteria can ferment to
-produce SCFAs may confer beneficial effects against colorectal cancer.
-Regardless, there is a lack of evidence that increasing SCFA
-concentrations can protect against colorectal cancer in humans. Based on
-similar observations, many microbiome studies use the concentrations of
-SCFAs and the presence of 16S rRNA gene sequences from organisms and the
-genes involved in producing them as a biomarker of a healthy microbiota
-(27--30). Case-control studies that have investigated SCFA
-concentrations in colorectal cancer found that patients with carcinomas
-had lower concentrations of SCFAs versus patients with adenomas or
-individuals without colon tumors (31). Although this would argue that
-increasing SCFA concentrations could be protective against
+SCFAs have have anti-inflammatory and anti-proliferative activities (9).
+Manipulation of SCFAs in mouse models of colorectal cancer by direct
+supplementation or feeding of fiber caused an overall reduction in tumor
+burden (10). These results suggest that supplementation with substrates
+that bacteria can ferment to produce SCFAs may confer beneficial effects
+against colorectal cancer. Regardless, there is a lack of evidence that
+increasing SCFA concentrations can protect against colorectal cancer in
+humans. Based on similar observations, many microbiome studies use the
+concentrations of SCFAs and the presence of 16S rRNA gene sequences from
+organisms and the genes involved in producing them as a biomarker of a
+healthy microbiota (11, 12). Case-control studies that have investigated
+SCFA concentrations in colorectal cancer found that patients with
+carcinomas had lower concentrations of SCFAs versus patients with
+adenomas or individuals without colon tumors (13). Although this would
+argue that increasing SCFA concentrations could be protective against
 tumorigenesis, in randomized controlled trials fiber supplementation has
 been inconsistently associated with protection against tumor formation
-and recurrence (32--36). These findings temper enthusiasm for treatments
+and recurrence (14, 15). These findings temper enthusiasm for treatments
 that aim to use SCFAs as biomarkers or protection against tumorigenesis.
@@
 characterized individuals with normal colons (N=172) and those with
-colonic adenomas (N=198) or carcinomas (N=120) (14). The only SCFA that
+colonic adenomas (N=198) or carcinomas (N=120) (16). The only SCFA that
 had a significantly different concentration across the diagnoses was
-isobutyrate (P=0.0091; Figure 1A). The median concentration of
-isobutyrate was 3.30 mmol/kg in people with normal colons and it was
-3.00 and 3.84 mmol/kg in people with colonic adenomas or carcinomas. The
-difference in isobutyrate concentration between people with adenomas and
-carcinomas was significantly different (P=0.0065); however, the
-differences between people with normal colons and those with adenomas or
-carcinomas was not significant (P=0.19 and P=0.11). The median
-concentration of isobutyrate in people with normal colons (3.30 mmol/kg)
-was between those with adenomas (3.00 mmol/kg) and carcinomas (3.84
-mmol/kg). Among the subjects with adenomas and carcinomas, a subset
+isobutyrate (P=0.0091; Figure 1A). Interestingly, the median
+concentration of isobutyrate was 3.30 mmol/kg in people with normal
+colons and it was 3.00 and 3.84 mmol/kg in people with colonic adenomas
+or carcinomas, respectively. The difference in isobutyrate concentration
+between people with adenomas and carcinomas was significantly different
+(P=0.0065); however, the differences between people with normal colons
+and those with adenomas or carcinomas was not significant (P=0.19 and
+P=0.11). Among the subjects with adenomas and carcinomas, a subset
 (N\textsubscript{adenoma}=41, N\textsubscript{carcinoma}=26) were
-treated and sampled a year later (37). The only SCFA that changed
+treated and sampled a year later (17). The only SCFA that changed
 following treatment was isobutyrate, which decreased by 0.99 mmol/kg
@@
 classify individuals as having adenomas or carcinomas using Random
-Forest machine learning models (13, 14). We repeated that analysis but
+Forest machine learning models (8, 16). We repeated that analysis but
 added the concentration of the individual or pooled SCFAs as possible
@@
 of our cohort (N\textsubscript{normal}=27, N\textsubscript{adenoma}=25,
-and N\textsubscript{cancer}=26) (16). We binned genes extracted from the
+and N\textsubscript{cancer}=26) (18). We binned genes extracted from the
 assembled metagenomes into operational protein families (OPFs) or KEGG
@@
 into metabolic pathways, enzyme commission numbers, or KEGG categories
-using picrust2. Again, SCFA concentrations could not be predicted based
+using PICRUSt. Again, SCFA concentrations could not be predicted based
 on the imputed metagenomic data; the largest amount of variation that
@@
 \textbf{Study design and sampling.} The overall study design and the
-resulting sequence data have been previously described (14, 37). In
+resulting sequence data have been previously described (16, 17). In
 brief, fecal samples were obtained from 172 individuals with normal
@@
 diagnosis was made by colonoscopic examination and histopathological
-review of the biopsies (14). The University of Michigan Institutional
+review of the biopsies (16). The University of Michigan Institutional
 Review Board approved the studies that generated the samples and
@@
 propionate, isobutyrate, and butyrate used a previously published
-protocol that used High-Performance Liquid Chromatography (HPLC) (38).
+protocol that used High-Performance Liquid Chromatography (HPLC) (19).
 Two changes were made to the protocol. First, instead of using fecal
@@
 changes did not affect the range of concentrations of these SCFAs
-between the two studies (see Table 1 (38) and Figure 1 here).
+between the two studies.
 
 \textbf{16S rRNA gene sequence data analysis.} Sequence data from Baxter
-et al.~(14) and Sze et al.~(37) were obtained from the Sequence Read
+et al.~(16) and Sze et al.~(17) were obtained from the Sequence Read
 Archive (studies SRP062005 and SRP096978) and reprocessed using using
-mothur v.1.42 (39). The sequence generation workflow and processing
-steps have been previously described (39, 40). In brief, the original
-studies generated sequence data from V4 region of the 16S rRNA gene
-using paired 250 nt reads on an Illumina MiSeq sequencer (40). The
-resulting sequence data were assembled into contigs and screened for
-quality aberrations including chimeras, which were removed based on the
-use of UCHIME (41). Sequences were clustered into OTUs at 97\%
-similarity using OptiClust (42) and assigned to the closest possible
-genus using a naive Bayesian classifier with an 80\% confidence
-threshold trained on the reference collection from the Ribosomal
-Database Project (v.16) (43). We used PICRUSt (v.2.1.0-b) with the
+mothur v.1.42 (20). The original studies generated sequence data from V4
+region of the 16S rRNA gene using paired 250 nt reads on an Illumina
+MiSeq sequencer. The resulting sequence data were assembled into
+contigs, screened to remove low quality contigs and chimeras. The
+curated sequences were then clustered into OTUs at a 97\% similarity
+threshold and assigned to the closest possible genus with an 80\%
+confidence threshold trained on the reference collection from the
+Ribosomal Database Project (v.16). We used PICRUSt (v.2.1.0-b) with the
 recommended standard operating protocol to generate imputed metagenomes
 based on the expected metabolic pathways, KEGG categories, and enzyme
-commission numbers (44).
+commission numbers (21).
 
 \textbf{Metagenomic DNA sequence analysis.} A subset of the samples from
-the samples described by Baxter et al.~(14) were used to generate
+the samples described by Baxter et al.~(16) were used to generate
 metagenomic sequence data (N\textsubscript{normal}=27,
 N\textsubscript{adenoma}=25, and N\textsubscript{cancer}=26). These data
-were generated by Hannigan et al.~(16) and deposited into the Sequence
+were generated by Hannigan et al.~(18) and deposited into the Sequence
 Read Archive (study SRP108915). Fecal DNA was subjected to shotgun
@@
 human genome to remove contaminating sequence data. We downloaded the
-sequences and assembled them into contigs using MEGAHIT (45), which were
-used to identify open reading frames (ORFs) using Prodigal (46). We
+sequences and assembled them into contigs using MEGAHIT (22), which were
+used to identify open reading frames (ORFs) using Prodigal (23). We
 determined the abundance of each ORF by mapping the raw reads back to
-the ORFs using Diamond (47). We clustered the ORFs into operational
+the ORFs using Diamond (24). We clustered the ORFs into operational
 protein families (OPFs) in which the clustered ORFs were more than 40\%
-identical to each other using mmseq2 (48, 49). We also used mmseq2 to
-map the ORFs to the KEGG database and clustered the ORFs according to
-which category the ORFs mapped (50).
+identical to each other using mmseq2 (25). We also used mmseq2 to map
+the ORFs to the KEGG database and clustered the ORFs according to which
+category the ORFs mapped.
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Haggar2009}{}%
-1. \textbf{Haggar F}, \textbf{Boushey R}. 2009. Colorectal cancer
-epidemiology: Incidence, mortality, survival, and risk factors. Clinics
-in Colon and Rectal Surgery \textbf{22}:191--197.
-doi:\href{https://doi.org/10.1055/s-0029-1242458}{10.1055/s-0029-1242458}.
-
 \leavevmode\hypertarget{ref-Siegel2016}{}%
-2. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
+1. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
 Cancer statistics, 2016. CA: A Cancer Journal for Clinicians
@@
 \leavevmode\hypertarget{ref-Fearon1990}{}%
-3. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
+2. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
 colorectal tumorigenesis. Cell \textbf{61}:759--767.
@@
 \leavevmode\hypertarget{ref-FlissIsakov2017}{}%
-4. \textbf{Fliss-Isakov N}, \textbf{Zelber-Sagi S}, \textbf{Webb M},
+3. \textbf{Fliss-Isakov N}, \textbf{Zelber-Sagi S}, \textbf{Webb M},
 \textbf{Halpern Z}, \textbf{Kariv R}. 2017. Smoking habits are strongly
@@
 \leavevmode\hypertarget{ref-Lee2015}{}%
-5. \textbf{Lee J}, \textbf{Jeon JY}, \textbf{Meyerhardt JA}. 2015. Diet
+4. \textbf{Lee J}, \textbf{Jeon JY}, \textbf{Meyerhardt JA}. 2015. Diet
 and lifestyle in survivors of colorectal cancer. Hematology/Oncology
@@
 
-\leavevmode\hypertarget{ref-Flynn2016}{}%
-6. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
-Metabolic and community synergy of oral bacteria in colorectal cancer.
-mSphere \textbf{1}.
-doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.
-
 \leavevmode\hypertarget{ref-Zackular2013}{}%
-7. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+5. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
 \textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
@@
 
-\leavevmode\hypertarget{ref-Baxter2014}{}%
-8. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
-colonization with human feces determines colonic tumor burden.
-Microbiome \textbf{2}:20.
-doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
-
-\leavevmode\hypertarget{ref-Zackular2015}{}%
-9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2015. Manipulation of the gut microbiota reveals
-role in colon tumorigenesis. mSphere \textbf{1}:e00001--15.
-doi:\href{https://doi.org/10.1128/msphere.00001-15}{10.1128/msphere.00001-15}.
-
 \leavevmode\hypertarget{ref-DeStefanoShields2016}{}%
-10. \textbf{Shields CED}, \textbf{Meerbeke SWV}, \textbf{Housseau F},
+6. \textbf{Shields CED}, \textbf{Meerbeke SWV}, \textbf{Housseau F},
 \textbf{Wang H}, \textbf{Huso DL}, \textbf{Casero RA}, \textbf{O'Hagan
 HM}, \textbf{Sears CL}. 2016. Reduction of murine colon tumorigenesis
-driven by EnterotoxigenicBacteroides fragilisUsing cefoxitin treatment.
-Journal of Infectious Diseases \textbf{214}:122--129.
+driven by Enterotoxigenic Bacteroides fragilis using cefoxitin
+treatment. Journal of Infectious Diseases \textbf{214}:122--129.
 doi:\href{https://doi.org/10.1093/infdis/jiw069}{10.1093/infdis/jiw069}.
@@
 \leavevmode\hypertarget{ref-Tomkovich2017}{}%
-11. \textbf{Tomkovich S}, \textbf{Yang Y}, \textbf{Winglee K},
+7. \textbf{Tomkovich S}, \textbf{Yang Y}, \textbf{Winglee K},
 \textbf{Gauthier J}, \textbf{Mühlbauer M}, \textbf{Sun X},
@@
 
-\leavevmode\hypertarget{ref-Kostic2011}{}%
-12. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
-\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
-AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
-\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
-\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
-\textbf{Garrett WS}, \textbf{Meyerson M}. 2011. Genomic analysis
-identifies association of fusobacterium with colorectal carcinoma.
-Genome Research \textbf{22}:292--298.
-doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.
-
 \leavevmode\hypertarget{ref-Sze2018}{}%
-13. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+8. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 
-\leavevmode\hypertarget{ref-Baxter2016}{}%
-14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\leavevmode\hypertarget{ref-Zackular2014}{}%
-15. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prevention Research \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.capr-14-0129}{10.1158/1940-6207.capr-14-0129}.
-
-\leavevmode\hypertarget{ref-Hannigan2017}{}%
-16. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
-\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
-\& the interactive dynamics of the colorectal cancer virome.
-doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
-
-\leavevmode\hypertarget{ref-Zeller2014}{}%
-17. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
-\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
-\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
-\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
-\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
-\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
-\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
-\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
-\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
-for early-stage detection of colorectal cancer. Molecular Systems
-Biology \textbf{10}:766--766.
-doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
-
-\leavevmode\hypertarget{ref-Shah2017}{}%
-18. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
-\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
-\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
-sequence-based faecal microbial community survey data to identify a
-composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.
-
-\leavevmode\hypertarget{ref-Dejea2018}{}%
-19. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
-\textbf{Boleij A}, \textbf{Taddese R}, \textbf{Geis AL}, \textbf{Wu X},
-\textbf{Shields CED}, \textbf{Hechenbleikner EM}, \textbf{Huso DL},
-\textbf{Anders RA}, \textbf{Giardiello FM}, \textbf{Wick EC},
-\textbf{Wang H}, \textbf{Wu S}, \textbf{Pardoll DM}, \textbf{Housseau
-F}, \textbf{Sears CL}. 2018. Patients with familial adenomatous
-polyposis harbor colonic biofilms containing tumorigenic bacteria.
-Science \textbf{359}:592--597.
-doi:\href{https://doi.org/10.1126/science.aah3648}{10.1126/science.aah3648}.
-
-\leavevmode\hypertarget{ref-Arthur2012}{}%
-20. \textbf{Arthur JC}, \textbf{Perez-Chanona E}, \textbf{Muhlbauer M},
-\textbf{Tomkovich S}, \textbf{Uronis JM}, \textbf{Fan T-J},
-\textbf{Campbell BJ}, \textbf{Abujamel T}, \textbf{Dogan B},
-\textbf{Rogers AB}, \textbf{Rhodes JM}, \textbf{Stintzi A},
-\textbf{Simpson KW}, \textbf{Hansen JJ}, \textbf{Keku TO}, \textbf{Fodor
-AA}, \textbf{Jobin C}. 2012. Intestinal inflammation targets
-cancer-inducing activity of the microbiota. Science
-\textbf{338}:120--123.
-doi:\href{https://doi.org/10.1126/science.1224820}{10.1126/science.1224820}.
-
-\leavevmode\hypertarget{ref-Encarnao2018}{}%
-21. \textbf{Encarnação JC}, \textbf{Pires AS}, \textbf{Amaral RA},
-\textbf{Gonçalves TJ}, \textbf{Laranjo M}, \textbf{Casalta-Lopes JE},
-\textbf{Gonçalves AC}, \textbf{Sarmento-Ribeiro AB}, \textbf{Abrantes
-AM}, \textbf{Botelho MF}. 2018. Butyrate, a dietary fiber derivative
-that improves irinotecan effect in colon cancer cells. The Journal of
-Nutritional Biochemistry \textbf{56}:183--192.
-doi:\href{https://doi.org/10.1016/j.jnutbio.2018.02.018}{10.1016/j.jnutbio.2018.02.018}.
-
-\leavevmode\hypertarget{ref-Verma2018}{}%
-22. \textbf{Verma MS}, \textbf{Fink MJ}, \textbf{Salmon GL},
-\textbf{Fornelos N}, \textbf{Ohara TE}, \textbf{Ryu SH},
-\textbf{Vlamakis H}, \textbf{Xavier RJ}, \textbf{Stappenbeck TS},
-\textbf{Whitesides GM}. 2018. A common mechanism links activities of
-butyrate in the colon. ACS Chemical Biology \textbf{13}:1291--1298.
-doi:\href{https://doi.org/10.1021/acschembio.8b00073}{10.1021/acschembio.8b00073}.
-
-\leavevmode\hypertarget{ref-Zheng2017}{}%
-23. \textbf{Zheng L}, \textbf{Kelly CJ}, \textbf{Battista KD},
-\textbf{Schaefer R}, \textbf{Lanis JM}, \textbf{Alexeev EE},
-\textbf{Wang RX}, \textbf{Onyiah JC}, \textbf{Kominsky DJ},
-\textbf{Colgan SP}. 2017. Microbial-derived butyrate promotes epithelial
-barrier function through IL-10 receptorDependent repression of
-claudin-2. The Journal of Immunology \textbf{199}:2976--2984.
-doi:\href{https://doi.org/10.4049/jimmunol.1700105}{10.4049/jimmunol.1700105}.
-
 \leavevmode\hypertarget{ref-OKeefe2016}{}%
-24. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+9. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
 metabolites and colon cancer. Nature Reviews Gastroenterology \&
@@
 \leavevmode\hypertarget{ref-Bishehsari2018}{}%
-25. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+10. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
 \textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
@@
 
-\leavevmode\hypertarget{ref-Tian2018}{}%
-26. \textbf{Tian Y}, \textbf{Xu Q}, \textbf{Sun L}, \textbf{Ye Y},
-\textbf{Ji G}. 2018. Short-chain fatty acids administration is
-protective in colitis-associated colorectal cancer development. The
-Journal of Nutritional Biochemistry \textbf{57}:103--109.
-doi:\href{https://doi.org/10.1016/j.jnutbio.2018.03.007}{10.1016/j.jnutbio.2018.03.007}.
-
-\leavevmode\hypertarget{ref-Vital2014}{}%
-27. \textbf{Vital M}, \textbf{Howe AC}, \textbf{Tiedje JM}. 2014.
-Revealing the bacterial butyrate synthesis pathways by analyzing
-(meta)genomic data. mBio \textbf{5}:e00889--14--e00889--14.
-doi:\href{https://doi.org/10.1128/mbio.00889-14}{10.1128/mbio.00889-14}.
-
 \leavevmode\hypertarget{ref-Sanna2019}{}%
-28. \textbf{Sanna S}, \textbf{Zuydam NR van}, \textbf{Mahajan A},
+11. \textbf{Sanna S}, \textbf{Zuydam NR van}, \textbf{Mahajan A},
 \textbf{Kurilshikov A}, \textbf{Vila AV}, \textbf{Võsa U},
@@
 
-\leavevmode\hypertarget{ref-Liu2019}{}%
-29. \textbf{Liu S}, \textbf{Li E}, \textbf{Sun Z}, \textbf{Fu D},
-\textbf{Duan G}, \textbf{Jiang M}, \textbf{Yu Y}, \textbf{Mei L},
-\textbf{Yang P}, \textbf{Tang Y}, \textbf{Zheng P}. 2019. Altered gut
-microbiota and short chain fatty acids in chinese children with autism
-spectrum disorder. Scientific Reports \textbf{9}.
-doi:\href{https://doi.org/10.1038/s41598-018-36430-z}{10.1038/s41598-018-36430-z}.
-
 \leavevmode\hypertarget{ref-Meisel2016}{}%
-30. \textbf{Meisel M}, \textbf{Mayassi T}, \textbf{Fehlner-Peach H},
+12. \textbf{Meisel M}, \textbf{Mayassi T}, \textbf{Fehlner-Peach H},
 \textbf{Koval JC}, \textbf{O'Brien SL}, \textbf{Hinterleitner R},
@@
 \leavevmode\hypertarget{ref-Ohigashi2013}{}%
-31. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
+13. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
 \textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
@@
 
-\leavevmode\hypertarget{ref-Schatzkin2000}{}%
-32. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
-\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
-\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
-\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
-\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
-of effect of a low-fat, high-fiber diet on the recurrence of colorectal
-adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
-doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.
-
 \leavevmode\hypertarget{ref-Yao2017}{}%
-33. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
+14. \textbf{Yao Y}, \textbf{Suo T}, \textbf{Andersson R}, \textbf{Cao
 Y}, \textbf{Wang C}, \textbf{Lu J}, \textbf{Chui E}. 2017. Dietary fibre
@@
 
-\leavevmode\hypertarget{ref-Kunzmann2015}{}%
-34. \textbf{Kunzmann AT}, \textbf{Coleman HG}, \textbf{Huang W-Y},
-\textbf{Kitahara CM}, \textbf{Cantwell MM}, \textbf{Berndt SI}. 2015.
-Dietary fiber intake and risk of colorectal cancer and incident and
-recurrent adenoma in the prostate, lung, colorectal, and ovarian cancer
-screening trial. The American Journal of Clinical Nutrition
-\textbf{102}:881--890.
-doi:\href{https://doi.org/10.3945/ajcn.115.113282}{10.3945/ajcn.115.113282}.
-
-\leavevmode\hypertarget{ref-Murphy2012}{}%
-35. \textbf{Murphy N}, \textbf{Norat T}, \textbf{Ferrari P},
-\textbf{Jenab M}, \textbf{Bueno-de-Mesquita B}, \textbf{Skeie G},
-\textbf{Dahm CC}, \textbf{Overvad K}, \textbf{Olsen A},
-\textbf{Tjønneland A}, \textbf{Clavel-Chapelon F},
-\textbf{Boutron-Ruault MC}, \textbf{Racine A}, \textbf{Kaaks R},
-\textbf{Teucher B}, \textbf{Boeing H}, \textbf{Bergmann MM},
-\textbf{Trichopoulou A}, \textbf{Trichopoulos D}, \textbf{Lagiou P},
-\textbf{Palli D}, \textbf{Pala V}, \textbf{Panico S}, \textbf{Tumino R},
-\textbf{Vineis P}, \textbf{Siersema P}, \textbf{Duijnhoven F van},
-\textbf{Peeters PHM}, \textbf{Hjartaker A}, \textbf{Engeset D},
-\textbf{González CA}, \textbf{Sánchez M-J}, \textbf{Dorronsoro M},
-\textbf{Navarro C}, \textbf{Ardanaz E}, \textbf{Quirós JR},
-\textbf{Sonestedt E}, \textbf{Ericson U}, \textbf{Nilsson L},
-\textbf{Palmqvist R}, \textbf{Khaw K-T}, \textbf{Wareham N}, \textbf{Key
-TJ}, \textbf{Crowe FL}, \textbf{Fedirko V}, \textbf{Wark PA},
-\textbf{Chuang S-C}, \textbf{Riboli E}. 2012. Dietary fibre intake and
-risks of cancers of the colon and rectum in the european prospective
-investigation into cancer and nutrition (EPIC). PLoS ONE
-\textbf{7}:e39361.
-doi:\href{https://doi.org/10.1371/journal.pone.0039361}{10.1371/journal.pone.0039361}.
-
 \leavevmode\hypertarget{ref-Gianfredi2018}{}%
-36. \textbf{Gianfredi V}, \textbf{Salvatori T}, \textbf{Villarini M},
+15. \textbf{Gianfredi V}, \textbf{Salvatori T}, \textbf{Villarini M},
 \textbf{Moretti M}, \textbf{Nucci D}, \textbf{Realdon S}. 2018. Is
@@
 
+\leavevmode\hypertarget{ref-Baxter2016}{}%
+16. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
 \leavevmode\hypertarget{ref-Sze2017}{}%
-37. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+17. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
 \textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
@@
 
+\leavevmode\hypertarget{ref-Hannigan2017}{}%
+18. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
+\& the interactive dynamics of the colorectal cancer virome.
+doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
+
 \leavevmode\hypertarget{ref-Venkataraman2016}{}%
-38. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+19. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
 \textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+20. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\leavevmode\hypertarget{ref-Kozich2013}{}%
-40. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
-
-\leavevmode\hypertarget{ref-Edgar2011}{}%
-41. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
-\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
-and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
-
-\leavevmode\hypertarget{ref-Westcott2017}{}%
-42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-improved method for assigning amplicon-based sequence data to
-operational taxonomic units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
-
-\leavevmode\hypertarget{ref-Wang2007}{}%
-43. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
-\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
-of rRNA sequences into the new bacterial taxonomy. Applied and
-Environmental Microbiology \textbf{73}:5261--5267.
-doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
-
 \leavevmode\hypertarget{ref-Langille2013}{}%
-44. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
+21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
 \textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
@@
 \leavevmode\hypertarget{ref-Li2015}{}%
-45. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
+22. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
 K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
@@
 \leavevmode\hypertarget{ref-Hyatt2010}{}%
-46. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
+23. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
 \textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
@@
 \leavevmode\hypertarget{ref-Buchfink2014}{}%
-47. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
+24. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
 and sensitive protein alignment using DIAMOND. Nature Methods
@@
 \leavevmode\hypertarget{ref-Steinegger2017}{}%
-48. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
+25. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
 sensitive protein sequence searching for the analysis of massive data
@@
 doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.
-
-\leavevmode\hypertarget{ref-Schloss2008}{}%
-49. \textbf{Schloss PD}, \textbf{Handelsman J}. 2008. A statistical
-toolbox for metagenomics: Assessing functional diversity in microbial
-communities. BMC Bioinformatics \textbf{9}.
-doi:\href{https://doi.org/10.1186/1471-2105-9-34}{10.1186/1471-2105-9-34}.
-
-\leavevmode\hypertarget{ref-Kanehisa2015}{}%
-50. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
-\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
-resource for gene and protein annotation. Nucleic Acids Research
-\textbf{44}:D457--D462.
-doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.
 

### 2019-03-20 submission/manuscript.Rmd
First pass at Abstract and Importance sections

@@
 ---
-title: "Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors"
 bibliography: references.bib
@@
 
-Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure
+# Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure
 
@@
 
+\vspace{10mm}
+
+## Observation format
+
 \newpage
@@
 
-## Abstract (250 words)
+## Abstract
 
-Something clever
+The gut microbiome is thought to have a role in the development of colorectal cancer by protecting against and exacerbating inflammation.
+Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria.
+We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing.
+We measured the fecal concentrations of acetate, propionate, isobutyrate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status.
+When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
+Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples.
+These models performed poorly and were only able to explain less than 10% of the observed variation in the SCFA concentrations.
+These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer.
+Although it is likely that bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited value.
 
-## Importance (150 words)
+\newpage
 
-Something clever
+## Importance
+
+Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors.
+Short chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties.
+We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors.
+We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status.
+Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations.
+These data indicate that there is not a meaningful link between the gut microbiome, SCFAs, and tumor burden.
 

### 2019-03-20 submission/manuscript.tex
First pass at Abstract and Importance sections

@@
 \hypersetup{unicode=true,
-            pdftitle={Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors},
             pdfborder={0 0 0},
@@
 
-  \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
-Microbiota, and Colorectal Tumors}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
+  \title{}
+    \pretitle{\vspace{\droptitle}}
+  \posttitle{}
     \author{}
@@
 \begin{document}
-\maketitle
 
@@
 
-Fecal short chain fatty acids are not predictive of colorectal cancer
-status and cannot be predicted based on bacterial community structure
+\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
+\section{Fecal short chain fatty acids are not predictive of colorectal
+cancer status and cannot be predicted based on bacterial community
+structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}
 
@@
 
+\vspace{10mm}
+
+\hypertarget{observation-format}{%
+\subsection{Observation format}\label{observation-format}}
+
 \newpage
@@
 
-\hypertarget{abstract-250-words}{%
-\subsection{Abstract (250 words)}\label{abstract-250-words}}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
-Something clever
+The gut microbiome is thought to have a role in the development of
+colorectal cancer by protecting against and exacerbating inflammation.
+Short chain fatty acids (SCFAs), including butyrate, have been shown to
+have anti-inflammatory properties and are produced in large quantities
+by colonic bacteria. We assessed whether there was an association
+between fecal SCFA concentrations and the presence of colonic adenomas
+or carcinomas in a cohort of individuals that was previously subjected
+to 16S rRNA gene and metagenomic shotgun sequencing. We measured the
+fecal concentrations of acetate, propionate, isobutyrate, and butyrate
+within the cohort and found that there were no meaningful differences in
+their concentration and disease status. When we incorporated these
+concentrations into Random Forest classification models trained to
+differentiate between people with normal colons and those with adenomas
+or carcinomas, we found that they did not significantly improve the
+ability of 16S rRNA gene or metagenomic gene sequence-based models to
+classify individuals. Finally, we generated Random Forest regression
+models trained to predict the concentration of each SCFA based on 16S
+rRNA gene or metagenomic gene sequence data from the same samples. These
+models performed poorly and were only able to explain less than 10\% of
+the observed variation in the SCFA concentrations. These results support
+the broader epidemiological data that questions the value of fiber
+consumption for reducing the risks of colorectal cancer. Although it is
+likely that bacterial metabolites may serve as biomarkers to detect
+adenomas or carcinomas, fecal SCFA concentrations have limited value.
 
-\hypertarget{importance-150-words}{%
-\subsection{Importance (150 words)}\label{importance-150-words}}
+\newpage
 
-Something clever
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
+
+Considering colorectal cancer is the third leading cancer-related cause
+of death within the United States, there is a great need to detect
+colorectal tumors early without invasive colonoscopy procedures and to
+prevent the formation of tumors. Short chain fatty acids (SCFAs) are
+often used as a surrogate for measuring gut health and for being
+anti-carcinogenic because of their anti-inflammatory properties. We
+evaluated the fecal SCFA concentration of a cohort of individuals with
+varying colonic tumor burden who were previously analyzed to identify
+microbiome-based biomarkers of tumors. We were unable to find an
+association between SCFA concentration and tumor burden or use SCFAs to
+improve our microbiome-based models of classifying people based on their
+tumor status. Furthermore, we were unable to find an association between
+the fecal community structure and SCFA concentrations. These data
+indicate that there is not a meaningful link between the gut microbiome,
+SCFAs, and tumor burden.
 

### 2019-03-22 code/figure_1.R
Convert violin plots to box plots

@@ -26,7 +26,8 @@ follow_up_metadata <- read_csv("data/metadata/follow_up.csv",
 cross_section <- inner_join(scfa_data, cross_section_metadata, by=c("study_id"="sample")) %>%
 	select(scfa, mmol_kg, dx) %>%
 	ggplot(aes(x=dx, y=mmol_kg, fill=dx)) +
-		geom_violin(show.legend=FALSE, draw_quantiles=0.5) +
+		# geom_violin(show.legend=FALSE, draw_quantiles=0.5) +
+		geom_boxplot(show.legend=FALSE) +
 		facet_wrap(~scfa, scales="free_y", nrow=1, labeller=labeller(.default=capitalize)) +
 		labs(x=NULL, y="mmol of SCFA/kg of feces") +
 		scale_x_discrete(labels = NULL) +
@@ -56,7 +57,8 @@ longitudinal <- inner_join(scfa_data, follow_up_metadata, by=c("study_id"="initi
 	mutate(ymax = max(abs(diff)), ymin = -1 * ymax) %>%
 	ungroup() %>%
 	ggplot(aes(x=dx, y=diff, fill=dx)) +
-		geom_violin(draw_quantiles=0.5) +
+		# geom_violin(draw_quantiles=0.5) +
+		geom_boxplot() +
 		geom_blank(aes(y = ymin)) + geom_blank(aes(y = ymax)) +
 		facet_wrap(~scfa, scales="free_y", nrow=1, labeller=labeller(.default=capitalize)) +
 		labs(x=NULL, y="Difference in SCFA concentration\nbefore and after treatment") +

### 2019-03-27 code/figure_2.R
Modify output files from plot_scfa_comparison.R script

@@ -1,0 +1,31 @@
+library(tidyverse)
+library(RColorBrewer)
+library(cowplot)
+
+### Helper function
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+data_file_name <- "data/rf/classification_data_pool.tsv"
+
+model_data <- read_tsv(data_file_name) %>%
+	mutate(scfa = str_detect(input, "scfa"),
+				microbiome=str_replace(input, "scfa_", ""),
+				microbiome=str_replace(microbiome, "_scfa", "")
+		) %>%
+	select(class, microbiome, scfa, cv_aucs, test_aucs) %>%
+	filter(class != "lesion")
+
+
+model_data %>%
+	filter(microbiome %in% c("scfa", "otu", "genus")) %>%
+	mutate(microbiome = factor(microbiome, levels=c("scfa", "genus", "otu"))) %>%
+	ggplot(aes(x=microbiome, fill=scfa, y=test_aucs)) +
+	 geom_boxplot(position=position_dodge2(preserve = "single"))+
+	#	geom_violin(draw_quantiles=0.5)+#, position=position_dodge(preserve = "single")) +
+		facet_wrap(~class) +
+		coord_cartesian(ylim=c(0.35,1.0)) +
+		ggsave("test.pdf")

### 2019-03-27 code/plot_scfa_comparisons.R

@@ -81,4 +81,4 @@ longitudinal <- inner_join(scfa_data, follow_up_metadata, by=c("study_id"="initi
 
 ### Merge the figures together
 plot_grid(cross_section, longitudinal, nrow=2, labels=c("A", "B"))
-ggsave("results/figures/scfa_abundance.pdf", width=6.875, height=6, units="in")
+ggsave("results/figures/scfa_comparisons.pdf", width=6.875, height=6, units="in")

### 2019-03-27 code/figure_2.R
Generalize to include regression data

@@ -1,31 +1,0 @@
-library(tidyverse)
-library(RColorBrewer)
-library(cowplot)
-
-### Helper function
-capitalize <- function(string) {
-  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
-  string
-}
-
-
-data_file_name <- "data/rf/classification_data_pool.tsv"
-
-model_data <- read_tsv(data_file_name) %>%
-	mutate(scfa = str_detect(input, "scfa"),
-				microbiome=str_replace(input, "scfa_", ""),
-				microbiome=str_replace(microbiome, "_scfa", "")
-		) %>%
-	select(class, microbiome, scfa, cv_aucs, test_aucs) %>%
-	filter(class != "lesion")
-
-
-model_data %>%
-	filter(microbiome %in% c("scfa", "otu", "genus")) %>%
-	mutate(microbiome = factor(microbiome, levels=c("scfa", "genus", "otu"))) %>%
-	ggplot(aes(x=microbiome, fill=scfa, y=test_aucs)) +
-	 geom_boxplot(position=position_dodge2(preserve = "single"))+
-	#	geom_violin(draw_quantiles=0.5)+#, position=position_dodge(preserve = "single")) +
-		facet_wrap(~class) +
-		coord_cartesian(ylim=c(0.35,1.0)) +
-		ggsave("test.pdf")

### 2019-03-27 code/plot_classification_regression.R

@@ -1,0 +1,52 @@
+library(tidyverse)
+library(RColorBrewer)
+library(cowplot)
+
+### Helper function
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+data_file_name <- "data/rf/classification_data_pool.tsv"
+
+model_data <- read_tsv(data_file_name) %>%
+	mutate(scfa = str_detect(input, "scfa"),
+				microbiome=str_replace(input, "scfa_", ""),
+				microbiome=str_replace(microbiome, "_scfa", "")
+		) %>%
+	select(class, microbiome, scfa, cv_aucs, test_aucs) %>%
+	filter(class != "lesion")
+
+
+classification_plot <- model_data %>%
+	filter(microbiome %in% c("scfa", "otu", "genus")) %>%
+	mutate(microbiome = factor(microbiome, levels=c("scfa", "genus", "otu"))) %>%
+	ggplot(aes(x=microbiome, fill=scfa, y=test_aucs)) +
+		geom_hline(yintercept=0.5, col="#888888") +
+		geom_boxplot(position=position_dodge2(preserve = "single"))+
+		facet_wrap(~class, labeller=labeller(.default=capitalize)) +
+		coord_cartesian(ylim=c(0.35,1.0)) +
+		scale_x_discrete(
+			breaks = c("scfa", "genus", "otu"),
+			labels = c("None", "Genus", "OTU")
+		) +
+		scale_fill_manual(
+			name = NULL,
+			breaks = c(FALSE, TRUE),
+			labels = c("Without SCFA data", "With SCFA data"),
+			values = brewer.pal("PRGn", n=3)[-2]
+		) +
+		labs(x="Microbiome data included", y="AUCROC for held out data") +
+		theme_classic() +
+		theme(
+			strip.background = element_blank(),
+			panel.spacing.x=unit(1, "lines"),
+			axis.ticks.x = element_blank(),
+			strip.text = element_text(size=10),
+			legend.position=c(0.15, 0.8)
+		)
+
+#acetate, propionate, isobutyrate, butyrate
+#Genus, OTU, KEGG, OPF, PC2Pathways, PC2KO

### 2019-03-27 code/rf_pool_optimum_mtry.R

@@ -2,7 +2,7 @@ library(tidyverse)
 
 read_optimum_mtry_file <- function(file) {
 
-	read_csv(file, col_types=cols(col_double(), col_double())) %>%
+	read_csv(file, col_types=cols(.default=col_double())) %>%
 		mutate(file=str_replace(file, "//", "/"),
 					iteration=as.numeric(str_replace(file, ".*optimum.mtry.(\\d*).csv", "\\1")),
 					condition=str_replace(file, "data/rf/(.*)/optimum.*", "\\1")

### 2019-03-27 code/rf_pool_pools.R

@@ -5,7 +5,7 @@ options(warn = 1)
 cv_test_compare_files <- function(file) {
 	print(file)
 
-	read_tsv(file, col_types=cols(col_character(), col_double(), col_double(), col_double())) %>%
+	read_tsv(file, col_types=cols(condition=col_character(), .default=col_double())) %>%
 		mutate(condition=str_replace(condition, "_", ".")) %>%
 		separate(condition, into=c("class", "input"), sep="\\.")
 

### 2019-03-27 code/plot_classification_regression.R
Generate figure depicting modelling performance

@@ -9,9 +9,14 @@ capitalize <- function(string) {
 }
 
 
-data_file_name <- "data/rf/classification_data_pool.tsv"
+classification_file_name <- "data/rf/classification_data_pool.tsv"
 
-model_data <- read_tsv(data_file_name) %>%
+classification_data <- read_tsv(classification_file_name,
+												col_types=cols(class=col_character(),
+																				input=col_character(),
+																				.default=col_double()
+																			)
+											) %>%
 	mutate(scfa = str_detect(input, "scfa"),
 				microbiome=str_replace(input, "scfa_", ""),
 				microbiome=str_replace(microbiome, "_scfa", "")
@@ -20,7 +25,7 @@ model_data <- read_tsv(data_file_name) %>%
 	filter(class != "lesion")
 
 
-classification_plot <- model_data %>%
+classification_plot <- classification_data %>%
 	filter(microbiome %in% c("scfa", "otu", "genus")) %>%
 	mutate(microbiome = factor(microbiome, levels=c("scfa", "genus", "otu"))) %>%
 	ggplot(aes(x=microbiome, fill=scfa, y=test_aucs)) +
@@ -35,18 +40,68 @@ classification_plot <- model_data %>%
 		scale_fill_manual(
 			name = NULL,
 			breaks = c(FALSE, TRUE),
-			labels = c("Without SCFA data", "With SCFA data"),
+			labels = c("Without SCFAs", "With SCFAs"),
 			values = brewer.pal("PRGn", n=3)[-2]
 		) +
-		labs(x="Microbiome data included", y="AUCROC for held out data") +
+		labs(x="Microbiome data included", y="AUCROC for\ndiagnosis classification") +
 		theme_classic() +
 		theme(
 			strip.background = element_blank(),
 			panel.spacing.x=unit(1, "lines"),
 			axis.ticks.x = element_blank(),
-			strip.text = element_text(size=10),
-			legend.position=c(0.15, 0.8)
+			strip.text = element_text(size=10)
 		)
 
-#acetate, propionate, isobutyrate, butyrate
-#Genus, OTU, KEGG, OPF, PC2Pathways, PC2KO
+
+regression_file_name <- "data/rf/regression_data_pool.tsv"
+
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+
[…]

### 2019-03-27 code/plot_classification_fit.R
Build supplementary figures

@@ -1,0 +1,57 @@
+library(tidyverse)
+library(RColorBrewer)
+
+### Helper function
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+classification_file_name <- "data/rf/classification_data_pool.tsv"
+
+classification_data <- read_tsv(classification_file_name,
+												col_types=cols(class=col_character(),
+																				input=col_character(),
+																				.default=col_double()
+																			)
+											) %>%
+	mutate(scfa = str_detect(input, "scfa"),
+				microbiome=str_replace(input, "scfa_", ""),
+				microbiome=str_replace(microbiome, "_scfa", "")
+		) %>%
+	select(class, microbiome, scfa, cv_aucs, test_aucs) %>%
+	filter(class != "lesion") %>%
+	filter(microbiome %in% c("scfa", "otu", "genus")) %>%
+	mutate(microbiome = factor(microbiome, levels=c("scfa", "genus", "otu"))) %>%
+	mutate(scfa = ifelse(scfa, "With SCFAs", "Without SCFAs")) %>%
+	gather(test, auc, cv_aucs, test_aucs)
+
+
+
+
+
+classification_data %>%
+	ggplot(aes(x=microbiome, fill=test, y=auc)) +
+		geom_hline(yintercept=0.5, col="#888888") +
+		geom_boxplot(position=position_dodge2(preserve = "single"))+
+		facet_grid(scfa~class, labeller=labeller(.default=capitalize)) +
+		coord_cartesian(ylim=c(0.35,1.0)) +
+		scale_x_discrete(
+			breaks = c("scfa", "genus", "otu"),
+			labels = c("None", "Genus", "OTU")
+		) +
+		scale_fill_manual(
+			name = NULL,
+			breaks = c("cv_aucs", "test_aucs"),
+			labels = c("Cross validation", "Held out data"),
+			values = brewer.pal("RdBu", n=3)[-2]
+		) +
+		labs(x="Microbiome data included", y="AUCROC for diagnosis classification") +
+		theme_classic() +
+		theme(
+			strip.background = element_blank(),
+			panel.spacing.x=unit(1, "lines"),
+			axis.ticks.x = element_blank(),
+			strip.text = element_text(size=10)
+		) + ggsave("results/figures/classification_testing.pdf", width=6.875, height=5)

### 2019-03-27 code/plot_regression_fit.R

@@ -1,0 +1,62 @@
+library(tidyverse)
+library(RColorBrewer)
+
+### Helper function
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+
+
+regression_file_name <- "data/rf/regression_data_pool.tsv"
+
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+
+regression_data <- read_tsv(regression_file_name,
+					col_types=cols(class=col_character(),
+													input=col_character(),
+													.default=col_double()
+												)
+				) %>%
+				mutate(microbiome=str_replace(input, "fit_", ""),
+							microbiome=str_replace(microbiome, "_fit", "")
+		) %>%
+	select(class, microbiome, train_Rsquared, test_Rsquared) %>%
+	filter(class %in% scfa_types) %>%
+	filter(microbiome %in% regression_features) %>%
+	mutate(class = factor(class, levels=scfa_types),
+					microbiome = factor(microbiome, levels=regression_features),
+					data_type = case_when(
+			       microbiome %in% c("genus", "otu")  ~ "16S rRNA gene",
+						 microbiome %in% c("kegg", "opf")  ~ "Metagenomics",
+						 microbiome %in% c("pc2ec", "pc2pathways")  ~ "PICRUSt"
+					 )
+				) %>%
+	gather(test, auc, train_Rsquared, test_Rsquared) %>%
+	mutate(test=factor(test, levels=c("train_Rsquared", "test_Rsquared")))
+
+regression_data %>%
+	ggplot(aes(x=microbiome, y=auc, fill=test)) +
+	geom_boxplot() +
+	facet_grid(class~data_type, labeller=labeller(.default=capitalize), scale="free_x") +
+	scale_x_discrete(
+		breaks = c("genus", "otu", "kegg", "opf", "pc2ec", "pc2pathways"),
+		labels = c("Genus", "OTU", "KEGG", "OPF", "KEGG", "Paths")
+	) +
+	scale_fill_manual(
+		name = NULL,
+		breaks = c("train_Rsquared", "test_Rsquared"),
+		labels = c("Cross validation", "Held out data"),
+		values = brewer.pal("RdBu", n=3)[-2]
+	) +
[…]

### 2019-03-28 submission/manuscript.Rmd
Add regression paragraph code

@@
 	min()
+```
 
-min_p_compare_metag <- w_wo_compare %>%
-	filter(class=="adenoma" | class == "cancer") %>%
-	filter(microbiome =="opf" | microbiome == "kegg") %>%
-	select(p.value) %>%
-	min()
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
+We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1).
+Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
+When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
+These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.
+
+```{r, regression}
+
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+
+regression <- read_tsv("../data/rf/regression_data_pool.tsv",
+												col_types=cols(class=col_character(),
+													input=col_character(),
+													.default=col_double()
+												)
+ 											) %>%
+							mutate(microbiome=str_replace(input, "fit_", ""),
+										microbiome=str_replace(microbiome, "_fit", "")
+								) %>%
+							select(class, microbiome, test_Rsquared) %>%
+							filter(class %in% scfa_types) %>%
+							filter(microbiome %in% regression_features) %>%
+							mutate(data_type = case_when(
+					       microbiome %in% c("genus", "otu")  ~ "16S",
+								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
+								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
+							 )) %>%
+							group_by(class, microbiome, data_type) %>%
+							summarize(median_Rsquared = median(test_Rsquared))
+
+top_16S <- regression %>% filter(data_type == "16S") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)
+
+top_metagenomics <- regression %>% filter(data_type == "metagenomics") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)
+
+top_picrust <- regression %>% filter(data_type == "picrust") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)
+
+convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ec = "KEGG", pc2pathways = "pathway")
 
@@
 
-**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
-We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model.
-Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
-When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
-We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
-We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories.
-Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A).
-These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.
-
-```{r, regression}
-```
-
-**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations.
-We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs.
-Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B).
-Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs.
-Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
-Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using PICRUSt.
-Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B).
-The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.
+**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis.
+We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2).
+The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared` observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B).
+We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
+We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic to predict the concentration of the SCFAs (Figure S2).
+Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
+Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2).
+SCFA concentrations could not be predicted based on the imputed metagenomic data.
+The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared` observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
+The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.
 
 **Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions.
-Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of individual or pooled SCFA concentrations.
+Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive SCFA concentrations.
 These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors.
@@
 
-\includegraphics{../results/figures/scfa_abundance.pdf}
+\includegraphics{../results/figures/scfa_comparisons.pdf}
 
@@
 
+\includegraphics{../results/figures/scfa_modelling.pdf}
+
 **Figure 2.
@@
 
+\includegraphics{../results/figures/classification_testing.pdf}
+
 **Figure S1. Comparison of training and testing results.**
+
+\newpage
+
+\includegraphics{../results/figures/regression_testing.pdf}
+
+**Figure S1. Comparison of training and testing results.**

### 2019-03-28 submission/manuscript.tex
Add regression paragraph code

@@
 Forest machine learning models (8, 16). We repeated that analysis but
-added the concentration of the individual or pooled SCFAs as possible
-features to train a model. Models trained using SCFAs to classify
-individuals as having adenomas or carcinomas rather than normal colons
-had median areas under the receiver operator characteristic curve
-(AUROC) that were significantly greater than 0.5
-(P\textsubscript{adenoma}\textless{}0.001 and
-P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
-to detect the presence of adenomas or carcinomas were only 0.54 and
-0.55, respectively (Figure 2A). When we trained the models with the
+added the concentration of the SCFAs as possible features to train the
+models (Figure S1). Models trained using SCFAs to classify individuals
+as having adenomas or carcinomas rather than normal colons had median
+areas under the receiver operator characteristic curve (AUROC) that were
+significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
+and P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC
+values to detect the presence of adenomas or carcinomas were only 0.54
+and 0.55, respectively (Figure 2A). When we trained the models with the
 SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
@@
 from the models trained without the SCFA concentrations
-(P\textgreater{}0.21; Figure 2A). We also trained models using a smaller
-dataset that generated shotgun metagenomic sequencing data from a subset
-of our cohort (N\textsubscript{normal}=27, N\textsubscript{adenoma}=25,
-and N\textsubscript{cancer}=26) (18). We binned genes extracted from the
-assembled metagenomes into operational protein families (OPFs) or KEGG
-categories. Again, the performance of the models trained with the
-meteagenomic data did not improve when the SCFA concentrations were
-added as possible features when training the model (P\textgreater{}0.24;
-Figure 2A). These data demonstrate that knowledge of the SCFA profile
-from a patient's fecal sample does not improve the ability to diagnose a
-colonic lesion.
+(P\textgreater{}0.21; Figure 2A). These data demonstrate that knowledge
+of the SCFA profile from a patient's fecal sample does not improve the
+ability to diagnose a colonic lesion.
 
 \textbf{Knowledge of microbial community structure does not predict SCFA
-concentrations.} Regardless of a person's diagnosis, we next asked
-whether the fecal community structure was predictive of fecal SCFA
-concentrations. We trained Random Forest regression models using 16S
-rRNA gene sequence data binned into OTUs and genera to predict the
-concentration of the individual and pooled SCFAs. Regardless of the
-binning method or SCFA, the largest amount of variation that the model
-could explain was \textbf{0.XXXX} (Figure 2B). Next, we trained Random
-Forest regression models using metagenomic sequence data binned into
-OPFs and KEGG categories to predict the concentration of the individual
-and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence
-data, the metagenomic data was not predictive of SCFA concentration; the
-largest amount of variation that the models could explain was
-\textbf{0.XXXX} (Figure 2B). Because of the limited number of samples
-that we were able to generate metagenomic sequence data from, we used
-our 16S rRNA gene sequence data to impute metagenomes that were binned
-into metabolic pathways, enzyme commission numbers, or KEGG categories
-using PICRUSt. Again, SCFA concentrations could not be predicted based
-on the imputed metagenomic data; the largest amount of variation that
-the models could explain was \textbf{0.XXXX} (Figure 2B). The inability
-to model SCFA concentrations from microbiome data indicates that the
-knowledge of the abundance of organisms and genes is insufficient to
-predict SCFA concentrations.
+concentrations.} We next asked whether the fecal community structure was
+predictive of fecal SCFA concentrations, regardless of a person's
+diagnosis. We trained Random Forest regression models using 16S rRNA
+gene sequence data binned into OTUs and genera to predict the
+concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
+between the observed SCFA concentrations and the modeled concentrations
+was 0.14 observed when using genus data to predict butyrate
+concentrations (Figure 2B). We also used a smaller dataset of shotgun
+metagenomic sequencing data generated from a subset of our cohort
+(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
+N\textsubscript{cancer}=26) (18). We binned genes extracted from the
+assembled metagenomes into operational protein families (OPFs) or KEGG
+categories and trained Random Forest regression models using metagenomic
+to predict the concentration of the SCFAs (Figure S2). Similar to the
+analysis using 16S rRNA gene sequence data, the metagenomic data was not
+predictive of SCFA concentration; the largest amount of variation that
+the models could explain was 0.055, which was observed when using KEGG
+data to predict propionate concentrations (Figure 2B). Because of the
+limited number of samples that we were able to generate metagenomic
+sequence data from, we used our 16S rRNA gene sequence data to impute
+metagenomes that were binned into metabolic pathways or KEGG categories
+using PICRUSt (Figure S2). SCFA concentrations could not be predicted
+based on the imputed metagenomic data. The largest amount of variation
+that the models could explain was 0.097 observed when using KEGG data to
+predict butyrate concentrations (Figure 2B). The inability to model SCFA
+concentrations from microbiome data indicates that the knowledge of the
+abundance of organisms and their genes was insufficient to predict SCFA
+concentrations.
 
@@
 someone with one of these lesions. Furthermore, knowledge of the
-taxonomic and genetic structure of gut microbiota was not predictive of
-individual or pooled SCFA concentrations. These results complement
-existing literature that suggest that fiber consumption and the
-production of SCFAs are unable to prevent the risk of developing colonic
-tumors. It is important to note that our analysis concerned fecal SCFA
-concentrations and microbiome characterization and that observations
-along the mucosa near the site of lesions may provide a stronger
-association. Regardless, given the growing literature in this area, it
-is unlikely that SCFAs are the primary mechanism that limits
-tumorigenesis. This may be a cautionary result to temper enthusiasm for
-SCFAs as a biomarker of gut health more generally. Going forward it is
-critical to develop additional hypotheses for how the microbiome and
-host interact to drive tumorigenesis to better understand the disease
-process and identify biomarkers that will allow early detection of
-tumors.
+taxonomic and genetic structure of gut microbiota was not predictive
+SCFA concentrations. These results complement existing literature that
+suggest that fiber consumption and the production of SCFAs are unable to
+prevent the risk of developing colonic tumors. It is important to note
+that our analysis concerned fecal SCFA concentrations and microbiome
+characterization and that observations along the mucosa near the site of
+lesions may provide a stronger association. Regardless, given the
+growing literature in this area, it is unlikely that SCFAs are the
+primary mechanism that limits tumorigenesis. This may be a cautionary
+result to temper enthusiasm for SCFAs as a biomarker of gut health more
+generally. Going forward it is critical to develop additional hypotheses
+for how the microbiome and host interact to drive tumorigenesis to
+better understand the disease process and identify biomarkers that will
+allow early detection of tumors.
 
@@
 
+\includegraphics{../results/figures/scfa_modelling.pdf}
+
 \textbf{Figure 2. SCFA concentrations do not improve models for
@@
 
+\includegraphics{../results/figures/classification_testing.pdf}
+
+\textbf{Figure S1. Comparison of training and testing results.}
+
+\newpage
+
+\includegraphics{../results/figures/regression_testing.pdf}
+
 \textbf{Figure S1. Comparison of training and testing results.}

### 2019-03-28 submission/manuscript.Rmd
Incorporate figure legends

@@
 							group_by(class, microbiome, data_type) %>%
-							summarize(median_Rsquared = median(test_Rsquared))
+							summarize(median_Rsquared = median(test_Rsquared)) %>%
+							ungroup()
 
@@
 
-**Random Forest models.**
+**Random Forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations.
+The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information.
+For classification and regression models, we pre-processed the features by scaling them to vary between zero and one.
+Features with no variance in the training set were removed from both the training and testing sets.
+We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data.
+The training set was used for hyperparameter selection and training the model, and the test set was used for evaluating prediction performance.
+For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations.
+Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected.
+We trained the random forest model using the selected mtry value and predicted the held-out test set.
+The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance.
+We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively.
+We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.
 
@@
 
-\includegraphics{../results/figures/scfa_comparisons.pdf}
+\includegraphics{figure_1.png}
 
 **Figure 1.
-SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) was for isobutyrate.
+SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`).
 (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.
@@
 
-\includegraphics{../results/figures/scfa_modelling.pdf}
+\includegraphics{figure_2.png}
 
 **Figure 2.
-SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions.
-16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**
+SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did improve performance of the models generated using 16S rRNA gene sequence data.
+(B) Regression models that were trained using 16S rRNA gene sequence, metagenomic, and PICRUSt data to predict the concentrations of SCFAs performed poorly (all median R^2^ values < `r max(regression$median_Rsquared)`).
+Regression models generated using 16S rRNA gene sequence and PICRUSt data included data from `r sum(n$n)` samples and those generated using metagenomic data included data from `r sum(n_s$n)` samples.
 
@@
 
-\includegraphics{../results/figures/classification_testing.pdf}
+\includegraphics{figure_s1.png}
 
-**Figure S1. Comparison of training and testing results.**
+**Figure S1.
+Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between subjects with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features.
+Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples.
+The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry.
+The mtry with the largest AUROC was selected to train the model, which was then tested on the held out 20% of the samples.
 
@@
 
-\includegraphics{../results/figures/regression_testing.pdf}
+\includegraphics{figure_s2.png}
 
-**Figure S1. Comparison of training and testing results.**
+**Figure S2.
+Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each subjects' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data.
+These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples.
+The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry.
+The mtry with the largest R^2^ was selected to train the model, which was then tested on the held out 20% of the samples.

### 2019-03-28 submission/manuscript.tex
Incorporate figure legends

@@
 
-\textbf{Random Forest models.}
+\textbf{Random Forest models.} The classification models were built to
+predict lesion type from microbiome information with or without SCFA
+concentrations. The regression models were built to predict the SCFA
+concentrations of acetate, butyrate, and propionate from microbiome
+information. For classification and regression models, we pre-processed
+the features by scaling them to vary between zero and one. Features with
+no variance in the training set were removed from both the training and
+testing sets. We randomly split the data into training and test sets so
+that the training set consisted of 80\% of the full dataset while the
+test set was composed of the remaining data. The training set was used
+for hyperparameter selection and training the model, and the test set
+was used for evaluating prediction performance. For each model, the best
+performing hyperparameter, mtry, was selected in an internal five-fold
+cross-validation of the training set with 100 randomizations. Six values
+of mtry were tested and the value that provided the largest AUROC or
+R\textsuperscript{2} was selected. We trained the random forest model
+using the selected mtry value and predicted the held-out test set. The
+data-split, hyperparameter selection, training and testing steps were
+repeated 100 times to get a reliable and robust reading of model
+prediction performance. We used AUROC and R\textsuperscript{2} as the
+prediction performance metric for classification and regression models,
+respectively. We used randomForest package implemented to the caret
+package (version 4.6-14) in R statistical software (version 6.0-81) for
+our models.
 
@@
 
-\includegraphics{../results/figures/scfa_comparisons.pdf}
+\includegraphics{figure_1.png}
 
@@
 individuals with normal colons (N=172) or those with adenoma (N=198) or
-carcinomas (N=120) was for isobutyrate. (B) A subset of individuals
-diagnosed with adenomas (N=41) or carcinomas (N=26) who underwent
-treatment were resampled a year after the initial sampling; one extreme
-propionate value (124.4 mmol/kg) was included in the adenoma analysis
-but censored from the visualization for clarity.
+carcinomas (N=120). (B) A subset of individuals diagnosed with adenomas
+(N=41) or carcinomas (N=26) who underwent treatment were resampled a
+year after the initial sampling; one extreme propionate value (124.4
+mmol/kg) was included in the adenoma analysis but censored from the
+visualization for clarity.
 
@@
 
-\includegraphics{../results/figures/scfa_modelling.pdf}
+\includegraphics{figure_2.png}
 
 \textbf{Figure 2. SCFA concentrations do not improve models for
-diagnosing the presence of adenomas, carcinomas, or all lesions. 16S
-rRNA gene and metagenomic sequence data do not predict SCFAs
-concentrations.}
+diagnosing the presence of adenomas, carcinomas, or all lesions and
+cannot be reliably predicted from 16S rRNA gene or metagenomic sequence
+data.} (A) The median AUROC for diagnosing individuals as having
+adenomas or carcinomas using SCFAs was slightly better than than chance
+(depicted by horizontal line at 0.50), but did improve performance of
+the models generated using 16S rRNA gene sequence data. (B) Regression
+models that were trained using 16S rRNA gene sequence, metagenomic, and
+PICRUSt data to predict the concentrations of SCFAs performed poorly
+(all median R\textsuperscript{2} values \textless{} 0.14). Regression
+models generated using 16S rRNA gene sequence and PICRUSt data included
+data from 490 samples and those generated using metagenomic data
+included data from 78 samples.
 
@@
 
-\includegraphics{../results/figures/classification_testing.pdf}
+\includegraphics{figure_s1.png}
 
-\textbf{Figure S1. Comparison of training and testing results.}
+\textbf{Figure S1. Comparison of training and testing results for
+classification models shows that the models are robust and are not
+overfit.} Random Forest classification models were generated to
+differentiate between subjects with normal colons and those with
+adenomas or carcinomas using 16S rRNA gene sequence data that were
+clustered into genera or OTUs with and without including the four SCFAs
+as additional features. Random Forest classification models were
+generated by partitioning the samples into a training set with 80\% of
+the data and a testing set with the remaining samples. The training data
+was used to fit the mtry hyperparameter using 100 five-fold cross
+validations for six values of mtry. The mtry with the largest AUROC was
+selected to train the model, which was then tested on the held out 20\%
+of the samples.
 
@@
 
-\includegraphics{../results/figures/regression_testing.pdf}
+\includegraphics{figure_s2.png}
 
-\textbf{Figure S1. Comparison of training and testing results.}
+\textbf{Figure S2. Comparison of training and testing results for
+regression models shows that the models are robust and are not overfit.}
+Random Forest regression models were generated to predict the
+concentration of each SCFA using each subjects' microbiome data
+generated using 16S rRNA gene sequence and metagenomic sequence data.
+These regression models were generated by partitioning the samples into
+a training set with 80\% of the data and a testing set with the
+remaining samples. The training data was used to fit the mtry
+hyperparameter using 100 five-fold cross validations for six values of
+mtry. The mtry with the largest R\textsuperscript{2} was selected to
+train the model, which was then tested on the held out 20\% of the
+samples.
 

### 2019-03-28 code/plot_classification_regression.R

@@ -104,4 +104,4 @@ regression_features_plot <- regression_data %>%
 
 plot_grid(classification_plot, regression_features_plot, nrow=2, labels=c("A", "B"), rel_heights=c(0.5, 1))
 
-ggsave("results/figures/scfa_modelling.pdf", width=6.875, height=6, units="in")
+ggsave("results/figures/scfa_modeling.pdf", width=6.875, height=6, units="in")

### 2019-03-29 submission/manuscript.Rmd
Edit manuscript before sending to co-authors

@@
 Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria.
-We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing.
+We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously individualed to 16S rRNA gene and metagenomic shotgun sequencing.
 We measured the fecal concentrations of acetate, propionate, isobutyrate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status.
@@
 Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples.
-These models performed poorly and were only able to explain less than 10% of the observed variation in the SCFA concentrations.
+These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations.
 These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer.
@@
 Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations.
-These data indicate that there is not a meaningful link between the gut microbiome, SCFAs, and tumor burden.
+These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden.
 
@@
 
-SCFAs have have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
-Manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018].
+SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
+Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burdenin mouse models of colorectal cancer [@Bishehsari2018].
 These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
+Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
 Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
-Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
-Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013].
-Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
+Case-control studies that have investigated a possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988].
+Based on these limited results, one would hypothesize increasing SCFA concentrations could limit tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
 These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.
@@
 cross_test <- read_tsv('../data/process/scfa_cross_section_stats.tsv')
-iso_p_cross <- cross_test %>% filter(p.value < 0.05)
+iso_p_cross <- cross_test %>% filter(!str_detect(scfa, "rel"), p.value < 0.05)
 stopifnot(nrow(iso_p_cross) == 1)
@@
 
-pooled_test <- min(c(cross_test %>% filter(scfa=="pooled") %>% pull(p.value), follow_test %>% filter(scfa=="pooled") %>% pull(p.value)))
+rel_test <- min(c(cross_test %>% filter(str_detect(scfa, "rel"), !str_detect(scfa, "iso")) %>% pull(p.value), follow_test %>% filter(str_detect(scfa, "rel"), !str_detect(scfa, "iso")) %>% pull(p.value)))
+
+pooled_test <- min(c(cross_test %>% filter(scfa=="pooled" | scfa == "total") %>% pull(p.value), follow_test %>% filter(scfa=="pooled" | scfa == "total") %>% pull(p.value)))
 
@@
 The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A).
-Interestingly, the median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas, respectively.
+Interestingly, the median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg of feces in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg of feces in people with colonic adenomas or carcinomas, respectively.
 The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`).
-Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
-The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg (P=`r iso_p_follow$p.value`; Figure 1B).
-For both the pre-treatment cross-sectional data and the pre/post treatment data, we pooled the SCFA concentrations on a per molecule of carbon basis and again failed to see any significant differences (P>`r pooled_test`).
-The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to this observation.
+Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
+The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg of feces (P=`r iso_p_follow$p.value`; Figure 1B).
+For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs other than isobutyrate (`r rel_test`).
+Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`).
+The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to the isobutyrate observations.
 
@@
 
-**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
 We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1).
 Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
-When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
-These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.
+When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
+These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample did not improve the ability to diagnose a colonic lesion.
 
@@
 We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2).
-The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared` observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B).
+The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B).
 We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
@@
 SCFA concentrations could not be predicted based on the imputed metagenomic data.
-The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared` observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
+The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
 The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.
@@
 This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally.
-Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis to better understand the disease process and identify biomarkers that will allow early detection of tumors.
+Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis so that we can better understand tumorigenesis and identify biomarkers that will allow early detection of lesions.
+
+## Acknowledgements
+
+The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+We would thank the University of Michigan Center for Microbial Systems for enabling our short-chain fatty acid analysis.
+Support for MAS came from the Canadian Institute of Health Research and the National Institutes of Health (UL1TR002240).
+This work was also supported by the National Institutes of Health (P30DK034933 and R01CA215574).
 
@@
 The original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer.
-The resulting sequence data were assembled into contigs, screened to remove low quality contigs and chimeras.
+The resulting sequence data were assembled into contigs and screened to remove low quality contigs and chimeras.
 The curated sequences were then clustered into OTUs at a 97% similarity threshold and assigned to the closest possible genus with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16).
-We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].
+We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways and KEGG categories [@Langille2013].
 
@@
 [@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915).
-Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads.
+Fecal DNA was individualed to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads.
 The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data.
@@
 We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data.
-The training set was used for hyperparameter selection and training the model, and the test set was used for evaluating prediction performance.
+The training set was used for hyperparameter selection and training the model and the test set was used for evaluating prediction performance.
 For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations.
@@
 
-## Acknowledgements
-
-The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-We would thank the University of Michigan Center for Microbial Systems for enabling our short-chain fatty acid analysis.
-Support for MAS came from the Canadian Institute of Health Research and the National Institutes of Health (UL1TR002240).
-Support for PDS came from the National Institutes of Health (P30DK034933 and R01CA215574).
-
-\newpage
-
 ## References
@@
 **Figure 1.
-SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`).
+SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) The concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`).
 (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.
@@
 **Figure S1.
-Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between subjects with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features.
-Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples.
-The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry.
-The mtry with the largest AUROC was selected to train the model, which was then tested on the held out 20% of the samples.
+Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features.
+Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.
 
@@
 **Figure S2.
-Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each subjects' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data.
-These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples.
-The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry.
-The mtry with the largest R^2^ was selected to train the model, which was then tested on the held out 20% of the samples.
+Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each individuals' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data.
+These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

### 2019-03-29 submission/manuscript.tex
Edit manuscript before sending to co-authors

@@
 between fecal SCFA concentrations and the presence of colonic adenomas
-or carcinomas in a cohort of individuals that was previously subjected
-to 16S rRNA gene and metagenomic shotgun sequencing. We measured the
-fecal concentrations of acetate, propionate, isobutyrate, and butyrate
-within the cohort and found that there were no meaningful differences in
-their concentration and disease status. When we incorporated these
-concentrations into Random Forest classification models trained to
-differentiate between people with normal colons and those with adenomas
-or carcinomas, we found that they did not significantly improve the
-ability of 16S rRNA gene or metagenomic gene sequence-based models to
-classify individuals. Finally, we generated Random Forest regression
-models trained to predict the concentration of each SCFA based on 16S
-rRNA gene or metagenomic gene sequence data from the same samples. These
-models performed poorly and were only able to explain less than 10\% of
-the observed variation in the SCFA concentrations. These results support
-the broader epidemiological data that questions the value of fiber
-consumption for reducing the risks of colorectal cancer. Although it is
-likely that bacterial metabolites may serve as biomarkers to detect
-adenomas or carcinomas, fecal SCFA concentrations have limited value.
+or carcinomas in a cohort of individuals that was previously
+individualed to 16S rRNA gene and metagenomic shotgun sequencing. We
+measured the fecal concentrations of acetate, propionate, isobutyrate,
+and butyrate within the cohort and found that there were no meaningful
+differences in their concentration and disease status. When we
+incorporated these concentrations into Random Forest classification
+models trained to differentiate between people with normal colons and
+those with adenomas or carcinomas, we found that they did not
+significantly improve the ability of 16S rRNA gene or metagenomic gene
+sequence-based models to classify individuals. Finally, we generated
+Random Forest regression models trained to predict the concentration of
+each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
+the same samples. These models performed poorly and were able to explain
+at most 14\% of the observed variation in the SCFA concentrations. These
+results support the broader epidemiological data that questions the
+value of fiber consumption for reducing the risks of colorectal cancer.
+Although it is likely that bacterial metabolites may serve as biomarkers
+to detect adenomas or carcinomas, fecal SCFA concentrations have limited
+value.
 
@@
 the fecal community structure and SCFA concentrations. These data
-indicate that there is not a meaningful link between the gut microbiome,
+indicate that there is no conclusive link between the gut microbiome,
 SCFAs, and tumor burden.
@@
 
-SCFAs have have anti-inflammatory and anti-proliferative activities (9).
-Manipulation of SCFAs in mouse models of colorectal cancer by direct
-supplementation or feeding of fiber caused an overall reduction in tumor
-burden (10). These results suggest that supplementation with substrates
-that bacteria can ferment to produce SCFAs may confer beneficial effects
-against colorectal cancer. Regardless, there is a lack of evidence that
-increasing SCFA concentrations can protect against colorectal cancer in
-humans. Based on similar observations, many microbiome studies use the
-concentrations of SCFAs and the presence of 16S rRNA gene sequences from
-organisms and the genes involved in producing them as a biomarker of a
-healthy microbiota (11, 12). Case-control studies that have investigated
-SCFA concentrations in colorectal cancer found that patients with
-carcinomas had lower concentrations of SCFAs versus patients with
-adenomas or individuals without colon tumors (13). Although this would
-argue that increasing SCFA concentrations could be protective against
-tumorigenesis, in randomized controlled trials fiber supplementation has
-been inconsistently associated with protection against tumor formation
-and recurrence (14, 15). These findings temper enthusiasm for treatments
-that aim to use SCFAs as biomarkers or protection against tumorigenesis.
+SCFAs have anti-inflammatory and anti-proliferative activities (9).
+Direct supplementation of SCFAs or feeding of fiber caused an overall
+reduction in tumor burdenin mouse models of colorectal cancer (10).
+These results suggest that supplementation with substrates that bacteria
+can ferment to produce SCFAs may confer beneficial effects against
+colorectal cancer. Based on similar observations, many microbiome
+studies use the concentrations of SCFAs and the presence of 16S rRNA
+gene sequences from organisms and the genes involved in producing them
+as a biomarker of a healthy microbiota (11, 12). Regardless, there is a
+lack of evidence that increasing SCFA concentrations can protect against
+colorectal cancer in humans. Case-control studies that have investigated
+a possible associations between SCFAs and colon tumor status have been
+plagued by relatively small numbers of subjects, but have reported
+increased total and relative fecal acetate levels and decreased relative
+fecal butyrate concentrations in subjects with colonic lesions (13).
+Based on these limited results, one would hypothesize increasing SCFA
+concentrations could limit tumorigenesis, in randomized controlled
+trials fiber supplementation has been inconsistently associated with
+protection against tumor formation and recurrence (14, 15). These
+findings temper enthusiasm for treatments that aim to use SCFAs as
+biomarkers or protection against tumorigenesis.
 
@@
 isobutyrate (P=0.0091; Figure 1A). Interestingly, the median
-concentration of isobutyrate was 3.30 mmol/kg in people with normal
-colons and it was 3.00 and 3.84 mmol/kg in people with colonic adenomas
-or carcinomas, respectively. The difference in isobutyrate concentration
-between people with adenomas and carcinomas was significantly different
-(P=0.0065); however, the differences between people with normal colons
-and those with adenomas or carcinomas was not significant (P=0.19 and
-P=0.11). Among the subjects with adenomas and carcinomas, a subset
-(N\textsubscript{adenoma}=41, N\textsubscript{carcinoma}=26) were
-treated and sampled a year later (17). The only SCFA that changed
-following treatment was isobutyrate, which decreased by 0.99 mmol/kg
-(P=0.002; Figure 1B). For both the pre-treatment cross-sectional data
-and the pre/post treatment data, we pooled the SCFA concentrations on a
-per molecule of carbon basis and again failed to see any significant
-differences (P\textgreater{}0.15). The low concentration of isobutyrate
-relative to the other SCFAs, inconsistent concentrations, and unexpected
-decrease in concentration with treatment makes it difficult to ascribe
-much biological relevance to this observation.
+concentration of isobutyrate was 3.30 mmol/kg of feces in people with
+normal colons and it was 3.00 and 3.84 mmol/kg of feces in people with
+colonic adenomas or carcinomas, respectively. The difference in
+isobutyrate concentration between people with adenomas and carcinomas
+was significantly different (P=0.0065); however, the differences between
+people with normal colons and those with adenomas or carcinomas was not
+significant (P=0.19 and P=0.11). Among the individuals with adenomas and
+carcinomas, a subset (N\textsubscript{adenoma}=41,
+N\textsubscript{carcinoma}=26) were treated and sampled a year later
+(17). The only SCFA that changed following treatment was isobutyrate,
+which decreased by 0.99 mmol/kg of feces (P=0.002; Figure 1B). For both
+the pre-treatment cross-sectional data and the pre/post treatment data,
+we also failed to detect any significant differences in the relative
+concentrations of any SCFAs other than isobutyrate (0.17). Finally, we
+pooled the SCFA concentrations on a total and per molecule of carbon
+basis and again failed to observe any significant differences
+(P\textgreater{}0.15). The low concentration of isobutyrate relative to
+the other SCFAs, inconsistent concentrations, and unexpected decrease in
+concentration with treatment makes it difficult to ascribe much
+biological relevance to the isobutyrate observations.
 
@@
 found that binning 16S rRNA gene sequence data into operational
-taxonomic units based on 97\% similarity or into genera enabled us to
-classify individuals as having adenomas or carcinomas using Random
+taxonomic units (OTUs) based on 97\% similarity or into genera enabled
+us to classify individuals as having adenomas or carcinomas using Random
 Forest machine learning models (8, 16). We repeated that analysis but
@@
 and 0.55, respectively (Figure 2A). When we trained the models with the
-SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
-relative abundances the AUROC values were not significantly different
-from the models trained without the SCFA concentrations
-(P\textgreater{}0.21; Figure 2A). These data demonstrate that knowledge
-of the SCFA profile from a patient's fecal sample does not improve the
-ability to diagnose a colonic lesion.
+SCFAs concentrations and OTU or genus-level relative abundances the
+AUROC values were not significantly different from the same models
+trained without the SCFA concentrations (P\textgreater{}0.21; Figure
+2A). These data demonstrate that knowledge of the SCFA profile from a
+patient's fecal sample did not improve the ability to diagnose a colonic
+lesion.
 
@@
 between the observed SCFA concentrations and the modeled concentrations
-was 0.14 observed when using genus data to predict butyrate
+was 0.14, which was observed when using genus data to predict butyrate
 concentrations (Figure 2B). We also used a smaller dataset of shotgun
@@
 based on the imputed metagenomic data. The largest amount of variation
-that the models could explain was 0.097 observed when using KEGG data to
-predict butyrate concentrations (Figure 2B). The inability to model SCFA
-concentrations from microbiome data indicates that the knowledge of the
-abundance of organisms and their genes was insufficient to predict SCFA
-concentrations.
+that the models could explain was 0.097, which was observed when using
+KEGG data to predict butyrate concentrations (Figure 2B). The inability
+to model SCFA concentrations from microbiome data indicates that the
+knowledge of the abundance of organisms and their genes was insufficient
+to predict SCFA concentrations.
 
@@
 generally. Going forward it is critical to develop additional hypotheses
-for how the microbiome and host interact to drive tumorigenesis to
-better understand the disease process and identify biomarkers that will
-allow early detection of tumors.
+for how the microbiome and host interact to drive tumorigenesis so that
+we can better understand tumorigenesis and identify biomarkers that will
+allow early detection of lesions.
+
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
+
+The authors thank the Great Lakes-New England Early Detection Research
+Network for providing the fecal samples that were used in this study. We
+would thank the University of Michigan Center for Microbial Systems for
+enabling our short-chain fatty acid analysis. Support for MAS came from
+the Canadian Institute of Health Research and the National Institutes of
+Health (UL1TR002240). This work was also supported by the National
+Institutes of Health (P30DK034933 and R01CA215574).
 
@@
 region of the 16S rRNA gene using paired 250 nt reads on an Illumina
-MiSeq sequencer. The resulting sequence data were assembled into
-contigs, screened to remove low quality contigs and chimeras. The
-curated sequences were then clustered into OTUs at a 97\% similarity
-threshold and assigned to the closest possible genus with an 80\%
-confidence threshold trained on the reference collection from the
-Ribosomal Database Project (v.16). We used PICRUSt (v.2.1.0-b) with the
+MiSeq sequencer. The resulting sequence data were assembled into contigs
+and screened to remove low quality contigs and chimeras. The curated
+sequences were then clustered into OTUs at a 97\% similarity threshold
+and assigned to the closest possible genus with an 80\% confidence
+threshold trained on the reference collection from the Ribosomal
+Database Project (v.16). We used PICRUSt (v.2.1.0-b) with the
 recommended standard operating protocol to generate imputed metagenomes
-based on the expected metabolic pathways, KEGG categories, and enzyme
-commission numbers (21).
+based on the expected metabolic pathways and KEGG categories (21).
 
@@
 were generated by Hannigan et al.~(18) and deposited into the Sequence
-Read Archive (study SRP108915). Fecal DNA was subjected to shotgun
+Read Archive (study SRP108915). Fecal DNA was individualed to shotgun
 sequencing on an Illumina HiSeq using 125 bp paired end reads. The
@@
 test set was composed of the remaining data. The training set was used
-for hyperparameter selection and training the model, and the test set
-was used for evaluating prediction performance. For each model, the best
+for hyperparameter selection and training the model and the test set was
+used for evaluating prediction performance. For each model, the best
 performing hyperparameter, mtry, was selected in an internal five-fold
@@
 available at \url{https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/}.
-
-\newpage
-
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
-
-The authors thank the Great Lakes-New England Early Detection Research
-Network for providing the fecal samples that were used in this study. We
-would thank the University of Michigan Center for Microbial Systems for
-enabling our short-chain fatty acid analysis. Support for MAS came from
-the Canadian Institute of Health Research and the National Institutes of
-Health (UL1TR002240). Support for PDS came from the National Institutes
-of Health (P30DK034933 and R01CA215574).
 
@@
 
-\leavevmode\hypertarget{ref-Ohigashi2013}{}%
-13. \textbf{Ohigashi S}, \textbf{Sudo K}, \textbf{Kobayashi D},
-\textbf{Takahashi O}, \textbf{Takahashi T}, \textbf{Asahara T},
-\textbf{Nomoto K}, \textbf{Onodera H}. 2013. Changes of the intestinal
-microbiota, short chain fatty acids, and fecal pH in patients with
-colorectal cancer. Digestive Diseases and Sciences
-\textbf{58}:1717--1726.
-doi:\href{https://doi.org/10.1007/s10620-012-2526-4}{10.1007/s10620-012-2526-4}.
+\leavevmode\hypertarget{ref-Weaver1988}{}%
+13. \textbf{Weaver GA}, \textbf{Krause JA}, \textbf{Miller TL},
+\textbf{Wolin MJ}. 1988. Short chain fatty acid distributions of enema
+samples from a sigmoidoscopy population: An association of high acetate
+and low butyrate ratios with adenomatous polyps and colon cancer. Gut
+\textbf{29}:1539--1543.
+doi:\href{https://doi.org/10.1136/gut.29.11.1539}{10.1136/gut.29.11.1539}.
 
@@
 diagnosis of colonic lesions or with treatment for adenomas or
-carcinomas.} (A) We measured the concentration of fecal SCFAs from
-individuals with normal colons (N=172) or those with adenoma (N=198) or
-carcinomas (N=120). (B) A subset of individuals diagnosed with adenomas
-(N=41) or carcinomas (N=26) who underwent treatment were resampled a
-year after the initial sampling; one extreme propionate value (124.4
-mmol/kg) was included in the adenoma analysis but censored from the
-visualization for clarity.
+carcinomas.} (A) The concentration of fecal SCFAs from individuals with
+normal colons (N=172) or those with adenoma (N=198) or carcinomas
+(N=120). (B) A subset of individuals diagnosed with adenomas (N=41) or
+carcinomas (N=26) who underwent treatment were resampled a year after
+the initial sampling; one extreme propionate value (124.4 mmol/kg) was
+included in the adenoma analysis but censored from the visualization for
+clarity.
 
@@
 overfit.} Random Forest classification models were generated to
-differentiate between subjects with normal colons and those with
+differentiate between individuals with normal colons and those with
 adenomas or carcinomas using 16S rRNA gene sequence data that were
@@
 generated by partitioning the samples into a training set with 80\% of
-the data and a testing set with the remaining samples. The training data
-was used to fit the mtry hyperparameter using 100 five-fold cross
-validations for six values of mtry. The mtry with the largest AUROC was
-selected to train the model, which was then tested on the held out 20\%
-of the samples.
+the data and a testing set with the remaining samples for 100
+randomizations.
 
@@
 Random Forest regression models were generated to predict the
-concentration of each SCFA using each subjects' microbiome data
+concentration of each SCFA using each individuals' microbiome data
 generated using 16S rRNA gene sequence and metagenomic sequence data.
@@
 a training set with 80\% of the data and a testing set with the
-remaining samples. The training data was used to fit the mtry
-hyperparameter using 100 five-fold cross validations for six values of
-mtry. The mtry with the largest R\textsuperscript{2} was selected to
-train the model, which was then tested on the held out 20\% of the
-samples.
+remaining samples for 100 randomizations.
 

### 2019-03-29 code/scfa_stat_analysis.R

@@ -1,7 +1,11 @@
 library(tidyverse)
 library(broom)
 
-scfa <- read_tsv("data/scfa/scfa_composite.tsv",  col_type = cols(study_id = col_character()))
+total_scfa <- read_tsv("data/scfa/scfa_composite.tsv",  col_type = cols(study_id = col_character())) %>% spread(scfa, mmol_kg) %>% mutate(total = acetate + butyrate + isobutyrate + propionate) %>% gather(scfa, mmol_kg, -study_id)
+
+rel_scfa <- total_scfa %>% spread(scfa, mmol_kg) %>% mutate(rel_acetate = acetate / total, rel_butyrate = butyrate/total, rel_isobutyrate = isobutyrate/total, rel_propionate = propionate / total) %>%  select(study_id, rel_acetate, rel_butyrate, rel_isobutyrate, rel_propionate) %>% gather(scfa, mmol_kg, -study_id)
+
+scfa <- rbind(total_scfa, rel_scfa)
 
 #cross section analysis
 get_mean_dx <- function(x){

### 2019-03-30 code/mothur.sh
Reduce number of processors to run earlier

@@ -35,7 +35,8 @@ done
 cut -d , -f 1,12 data/raw/16S/SRP0* | grep "SRR" | sed -E "s/(^SRR.*),([^_]*)_.*/\2\t\\1_1.fastq\t\\1_2.fastq/" > $RAWDIR/crc.files
 
 #Run mothur process
-mothur "#make.contigs(file=crc.files, inputdir=$RAWDIR, outputdir=$WORKDIR, processors=12);
+mothur "#set.seed(seed=19760620);
+	make.contigs(file=crc.files, inputdir=$RAWDIR, outputdir=$WORKDIR, processors=6);
 	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275);
 	unique.seqs(fasta=current);
 	count.seqs(name=current, group=current);

### 2019-04-03 submission/manuscript.Rmd
Remove isobutyrate since it's below the LOD

@@
 Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria.
-We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously individualed to 16S rRNA gene and metagenomic shotgun sequencing.
-We measured the fecal concentrations of acetate, propionate, isobutyrate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status.
+We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing.
+We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status.
 When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
@@
 cross_test <- read_tsv('../data/process/scfa_cross_section_stats.tsv')
-iso_p_cross <- cross_test %>% filter(!str_detect(scfa, "rel"), p.value < 0.05)
-stopifnot(nrow(iso_p_cross) == 1)
+p_cross_min <- cross_test %>% filter(scfa %in% c("acetate", "butyrate", "propionate")) %>% pull(p.value) %>% min
+stopifnot(p_cross_min > 0.05)
 
@@
 follow_test <- read_tsv('../data/process/scfa_pre_post_stats.tsv')
-iso_p_follow <- follow_test %>% filter(p.value < 0.05)
-stopifnot(nrow(iso_p_follow) == 1)
+p_follow_min <- follow_test %>% filter(scfa %in% c('acetate', 'propionate', 'butyrate')) %>% pull(p.value) %>% min
+stopifnot(p_follow_min > 0.05)
 
-rel_test <- min(c(cross_test %>% filter(str_detect(scfa, "rel"), !str_detect(scfa, "iso")) %>% pull(p.value), follow_test %>% filter(str_detect(scfa, "rel"), !str_detect(scfa, "iso")) %>% pull(p.value)))
+rel_test <- min(c(cross_test %>% filter(str_detect(scfa, "rel")) %>% pull(p.value), follow_test %>% filter(str_detect(scfa, "rel")) %>% pull(p.value)))
 
@@
 
-**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, isobutyrate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
-The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A).
-Interestingly, the median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg of feces in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg of feces in people with colonic adenomas or carcinomas, respectively.
-The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`).
+**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
+We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A).
 Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
-The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg of feces (P=`r iso_p_follow$p.value`; Figure 1B).
-For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs other than isobutyrate (`r rel_test`).
+None of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B).
+For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`).
 Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`).
-The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to the isobutyrate observations.
+These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.
 
@@
 regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
-scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+scfa_types <- c("acetate", "propionate", "butyrate")
 
@@
 These methodological changes did not affect the range of concentrations of these SCFAs between the two studies.
+The concentrations of isobutyrate were consistently at or below the limit of detection and were not included in our analysis.
 
@@
 [@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915).
-Fecal DNA was individualed to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads.
+Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads.
 The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data.

### 2019-04-03 code/plot_classification_regression.R

@@ -56,7 +56,7 @@ classification_plot <- classification_data %>%
 regression_file_name <- "data/rf/regression_data_pool.tsv"
 
 regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
-scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+scfa_types <- c("acetate", "propionate", "butyrate")
 
 regression_data <- read_tsv(regression_file_name,
 					col_types=cols(class=col_character(),

### 2019-04-03 code/plot_regression_fit.R

@@ -13,7 +13,7 @@ capitalize <- function(string) {
 regression_file_name <- "data/rf/regression_data_pool.tsv"
 
 regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
-scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")
+scfa_types <- c("acetate", "propionate", "butyrate")
 
 regression_data <- read_tsv(regression_file_name,
 					col_types=cols(class=col_character(),
@@ -59,4 +59,4 @@ regression_data %>%
 		panel.spacing.x=unit(1, "lines"),
 		axis.ticks.x = element_blank(),
 		strip.text = element_text(size=10)
-	) + ggsave("results/figures/regression_testing.pdf")
+	) + ggsave("results/figures/regression_testing.pdf", width=6.875, height=5)

### 2019-04-03 code/plot_scfa_comparisons.R

@@ -11,7 +11,8 @@ capitalize <- function(string) {
 
 ### Read in scfa data and metadata
 scfa_data <- read_tsv('data/scfa/scfa_composite.tsv') %>%
-	mutate(scfa=factor(scfa, levels=c("acetate", "propionate", "isobutyrate", "butyrate"))) %>%
+	filter(scfa != "isobutyrate") %>%
+	mutate(scfa=factor(scfa, levels=c("acetate", "propionate", "butyrate"))) %>%
 	filter(scfa != "pooled")
 
 cross_section_metadata <- read_csv("data/metadata/cross_section.csv",

### 2019-04-03 code/scfa_combine_plates.R

@@ -52,6 +52,5 @@ sapply(scfas,
 			summarize(mmol_kg = mean(mmol_kg)) %>%
 			ungroup() %>%
 			spread(scfa, mmol_kg, fill=0) %>%
-			mutate(pooled = 4*butyrate + 4*isobutyrate + 3*propionate + 2*acetate) %>%
 			gather(scfa, mmol_kg, -study_id) %>%
 			write_tsv("data/scfa/scfa_composite.tsv")

### 2019-04-03 code/scfa_stat_analysis.R

@@ -1,9 +1,11 @@
 library(tidyverse)
 library(broom)
 
-total_scfa <- read_tsv("data/scfa/scfa_composite.tsv",  col_type = cols(study_id = col_character())) %>% spread(scfa, mmol_kg) %>% mutate(total = acetate + butyrate + isobutyrate + propionate) %>% gather(scfa, mmol_kg, -study_id)
+#remove isobutyrate because it was below the limit of detection
+total_scfa <- read_tsv("data/scfa/scfa_composite.tsv",  col_type = cols(study_id = col_character())) %>% filter(scfa != "isobutyrate") %>% spread(scfa, mmol_kg) %>% mutate(pooled = acetate + 2*propionate + 3*butyrate,
+	total = acetate + butyrate + propionate) %>% gather(scfa, mmol_kg, -study_id)
 
-rel_scfa <- total_scfa %>% spread(scfa, mmol_kg) %>% mutate(rel_acetate = acetate / total, rel_butyrate = butyrate/total, rel_isobutyrate = isobutyrate/total, rel_propionate = propionate / total) %>%  select(study_id, rel_acetate, rel_butyrate, rel_isobutyrate, rel_propionate) %>% gather(scfa, mmol_kg, -study_id)
+rel_scfa <- total_scfa %>% spread(scfa, mmol_kg) %>% mutate(rel_acetate = acetate / total, rel_butyrate = butyrate/total, rel_propionate = propionate / total) %>%  select(study_id, rel_acetate, rel_butyrate, rel_propionate) %>% gather(scfa, mmol_kg, -study_id)
 
 scfa <- rbind(total_scfa, rel_scfa)
 

### 2019-04-03 submission/manuscript.Rmd
Add MAS comments

@@
 Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990].
-This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015].
-Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
-Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis.
+This leaves a significant role for environmental and behavioral factors and diet [@FlissIsakov2017; @Lee2015].
+Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
+Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to create mutations, induce inflammation, and accelerate tumorigenesis.
 Numerous studies in murine models have supported this model [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
@@
 SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
-Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burdenin mouse models of colorectal cancer [@Bishehsari2018].
+Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018].
 These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
@@
 Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
-Case-control studies that have investigated a possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988].
+Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988].
 Based on these limited results, one would hypothesize increasing SCFA concentrations could limit tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
@@
 We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
-We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic to predict the concentration of the SCFAs (Figure S2).
+We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2).
 Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
@@
 **Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions.
-Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive SCFA concentrations.
+Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of SCFA concentrations.
 These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors.
@@
 [@Baxter2016] and Sze et al.
-[@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009].
+[@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using mothur v.1.42 [@Schloss2009].
 The original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer.

### 2019-04-03 submission/manuscript.Rmd
Add NAL comments

@@
 
-The gut microbiome is thought to have a role in the development of colorectal cancer by protecting against and exacerbating inflammation.
-Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria.
-We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing.
-We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status.
+Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation.
+Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber.
+We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data.
+We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful association between SCFA concentration and disease status.
 When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
@@
 These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer.
-Although it is likely that bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited value.
+Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.
 
@@
 Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
-Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to create mutations, induce inflammation, and accelerate tumorigenesis.
-Numerous studies in murine models have supported this model [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
+Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
 Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018].
@@
 
+Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
 SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
 Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018].
-These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
-Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
+These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
 Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
 Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988].
-Based on these limited results, one would hypothesize increasing SCFA concentrations could limit tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
+In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
 These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.
@@
 **Figure S1.
-Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features.
+Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features.
 Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

### 2019-04-03 code/plot_classification_fit.R
Adjust figures to remove isob and faceting

@@ -35,7 +35,7 @@ classification_data %>%
 	ggplot(aes(x=microbiome, fill=test, y=auc)) +
 		geom_hline(yintercept=0.5, col="#888888") +
 		geom_boxplot(position=position_dodge2(preserve = "single"))+
-		facet_grid(scfa~class, labeller=labeller(.default=capitalize)) +
+		facet_grid(class~scfa, labeller=labeller(.default=capitalize)) +
 		coord_cartesian(ylim=c(0.35,1.0)) +
 		scale_x_discrete(
 			breaks = c("scfa", "genus", "otu"),
@@ -47,7 +47,7 @@ classification_data %>%
 			labels = c("Cross validation", "Held out data"),
 			values = brewer.pal("RdBu", n=3)[-2]
 		) +
-		labs(x="Microbiome data included", y="AUCROC for diagnosis classification") +
+		labs(x="Microbiome data included", y="AUROC for diagnosis classification") +
 		theme_classic() +
 		theme(
 			strip.background = element_blank(),

### 2019-04-03 code/plot_classification_regression.R

@@ -43,7 +43,7 @@ classification_plot <- classification_data %>%
 			labels = c("Without SCFAs", "With SCFAs"),
 			values = brewer.pal("PRGn", n=3)[-2]
 		) +
-		labs(x="Microbiome data included", y="AUCROC for\ndiagnosis classification") +
+		labs(x="Microbiome data included", y="AUROC for classification") +
 		theme_classic() +
 		theme(
 			strip.background = element_blank(),
@@ -82,7 +82,7 @@ regression_data <- read_tsv(regression_file_name,
 regression_features_plot <- regression_data %>%
 	ggplot(aes(x=microbiome, y=test_Rsquared, fill=data_type)) +
 	geom_boxplot() +
-	facet_wrap(~class, nrow=2, labeller=labeller(.default=capitalize)) +
+	facet_wrap(~class, nrow=1, labeller=labeller(.default=capitalize)) +
 	scale_x_discrete(
 		breaks = c("genus", "otu", "kegg", "opf", "pc2ec", "pc2pathways"),
 		labels = c("Genus", "OTU", "KEGG", "OPF", "KEGG", "Paths")
@@ -93,15 +93,17 @@ regression_features_plot <- regression_data %>%
 		labels = c("16S rRNA gene", "Metagenomic", "PICRUSt"),
 		values = brewer.pal("RdYlBu", n=3)
 	) +
-	labs(x="Microbiome data included", y=bquote(R^{2}~'for regression of SCFA concentrations')) +
+	labs(x="Microbiome data included", y=bquote(R^{2}~"for regression")) +
 	theme_classic() +
 	theme(
 		strip.background = element_blank(),
 		panel.spacing.x=unit(1, "lines"),
 		axis.ticks.x = element_blank(),
-		strip.text = element_text(size=10)
+		strip.text = element_text(size=10),
+		axis.text.x = element_text(angle = 90, hjust = 1,  vjust = 0.5)
 	)
 
-plot_grid(classification_plot, regression_features_plot, nrow=2, labels=c("A", "B"), rel_heights=c(0.5, 1))
+plot_grid(classification_plot, regression_features_plot, nrow=2, labels=c("A", "B"), rel_heights=c(1, 1))
 
-ggsave("results/figures/scfa_modeling.pdf", width=6.875, height=6, units="in")
+ggsave("results/figures/scfa_modeling.pdf", width=6.875, height=5.5, units="in")
[…]

### 2019-04-03 submission/manuscript.Rmd
Render manuscript

@@
 
-# Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure
+# Fecal short chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure
 
@@
 
-Marc A. Sze${^1}$, Begüm Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
+Marc A. Sze${^1}$, Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$
 
@@
 We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data.
-We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful association between SCFA concentration and disease status.
-When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
-Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples.
+We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and disease status.
+When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
+Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples.
 These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations.
@@
 
-**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
+**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
 We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A).
@@
 
-**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018].
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018].
 We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1).
-Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A).
+Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001).
+However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A).
 When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
@@
 **Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis.
-We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2).
+We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2).
 The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B).
 We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017].
-We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2).
-Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
+We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2).
+Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r 100*top_metagenomics$median_Rsquared`%, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
 Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2).
 SCFA concentrations could not be predicted based on the imputed metagenomic data.
-The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
+The largest amount of variation that the models could explain was `r 100*top_picrust$median_Rsquared`%, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
 The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.
@@
 
-**Random Forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations.
+**random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations.
 The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information.
@@
 To compare the median AUCROC for the held out data for the model generated using only the SCFAs, we compared the distribution of the data to the expected median of 0.5 using the Wilcoxon rank sum test to test whether the model performed better than would be achieved by randomly assigning the data to each diagnosis.
-When we compared the Random Forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.
+When we compared the random forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.
 
@@
 **Figure 2.
-SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did improve performance of the models generated using 16S rRNA gene sequence data.
+SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did not improve performance of the models generated using 16S rRNA gene sequence data.
 (B) Regression models that were trained using 16S rRNA gene sequence, metagenomic, and PICRUSt data to predict the concentrations of SCFAs performed poorly (all median R^2^ values < `r max(regression$median_Rsquared)`).
@@
 **Figure S1.
-Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features.
-Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.
+Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** random forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features.
+random forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.
 
@@
 **Figure S2.
-Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each individuals' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data.
+Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** random forest regression models were generated to predict the concentration of each SCFA using each individuals' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data.
 These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

### 2019-04-03 submission/manuscript.tex
Render manuscript

@@
 
-\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
-\section{Fecal short chain fatty acids are not predictive of colorectal
-cancer status and cannot be predicted based on bacterial community
-structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}
+\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
+\section{Fecal short chain fatty acids are not predictive of colonic
+tumor status and cannot be predicted based on bacterial community
+structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}
 
@@
 
-Marc A. Sze\({^1}\), Begüm Topçuoğlu\({^1}\), Nicholas A.
+Marc A. Sze\({^1}\), Begüm D. Topçuoğlu\({^1}\), Nicholas A.
 Lesniak\({^1}\), Mack T. Ruffin IV\({^2}\), Patrick D.
@@
 
-The gut microbiome is thought to have a role in the development of
-colorectal cancer by protecting against and exacerbating inflammation.
-Short chain fatty acids (SCFAs), including butyrate, have been shown to
-have anti-inflammatory properties and are produced in large quantities
-by colonic bacteria. We assessed whether there was an association
-between fecal SCFA concentrations and the presence of colonic adenomas
-or carcinomas in a cohort of individuals that was previously
-individualed to 16S rRNA gene and metagenomic shotgun sequencing. We
-measured the fecal concentrations of acetate, propionate, isobutyrate,
-and butyrate within the cohort and found that there were no meaningful
-differences in their concentration and disease status. When we
-incorporated these concentrations into Random Forest classification
+Colonic bacterial populations are thought to have a role in the
+development of colorectal cancer with some protecting against
+inflammation and others exacerbating inflammation. Short chain fatty
+acids (SCFAs), including butyrate, have been shown to have
+anti-inflammatory properties and are produced in large quantities by
+colonic bacteria which produce SCFAs by fermenting fiber. We assessed
+whether there was an association between fecal SCFA concentrations and
+the presence of colonic adenomas or carcinomas in a cohort of
+individuals using 16S rRNA gene and metagenomic shotgun sequence data.
+We measured the fecal concentrations of acetate, propionate, and
+butyrate within the cohort and found that there were no significant
+associations between SCFA concentration and disease status. When we
+incorporated these concentrations into random forest classification
 models trained to differentiate between people with normal colons and
@@
 sequence-based models to classify individuals. Finally, we generated
-Random Forest regression models trained to predict the concentration of
+random forest regression models trained to predict the concentration of
 each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
@@
 value of fiber consumption for reducing the risks of colorectal cancer.
-Although it is likely that bacterial metabolites may serve as biomarkers
-to detect adenomas or carcinomas, fecal SCFA concentrations have limited
-value.
+Although other bacterial metabolites may serve as biomarkers to detect
+adenomas or carcinomas, fecal SCFA concentrations have limited
+predictive power.
 
@@
 to genetic risk factors (2). This leaves a significant role for
-environmental, behavioral, and other factors such as smoking and diet
-(3, 4). Colorectal cancer is thought to be initiated by a series of
-mutations that accumulate as mutated cells proliferate leading to
-adenomatous lesions, which are succeeded by carcinomas (2). Throughout
-this progression, there are ample opportunities for bacterial
-populations to create mutations, induce inflammation, and accelerate
-tumorigenesis. Numerous studies in murine models have supported this
-model (5--7). Additional cross sectional studies in humans have
-identified microbiome-based biomarkers of disease (8). These studies
-suggest that in some cases, it is the loss of bacterial populations that
-produce short-chain fatty acids (SCFAs) that results in increased
-inflammation and tumorigenesis.
+environmental and behavioral factors and diet (3, 4). Colorectal cancer
+is thought to be initiated by a series of mutations that accumulate as
+the mutated cells begin to proliferate leading to adenomatous lesions,
+which are succeeded by carcinomas (2). Throughout this progression,
+there are ample opportunities for bacterial populations to have a role
+as bacteria are known to cause mutations, induce inflammation, and
+accelerate tumorigenesis (5--7). Additional cross sectional studies in
+humans have identified microbiome-based biomarkers of disease (8). These
+studies suggest that in some cases, it is the loss of bacterial
+populations that produce short-chain fatty acids (SCFAs) that results in
+increased inflammation and tumorigenesis.
 
-SCFAs have anti-inflammatory and anti-proliferative activities (9).
-Direct supplementation of SCFAs or feeding of fiber caused an overall
-reduction in tumor burdenin mouse models of colorectal cancer (10).
-These results suggest that supplementation with substrates that bacteria
-can ferment to produce SCFAs may confer beneficial effects against
-colorectal cancer. Based on similar observations, many microbiome
-studies use the concentrations of SCFAs and the presence of 16S rRNA
-gene sequences from organisms and the genes involved in producing them
-as a biomarker of a healthy microbiota (11, 12). Regardless, there is a
-lack of evidence that increasing SCFA concentrations can protect against
-colorectal cancer in humans. Case-control studies that have investigated
-a possible associations between SCFAs and colon tumor status have been
-plagued by relatively small numbers of subjects, but have reported
-increased total and relative fecal acetate levels and decreased relative
-fecal butyrate concentrations in subjects with colonic lesions (13).
-Based on these limited results, one would hypothesize increasing SCFA
-concentrations could limit tumorigenesis, in randomized controlled
-trials fiber supplementation has been inconsistently associated with
-protection against tumor formation and recurrence (14, 15). These
-findings temper enthusiasm for treatments that aim to use SCFAs as
-biomarkers or protection against tumorigenesis.
+Many microbiome studies use the concentrations of SCFAs and the presence
+of 16S rRNA gene sequences from organisms and the genes involved in
+producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
+have anti-inflammatory and anti-proliferative activities (11). Direct
+supplementation of SCFAs or feeding of fiber caused an overall reduction
+in tumor burden in mouse models of colorectal cancer (12). These results
+suggest that supplementation with fiber, which many colonic bacteria
+ferment to produce SCFAs may confer beneficial effects against
+colorectal cancer. Regardless, there is a lack of evidence that
+increasing SCFA concentrations can protect against colorectal cancer in
+humans. Case-control studies that have investigated possible
+associations between SCFAs and colon tumor status have been plagued by
+relatively small numbers of subjects, but have reported increased total
+and relative fecal acetate levels and decreased relative fecal butyrate
+concentrations in subjects with colonic lesions (13). In randomized
+controlled trials fiber supplementation has been inconsistently
+associated with protection against tumor formation and recurrence (14,
+15). These findings temper enthusiasm for treatments that aim to use
+SCFAs as biomarkers or protection against tumorigenesis.
 
-\textbf{SCFA concentrations do not meaningfully vary with diagnosis or
+\textbf{Fecal SCFA concentrations did not vary with diagnosis or
 treatment.} To quantify the associations between colorectal cancer, the
 microbiome, and SCFAs, we quantified the concentration of acetate,
-propionate, isobutyrate, and butyrate in feces of previously
-characterized individuals with normal colons (N=172) and those with
-colonic adenomas (N=198) or carcinomas (N=120) (16). The only SCFA that
-had a significantly different concentration across the diagnoses was
-isobutyrate (P=0.0091; Figure 1A). Interestingly, the median
-concentration of isobutyrate was 3.30 mmol/kg of feces in people with
-normal colons and it was 3.00 and 3.84 mmol/kg of feces in people with
-colonic adenomas or carcinomas, respectively. The difference in
-isobutyrate concentration between people with adenomas and carcinomas
-was significantly different (P=0.0065); however, the differences between
-people with normal colons and those with adenomas or carcinomas was not
-significant (P=0.19 and P=0.11). Among the individuals with adenomas and
-carcinomas, a subset (N\textsubscript{adenoma}=41,
+propionate, and butyrate in feces of previously characterized
+individuals with normal colons (N=172) and those with colonic adenomas
+(N=198) or carcinomas (N=120) (16). We were unable to detect a
+significant difference in any SCFA concentration across the diagnoses
+groups (all P\textgreater{}0.15; Figure 1A). Among the individuals with
+adenomas and carcinomas, a subset (N\textsubscript{adenoma}=41,
 N\textsubscript{carcinoma}=26) were treated and sampled a year later
-(17). The only SCFA that changed following treatment was isobutyrate,
-which decreased by 0.99 mmol/kg of feces (P=0.002; Figure 1B). For both
-the pre-treatment cross-sectional data and the pre/post treatment data,
-we also failed to detect any significant differences in the relative
-concentrations of any SCFAs other than isobutyrate (0.17). Finally, we
-pooled the SCFA concentrations on a total and per molecule of carbon
-basis and again failed to observe any significant differences
-(P\textgreater{}0.15). The low concentration of isobutyrate relative to
-the other SCFAs, inconsistent concentrations, and unexpected decrease in
-concentration with treatment makes it difficult to ascribe much
-biological relevance to the isobutyrate observations.
+(17). None of the SCFAs exhibited a significant change with treatment
+(all P\textgreater{}0.058; Figure 1B). For both the pre-treatment
+cross-sectional data and the pre/post treatment data, we also failed to
+detect any significant differences in the relative concentrations of any
+SCFAs (P\textgreater{}0.16). Finally, we pooled the SCFA concentrations
+on a total and per molecule of carbon basis and again failed to observe
+any significant differences (P\textgreater{}0.077). These results
+demonstrated that there were no significant associations between fecal
+SCFA concentration and diagnosis or treatment.
 
@@
 taxonomic units (OTUs) based on 97\% similarity or into genera enabled
-us to classify individuals as having adenomas or carcinomas using Random
-Forest machine learning models (8, 16). We repeated that analysis but
+us to classify individuals as having adenomas or carcinomas using random
+forest machine learning models (8, 16). We repeated that analysis but
 added the concentration of the SCFAs as possible features to train the
@@
 significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
-and P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC
+and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
 values to detect the presence of adenomas or carcinomas were only 0.54
-and 0.55, respectively (Figure 2A). When we trained the models with the
-SCFAs concentrations and OTU or genus-level relative abundances the
-AUROC values were not significantly different from the same models
-trained without the SCFA concentrations (P\textgreater{}0.21; Figure
-2A). These data demonstrate that knowledge of the SCFA profile from a
-patient's fecal sample did not improve the ability to diagnose a colonic
-lesion.
+and 0.55, respectively, indicating that SCFAs had poor predictive power
+on their own (Figure 2A). When we trained the models with the SCFAs
+concentrations and OTU or genus-level relative abundances the AUROC
+values were not significantly different from the same models trained
+without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
+data demonstrate that knowledge of the SCFA profile from a patient's
+fecal sample did not improve the ability to diagnose a colonic lesion.
 
@@
 predictive of fecal SCFA concentrations, regardless of a person's
-diagnosis. We trained Random Forest regression models using 16S rRNA
+diagnosis. We trained random forest regression models using 16S rRNA
 gene sequence data binned into OTUs and genera to predict the
@@
 assembled metagenomes into operational protein families (OPFs) or KEGG
-categories and trained Random Forest regression models using metagenomic
-to predict the concentration of the SCFAs (Figure S2). Similar to the
-analysis using 16S rRNA gene sequence data, the metagenomic data was not
-predictive of SCFA concentration; the largest amount of variation that
-the models could explain was 0.055, which was observed when using KEGG
-data to predict propionate concentrations (Figure 2B). Because of the
-limited number of samples that we were able to generate metagenomic
-sequence data from, we used our 16S rRNA gene sequence data to impute
-metagenomes that were binned into metabolic pathways or KEGG categories
-using PICRUSt (Figure S2). SCFA concentrations could not be predicted
-based on the imputed metagenomic data. The largest amount of variation
-that the models could explain was 0.097, which was observed when using
-KEGG data to predict butyrate concentrations (Figure 2B). The inability
-to model SCFA concentrations from microbiome data indicates that the
-knowledge of the abundance of organisms and their genes was insufficient
-to predict SCFA concentrations.
+categories and trained random forest regression models using metagenomic
+sequence data to predict the concentration of the SCFAs (Figure S2).
+Similar to the analysis using 16S rRNA gene sequence data, the
+metagenomic data was not predictive of SCFA concentration; the largest
+amount of variation that the models could explain was 5.45\%, which was
+observed when using KEGG data to predict propionate concentrations
+(Figure 2B). Because of the limited number of samples that we were able
+to generate metagenomic sequence data from, we used our 16S rRNA gene
+sequence data to impute metagenomes that were binned into metabolic
+pathways or KEGG categories using PICRUSt (Figure S2). SCFA
+concentrations could not be predicted based on the imputed metagenomic
+data. The largest amount of variation that the models could explain was
+9.67\%, which was observed when using KEGG data to predict butyrate
+concentrations (Figure 2B). The inability to model SCFA concentrations
+from microbiome data indicates that the knowledge of the abundance of
+organisms and their genes was insufficient to predict SCFA
+concentrations.
 
@@
 someone with one of these lesions. Furthermore, knowledge of the
-taxonomic and genetic structure of gut microbiota was not predictive
+taxonomic and genetic structure of gut microbiota was not predictive of
 SCFA concentrations. These results complement existing literature that
@@
 changes did not affect the range of concentrations of these SCFAs
-between the two studies.
+between the two studies. The concentrations of isobutyrate were
+consistently at or below the limit of detection and were not included in
+our analysis.
 
@@
 et al.~(16) and Sze et al.~(17) were obtained from the Sequence Read
-Archive (studies SRP062005 and SRP096978) and reprocessed using using
-mothur v.1.42 (20). The original studies generated sequence data from V4
-region of the 16S rRNA gene using paired 250 nt reads on an Illumina
-MiSeq sequencer. The resulting sequence data were assembled into contigs
-and screened to remove low quality contigs and chimeras. The curated
+Archive (studies SRP062005 and SRP096978) and reprocessed using mothur
+v.1.42 (20). The original studies generated sequence data from V4 region
+of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq
+sequencer. The resulting sequence data were assembled into contigs and
+screened to remove low quality contigs and chimeras. The curated
 sequences were then clustered into OTUs at a 97\% similarity threshold
@@
 were generated by Hannigan et al.~(18) and deposited into the Sequence
-Read Archive (study SRP108915). Fecal DNA was individualed to shotgun
+Read Archive (study SRP108915). Fecal DNA was subjected to shotgun
 sequencing on an Illumina HiSeq using 125 bp paired end reads. The
@@
 
-\textbf{Random Forest models.} The classification models were built to
+\textbf{random forest models.} The classification models were built to
 predict lesion type from microbiome information with or without SCFA
@@
 randomly assigning the data to each diagnosis. When we compared the
-Random Forest models generated without and with SCFA data included, we
+random forest models generated without and with SCFA data included, we
 used Wilcoxon rank sum tests to determine whether the models with the
@@
 
-\leavevmode\hypertarget{ref-OKeefe2016}{}%
-9. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
-metabolites and colon cancer. Nature Reviews Gastroenterology \&
-Hepatology \textbf{13}:691--706.
-doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
-
-\leavevmode\hypertarget{ref-Bishehsari2018}{}%
-10. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
-\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
-M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
-\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
-\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
-composition of gut microbiota, promotes SCFA production, and suppresses
-colon carcinogenesis. Genes \textbf{9}:102.
-doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
-
 \leavevmode\hypertarget{ref-Sanna2019}{}%
-11. \textbf{Sanna S}, \textbf{Zuydam NR van}, \textbf{Mahajan A},
+9. \textbf{Sanna S}, \textbf{Zuydam NR van}, \textbf{Mahajan A},
 \textbf{Kurilshikov A}, \textbf{Vila AV}, \textbf{Võsa U},
@@
 \leavevmode\hypertarget{ref-Meisel2016}{}%
-12. \textbf{Meisel M}, \textbf{Mayassi T}, \textbf{Fehlner-Peach H},
+10. \textbf{Meisel M}, \textbf{Mayassi T}, \textbf{Fehlner-Peach H},
 \textbf{Koval JC}, \textbf{O'Brien SL}, \textbf{Hinterleitner R},
@@
 doi:\href{https://doi.org/10.1038/ismej.2016.114}{10.1038/ismej.2016.114}.
+
+\leavevmode\hypertarget{ref-OKeefe2016}{}%
+11. \textbf{O'Keefe SJD}. 2016. Diet, microorganisms and their
+metabolites and colon cancer. Nature Reviews Gastroenterology \&
+Hepatology \textbf{13}:691--706.
+doi:\href{https://doi.org/10.1038/nrgastro.2016.165}{10.1038/nrgastro.2016.165}.
+
+\leavevmode\hypertarget{ref-Bishehsari2018}{}%
+12. \textbf{Bishehsari F}, \textbf{Engen P}, \textbf{Preite N},
+\textbf{Tuncil Y}, \textbf{Naqib A}, \textbf{Shaikh M}, \textbf{Rossi
+M}, \textbf{Wilber S}, \textbf{Green S}, \textbf{Hamaker B},
+\textbf{Khazaie K}, \textbf{Voigt R}, \textbf{Forsyth C},
+\textbf{Keshavarzian A}. 2018. Dietary fiber treatment corrects the
+composition of gut microbiota, promotes SCFA production, and suppresses
+colon carcinogenesis. Genes \textbf{9}:102.
+doi:\href{https://doi.org/10.3390/genes9020102}{10.3390/genes9020102}.
 
@@
 adenomas or carcinomas using SCFAs was slightly better than than chance
-(depicted by horizontal line at 0.50), but did improve performance of
-the models generated using 16S rRNA gene sequence data. (B) Regression
-models that were trained using 16S rRNA gene sequence, metagenomic, and
-PICRUSt data to predict the concentrations of SCFAs performed poorly
-(all median R\textsuperscript{2} values \textless{} 0.14). Regression
-models generated using 16S rRNA gene sequence and PICRUSt data included
-data from 490 samples and those generated using metagenomic data
-included data from 78 samples.
+(depicted by horizontal line at 0.50), but did not improve performance
+of the models generated using 16S rRNA gene sequence data. (B)
+Regression models that were trained using 16S rRNA gene sequence,
+metagenomic, and PICRUSt data to predict the concentrations of SCFAs
+performed poorly (all median R\textsuperscript{2} values \textless{}
+0.14). Regression models generated using 16S rRNA gene sequence and
+PICRUSt data included data from 490 samples and those generated using
+metagenomic data included data from 78 samples.
 
@@
 classification models shows that the models are robust and are not
-overfit.} Random Forest classification models were generated to
+overfit.} random forest classification models were generated to
 differentiate between individuals with normal colons and those with
 adenomas or carcinomas using 16S rRNA gene sequence data that were
-clustered into genera or OTUs with and without including the four SCFAs
-as additional features. Random Forest classification models were
+clustered into genera or OTUs with and without including the three SCFAs
+as additional features. random forest classification models were
 generated by partitioning the samples into a training set with 80\% of
@@
 regression models shows that the models are robust and are not overfit.}
-Random Forest regression models were generated to predict the
+random forest regression models were generated to predict the
 concentration of each SCFA using each individuals' microbiome data

### 2019-04-03 submission/manuscript.Rmd
Edits

@@
 
-# Fecal short chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure
+# Fecal short-chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure
 
@@
 Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation.
-Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber.
+Short-chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber.
 We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data.
-We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and disease status.
+We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and tumor status.
 When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals.
@@
 Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors.
-Short chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties.
+Short-chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties.
 We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors.
@@
 Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990].
-This leaves a significant role for environmental and behavioral factors and diet [@FlissIsakov2017; @Lee2015].
+This leaves a significant role for environmental, behavioral, and dietary factors [@FlissIsakov2017; @Lee2015].
 Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
@@
 In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
-These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.
+These findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.
 
@@
 When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A).
-These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample did not improve the ability to diagnose a colonic lesion.
+These data demonstrate that knowledge of the SCFA profile from a subject's fecal sample did not improve the ability to diagnose a colonic lesion.
 
@@
 We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2).
-Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r 100*top_metagenomics$median_Rsquared`%, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
+Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration.
+The largest R^2^ was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B).
 Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2).
 SCFA concentrations could not be predicted based on the imputed metagenomic data.
-The largest amount of variation that the models could explain was `r 100*top_picrust$median_Rsquared`%, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
+The largest R^2^ was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
 The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.
@@
 These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors.
-It is important to note that our analysis concerned fecal SCFA concentrations and microbiome characterization and that observations along the mucosa near the site of lesions may provide a stronger association.
-Regardless, given the growing literature in this area, it is unlikely that SCFAs are the primary mechanism that limits tumorigenesis.
+It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples and that observations along the mucosa near the site of lesions may provide a stronger association.
 This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally.

### 2019-04-03 submission/manuscript.tex
Edits

@@
 \hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
-\section{Fecal short chain fatty acids are not predictive of colonic
+\section{Fecal short-chain fatty acids are not predictive of colonic
 tumor status and cannot be predicted based on bacterial community
@@
 development of colorectal cancer with some protecting against
-inflammation and others exacerbating inflammation. Short chain fatty
+inflammation and others exacerbating inflammation. Short-chain fatty
 acids (SCFAs), including butyrate, have been shown to have
@@
 butyrate within the cohort and found that there were no significant
-associations between SCFA concentration and disease status. When we
+associations between SCFA concentration and tumor status. When we
 incorporated these concentrations into random forest classification
@@
 colorectal tumors early without invasive colonoscopy procedures and to
-prevent the formation of tumors. Short chain fatty acids (SCFAs) are
+prevent the formation of tumors. Short-chain fatty acids (SCFAs) are
 often used as a surrogate for measuring gut health and for being
@@
 to genetic risk factors (2). This leaves a significant role for
-environmental and behavioral factors and diet (3, 4). Colorectal cancer
+environmental, behavioral, and dietary factors (3, 4). Colorectal cancer
 is thought to be initiated by a series of mutations that accumulate as
@@
 associated with protection against tumor formation and recurrence (14,
-15). These findings temper enthusiasm for treatments that aim to use
-SCFAs as biomarkers or protection against tumorigenesis.
+15). These findings temper enthusiasm for treatments that target the
+production of SCFAs or for using them as biomarkers for protection
+against tumorigenesis.
 
@@
 without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
-data demonstrate that knowledge of the SCFA profile from a patient's
+data demonstrate that knowledge of the SCFA profile from a subject's
 fecal sample did not improve the ability to diagnose a colonic lesion.
@@
 Similar to the analysis using 16S rRNA gene sequence data, the
-metagenomic data was not predictive of SCFA concentration; the largest
-amount of variation that the models could explain was 5.45\%, which was
-observed when using KEGG data to predict propionate concentrations
-(Figure 2B). Because of the limited number of samples that we were able
-to generate metagenomic sequence data from, we used our 16S rRNA gene
-sequence data to impute metagenomes that were binned into metabolic
-pathways or KEGG categories using PICRUSt (Figure S2). SCFA
-concentrations could not be predicted based on the imputed metagenomic
-data. The largest amount of variation that the models could explain was
-9.67\%, which was observed when using KEGG data to predict butyrate
+metagenomic data was not predictive of SCFA concentration. The largest
+R\textsuperscript{2} was 0.055, which was observed when using KEGG data
+to predict propionate concentrations (Figure 2B). Because of the limited
+number of samples that we were able to generate metagenomic sequence
+data from, we used our 16S rRNA gene sequence data to impute metagenomes
+that were binned into metabolic pathways or KEGG categories using
+PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
+the imputed metagenomic data. The largest R\textsuperscript{2} was
+0.097, which was observed when using KEGG data to predict butyrate
 concentrations (Figure 2B). The inability to model SCFA concentrations
@@
 prevent the risk of developing colonic tumors. It is important to note
-that our analysis concerned fecal SCFA concentrations and microbiome
-characterization and that observations along the mucosa near the site of
-lesions may provide a stronger association. Regardless, given the
-growing literature in this area, it is unlikely that SCFAs are the
-primary mechanism that limits tumorigenesis. This may be a cautionary
-result to temper enthusiasm for SCFAs as a biomarker of gut health more
-generally. Going forward it is critical to develop additional hypotheses
-for how the microbiome and host interact to drive tumorigenesis so that
-we can better understand tumorigenesis and identify biomarkers that will
-allow early detection of lesions.
+that our analysis was based on characterizations of SCFA and microbiome
+profiles using fecal samples and that observations along the mucosa near
+the site of lesions may provide a stronger association. This may be a
+cautionary result to temper enthusiasm for SCFAs as a biomarker of gut
+health more generally. Going forward it is critical to develop
+additional hypotheses for how the microbiome and host interact to drive
+tumorigenesis so that we can better understand tumorigenesis and
+identify biomarkers that will allow early detection of lesions.
 

### 2019-04-07 code/picrust1.sh
Fix dependencies

@@ -1,8 +1,7 @@
 #!bash
 
-# module load python-anaconda2/latest
-
-# module load R/3.5.1
+module load python-anaconda2/latest
+module load R/3.5.1
 
 #Set local variables
 REF=data/references

### 2019-04-07 code/picrust2.sh

@@ -1,5 +1,8 @@
 THREADS=6
 
+module load python-anaconda3/latest
+module load R/3.5.1
+
 WORKDIR=data/picrust2
 rm -rf $WORKDIR
 mkdir -p $WORKDIR
@@ -14,7 +17,7 @@ mothur "#make.biom(shared=data/asv/crc.asv.shared, outputdir=$WORKDIR)"
 
 mv $WORKDIR/crc.asv.ASV.biom $BIOM
 
-
+conda activate
 conda activate picrust2
 
 # Comments on whether this is an acceptable biom format

### 2019-04-08 submission/manuscript.Rmd
Update data and render prior to submission

@@
 In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
-These findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.
+Such studies are plagued by difficulties insuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation.
+Together, these findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.
 
@@
 
-regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ko")
 scfa_types <- c("acetate", "propionate", "butyrate")
@@
 								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
-								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
+								 microbiome %in% c("pc2ko", "pc2pathways")  ~ "picrust"
 							 )) %>%
@@
 
-convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ec = "KEGG", pc2pathways = "pathway")
+convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ko = "KEGG", pc2pathways = "pathway")
 

### 2019-04-08 submission/manuscript.tex
Update data and render prior to submission

@@
 associated with protection against tumor formation and recurrence (14,
-15). These findings temper enthusiasm for treatments that target the
-production of SCFAs or for using them as biomarkers for protection
-against tumorigenesis.
+15). Such studies are plagued by difficulties insuring subjects took the
+proper dose and using subjects with prior polyp history who may be
+beyond a point of benefiting from fiber supplementation. Together, these
+findings temper enthusiasm for treatments that target the production of
+SCFAs or for using them as biomarkers for protection against
+tumorigenesis.
 
@@
 values were not significantly different from the same models trained
-without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
+without the SCFA concentrations (P\textgreater{}0.15; Figure 2A). These
 data demonstrate that knowledge of the SCFA profile from a subject's
@@
 the imputed metagenomic data. The largest R\textsuperscript{2} was
-0.097, which was observed when using KEGG data to predict butyrate
+0.085, which was observed when using KEGG data to predict propionate
 concentrations (Figure 2B). The inability to model SCFA concentrations

### 2019-04-08 code/plot_classification_regression.R

@@ -55,7 +55,7 @@ classification_plot <- classification_data %>%
 
 regression_file_name <- "data/rf/regression_data_pool.tsv"
 
-regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ko")
 scfa_types <- c("acetate", "propionate", "butyrate")
 
 regression_data <- read_tsv(regression_file_name,
@@ -75,7 +75,7 @@ regression_data <- read_tsv(regression_file_name,
 					data_type = case_when(
 			       microbiome %in% c("genus", "otu")  ~ "16S",
 						 microbiome %in% c("kegg", "opf")  ~ "metagenome",
-						 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
+						 microbiome %in% c("pc2ko", "pc2pathways")  ~ "picrust"
 					 )
 				)
 
@@ -84,7 +84,7 @@ regression_features_plot <- regression_data %>%
 	geom_boxplot() +
 	facet_wrap(~class, nrow=1, labeller=labeller(.default=capitalize)) +
 	scale_x_discrete(
-		breaks = c("genus", "otu", "kegg", "opf", "pc2ec", "pc2pathways"),
+		breaks = c("genus", "otu", "kegg", "opf", "pc2ko", "pc2pathways"),
 		labels = c("Genus", "OTU", "KEGG", "OPF", "KEGG", "Paths")
 	) +
 	scale_fill_manual(

### 2019-04-08 code/plot_regression_fit.R

@@ -12,7 +12,7 @@ capitalize <- function(string) {
 
 regression_file_name <- "data/rf/regression_data_pool.tsv"
 
-regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
+regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ko")
 scfa_types <- c("acetate", "propionate", "butyrate")
 
 regression_data <- read_tsv(regression_file_name,
@@ -32,7 +32,7 @@ regression_data <- read_tsv(regression_file_name,
 					data_type = case_when(
 			       microbiome %in% c("genus", "otu")  ~ "16S rRNA gene",
 						 microbiome %in% c("kegg", "opf")  ~ "Metagenomics",
-						 microbiome %in% c("pc2ec", "pc2pathways")  ~ "PICRUSt"
+						 microbiome %in% c("pc2ko", "pc2pathways")  ~ "PICRUSt"
 					 )
 				) %>%
 	gather(test, auc, train_Rsquared, test_Rsquared) %>%
@@ -43,7 +43,7 @@ regression_data %>%
 	geom_boxplot() +
 	facet_grid(class~data_type, labeller=labeller(.default=capitalize), scale="free_x") +
 	scale_x_discrete(
-		breaks = c("genus", "otu", "kegg", "opf", "pc2ec", "pc2pathways"),
+		breaks = c("genus", "otu", "kegg", "opf", "pc2ko", "pc2pathways"),
 		labels = c("Genus", "OTU", "KEGG", "OPF", "KEGG", "Paths")
 	) +
 	scale_fill_manual(

### 2019-05-30 submission/response_to_reviewers.md
Respond to reviewers' comments

We appreciate the comments provided by the two reviewers and are grateful for their suggestions and overall positive view of the manuscript. We have done our best to address their comments below. Where appropriate, we have referenced line numbers in the marked up version of the manuscript.

**Reviewer #1 (Comments for the Author):**

**This paper investigates associations between colon cancer and fecal microbiota/SCFAs. This is an important study because intersections between each of these variables are current active areas of research in the field. Overall, I found this paper concise and easy to read. I also found the analyses clear. My primary comments (described below) center on the interpretation of those analyses.**

**Major points:**

**This manuscript deals with the tricky case of how to report non-significant (i.e. p>0.05) results. The approach the authors take is to argue for a lack of association between the measured variables (e.g. fecal SCFAs & tumor status). I have two concerns about this interpretation. First, statisticians are increasingly clarifying that the lack of a significant statistical result does not mean there is no association between two variables (see Amrhein et al., Nature 2019). Second, p-values reflect the odds that a given statistical result was arrived at by chance, assuming the null hypothesis is true. Those odds, in my opinion, are not fundamentally different between 0.05 and say 0.058 or even 0.077 (L66; L70). Additionally, the authors even report an instance where a model associating SCFAs and tumor status has a p-value below the conventional cutoff (p < 0.0001; L79,80).**

**Still, I don't mean to encourage alternative problematic interpretations such as trumpeting significant associations when p-values are marginal. I think instead what might be merited here is nuance. Perhaps rather than rejecting an association, the authors could consider simply reporting the p-values, as well as a measure of effect (e.g. difference of means or odds ratios). Those effect sizes, I suspect, will support the conclusion that even if associations exist, they are weak (which is how the random forest models in L72-85 are presented).**

> We agree that relying heavily on an arbitrary threshold for establishing significance or lack of significance is problematic. This was why we also showed the data in Figure 1. These box plots hopefully make it clear that although there may be a couple of nearly significant P-values, the effect sizes are small and are inconsistent. We have added a sentence to emphasize this point at L75-77. We have considered different ways of reporting the P-values and effect sizes and have concluded that the revised approach is the most suitable given the word length restrictions on an Observation format manuscript.

> For the reviewer's benefit in the paragraph from L63 to L78...

> * The "all P>0.148" set of P-values for the cross sectional data included values of ***0.148***, 0.222, and 0.655 for acetate, butyrate, and propionate, respectively. The largest effect sizes for these three comparisons were 2.45, 1.72, and 0.352 mmol/kg
> * The "all P>0.058" set of P-values for the follow-up data included values of 0.508 [cancer] and 0.350 [adenoma], 0.916 [cancer] and 0.101 [adenoma], 0.853 [cancer] and ***0.058*** [adenoma] for acetate, butyrate, and propionate, respectively. The corresponding differences between the before and after samples were 3.00 [cancer] and -3.84 [adenoma], 0.312 [cancer] and -1.75 [adenoma], -0.396 [cancer] and -1.68 [adenoma] for acetate, butyrate, and propionate, respectively.
> * The "all P>0.16" set of P-values for the relative concentrations of SCFA included values of 0.571, 0.765, and 0.245 for acetate, butyrate, and propionate, respectively from the cross sectional data. They were ***0.164*** [cancer] and 0.322 [adenoma], 0.692 [cancer] and 0.432 [adenoma], and 0.210 [cancer] and 0.365 [adenoma] for acetate, butyrate, and propionate follow-up data, respectively. The largest effect sizes for the cross-sectional data were 0.006, 0.004, and 0.011  for acetate, butyrate, and propionate, respectively. The differences in relative SCFA concentrations in the follow-up data were 0.029 [cancer] and 0.015 [adenoma], -0.006 [cancer] and -0.007 [adenoma], and -0.020 [cancer] and -0.011 [adenoma] for acetate, butyrate, and propionate, respectively.
> * The "all P>0.077" set of P-values for the total concentration of SCFA included values of 0.189 for the cross sectional data and 0.833 for the follow-up cancer data and 0.150 for the follow-up adenoma data. In addition the set included values for the per-molecule of carbon basis which were 0.216 for the cross sectional data and 0.916 for the follow-up cancer data and ***0.077*** for the follow-up adenoma data.

> Again, aside from cluttering the figure with P-values or the text with extra effect sizes we think it is best to leave it to the reader to see that the effect sizes in Figure 1 are quite small.

> Finally, the reviewer points out that "a model associating SCFAs and tumor status has a p-value below the conventional cutoff (p < 0.0001; L79,80)". We point out at that location in the manuscript that the comparison is to a random classification and that the actual AUROC was only 0.54 and 0.55, which was unimpressive. We feel that this was sufficient to explain the P-value.

**I encourage the authors to consider tempering their language regarding the association between fiber intake, SCFAs and protection tumor risk (e.g. L18). I am aware that butyrate and other SCFAs may be over-hyped as a panacea for GI disease. Still, if protection is mediated by SCFA uptake (to me the most reasonable model) then excreted (i.e. fecal SCFAs) should not necessarily be correlated with disease. Additionally, the authors are examining a single point measurement of SCFAs, without consideration of how noisy they may be. It's possible that more frequent longitudinal measurements would reveal a stronger association.**

> We appreciate the reviewer's concern that we are over interpreting negative results. In various places where we provide background to the problem, including L18, we are pointing to epidemiological data which provide a confused picture of the value of fiber and SCFAs in preventing colorectal cancer (CRC). We have gone back through the reporting of our results to emphasize the use of fecal SCFAs, a single time point, and the goal of classification. The most significant edits were made at L31-32, L114, and L118. For example, at L31-33, we rewrote, *"These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden."* as *"Our results indicate that the association between fecal SCFAs, the gut microbiome, and tumor burden is weak."*

> It should be noted that our manuscript is in the context of diagnosis, not necessarily identifying mechanisms of tumorigenesis that are mediated by microorganisms. We are limited in the amount of discussion we can provide because of word limitations, but it is important to note that it is technically challenging/impossible to obtain mucosal SCFA concentrations since the typical bowel preparation strips away much of the mucosa limiting the ability to sample the SCFAs at the interface. Furthermore, considering our goal is to develop a diagnostic to detect colonic lesions, longitudinal sampling is unlikely to be possible.

**Minor points: **

**L38: This line could be interpreted as saying any bacterium could cause mutations, inflammations, etc. But, to date, only select bacteria are known to cause such events.**

> We have revised the text from "...role as bacteria are known to..." to "...role as some bacteria are known to..."

**L45: "SCFA have ... anti-proliferative activities." This is true but only contextually. Butyrate in particular has proliferative activity in colonocytes. (See Donohoe et al. Mol Cell. The Warburg Effect Dictates the Mehcanism of Butyrate Mediated Histone Acetylation and Cell Proliferation. 2012)**

> This sentence has been edited to highlight the contextual nature of this observation.

**L55: Insuring should be 'ensuring'**

> This as been corrected as suggested.

**L65: Is it really surprising that there are no effects 1 year after treatment? Also, what exactly were the treatments?**

> We have added text to the sentence to make this more clear at L70 - "None of the *individuals showed signs of recurrence and yet none of the* SCFAs exhibited a significant change with treatment (all P>0.058; Figure 1B)". We previously showed that the microbiome of those individuals diagnosed with carcinomas returned to a more normal state following treatment. If SCFAs were associated with disease, then the prediction would be that the concentrations would change with a return to health. That we did not see such a change further supports our assertion that there is a tenuous link between fecal SCFA concentrations and health status.

> Because of the limits to the number of words available in the *mBio* Observation Format, we do not have the space to recap the treatments. Patients underwent surgical resection and/or chemotherapy. This is more extensively covered in the referenced paper by Sze et al. (2017) [citation 17].

**L92: To be consistent with other statistical tests reported in this manuscript, can the authors provide some measure of how often an R2 of 0.14 would be achieved by chance?**

> The R2 values that we report are the median observed values for our 80/20 split hold out permutation test based on 100 permutations. The distribution of observed values are provided in Supplemental Figure 2. If the association were truly random, the R2 value would be near zero. Our point is that the observed R2 values are quite low (e.g. at most 0.14) and although they are significantly different from zero, the models do a poor job of predicting SCFA concentrations. Because we are not comparing models to each other (e.g. an OTU-based model to an OPF-based model), there is no P-value to compare across R2 values.

**L135: The authors use an aqueous extraction and aqueous injection method for measurement of SCFA by HPLC, and do not acidify the aqueous solution to a pH of ~3 before filtration, as has previously been described as essential for accurate quantitation by GC (see Zhao, Nyman, Jönsson. Biomedical Chromatorgraphy (2006)), and remains a standard practice in GC (see Teixeira et al, British Journal of Nutrition (2013)). Why would acidification of the sample to ensure protonation of the SCFA not be necessary for efficient extraction, as the anionic form of the conjugate base is extracted less efficiently from the stool matrix? The literature often does not provide a rationale for the lack of a need for acidification for quantitation of SCFA from feces. If such a rationale has been published, can the authors please provide this reference?**

> We are unaware of a specific reference as to why SCFAs are extracted differently when using GC or HPLC. The protocol we used is based on that used by Venkataraman et al. (2016)[citation 19], which developed their methods based on the existing literature. The approach is effective and has worked well for a variety of fecal samples.

**L173: Can the authors define 'mtry'? Some readers might not be familiar with this term.**

> We have added a sentence at L183-185 to clarify the meaning of this parameter, *"The mtry parameter represents the number of features randomly sampled from the available features at a question point in the classification tree (i.e. called splits of nodes) that, when answered, lead to the greatest improvement in classification"*.

**Fig. 1B: Are the units here the same as in Fig 1A? Can these be specified?**

> They are the same units and we've inserted them with the y-axis label for Figure 1B.

**Reviewer #2 (Comments for the Author):**

**This manuscript by Sze et al., carries out testing on fecal SCFAs, fecal 16S/metagenomics, in order to assess the value of SCFAs to a colorectal cancer classifier. In this aspect, it is a fine contribution and of interest to the readers of mBio. On that basis I like it, with one major caveat:**

**The language of the claims is problematic. It is very hard to prove that something cannot be done. I would have less of a problem with these statements if there were multiple methods or an attempt to quantify the overall amount of information embedded in SCFAs to try to find an upper bound for how predictive they could be. Instead, the authors try one or two tests in each case and declare that SCFAs cannot be used for CRC classification. In other words, the scientific diligence is not commensurate with the broad nature of the claims.**

> This reviewer shares the concerns of the first reviewer, which we have hopefully addressed. We agree that it is impossible to prove that something cannot be done - scientifically we can't *prove* anything. We can only propose hypotheses and either reject them or find support for the hypothesis.

> We disagree with the reviewer's comment that we haven't pursued multiple methods. First of all, we used a large collection of samples from a cross sectional study and a collection of samples from people with a significant colonic lesion who underwent treatment. Second, our analysis compared SCFA concentrations across diagnosis groups with straightforward statistical tests and we used those data to see if they would improve the ability to diagnose individuals using 16S rRNA gene sequence data, metagenomic data, and imputed metagenomic sequence data. Finally, we attempted to use the 16S rRNA gene sequence data, metagenomic data, and imputed metagenomic sequence data to predict the fecal SCFA concentrations. For these comparisons we tested for an association using the individual concentrations of each SCFA, the relative concentration of each SCFA, the total concentration of SCFAs, and the total amount of SCFA on a per carbon molecule basis. Although, the only classification modeling approach we used was random forest; however, this method and our testing framework are considered to be less sensitive to overfitting, able to incorporate non-linearities in the data, and treat the data in a context-dependent manner. Other methods are available (e.g. logistic regression, SVM, decision tree, etc.), but at least in our hands with similar data do not perform as well as random forest.

**More specifically, I would suggest the following edits to reign this in:**

**1. "Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas" should be changed to "Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas in a random forest machine learning model."**

**2. "Knowledge of microbial community structure does not predict SCFA concentrations in a random forest regression model."**

> Both of these section headings have been edited as suggested by the reviewer.

### 2019-05-30 submission/manuscript.Rmd
Fix email address
Make figures an explicit dependency and remove from body of manuscript
Respond to reviewers' comments

@@
 
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
+$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
 
@@
 Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation.
-Short-chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber.
+Short-chain fatty acids (SCFAs) have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber.
 We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data.
@@
 
-Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors.
+Considering colorectal cancer is the third leading cancer-related cause of death within the United States, it is important to detect colorectal tumors early and to prevent the formation of tumors.
 Short-chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties.
@@
 Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations.
-These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden.
+Our results indicate that the association between fecal SCFAs, the gut microbiome, and tumor burden is weak.
 
@@
 Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990].
-Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
+Throughout this progression, there are ample opportunities for bacterial populations to have a role as some bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017].
 Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018].
@@
 Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016].
-SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016].
+Depending on the concentrations, SCFAs can have proliferative activities at low concentrations or anti-proliferative activities at higher concentrations; they can also have anti-inflammatory activities [@OKeefe2016].
 Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018].
-These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer.
-Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
+These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs, may confer beneficial effects against colorectal cancer.
+Regardless, there is a lack of consistent evidence that increasing SCFA concentrations can protect against colorectal cancer in humans.
 Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988].
 In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018].
-Such studies are plagued by difficulties insuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation.
+Such studies are confounded by difficulties ensuring subjects took the proper dose and using subjects with prior polyp history who may be beyond a point of benefiting from fiber supplementation.
 Together, these findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.
@@
 
-**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
+**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To test for a significant association between colorectal cancer and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016].
 We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A).
 Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017].
-None of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B).
+None of the individuals showed signs of recurrence and yet none of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B).
 For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`).
 Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`).
+Although some of the P-values from our analyses were close to 0.05, the effect sizes were all relatively small and inconsistent given the disease progression (Figure 1).
 These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.
@@
 
-**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018].
+**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas using a random forest model.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018].
 We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1).
@@
 
-**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis.
+**Knowledge of microbial community structure does not predict SCFA concentrations using a random forest model.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis.
 We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2).
@@
 The largest R^2^ was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B).
-The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.
+The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict fecal SCFA concentrations.
 
 **Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions.
-Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of SCFA concentrations.
+Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not meaningfully predictive of SCFA concentrations.
 These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors.
-It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples and that observations along the mucosa near the site of lesions may provide a stronger association.
+It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples at a single time point.
+Furthermore, observations along the mucosa near the site of lesions may provide a stronger association.
 This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally.
@@
 
-**random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations.
+**Random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations.
 The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information.
@@
 For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations.
+The mtry parameter represents the number of features randomly sampled from the available features at a question point in the classification tree (i.e.
+called splits of nodes) that, when answered, lead to the greatest improvement in classification.
 Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected.
@@
 We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively.
-We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.
+We used the randomForest R package (version `r installed.packages()["randomForest" ,"Version"]`) as implemented in the caret R package (version `r installed.packages()["caret" ,"Version"]`) for developing and testing our models.
 
@@
 
-**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/.
+**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_mBio_2019/.
 
@@
 
-\includegraphics{figure_1.png}
+## Figures
 
@@
 (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.
-
-\newpage
-
-\includegraphics{figure_2.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s1.png}
-
 **Figure S1.
@@
 
-\newpage
-
-\includegraphics{figure_s2.png}
-
 **Figure S2.

### 2019-05-30 submission/manuscript.tex
Fix email address
Make figures an explicit dependency and remove from body of manuscript
Respond to reviewers' comments

@@
 % Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
+\providecommand{\subtitle}[1]{
   \posttitle{
@@
 \(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
 
@@
 inflammation and others exacerbating inflammation. Short-chain fatty
-acids (SCFAs), including butyrate, have been shown to have
-anti-inflammatory properties and are produced in large quantities by
-colonic bacteria which produce SCFAs by fermenting fiber. We assessed
-whether there was an association between fecal SCFA concentrations and
-the presence of colonic adenomas or carcinomas in a cohort of
-individuals using 16S rRNA gene and metagenomic shotgun sequence data.
-We measured the fecal concentrations of acetate, propionate, and
-butyrate within the cohort and found that there were no significant
-associations between SCFA concentration and tumor status. When we
-incorporated these concentrations into random forest classification
-models trained to differentiate between people with normal colons and
-those with adenomas or carcinomas, we found that they did not
-significantly improve the ability of 16S rRNA gene or metagenomic gene
-sequence-based models to classify individuals. Finally, we generated
-random forest regression models trained to predict the concentration of
-each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
-the same samples. These models performed poorly and were able to explain
-at most 14\% of the observed variation in the SCFA concentrations. These
-results support the broader epidemiological data that questions the
-value of fiber consumption for reducing the risks of colorectal cancer.
-Although other bacterial metabolites may serve as biomarkers to detect
-adenomas or carcinomas, fecal SCFA concentrations have limited
-predictive power.
+acids (SCFAs) have been shown to have anti-inflammatory properties and
+are produced in large quantities by colonic bacteria which produce SCFAs
+by fermenting fiber. We assessed whether there was an association
+between fecal SCFA concentrations and the presence of colonic adenomas
+or carcinomas in a cohort of individuals using 16S rRNA gene and
+metagenomic shotgun sequence data. We measured the fecal concentrations
+of acetate, propionate, and butyrate within the cohort and found that
+there were no significant associations between SCFA concentration and
+tumor status. When we incorporated these concentrations into random
+forest classification models trained to differentiate between people
+with normal colons and those with adenomas or carcinomas, we found that
+they did not significantly improve the ability of 16S rRNA gene or
+metagenomic gene sequence-based models to classify individuals. Finally,
+we generated random forest regression models trained to predict the
+concentration of each SCFA based on 16S rRNA gene or metagenomic gene
+sequence data from the same samples. These models performed poorly and
+were able to explain at most 14\% of the observed variation in the SCFA
+concentrations. These results support the broader epidemiological data
+that questions the value of fiber consumption for reducing the risks of
+colorectal cancer. Although other bacterial metabolites may serve as
+biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations
+have limited predictive power.
 
@@
 Considering colorectal cancer is the third leading cancer-related cause
-of death within the United States, there is a great need to detect
-colorectal tumors early without invasive colonoscopy procedures and to
-prevent the formation of tumors. Short-chain fatty acids (SCFAs) are
-often used as a surrogate for measuring gut health and for being
-anti-carcinogenic because of their anti-inflammatory properties. We
-evaluated the fecal SCFA concentration of a cohort of individuals with
-varying colonic tumor burden who were previously analyzed to identify
-microbiome-based biomarkers of tumors. We were unable to find an
-association between SCFA concentration and tumor burden or use SCFAs to
-improve our microbiome-based models of classifying people based on their
-tumor status. Furthermore, we were unable to find an association between
-the fecal community structure and SCFA concentrations. These data
-indicate that there is no conclusive link between the gut microbiome,
-SCFAs, and tumor burden.
+of death within the United States, it is important to detect colorectal
+tumors early and to prevent the formation of tumors. Short-chain fatty
+acids (SCFAs) are often used as a surrogate for measuring gut health and
+for being anti-carcinogenic because of their anti-inflammatory
+properties. We evaluated the fecal SCFA concentration of a cohort of
+individuals with varying colonic tumor burden who were previously
+analyzed to identify microbiome-based biomarkers of tumors. We were
+unable to find an association between SCFA concentration and tumor
+burden or use SCFAs to improve our microbiome-based models of
+classifying people based on their tumor status. Furthermore, we were
+unable to find an association between the fecal community structure and
+SCFA concentrations. Our results indicate that the association between
+fecal SCFAs, the gut microbiome, and tumor burden is weak.
 
@@
 there are ample opportunities for bacterial populations to have a role
-as bacteria are known to cause mutations, induce inflammation, and
+as some bacteria are known to cause mutations, induce inflammation, and
 accelerate tumorigenesis (5--7). Additional cross sectional studies in
@@
 of 16S rRNA gene sequences from organisms and the genes involved in
-producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
-have anti-inflammatory and anti-proliferative activities (11). Direct
-supplementation of SCFAs or feeding of fiber caused an overall reduction
-in tumor burden in mouse models of colorectal cancer (12). These results
-suggest that supplementation with fiber, which many colonic bacteria
-ferment to produce SCFAs may confer beneficial effects against
-colorectal cancer. Regardless, there is a lack of evidence that
-increasing SCFA concentrations can protect against colorectal cancer in
-humans. Case-control studies that have investigated possible
-associations between SCFAs and colon tumor status have been plagued by
-relatively small numbers of subjects, but have reported increased total
-and relative fecal acetate levels and decreased relative fecal butyrate
-concentrations in subjects with colonic lesions (13). In randomized
-controlled trials fiber supplementation has been inconsistently
-associated with protection against tumor formation and recurrence (14,
-15). Such studies are plagued by difficulties insuring subjects took the
-proper dose and using subjects with prior polyp history who may be
-beyond a point of benefiting from fiber supplementation. Together, these
-findings temper enthusiasm for treatments that target the production of
-SCFAs or for using them as biomarkers for protection against
-tumorigenesis.
+producing them as a biomarker of a healthy microbiota (9, 10). Depending
+on the concentrations, SCFAs can have proliferative activities at low
+concentrations or anti-proliferative activities at higher
+concentrations; they can also have anti-inflammatory activities (11).
+Direct supplementation of SCFAs or feeding of fiber caused an overall
+reduction in tumor burden in mouse models of colorectal cancer (12).
+These results suggest that supplementation with fiber, which many
+colonic bacteria ferment to produce SCFAs, may confer beneficial effects
+against colorectal cancer. Regardless, there is a lack of consistent
+evidence that increasing SCFA concentrations can protect against
+colorectal cancer in humans. Case-control studies that have investigated
+possible associations between SCFAs and colon tumor status have been
+plagued by relatively small numbers of subjects, but have reported
+increased total and relative fecal acetate levels and decreased relative
+fecal butyrate concentrations in subjects with colonic lesions (13). In
+randomized controlled trials fiber supplementation has been
+inconsistently associated with protection against tumor formation and
+recurrence (14, 15). Such studies are confounded by difficulties
+ensuring subjects took the proper dose and using subjects with prior
+polyp history who may be beyond a point of benefiting from fiber
+supplementation. Together, these findings temper enthusiasm for
+treatments that target the production of SCFAs or for using them as
+biomarkers for protection against tumorigenesis.
 
 \textbf{Fecal SCFA concentrations did not vary with diagnosis or
-treatment.} To quantify the associations between colorectal cancer, the
-microbiome, and SCFAs, we quantified the concentration of acetate,
+treatment.} To test for a significant association between colorectal
+cancer and SCFAs, we quantified the concentration of acetate,
 propionate, and butyrate in feces of previously characterized
@@
 N\textsubscript{carcinoma}=26) were treated and sampled a year later
-(17). None of the SCFAs exhibited a significant change with treatment
-(all P\textgreater{}0.058; Figure 1B). For both the pre-treatment
+(17). None of the individuals showed signs of recurrence and yet none of
+the SCFAs exhibited a significant change with treatment (all
+P\textgreater{}0.058; Figure 1B). For both the pre-treatment
 cross-sectional data and the pre/post treatment data, we also failed to
@@
 on a total and per molecule of carbon basis and again failed to observe
-any significant differences (P\textgreater{}0.077). These results
-demonstrated that there were no significant associations between fecal
-SCFA concentration and diagnosis or treatment.
+any significant differences (P\textgreater{}0.077). Although some of the
+P-values from our analyses were close to 0.05, the effect sizes were all
+relatively small and inconsistent given the disease progression (Figure
+1). These results demonstrated that there were no significant
+associations between fecal SCFA concentration and diagnosis or
+treatment.
 
 \textbf{Combining SCFA and microbiome data does not improve the ability
-to diagnose individual as having adenomas or carcinomas.} We previously
-found that binning 16S rRNA gene sequence data into operational
-taxonomic units (OTUs) based on 97\% similarity or into genera enabled
-us to classify individuals as having adenomas or carcinomas using random
-forest machine learning models (8, 16). We repeated that analysis but
-added the concentration of the SCFAs as possible features to train the
-models (Figure S1). Models trained using SCFAs to classify individuals
-as having adenomas or carcinomas rather than normal colons had median
-areas under the receiver operator characteristic curve (AUROC) that were
-significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
-and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
-values to detect the presence of adenomas or carcinomas were only 0.54
-and 0.55, respectively, indicating that SCFAs had poor predictive power
-on their own (Figure 2A). When we trained the models with the SCFAs
+to diagnose individual as having adenomas or carcinomas using a random
+forest model.} We previously found that binning 16S rRNA gene sequence
+data into operational taxonomic units (OTUs) based on 97\% similarity or
+into genera enabled us to classify individuals as having adenomas or
+carcinomas using random forest machine learning models (8, 16). We
+repeated that analysis but added the concentration of the SCFAs as
+possible features to train the models (Figure S1). Models trained using
+SCFAs to classify individuals as having adenomas or carcinomas rather
+than normal colons had median areas under the receiver operator
+characteristic curve (AUROC) that were significantly greater than 0.5
+(P\textsubscript{adenoma}\textless{}0.001 and
+P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC values
+to detect the presence of adenomas or carcinomas were only 0.54 and
+0.55, respectively, indicating that SCFAs had poor predictive power on
+their own (Figure 2A). When we trained the models with the SCFAs
 concentrations and OTU or genus-level relative abundances the AUROC
@@
 \textbf{Knowledge of microbial community structure does not predict SCFA
-concentrations.} We next asked whether the fecal community structure was
-predictive of fecal SCFA concentrations, regardless of a person's
-diagnosis. We trained random forest regression models using 16S rRNA
-gene sequence data binned into OTUs and genera to predict the
-concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
-between the observed SCFA concentrations and the modeled concentrations
-was 0.14, which was observed when using genus data to predict butyrate
-concentrations (Figure 2B). We also used a smaller dataset of shotgun
-metagenomic sequencing data generated from a subset of our cohort
-(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
-N\textsubscript{cancer}=26) (18). We binned genes extracted from the
-assembled metagenomes into operational protein families (OPFs) or KEGG
-categories and trained random forest regression models using metagenomic
-sequence data to predict the concentration of the SCFAs (Figure S2).
-Similar to the analysis using 16S rRNA gene sequence data, the
-metagenomic data was not predictive of SCFA concentration. The largest
-R\textsuperscript{2} was 0.055, which was observed when using KEGG data
-to predict propionate concentrations (Figure 2B). Because of the limited
-number of samples that we were able to generate metagenomic sequence
-data from, we used our 16S rRNA gene sequence data to impute metagenomes
-that were binned into metabolic pathways or KEGG categories using
-PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
-the imputed metagenomic data. The largest R\textsuperscript{2} was
-0.085, which was observed when using KEGG data to predict propionate
-concentrations (Figure 2B). The inability to model SCFA concentrations
-from microbiome data indicates that the knowledge of the abundance of
-organisms and their genes was insufficient to predict SCFA
-concentrations.
+concentrations using a random forest model.} We next asked whether the
+fecal community structure was predictive of fecal SCFA concentrations,
+regardless of a person's diagnosis. We trained random forest regression
+models using 16S rRNA gene sequence data binned into OTUs and genera to
+predict the concentration of the SCFAs (Figure S2). The largest
+R\textsuperscript{2} between the observed SCFA concentrations and the
+modeled concentrations was 0.14, which was observed when using genus
+data to predict butyrate concentrations (Figure 2B). We also used a
+smaller dataset of shotgun metagenomic sequencing data generated from a
+subset of our cohort (N\textsubscript{normal}=27,
+N\textsubscript{adenoma}=25, and N\textsubscript{cancer}=26) (18). We
+binned genes extracted from the assembled metagenomes into operational
+protein families (OPFs) or KEGG categories and trained random forest
+regression models using metagenomic sequence data to predict the
+concentration of the SCFAs (Figure S2). Similar to the analysis using
+16S rRNA gene sequence data, the metagenomic data was not predictive of
+SCFA concentration. The largest R\textsuperscript{2} was 0.055, which
+was observed when using KEGG data to predict propionate concentrations
+(Figure 2B). Because of the limited number of samples that we were able
+to generate metagenomic sequence data from, we used our 16S rRNA gene
+sequence data to impute metagenomes that were binned into metabolic
+pathways or KEGG categories using PICRUSt (Figure S2). SCFA
+concentrations could not be predicted based on the imputed metagenomic
+data. The largest R\textsuperscript{2} was 0.085, which was observed
+when using KEGG data to predict propionate concentrations (Figure 2B).
+The inability to model SCFA concentrations from microbiome data
+indicates that the knowledge of the abundance of organisms and their
+genes was insufficient to predict fecal SCFA concentrations.
 
@@
 someone with one of these lesions. Furthermore, knowledge of the
-taxonomic and genetic structure of gut microbiota was not predictive of
-SCFA concentrations. These results complement existing literature that
-suggest that fiber consumption and the production of SCFAs are unable to
-prevent the risk of developing colonic tumors. It is important to note
-that our analysis was based on characterizations of SCFA and microbiome
-profiles using fecal samples and that observations along the mucosa near
-the site of lesions may provide a stronger association. This may be a
-cautionary result to temper enthusiasm for SCFAs as a biomarker of gut
-health more generally. Going forward it is critical to develop
-additional hypotheses for how the microbiome and host interact to drive
-tumorigenesis so that we can better understand tumorigenesis and
-identify biomarkers that will allow early detection of lesions.
+taxonomic and genetic structure of gut microbiota was not meaningfully
+predictive of SCFA concentrations. These results complement existing
+literature that suggest that fiber consumption and the production of
+SCFAs are unable to prevent the risk of developing colonic tumors. It is
+important to note that our analysis was based on characterizations of
+SCFA and microbiome profiles using fecal samples at a single time point.
+Furthermore, observations along the mucosa near the site of lesions may
+provide a stronger association. This may be a cautionary result to
+temper enthusiasm for SCFAs as a biomarker of gut health more generally.
+Going forward it is critical to develop additional hypotheses for how
+the microbiome and host interact to drive tumorigenesis so that we can
+better understand tumorigenesis and identify biomarkers that will allow
+early detection of lesions.
 
@@
 
-\textbf{random forest models.} The classification models were built to
+\textbf{Random forest models.} The classification models were built to
 predict lesion type from microbiome information with or without SCFA
@@
 performing hyperparameter, mtry, was selected in an internal five-fold
-cross-validation of the training set with 100 randomizations. Six values
-of mtry were tested and the value that provided the largest AUROC or
-R\textsuperscript{2} was selected. We trained the random forest model
-using the selected mtry value and predicted the held-out test set. The
-data-split, hyperparameter selection, training and testing steps were
-repeated 100 times to get a reliable and robust reading of model
-prediction performance. We used AUROC and R\textsuperscript{2} as the
-prediction performance metric for classification and regression models,
-respectively. We used randomForest package implemented to the caret
-package (version 4.6-14) in R statistical software (version 6.0-81) for
-our models.
+cross-validation of the training set with 100 randomizations. The mtry
+parameter represents the number of features randomly sampled from the
+available features at a question point in the classification tree
+(i.e.~called splits of nodes) that, when answered, lead to the greatest
+improvement in classification. Six values of mtry were tested and the
+value that provided the largest AUROC or R\textsuperscript{2} was
+selected. We trained the random forest model using the selected mtry
+value and predicted the held-out test set. The data-split,
+hyperparameter selection, training and testing steps were repeated 100
+times to get a reliable and robust reading of model prediction
+performance. We used AUROC and R\textsuperscript{2} as the prediction
+performance metric for classification and regression models,
+respectively. We used the randomForest R package (version 4.6-14) as
+implemented in the caret R package (version 6.0-81) for developing and
+testing our models.
 
@@
 analysis steps including an Rmarkdown version of this manuscript is
-available at \url{https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/}.
+available at \url{https://github.com/SchlossLab/Sze_SCFACRC_mBio_2019/}.
 
@@
 
-\includegraphics{figure_1.png}
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
 
@@
 clarity.
-
-\newpage
-
-\includegraphics{figure_2.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s1.png}
-
 \textbf{Figure S1. Comparison of training and testing results for
@@
 
-\newpage
-
-\includegraphics{figure_s2.png}
-
 \textbf{Figure S2. Comparison of training and testing results for

### 2019-05-30 code/plot_scfa_comparisons.R

@@ -62,7 +62,7 @@ longitudinal <- inner_join(scfa_data, follow_up_metadata, by=c("study_id"="initi
 		geom_boxplot() +
 		geom_blank(aes(y = ymin)) + geom_blank(aes(y = ymax)) +
 		facet_wrap(~scfa, scales="free_y", nrow=1, labeller=labeller(.default=capitalize)) +
-		labs(x=NULL, y="Difference in SCFA concentration\nbefore and after treatment") +
+		labs(x=NULL, y="Difference in SCFA concentration\nbefore and after treatment\n(mmol of SCFA/kg of feces)") +
 		scale_x_discrete(labels = NULL) +
 		scale_fill_manual(
 			breaks=c("normal", "adenoma", "cancer"),
